---
title: "Panel Degradation Pathologies: Volatility, Level Bias, and Composition Effects"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
execute:
  warning: false
  message: false
---

Four analyses examine what V-Dem panel degradation does to measurement quality,
motivating the panel-member paper's augmentation experiment. Analyses 1–3 test for
bias in published point estimates using volatility, TWFE levels, and first differences
respectively; they find small, regime-conditional effects but no systematic crisis in
the raw scores. Analysis 4 is the primary result: using V-Dem's own published posterior
standard deviations, it shows that panel size directly and robustly predicts IRT
uncertainty, and that post-2013 attrition has raised average posterior SD by
approximately 12% relative to the 2008–2010 peak. The case for augmentation rests on
uncertainty, not bias.

Indicator selection follows bridge-coder QMD 07: all 0–4 ordinal indicators in 2020 with
at least 50 ratings, direction-coded from their correlation with v2x_polyarchy, with
ambiguous indicators (|r| ≤ 0.10) dropped. The period is 1990–2024. Starting at 1990
avoids the retrospective-coding artefact: pre-1990 coders assigned values to historical
years all at once, so year-over-year "changes" reflect a single expert's interpolation
rather than responses to real conditions.

---

## 1. Setup

```{r setup}
library(tidyverse)
library(fixest)
library(modelsummary)
library(gt)
library(vdemdata)
library(here)

anchor_vars <- c(
  "v2zzdemswe", "v2zzdemus",  "v2zzdemru",  "v2zzdemnk",
  "v2zzdemuk",  "v2zzdemin",  "v2zzdemcr",  "v2zzdemcu",
  "v2zzdemni",  "v2zzdemsar", "v2zzdemsaf", "v2zzdemswz", "v2zzdemvz"
)

global_anchor_means <- c(
  v2zzdemswe = 91.07, v2zzdemus  = 78.74, v2zzdemru  = 26.59,
  v2zzdemnk  =  6.92, v2zzdemuk  = 84.38, v2zzdemin  = 63.78,
  v2zzdemcr  = 69.11, v2zzdemcu  = 24.38, v2zzdemni  = 42.75,
  v2zzdemsar = 14.14, v2zzdemsaf = 67.36, v2zzdemswz = 90.39,
  v2zzdemvz  = 27.78
)

region_labels <- c(
  "1" = "E. Europe & C. Asia",
  "2" = "Latin America",
  "3" = "MENA",
  "4" = "Sub-Saharan Africa",
  "5" = "W. Europe & N. America",
  "6" = "Asia & Pacific"
)

coder_path <- file.path(
  here::here(".."),
  "initial-exploration/V-Dem-Coder-Level-v15_rds/Coder-Level-Dataset-v15.rds"
)
```

---

## 2. Data loading and preparation

```{r load-full}
#| cache: true
#| cache.lazy: false
d <- readRDS(coder_path)
d$year <- as.integer(format(as.Date(d[["historical_date"]], origin = "1970-01-01"), "%Y"))

cat("Full dataset rows:", nrow(d), "\n")
cat("Year range:", min(d$year), "–", max(d$year), "\n")
```

```{r coder-mad}
#| cache: true
#| cache.lazy: false

# Anchor ratings and coder-level MAD — computed from full dataset so that
# coders who are active only in early years still get an anchor MAD estimate
coder_anchors_raw <- d |>
  group_by(coder_id) |>
  summarise(across(all_of(anchor_vars), ~ first(na.omit(.))), .groups = "drop")

coder_mad <- coder_anchors_raw |>
  rowwise() |>
  mutate(
    n_anchors  = sum(!is.na(c_across(all_of(anchor_vars)))),
    anchor_mad = mean(abs(c_across(all_of(anchor_vars)) - global_anchor_means), na.rm = TRUE)
  ) |>
  ungroup() |>
  filter(n_anchors >= 5) |>
  select(coder_id, anchor_mad)

cat("Coders with anchor MAD (≥ 5 anchors):", nrow(coder_mad), "\n")
```

```{r indicator-selection}
#| cache: true
#| cache.lazy: false

# Direction-code from the 2020 cross-section — consistent with bridge-coder QMD 07
d2020 <- d[d$year == 2020, ]

vdem_2020 <- vdem |>
  filter(year == 2020) |>
  select(country_text_id, v2x_polyarchy)

v2_cols <- grep("^v2[a-z]", names(d2020), value = TRUE)
v2_cols <- v2_cols[!grepl(
  "_conf$|_codelow|_codehigh|_mean|_nr|_ord|_osp|_rec|_sd|_vote|_ccp|casoe|3C|4C|5C|6C|7C|8C",
  v2_cols
)]
v2_cols <- v2_cols[sapply(v2_cols, \(col) is.numeric(d2020[[col]]))]

full_scale_cols <- Filter(function(col) {
  x <- na.omit(d2020[[col]])
  length(x) >= 50 &&
    all(x == floor(x)) &&
    max(x) == 4 &&
    min(x) <= 1 &&
    length(unique(x)) >= 4
}, v2_cols)

cat("Candidate 0–4 indicators:", length(full_scale_cols), "\n")

direction_df <- lapply(full_scale_cols, function(col) {
  means <- d2020 |>
    group_by(country_text_id) |>
    summarise(m = mean(.data[[col]], na.rm = TRUE), .groups = "drop")
  joined <- left_join(means, vdem_2020, by = "country_text_id") |>
    filter(!is.na(v2x_polyarchy), !is.na(m))
  if (nrow(joined) < 10) return(NULL)
  data.frame(
    indicator = col,
    r = cor(joined$m, joined$v2x_polyarchy, use = "complete.obs"),
    n = nrow(joined)
  )
}) |> Filter(Negate(is.null), x = _) |> bind_rows()

indicators_keep     <- filter(direction_df, r >  0.10) |> pull(indicator)
indicators_flip     <- filter(direction_df, r < -0.10) |> pull(indicator)
working_indicators  <- c(indicators_keep, indicators_flip)

cat("Keep (positive direction):", length(indicators_keep), "\n")
cat("Flip (negative direction):", length(indicators_flip), "\n")
cat("Working set:", length(working_indicators), "\n")

rm(d2020, v2_cols, full_scale_cols, direction_df)
```

```{r bridge-coders}
#| cache: true
#| cache.lazy: false

# Bridge coder status: year-specific. A coder is a bridge coder in year t if they
# provided at least one rating on any working indicator for more than one country
# in that year. Uses a rowSums check on the wide data to avoid a full pivot.
bridge_by_year <- d |>
  filter(year >= 1990, year <= 2024) |>
  mutate(has_rating = rowSums(!is.na(across(any_of(working_indicators)))) > 0) |>
  filter(has_rating) |>
  group_by(coder_id, year) |>
  summarise(n_countries = n_distinct(country_text_id), .groups = "drop") |>
  mutate(is_bridge = n_countries > 1) |>
  select(coder_id, year, is_bridge)

cat("Coder-year rows with bridge status:", nrow(bridge_by_year), "\n")
cat("Bridge coder-years:",
    sum(bridge_by_year$is_bridge), "/", nrow(bridge_by_year), "\n")
```

```{r panel-metrics}
#| cache: true
#| cache.lazy: false

# Long-format ratings: 1990–2024, working indicators only, direction-corrected
d_long <- d |>
  filter(year >= 1990, year <= 2024) |>
  select(coder_id, country_text_id, year, all_of(working_indicators)) |>
  pivot_longer(all_of(working_indicators),
               names_to = "indicator", values_to = "rating") |>
  filter(!is.na(rating)) |>
  mutate(
    rating = if_else(indicator %in% indicators_flip, 4 - rating, rating)
  )

rm(d)

# Aggregate to country × indicator × year with composition metrics
panel_metrics <- d_long |>
  left_join(coder_mad,     by = "coder_id") |>
  left_join(bridge_by_year, by = c("coder_id", "year")) |>
  mutate(is_bridge = replace_na(is_bridge, FALSE)) |>
  group_by(country_text_id, indicator, year) |>
  summarise(
    raw_mean        = mean(rating, na.rm = TRUE),
    n_coders        = n_distinct(coder_id),
    bridge_frac     = mean(is_bridge),
    mean_anchor_mad = mean(anchor_mad, na.rm = TRUE),
    n_with_mad      = sum(!is.na(anchor_mad)),
    .groups         = "drop"
  )

rm(d_long, bridge_by_year)

cat("Panel metrics rows:", nrow(panel_metrics), "\n")
```

```{r country-covariates}
#| cache: true

vdem_cov <- vdem |>
  filter(year >= 1990, year <= 2024) |>
  select(country_text_id, year, v2x_polyarchy, v2x_regime, e_regionpol_6C)

poly_global_mean <- mean(vdem_cov$v2x_polyarchy, na.rm = TRUE)

country_summary <- vdem_cov |>
  group_by(country_text_id) |>
  summarise(
    mean_polyarchy_c = mean(v2x_polyarchy, na.rm = TRUE) - poly_global_mean,
    region_modal     = as.integer(names(sort(table(e_regionpol_6C), decreasing = TRUE)[1])),
    .groups          = "drop"
  )

panel_full <- panel_metrics |>
  left_join(vdem_cov, by = c("country_text_id", "year")) |>
  left_join(country_summary, by = "country_text_id") |>
  filter(!is.na(v2x_polyarchy)) |>
  mutate(
    post2013    = as.integer(year > 2013),
    poly_c      = v2x_polyarchy - poly_global_mean,
    region      = factor(e_regionpol_6C, levels = 1:6, labels = region_labels),
    region_modal = factor(region_modal, levels = 1:6, labels = region_labels)
  )

cat("Analysis dataset rows:", nrow(panel_full), "\n")
cat("Countries:", n_distinct(panel_full$country_text_id), "\n")
cat("Indicators:", n_distinct(panel_full$indicator), "\n")
cat("Years:", range(panel_full$year), "\n")
```

---

## 3. Context: Panel size by year and regime type

Before the regressions, two descriptive figures establish the structural problem.

```{r context-pyramid, fig.width = 9, fig.height = 4.5}
panel_full |>
  group_by(year) |>
  summarise(
    mean_n   = mean(n_coders),
    median_n = median(n_coders),
    .groups  = "drop"
  ) |>
  ggplot(aes(x = year)) +
  geom_line(aes(y = mean_n),   colour = "steelblue", linewidth = 1,   linetype = "solid") +
  geom_line(aes(y = median_n), colour = "tomato",    linewidth = 0.8, linetype = "dashed") +
  geom_vline(xintercept = 2005, linetype = "dotted", colour = "grey40") +
  geom_vline(xintercept = 2013, linetype = "dotted", colour = "grey40") +
  annotate("text", x = 2005, y = Inf, label = "2005\n(post-2013 recruit floor)",
           hjust = -0.1, vjust = 1.5, size = 3, colour = "grey40") +
  annotate("text", x = 2013, y = Inf, label = "2013\n(attrition begins)",
           hjust = -0.1, vjust = 1.5, size = 3, colour = "grey40") +
  labs(
    title    = "Mean panel size by year, pooled across all country-indicator pairs (1990–2024)",
    subtitle = "Blue = mean, red dashed = median. Pooled across all working indicators.",
    x = "Year", y = "Coders per country-year"
  ) +
  theme_minimal(base_size = 12)
```

```{r context-regime, fig.width = 8, fig.height = 4.5}
panel_full |>
  filter(!is.na(v2x_regime)) |>
  mutate(
    regime_label = factor(v2x_regime, 0:3,
                          labels = c("Closed\nautocracy", "Electoral\nautocracy",
                                     "Electoral\ndemocracy", "Liberal\ndemocracy"))
  ) |>
  group_by(year, regime_label) |>
  summarise(mean_n = mean(n_coders), .groups = "drop") |>
  ggplot(aes(x = year, y = mean_n, colour = regime_label)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = 2013, linetype = "dotted", colour = "grey40") +
  scale_colour_brewer(palette = "Set1", name = NULL) +
  labs(
    title    = "Mean panel size by year and regime type (1990–2024)",
    subtitle = "Autocracies have persistently thinner panels; all regime types show post-2013 attrition.",
    x = "Year", y = "Mean coders per country-year"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

## 4. Analysis 1: Score Volatility

Country-indicator pairs with chronically thin panels produce more variable scores over
time. This is the most direct illustration of measurement unreliability: even if a thin
panel is unbiased on average, its scores jump around more from year to year as a function
of small-sample noise rather than real political change.

**Design**: one row per country × indicator. Outcome = SD of the annual raw panel mean
across 1990–2024 (minimum 5 years required). Indicator FEs absorb indicator-specific
baseline variance; region FEs absorb regional patterns in political volatility.
M3 adds an interaction with the country's mean polyarchy (time-invariant, not absorbed
by country FEs in a cross-sectional model) to test whether the volatility effect is
concentrated in autocracies. M4 splits the panel-size predictor pre/post 2013 to ask
whether the attrition era specifically drives the result.

```{r vol-data}
#| cache: true

volatility_data <- panel_full |>
  group_by(country_text_id, indicator) |>
  summarise(
    sd_raw_mean      = sd(raw_mean, na.rm = TRUE),
    mean_n_coders    = mean(n_coders, na.rm = TRUE),
    mean_n_pre2013   = mean(n_coders[year <= 2013], na.rm = TRUE),
    mean_n_post2013  = mean(n_coders[year >  2013], na.rm = TRUE),
    mean_polyarchy_c = first(mean_polyarchy_c),
    region_modal     = first(region_modal),
    n_years          = n_distinct(year),
    .groups          = "drop"
  ) |>
  filter(n_years >= 5, !is.na(sd_raw_mean))

cat("Country-indicator pairs:", nrow(volatility_data), "\n")
cat("Median years per pair:", median(volatility_data$n_years), "\n")
```

```{r vol-regressions}
#| cache: true

m1_vol <- feols(sd_raw_mean ~ mean_n_coders,
                data = volatility_data, cluster = ~country_text_id)

m2_vol <- feols(sd_raw_mean ~ mean_n_coders + n_years | indicator + region_modal,
                data = volatility_data, cluster = ~country_text_id)

m3_vol <- feols(sd_raw_mean ~ mean_n_coders * mean_polyarchy_c + n_years |
                  indicator + region_modal,
                data = volatility_data, cluster = ~country_text_id)

m4_vol <- feols(sd_raw_mean ~ mean_n_pre2013 + mean_n_post2013 + n_years |
                  indicator + region_modal,
                data = volatility_data, cluster = ~country_text_id)
```

```{r vol-table}
coef_map_vol <- c(
  "mean_n_coders"              = "Mean panel size",
  "mean_n_pre2013"             = "Mean panel size (1990–2013)",
  "mean_n_post2013"            = "Mean panel size (2014–2024)",
  "mean_polyarchy_c"           = "Mean polyarchy (centered)",
  "mean_n_coders:mean_polyarchy_c" = "Panel size × Mean polyarchy",
  "n_years"                    = "Years of coverage"
)

modelsummary(
  list(M1 = m1_vol, M2 = m2_vol, M3 = m3_vol, M4 = m4_vol),
  coef_map  = coef_map_vol,
  statistic = "({std.error})",
  stars     = c("*" = 0.05, "**" = 0.01, "***" = 0.001),
  gof_map   = list(
    list(raw = "nobs",      clean = "N",  fmt = 0),
    list(raw = "r.squared", clean = "R²", fmt = 3)
  ),
  notes  = "OLS. M2–M4 absorb indicator and region FEs. SEs clustered by country.",
  title  = "Score volatility (SD of annual raw panel mean, 1990–2024) ~ panel size",
  output = "gt"
)
```

```{r vol-scatter, fig.width = 9, fig.height = 5}
volatility_data |>
  mutate(
    regime_group = case_when(
      mean_polyarchy_c < -0.15 ~ "Autocracy",
      mean_polyarchy_c <  0.15 ~ "Hybrid",
      TRUE                     ~ "Democracy"
    ),
    regime_group = factor(regime_group, levels = c("Autocracy", "Hybrid", "Democracy"))
  ) |>
  ggplot(aes(x = mean_n_coders, y = sd_raw_mean, colour = regime_group)) +
  geom_point(alpha = 0.15, size = 0.7) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9) +
  scale_colour_manual(
    values = c(Autocracy = "firebrick", Hybrid = "goldenrod", Democracy = "steelblue"),
    name   = NULL
  ) +
  labs(
    x        = "Mean panel size (distinct coders per year, 1990–2024)",
    y        = "SD of annual raw panel mean",
    title    = "Larger panels produce more stable scores",
    subtitle = "Each point = one country × indicator. Effect strongest in autocratic contexts."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

## 5. Analysis 2: Level Bias (TWFE)

If thin panels produce biased scores, the mechanism is regional reference frame
compression: panels dominated by coders from authoritarian contexts inflate raw means.
This produces a systematic level shift — thin-panel autocracies score higher than
same-level peers — distinct from the variance increase in Analysis 1.

**Design**: standard TWFE, one row per country × indicator × year. Country FEs absorb
time-invariant country characteristics. Year FEs absorb global annual shocks (real
democratic backsliding, V-Dem methodology changes). Indicator FEs absorb
indicator-level baselines. Identification comes from within-country variation in panel
size over time — primarily the post-2013 attrition.

`mean_polyarchy_c` is time-invariant (country mean over 1990–2024), so its main effect
is absorbed by the country FE; only its interaction with `n_coders` is estimated. This
interaction tests whether the within-country panel-size effect depends on a country's
typical democracy level. The `n_coders:post2013` interaction tests whether the panel-size
effect strengthened after attrition began, without dropping year FEs (the interaction
varies cross-sectionally within years and is not collinear with year FEs).

```{r lev-regressions}
#| cache: true

m1_lev <- feols(
  raw_mean ~ n_coders | country_text_id + indicator + year,
  data = panel_full, cluster = ~country_text_id
)

m2_lev <- feols(
  raw_mean ~ n_coders + n_coders:mean_polyarchy_c | country_text_id + indicator + year,
  data = panel_full, cluster = ~country_text_id
)

m3_lev <- feols(
  raw_mean ~ n_coders + n_coders:post2013 | country_text_id + indicator + year,
  data = panel_full, cluster = ~country_text_id
)

m4_lev <- feols(
  raw_mean ~ n_coders + n_coders:mean_polyarchy_c + n_coders:post2013 |
    country_text_id + indicator + year,
  data = panel_full, cluster = ~country_text_id
)
```

```{r lev-table}
coef_map_lev <- c(
  "n_coders"                = "Panel size",
  "n_coders:mean_polyarchy_c" = "Panel size × Mean polyarchy",
  "n_coders:post2013"       = "Panel size × Post-2013"
)

modelsummary(
  list(M1 = m1_lev, M2 = m2_lev, M3 = m3_lev, M4 = m4_lev),
  coef_map  = coef_map_lev,
  statistic = "({std.error})",
  stars     = c("*" = 0.05, "**" = 0.01, "***" = 0.001),
  gof_map   = list(
    list(raw = "nobs",       clean = "N",        fmt = 0),
    list(raw = "r.squared",  clean = "R²",        fmt = 3),
    list(raw = "r2.within",  clean = "Within R²", fmt = 3)
  ),
  notes  = "TWFE: country + indicator + year FEs absorbed. SEs clustered by country.",
  title  = "Level bias: raw_mean ~ panel size, 1990–2024",
  output = "gt"
)
```

```{r lev-interaction-plot, fig.width = 8, fig.height = 5}
# Effective slope of n_coders as a function of country mean polyarchy (from M2)
poly_seq   <- seq(-0.5, 0.5, by = 0.01)
slope_seq  <- coef(m2_lev)["n_coders"] +
              coef(m2_lev)["n_coders:mean_polyarchy_c"] * poly_seq
slope_lo   <- (coef(m2_lev)["n_coders"] - 1.96 * se(m2_lev)["n_coders"]) +
              (coef(m2_lev)["n_coders:mean_polyarchy_c"] -
               1.96 * se(m2_lev)["n_coders:mean_polyarchy_c"]) * poly_seq
slope_hi   <- (coef(m2_lev)["n_coders"] + 1.96 * se(m2_lev)["n_coders"]) +
              (coef(m2_lev)["n_coders:mean_polyarchy_c"] +
               1.96 * se(m2_lev)["n_coders:mean_polyarchy_c"]) * poly_seq

ggplot(
  data.frame(poly = poly_seq + poly_global_mean, slope = slope_seq,
             lo = slope_lo, hi = slope_hi),
  aes(x = poly, y = slope)
) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, fill = "firebrick") +
  geom_line(colour = "firebrick", linewidth = 1.2) +
  labs(
    x = "Country mean polyarchy (v2x_polyarchy, 1990–2024 average)",
    y = "Within-country effect of one additional coder on raw_mean",
    title = "Thin panels push autocratic scores toward the floor",
    subtitle = paste0(
      "More coders raise raw means in autocracies and lower them in democracies, ",
      "compressing scores toward the global center. Attrition has the opposite effect."
    )
  ) +
  theme_minimal(base_size = 12)
```

---

## 6. Analysis 3: Generalized Afghanistan (First Differences)

The level-bias result is consistent with a compositional confound: autocratic countries
may systematically attract less-calibrated coders regardless of panel size, producing an
omitted-variables bias even within TWFE. The first-difference design eliminates this by
identifying from *changes* in composition within the same country-indicator over time.

The direct claim: year-over-year changes in panel calibration quality (`Δ mean anchor MAD`)
and bridge coder fraction (`Δ bridge_frac`) predict year-over-year score changes, net of
indicator-level change rates (indicator FEs) and global annual shocks (year FEs). Country
FEs drop out through differencing. Only consecutive year pairs are used.

The regime interaction in M3 tests whether the composition effect is concentrated in
autocracies — the direct generalisation of the Afghanistan mechanism.

```{r fd-data}
#| cache: true

fd_data <- panel_full |>
  filter(!is.na(mean_anchor_mad)) |>
  arrange(country_text_id, indicator, year) |>
  group_by(country_text_id, indicator) |>
  mutate(
    d_raw_mean    = raw_mean        - lag(raw_mean),
    d_n_coders    = n_coders        - lag(n_coders),
    d_anchor_mad  = mean_anchor_mad - lag(mean_anchor_mad),
    d_bridge_frac = bridge_frac     - lag(bridge_frac),
    regime_lag    = lag(v2x_regime),
    consec        = (year - lag(year)) == 1L
  ) |>
  ungroup() |>
  filter(!is.na(d_raw_mean), consec)

cat("First-difference observations:", nrow(fd_data), "\n")
cat("Countries:", n_distinct(fd_data$country_text_id), "\n")
cat("Years covered:", range(fd_data$year), "\n")
```

```{r fd-regressions}
#| cache: true

m1_fd <- feols(
  d_raw_mean ~ d_n_coders | indicator + year,
  data = fd_data, cluster = ~country_text_id
)

m2_fd <- feols(
  d_raw_mean ~ d_n_coders + d_anchor_mad + d_bridge_frac | indicator + year,
  data = fd_data, cluster = ~country_text_id
)

m3_fd <- feols(
  d_raw_mean ~ d_n_coders   * regime_lag +
               d_anchor_mad  * regime_lag +
               d_bridge_frac * regime_lag | indicator + year,
  data = fd_data, cluster = ~country_text_id
)
```

```{r fd-table}
coef_map_fd <- c(
  "d_n_coders"                   = "Δ Panel size",
  "d_anchor_mad"                 = "Δ Mean anchor MAD",
  "d_bridge_frac"                = "Δ Bridge coder fraction",
  "regime_lag"                   = "Regime type t−1 (0–3)",
  "d_n_coders:regime_lag"        = "Δ Panel size × Regime",
  "d_anchor_mad:regime_lag"      = "Δ Anchor MAD × Regime",
  "d_bridge_frac:regime_lag"     = "Δ Bridge frac × Regime"
)

modelsummary(
  list(M1 = m1_fd, M2 = m2_fd, M3 = m3_fd),
  coef_map  = coef_map_fd,
  statistic = "({std.error})",
  stars     = c("*" = 0.05, "**" = 0.01, "***" = 0.001),
  gof_map   = list(
    list(raw = "nobs",       clean = "N",        fmt = 0),
    list(raw = "r.squared",  clean = "R²",        fmt = 3),
    list(raw = "r2.within",  clean = "Within R²", fmt = 3)
  ),
  notes  = paste0(
    "First differences. Country FEs removed by differencing. ",
    "Indicator + year FEs absorbed. Consecutive year pairs only. ",
    "SEs clustered by country."
  ),
  title  = "Generalized Afghanistan: composition changes predict score changes (1990–2024)",
  output = "gt"
)
```

```{r fd-coef-plot, fig.width = 9, fig.height = 5.5}
bind_rows(
  broom::tidy(m2_fd, conf.int = TRUE) |> mutate(model = "M2: Composition"),
  broom::tidy(m3_fd, conf.int = TRUE) |> mutate(model = "M3: × Regime")
) |>
  filter(term %in% names(coef_map_fd)) |>
  mutate(term = coef_map_fd[term],
         term = factor(term, levels = rev(coef_map_fd))) |>
  ggplot(aes(x = estimate, y = term, colour = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.25, linewidth = 0.6,
                 position = position_dodge(0.5)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  scale_colour_manual(
    values = c("M2: Composition" = "steelblue", "M3: × Regime" = "firebrick"),
    name   = NULL
  ) +
  labs(
    x        = "Coefficient (95% CI, SEs clustered by country)",
    y        = NULL,
    title    = "Composition changes drive score changes — concentrated in autocracies",
    subtitle = "First-difference estimates. Indicator + year FEs. Consecutive year pairs only."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r fd-country-scan}
#| cache: true

# For each country: R² from regressing d_raw_mean on d_anchor_mad,
# pooled across indicators and years. Countries where R² is highest are
# those where composition changes most reliably predict score changes —
# generalisations of the Afghanistan pattern.
country_r2 <- fd_data |>
  filter(!is.na(d_anchor_mad)) |>
  group_by(country_text_id) |>
  filter(n() >= 20) |>
  group_modify(~ {
    m <- lm(d_raw_mean ~ d_anchor_mad, data = .x)
    tibble(
      r2          = summary(m)$r.squared,
      coef_mad    = coef(m)["d_anchor_mad"],
      n_obs       = nrow(.x),
      mean_n      = round(mean(.x$n_coders, na.rm = TRUE), 1),
      mean_regime = round(mean(.x$regime_lag, na.rm = TRUE), 2)
    )
  }) |>
  ungroup() |>
  arrange(desc(r2))

cat("Countries with ≥ 20 FD observations:", nrow(country_r2), "\n")
cat("Median R²:", round(median(country_r2$r2), 3), "\n")
cat("Share with positive coefficient:", round(mean(country_r2$coef_mad > 0), 2), "\n")
```

```{r fd-country-table}
country_r2 |>
  slice_head(n = 20) |>
  mutate(across(c(r2, coef_mad), ~ round(., 3))) |>
  select(country_text_id, r2, coef_mad, n_obs, mean_n, mean_regime) |>
  rename(
    Country             = country_text_id,
    `R²`                = r2,
    `Coef (Δ MAD)`      = coef_mad,
    `N obs`             = n_obs,
    `Mean panel size`   = mean_n,
    `Mean regime (0–3)` = mean_regime
  ) |>
  knitr::kable(
    caption = paste0(
      "Top 20 countries where Δ mean anchor MAD explains Δ raw_mean. ",
      "R² from bivariate regression of Δraw_mean ~ Δanchor_MAD, pooled across ",
      "all indicators and consecutive year pairs. ",
      "Positive coefficient = less-calibrated panel → higher scores (floor inflation). ",
      "Mean regime: 0 = closed autocracy, 3 = liberal democracy."
    )
  )
```

---

## 7. Analysis 4: Panel Size and IRT Posterior Uncertainty

The V-Dem codebook documents a `_nr` suffix (number of coders per country-variable-year)
and a `_sd` suffix (posterior standard deviation from the measurement model) for every
C-type indicator. These are available directly via `vdemdata::vdem`, so no coder-level
data is needed here. The codebook notes explicitly: *"We strongly advise against using
observations based on three or fewer coders."*

```{r uncertainty-data}
nr_cols <- paste0(working_indicators, "_nr")
sd_cols <- paste0(working_indicators, "_sd")

vdem_raw <- vdemdata::vdem |>
  select(country_text_id, year, any_of(c(nr_cols, sd_cols))) |>
  filter(year >= 1990, year <= 2024)

unc_long <- vdem_raw |>
  pivot_longer(
    cols          = -c(country_text_id, year),
    names_to      = c("indicator", ".value"),
    names_pattern = "^(.+)_(nr|sd)$"
  ) |>
  rename(n_coders = nr, posterior_sd = sd) |>
  filter(!is.na(n_coders), !is.na(posterior_sd), n_coders >= 1,
         indicator %in% working_indicators)

cat("Country-indicator-years:", nrow(unc_long), "\n")
cat("Panel size range:", paste(range(unc_long$n_coders), collapse = "–"), "\n")
cat("Posterior SD range:", paste(round(range(unc_long$posterior_sd), 3), collapse = "–"), "\n")
```

```{r uncertainty-binned-plot}
unc_binned <- unc_long |>
  mutate(n_bin = case_when(
    n_coders <= 3  ~ "1–3\n(V-Dem\nwarns)",
    n_coders <= 5  ~ "4–5",
    n_coders <= 8  ~ "6–8",
    n_coders <= 12 ~ "9–12",
    TRUE           ~ "13+"
  ),
  n_bin = factor(n_bin, levels = c("1–3\n(V-Dem\nwarns)", "4–5", "6–8", "9–12", "13+"))) |>
  group_by(n_bin) |>
  summarise(
    median_sd = median(posterior_sd),
    q25_sd    = quantile(posterior_sd, 0.25),
    q75_sd    = quantile(posterior_sd, 0.75),
    n         = n(),
    .groups   = "drop"
  )

ggplot(unc_binned, aes(x = n_bin, y = median_sd)) +
  geom_col(fill = "steelblue", alpha = 0.75, width = 0.6) +
  geom_errorbar(aes(ymin = q25_sd, ymax = q75_sd), width = 0.25) +
  labs(
    title    = "Fewer coders → wider IRT posterior intervals",
    subtitle = "Median posterior SD by panel size bin. Error bars = IQR. Pooled across ~150 working indicators, 1990–2024.",
    x        = "Panel size (number of coders)",
    y        = "Posterior SD (V-Dem measurement model)"
  ) +
  theme_minimal(base_size = 12)
```

```{r uncertainty-time-plot}
unc_year <- unc_long |>
  group_by(year) |>
  summarise(
    mean_sd  = mean(posterior_sd, na.rm = TRUE),
    mean_nr  = mean(n_coders,     na.rm = TRUE),
    .groups  = "drop"
  )

ggplot(unc_year, aes(x = year, y = mean_sd)) +
  geom_line(colour = "steelblue", linewidth = 0.9) +
  geom_smooth(method = "loess", span = 0.35, colour = "darkred", se = FALSE) +
  geom_vline(xintercept = 2013, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 2014, y = max(unc_year$mean_sd) * 0.98,
           label = "Post-2013 attrition", hjust = 0, size = 3.5, colour = "grey30") +
  labs(
    title    = "IRT uncertainty has risen with post-2013 panel attrition",
    subtitle = "Mean posterior SD across all country-indicator-years, 1990–2024.",
    x        = NULL,
    y        = "Mean posterior SD"
  ) +
  theme_minimal(base_size = 12)
```

```{r uncertainty-regression}
m_unc1 <- feols(posterior_sd ~ n_coders,
                data = unc_long, cluster = ~country_text_id)
m_unc2 <- feols(posterior_sd ~ n_coders | indicator,
                data = unc_long, cluster = ~country_text_id)
m_unc3 <- feols(posterior_sd ~ n_coders | indicator + country_text_id,
                data = unc_long, cluster = ~country_text_id)
m_unc4 <- feols(posterior_sd ~ n_coders | indicator + country_text_id + year,
                data = unc_long, cluster = ~country_text_id)

modelsummary(
  list(M1 = m_unc1, M2 = m_unc2, M3 = m_unc3, M4 = m_unc4),
  stars      = TRUE,
  coef_rename = c(n_coders = "Panel size"),
  gof_map    = c("nobs", "r.squared", "r2.within"),
  notes      = "OLS. M2 absorbs indicator FE. M3 adds country FE. M4 adds year FE. SEs clustered by country.",
  output     = "gt",
  title      = "Panel size and IRT posterior uncertainty (posterior SD ~ panel size)"
)
```

---

## 8. Conclusion

The four analyses tell a consistent but hierarchically ordered story about what panel
degradation does to V-Dem measurement.

**Analyses 1–3 are honest nulls.** Score volatility (Analysis 1) is driven almost
entirely by indicator-level differences — which indicators are inherently noisier — not
by panel size. The TWFE level-bias analysis (Analysis 2) finds a statistically
significant but small interaction: additional coders compress raw means toward the
global center, raising scores in autocracies and lowering them in democracies, with an
effect that is regime-conditional and robust but tiny in explained variance (within
R² = 0.002). The first-difference analysis (Analysis 3) is broadly null until regime
type is conditioned on, at which point composition changes — particularly bridge coder
entry — produce marginally significant score shifts in autocracies. None of these
results constitutes a crisis. Panel degradation does not produce dramatic or consistent
bias in V-Dem's published point estimates.

**Analysis 4 is where the consequence is measurable and direct.** V-Dem's own
posterior standard deviations — the uncertainty the measurement model places around
each country-indicator-year estimate — are strongly and robustly predicted by panel
size. Each additional coder reduces posterior SD by 0.012–0.013 units (p < 0.001),
and this effect survives indicator, country, and year fixed effects (within R² = 0.046
in the most saturated specification). The effect is structural: more observations give
the IRT model more information, producing tighter posteriors. This is not a finding
that could go the other way.

The time series makes the practical consequence visible. Mean posterior SD fell
continuously from 1990 to a minimum around 2008–2010 — the peak of the panel-size
pyramid — then reversed sharply after 2013 as the initial coder cohort began to
attrite. By 2024, average IRT uncertainty has risen by approximately 0.05 SD units,
roughly 12% above the 2008–2010 minimum. The regression coefficient almost exactly
predicts this observed rise: the ~4–5 coder reduction in mean panel size since 2013
implies a +0.05–0.06 increase in posterior SD at the estimated −0.012 per coder.

The argument for panel augmentation therefore rests on uncertainty, not bias. V-Dem
expert panels were built to provide the IRT model with sufficient information to
estimate latent democracy scores with reasonable precision. Post-2013 attrition has
eroded that precision in a way that is measurable, monotone, and ongoing. The
replacement experiment tests whether AI coders can rebuild effective panel size and
reverse the uncertainty trend — using V-Dem's own posterior SD as the direct measure
of whether the augmentation is working.
