---
title: "Panel Degradation Pathologies: Volatility, Level Bias, and Composition Effects"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
execute:
  warning: false
  message: false
---

Four analyses examine what V-Dem panel degradation does to measurement quality,
motivating the panel-member paper's augmentation experiment. Analyses 1–3 test for
bias in published point estimates using volatility, TWFE levels, and first differences
respectively; they find small, regime-conditional effects but no systematic crisis in
the raw scores. Analysis 4 is the primary result: using V-Dem's own published posterior
standard deviations, it shows that panel size directly and robustly predicts IRT
uncertainty, and that average posterior SD in the years V-Dem reports today runs about
12% above its 2008–2010 low. Analysis 14 shows that gap is back-coding that has not
arrived yet rather than a shrinking workforce, which makes it a standing feature of the
leading edge rather than a decline. The case for augmentation rests on uncertainty, not
bias.

Indicator selection follows bridge-coder QMD 07: all 0–4 ordinal indicators in 2020 with
at least 50 ratings, direction-coded from their correlation with v2x_polyarchy, with
ambiguous indicators (|r| ≤ 0.10) dropped. The period is 1990–2024. Starting at 1990
avoids the retrospective-coding artefact: pre-1990 coders assigned values to historical
years all at once, so year-over-year "changes" reflect a single expert's interpolation
rather than responses to real conditions.

---

## 1. Setup

```{r setup}
library(tidyverse)
library(fixest)
library(modelsummary)
library(gt)
library(vdemdata)
library(here)

anchor_vars <- c(
  "v2zzdemswe", "v2zzdemus",  "v2zzdemru",  "v2zzdemnk",
  "v2zzdemuk",  "v2zzdemin",  "v2zzdemcr",  "v2zzdemcu",
  "v2zzdemni",  "v2zzdemsar", "v2zzdemsaf", "v2zzdemswz", "v2zzdemvz"
)

global_anchor_means <- c(
  v2zzdemswe = 91.07, v2zzdemus  = 78.74, v2zzdemru  = 26.59,
  v2zzdemnk  =  6.92, v2zzdemuk  = 84.38, v2zzdemin  = 63.78,
  v2zzdemcr  = 69.11, v2zzdemcu  = 24.38, v2zzdemni  = 42.75,
  v2zzdemsar = 14.14, v2zzdemsaf = 67.36, v2zzdemswz = 90.39,
  v2zzdemvz  = 27.78
)

region_labels <- c(
  "1" = "E. Europe & C. Asia",
  "2" = "Latin America",
  "3" = "MENA",
  "4" = "Sub-Saharan Africa",
  "5" = "W. Europe & N. America",
  "6" = "Asia & Pacific"
)

coder_path <- file.path(
  here::here(".."),
  "initial-exploration/V-Dem-Coder-Level-v15_rds/Coder-Level-Dataset-v15.rds"
)
```

---

## 2. Data loading and preparation

```{r load-full}
#| cache: true
#| cache.lazy: false
d <- readRDS(coder_path)
d$year <- as.integer(format(as.Date(d[["historical_date"]], origin = "1970-01-01"), "%Y"))

cat("Full dataset rows:", nrow(d), "\n")
cat("Year range:", min(d$year), "–", max(d$year), "\n")
```

```{r coder-mad}
#| cache: true
#| cache.lazy: false

# Anchor ratings and coder-level MAD — computed from full dataset so that
# coders who are active only in early years still get an anchor MAD estimate
coder_anchors_raw <- d |>
  group_by(coder_id) |>
  summarise(across(all_of(anchor_vars), ~ first(na.omit(.))), .groups = "drop")

coder_mad <- coder_anchors_raw |>
  rowwise() |>
  mutate(
    n_anchors  = sum(!is.na(c_across(all_of(anchor_vars)))),
    anchor_mad = mean(abs(c_across(all_of(anchor_vars)) - global_anchor_means), na.rm = TRUE)
  ) |>
  ungroup() |>
  filter(n_anchors >= 5) |>
  select(coder_id, anchor_mad)

cat("Coders with anchor MAD (≥ 5 anchors):", nrow(coder_mad), "\n")
```

```{r indicator-selection}
#| cache: true
#| cache.lazy: false

# Direction-code from the 2020 cross-section — consistent with bridge-coder QMD 07
d2020 <- d[d$year == 2020, ]

vdem_2020 <- vdem |>
  filter(year == 2020) |>
  select(country_text_id, v2x_polyarchy)

v2_cols <- grep("^v2[a-z]", names(d2020), value = TRUE)
v2_cols <- v2_cols[!grepl(
  "_conf$|_codelow|_codehigh|_mean|_nr|_ord|_osp|_rec|_sd|_vote|_ccp|casoe|3C|4C|5C|6C|7C|8C",
  v2_cols
)]
v2_cols <- v2_cols[sapply(v2_cols, \(col) is.numeric(d2020[[col]]))]

full_scale_cols <- Filter(function(col) {
  x <- na.omit(d2020[[col]])
  length(x) >= 50 &&
    all(x == floor(x)) &&
    max(x) == 4 &&
    min(x) <= 1 &&
    length(unique(x)) >= 4
}, v2_cols)

cat("Candidate 0–4 indicators:", length(full_scale_cols), "\n")

direction_df <- lapply(full_scale_cols, function(col) {
  means <- d2020 |>
    group_by(country_text_id) |>
    summarise(m = mean(.data[[col]], na.rm = TRUE), .groups = "drop")
  joined <- left_join(means, vdem_2020, by = "country_text_id") |>
    filter(!is.na(v2x_polyarchy), !is.na(m))
  if (nrow(joined) < 10) return(NULL)
  data.frame(
    indicator = col,
    r = cor(joined$m, joined$v2x_polyarchy, use = "complete.obs"),
    n = nrow(joined)
  )
}) |> Filter(Negate(is.null), x = _) |> bind_rows()

indicators_keep     <- filter(direction_df, r >  0.10) |> pull(indicator)
indicators_flip     <- filter(direction_df, r < -0.10) |> pull(indicator)
working_indicators  <- c(indicators_keep, indicators_flip)

cat("Keep (positive direction):", length(indicators_keep), "\n")
cat("Flip (negative direction):", length(indicators_flip), "\n")
cat("Working set:", length(working_indicators), "\n")

rm(d2020, v2_cols, full_scale_cols, direction_df)

# The 205 indicators the paper's results actually rest on, from helpers/
# build_analysis_indicators.R. `working_indicators` above is a direction-coded SUBSET
# (152) built for Analyses 1-3, which need a sign relative to polyarchy; its 0-4
# full-range filter drops 64 indicators the paper does score. Descriptive panel-size
# figures should describe the analytic sample, so they use `analysis_indicators`.
analysis_indicators <- readRDS(here::here("data", "derived",
                                          "analysis_indicators.rds"))$with_nr
cat("Analytic set (scored by the paper):", length(analysis_indicators), "\n")
cat("  also in the direction-coded working set:",
    length(intersect(analysis_indicators, working_indicators)), "\n")
cat("  scored but outside it:",
    length(setdiff(analysis_indicators, working_indicators)), "\n")
```

```{r bridge-coders}
#| cache: true
#| cache.lazy: false

# Bridge coder status: year-specific. A coder is a bridge coder in year t if they
# provided at least one rating on any working indicator for more than one country
# in that year. Uses a rowSums check on the wide data to avoid a full pivot.
bridge_by_year <- d |>
  filter(year >= 1990, year <= 2024) |>
  mutate(has_rating = rowSums(!is.na(across(any_of(working_indicators)))) > 0) |>
  filter(has_rating) |>
  group_by(coder_id, year) |>
  summarise(n_countries = n_distinct(country_text_id), .groups = "drop") |>
  mutate(is_bridge = n_countries > 1) |>
  select(coder_id, year, is_bridge)

cat("Coder-year rows with bridge status:", nrow(bridge_by_year), "\n")
cat("Bridge coder-years:",
    sum(bridge_by_year$is_bridge), "/", nrow(bridge_by_year), "\n")
```

```{r panel-metrics}
#| cache: true
#| cache.lazy: false

# Long-format ratings: 1990–2024, working indicators only, direction-corrected
d_long <- d |>
  filter(year >= 1990, year <= 2024) |>
  select(coder_id, country_text_id, year, all_of(working_indicators)) |>
  pivot_longer(all_of(working_indicators),
               names_to = "indicator", values_to = "rating") |>
  filter(!is.na(rating)) |>
  mutate(
    rating = if_else(indicator %in% indicators_flip, 4 - rating, rating)
  )

rm(d)

# Aggregate to country × indicator × year with composition metrics
panel_metrics <- d_long |>
  left_join(coder_mad,     by = "coder_id") |>
  left_join(bridge_by_year, by = c("coder_id", "year")) |>
  mutate(is_bridge = replace_na(is_bridge, FALSE)) |>
  group_by(country_text_id, indicator, year) |>
  summarise(
    raw_mean        = mean(rating, na.rm = TRUE),
    n_coders        = n_distinct(coder_id),
    bridge_frac     = mean(is_bridge),
    mean_anchor_mad = mean(anchor_mad, na.rm = TRUE),
    n_with_mad      = sum(!is.na(anchor_mad)),
    .groups         = "drop"
  )

rm(d_long, bridge_by_year)

cat("Panel metrics rows:", nrow(panel_metrics), "\n")
```

```{r country-covariates}
#| cache: true

vdem_cov <- vdem |>
  filter(year >= 1990, year <= 2024) |>
  select(country_text_id, year, v2x_polyarchy, v2x_regime, e_regionpol_6C)

poly_global_mean <- mean(vdem_cov$v2x_polyarchy, na.rm = TRUE)

country_summary <- vdem_cov |>
  group_by(country_text_id) |>
  summarise(
    mean_polyarchy_c = mean(v2x_polyarchy, na.rm = TRUE) - poly_global_mean,
    region_modal     = as.integer(names(sort(table(e_regionpol_6C), decreasing = TRUE)[1])),
    .groups          = "drop"
  )

panel_full <- panel_metrics |>
  left_join(vdem_cov, by = c("country_text_id", "year")) |>
  left_join(country_summary, by = "country_text_id") |>
  filter(!is.na(v2x_polyarchy)) |>
  mutate(
    post2013    = as.integer(year > 2013),
    poly_c      = v2x_polyarchy - poly_global_mean,
    region      = factor(e_regionpol_6C, levels = 1:6, labels = region_labels),
    region_modal = factor(region_modal, levels = 1:6, labels = region_labels)
  )

cat("Analysis dataset rows:", nrow(panel_full), "\n")
cat("Countries:", n_distinct(panel_full$country_text_id), "\n")
cat("Indicators:", n_distinct(panel_full$indicator), "\n")
cat("Years:", range(panel_full$year), "\n")
```

---

## 3. Context: Panel size by year and regime type

Before the regressions, two descriptive figures establish the structural problem.

```{r context-pyramid, fig.width = 9, fig.height = 4.5}
# Built from V-Dem's published _nr counts on the 205-indicator analytic set, so this
# figure and the band chart below describe the same sample. (panel_full, used by
# Analyses 1-3, stays on the 152 direction-coded indicators.)
pyramid_dat <- vdemdata::vdem |>
  select(country_text_id, year, any_of(paste0(analysis_indicators, "_nr"))) |>
  filter(year >= 1990, year <= 2024) |>
  pivot_longer(-c(country_text_id, year), names_to = "indicator", values_to = "n_coders") |>
  filter(!is.na(n_coders))

pyramid_dat |>
  group_by(year) |>
  summarise(Mean = mean(n_coders), Median = median(n_coders), .groups = "drop") |>
  pivot_longer(c(Mean, Median), names_to = "stat", values_to = "n") |>
  ggplot(aes(x = year, y = n, colour = stat, linetype = stat)) +
  geom_line(linewidth = 0.95) +
  geom_vline(xintercept = 2005, linetype = "dotted", colour = "grey40") +
  geom_vline(xintercept = 2013, linetype = "dotted", colour = "grey40") +
  annotate("text", x = 2005, y = Inf, label = "2005\n(recruit back-coding floor)",
           hjust = -0.1, vjust = 1.5, size = 3, colour = "grey40") +
  annotate("text", x = 2013, y = Inf, label = "2013\n(original round ended 2012)",
           hjust = -0.1, vjust = 1.5, size = 3, colour = "grey40") +
  scale_colour_manual(values = c(Mean = "#0072B2", Median = "#D55E00"), name = NULL) +
  scale_linetype_manual(values = c(Mean = "solid", Median = "22"), name = NULL) +
  labs(
    title    = "Mean panel size by year, pooled across all country-indicator pairs (1990\u20132024)",
    subtitle = paste(length(analysis_indicators),
                     "indicators \u2014 the set scored in the paper's runs."),
    x = "Year", y = "Coders per country-year"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        plot.title.position = "plot")
```

Both vertical lines mark features of V-Dem's coding schedule rather than of its workforce.

**2005** is where contemporary recruits stop back-coding. A newly recruited expert codes
their country from 2005 to the present, so every later cohort adds to 2005 and after but
not before it. That is the step in the line, not a change in how many people were working
in 2005.

**2013** is where V-Dem's original coding round ended. The codebook describes the
contemporary period as "originally coded to 2012," with updates covering 2013-2024 added
afterwards. 709 coders in v15 have 2012 as their last rating-year: they completed the
original round and did not continue into the annual updates. They account for most of the
step. Mean panel size falls 10.31 to 9.45 across that boundary; excluding those 709 coders
it falls only 9.63 to 9.45.

Analysis 14 takes this up directly: comparing successive releases at each one's own leading
edge shows the active roster has been roughly flat since 2020, so the decline to the right
of 2013 is years that have not yet accumulated their back-coders rather than a workforce
that is shrinking.

The figure above is a mean, which hides the shape of the distribution behind it: a mean of
five is produced equally by every panel having five coders and by half having two and half
having eight. The companion below decomposes the same quantity into the share of
country-indicator-years at each panel size, across the full V-Dem time range rather than
just 1990 onward.

Bands are cut at V-Dem's own published thresholds (codebook p. 35): oranges are at or below
four coders, which is the zone V-Dem either warns against (1-3) or leaves below its
recommended floor (4); blues are at or above the floor of five.

```{r context-bands, fig.width = 9.5, fig.height = 5.4}
band_periods <- c(-Inf, 1899, 1944, 1989, 2004, Inf)
band_labels  <- c("19th century\n1789–1899", "Pre-war\n1900–44",
                  "Cold War\n1945–89", "Post-Cold War\n1990–2004",
                  "Contemporary\n2005–2025")

# Separate pull from unc_long, which is restricted to 1990-2024.
band_dat <- vdemdata::vdem |>
  select(country_text_id, year, any_of(paste0(analysis_indicators, "_nr"))) |>
  pivot_longer(-c(country_text_id, year), names_to = "indicator", values_to = "n_coders") |>
  filter(!is.na(n_coders)) |>
  mutate(
    period = cut(year, band_periods, labels = band_labels),
    band   = cut(n_coders, c(0, 1, 2, 3, 4, 8, Inf),
                 labels = c("1", "2", "3", "4", "5–8", "9+"))
  )

band_summary <- band_dat |>
  count(period, band) |>
  group_by(period) |>
  mutate(prop = n / sum(n)) |>
  ungroup()

band_labels_lt5 <- band_dat |>
  group_by(period) |>
  summarise(cells = n(), pct_lt5 = mean(n_coders < 5) * 100, .groups = "drop")

# Panel size is ordered, so a categorical palette would throw the ordering away. These
# are ramps anchored on the repo's two Okabe-Ito hues instead: vermillion (#D55E00)
# stepped dark-to-light below V-Dem's floor, blue (#0072B2) at or above it. Adjacent-pair
# separation validated: worst normal-vision dE 15.6, worst CVD dE 10.2.
band_pal <- c("1"     = "#5A2800", "2"  = "#C25600", "3"  = "#F09A4A",
              "4"     = "#FBD9BE", "5–8" = "#8ECFEE", "9+" = "#0072B2")

ggplot(band_summary, aes(period, prop, fill = band)) +
  geom_col(width = 0.72, colour = "white", linewidth = 0.4,
           position = position_fill(reverse = TRUE)) +
  geom_text(data = band_labels_lt5,
            aes(period, y = 1.04, label = sprintf("%.1f%%", pct_lt5)),
            inherit.aes = FALSE, size = 3.2, colour = "grey25", fontface = "bold") +
  scale_fill_manual(values = band_pal, name = "Coders per\ncountry-indicator-year",
                    guide = guide_legend(reverse = FALSE, nrow = 1)) +
  scale_y_continuous(labels = scales::percent, breaks = seq(0, 1, 0.25),
                     expand = expansion(mult = c(0.01, 0.09))) +
  labs(
    title    = "Panel size is not one story: the distribution behind the mean",
    subtitle = paste("Share of country-indicator-years at each panel size; bold figures are",
                     "the share below five.\nOranges fall below V-Dem's recommended floor."),
    x = NULL, y = "Share of country-indicator-years",
    caption = paste0(length(analysis_indicators), " indicators scored in the paper's runs; ",
                     n_distinct(band_dat$indicator[band_dat$year < 1900]),
                     " of them carry pre-1900 data, so the 19th-century column rests on a ",
                     "narrower indicator set.")
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(), plot.title.position = "plot",
        plot.caption = element_text(colour = "grey40", hjust = 0))
```

```{r context-bands-table}
band_summary |>
  select(period, band, prop) |>
  mutate(prop = round(prop * 100, 2)) |>
  pivot_wider(names_from = band, values_from = prop, values_fill = 0) |>
  left_join(band_labels_lt5 |> select(period, cells, pct_lt5) |>
              mutate(pct_lt5 = round(pct_lt5, 2)), by = "period") |>
  mutate(period = gsub("\n", " ", period)) |>
  rename(Period = period, Cells = cells, `% < 5` = pct_lt5) |>
  gt::gt() |>
  gt::tab_header(title = "Percent of country-indicator-years at each panel size",
                 subtitle = "Rows sum to 100 across the six bands.")
```

### Where the variation comes from

The era figures above treat time as the source of thin panels. It is not the only one.
Within the contemporary period, panel size varies across countries and across V-Dem's
surveys as well, and these are separable: a country-indicator-year's panel size depends on
who was recruited for that country, which survey the indicator belongs to, and how long the
year has had to accumulate back-coders.

V-Dem recruits at the **survey** level, not the indicator level. C-indicators are organised
into 16 topical surveys grouped into five clusters, and the methodology notes that experts
"code at least one cluster" and "most code one to two clusters" (methodology p. 16). So
indicator-level variation is largely survey-level variation, and the survey is the unit that
corresponds to an actual recruitment decision.

```{r variation-country, fig.width = 10, fig.height = 5.8}
library(ggrepel)

var_dat <- vdemdata::vdem |>
  select(country_text_id, year, any_of(paste0(analysis_indicators, "_nr"))) |>
  filter(year >= 2005, year <= 2024) |>
  pivot_longer(-c(country_text_id, year), names_to = "indicator", values_to = "n_coders") |>
  filter(!is.na(n_coders)) |>
  mutate(survey = substr(sub("_nr$", "", indicator), 1, 4))

country_names_var <- read_csv(here::here("data", "processed", "ert.csv"),
                              show_col_types = FALSE) |>
  distinct(country_text_id, country_name)

# Mean panel size and the share of cells under V-Dem's floor are related but not
# redundant (r = -0.54), so the ranking carries the level and the colour carries the tail.
country_var <- var_dat |>
  group_by(country_text_id) |>
  summarise(mean_n = mean(n_coders), pct_lt5 = mean(n_coders < 5) * 100, .groups = "drop") |>
  left_join(country_names_var, by = "country_text_id") |>
  mutate(country_name = coalesce(country_name, country_text_id)) |>
  arrange(mean_n) |>
  mutate(rank  = row_number(),
         label = if_else(rank <= 6 | rank > n() - 6, country_name, NA_character_),
         lt5_bin = cut(pct_lt5, c(-Inf, 1, 5, 10, 20, Inf),
                       labels = c("<1%", "1–5%", "5–10%", "10–20%", ">20%")))

lt5_pal <- c("<1%" = "#FDDBC7", "1–5%" = "#F4A582", "5–10%" = "#D55E00",
             "10–20%" = "#A03F00", ">20%" = "#6B2900")

ggplot(country_var, aes(rank, mean_n)) +
  geom_hline(yintercept = 5, linetype = "22", colour = "grey55") +
  annotate("text", x = nrow(country_var), y = 5,
           label = "V-Dem's recommended floor of five", hjust = 1, vjust = -0.7,
           size = 3, colour = "grey45") +
  geom_point(aes(colour = lt5_bin), size = 1.9) +
  geom_text_repel(aes(label = label), size = 3, colour = "grey20", na.rm = TRUE,
                  nudge_x = ifelse(country_var$rank <= 6,  18, -18),
                  nudge_y = ifelse(country_var$rank <= 6, 3.6, -3.6),
                  direction = "y", hjust = ifelse(country_var$rank <= 6, 0, 1),
                  segment.colour = "grey65", segment.size = 0.25,
                  segment.alpha = 0.8, min.segment.length = 0, max.overlaps = Inf,
                  box.padding = 0.5, point.padding = 0.3, seed = 1) +
  scale_colour_manual(values = lt5_pal, name = "Share of that country's\ncells below five",
                      guide = guide_legend(nrow = 1)) +
  scale_y_continuous(breaks = seq(6, 16, 2)) +
  # The label nudges expand the range downward; trim the empty band back off.
  coord_cartesian(ylim = c(4.8, max(country_var$mean_n) + 0.4)) +
  labs(title = "Mean panel size by country, 2005–2024",
       subtitle = paste(nrow(country_var),
                        "countries, ranked. Labels: the 6 thinnest and 6 thickest."),
       x = "Country (ranked)", y = "Mean coders per country-indicator-year") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        legend.position = "bottom", plot.title.position = "plot")
```

No country's *mean* falls below five — the United Arab Emirates sits lowest at 5.0, on the
line. Sub-floor cells are therefore within-country variation rather than a set of
permanently deficient countries, though the colouring shows the two are related: thin
countries carry a much larger share of sub-floor cells.

```{r variation-survey, fig.width = 11, fig.height = 5}
library(patchwork)

survey_labels <- c(
  v2el = "Elections",       v2ps = "Political Parties", v2ex = "Executive",
  v2lg = "Legislature",     v2dl = "Deliberation",      v2ju = "Judiciary",
  v2cl = "Civil Liberty",   v2sv = "Sovereignty",       v2st = "The State",
  v2cs = "Civil Society",   v2me = "Media",             v2pe = "Political Equality",
  v2ca = "Civic & Academic Space", v2sm = "Digital Society"
)

# A survey's first year matters: one introduced in 2000 has had fewer recruitment rounds
# to accumulate coders than one running from 1789, so youth and thinness are confounded.
survey_start <- vdemdata::vdem |>
  select(country_text_id, year, any_of(paste0(analysis_indicators, "_nr"))) |>
  pivot_longer(-c(country_text_id, year), names_to = "indicator", values_to = "n_coders") |>
  filter(!is.na(n_coders)) |>
  mutate(survey = substr(sub("_nr$", "", indicator), 1, 4)) |>
  group_by(survey) |>
  summarise(first_year = min(year), .groups = "drop")

survey_var <- var_dat |>
  group_by(survey) |>
  summarise(mean_n = mean(n_coders), n_ind = n_distinct(indicator), .groups = "drop") |>
  left_join(survey_start, by = "survey") |>
  mutate(name = paste0(survey_labels[survey], " (", n_ind, ")"),
         later_start = first_year > 1789) |>
  arrange(mean_n) |>
  mutate(name = factor(name, levels = name))

p_survey <- ggplot(survey_var, aes(mean_n, name)) +
  geom_vline(xintercept = 5, linetype = "22", colour = "grey55") +
  geom_segment(aes(x = 5, xend = mean_n, yend = name), colour = "grey75", linewidth = 0.5) +
  geom_point(aes(colour = later_start), size = 3) +
  geom_text(aes(label = sprintf("%.1f", mean_n)), hjust = -0.55, size = 2.9, colour = "grey25") +
  scale_colour_manual(values = c(`FALSE` = "#0072B2", `TRUE` = "#D55E00"), name = NULL,
                      labels = c(`FALSE` = "series starts 1789", `TRUE` = "series starts later")) +
  scale_x_continuous(limits = c(5, 12.2), breaks = seq(5, 12, 1)) +
  labs(title = "By survey", x = "Mean coders", y = NULL,
       subtitle = "Mean panel size, 2005–2024 (indicators in parentheses)") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank(),
        legend.position = "bottom", plot.title.position = "plot")

var_r2 <- tibble(
  source = c("Country", "Year", "Survey", "Indicator", "Country + indicator", "+ year"),
  r2 = c(r2(feols(n_coders ~ 1 | country_text_id, var_dat), "r2"),
         r2(feols(n_coders ~ 1 | year,            var_dat), "r2"),
         r2(feols(n_coders ~ 1 | survey,          var_dat), "r2"),
         r2(feols(n_coders ~ 1 | indicator,       var_dat), "r2"),
         r2(feols(n_coders ~ 1 | country_text_id + indicator, var_dat), "r2"),
         r2(feols(n_coders ~ 1 | country_text_id + indicator + year, var_dat), "r2")),
  kind = c(rep("one predictor", 4), rep("cumulative", 2))
) |> mutate(source = factor(source, levels = rev(source)))

p_var <- ggplot(var_r2, aes(r2, source, fill = kind)) +
  geom_col(width = 0.62) +
  geom_text(aes(label = sprintf("%.3f", r2)), hjust = -0.2, size = 2.9, colour = "grey25") +
  scale_fill_manual(values = c(`one predictor` = "#0072B2", cumulative = "grey70"), name = NULL) +
  scale_x_continuous(limits = c(0, 0.78), expand = expansion(mult = c(0, 0.02))) +
  labs(title = "What explains panel size?", x = "R²", y = NULL,
       subtitle = "Share of variance in coders per cell, 2005–2024") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank(),
        legend.position = "bottom", plot.title.position = "plot")

(p_survey | p_var) + plot_layout(widths = c(1.25, 1))
```

The two thinnest surveys are also the two whose series begin after 1789 — Digital Society
(from 2000) and Civic & Academic Space (from 1900), marked in vermillion. A survey
introduced later has had fewer recruitment rounds to accumulate coders, so its youth and
its thinness are not separable from this figure. Digital Society is also the largest survey
by indicator count, at 32.



```{r context-regime, fig.width = 8, fig.height = 4.5}
panel_full |>
  filter(!is.na(v2x_regime)) |>
  mutate(
    regime_label = factor(v2x_regime, 0:3,
                          labels = c("Closed\nautocracy", "Electoral\nautocracy",
                                     "Electoral\ndemocracy", "Liberal\ndemocracy"))
  ) |>
  group_by(year, regime_label) |>
  summarise(mean_n = mean(n_coders), .groups = "drop") |>
  ggplot(aes(x = year, y = mean_n, colour = regime_label)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = 2013, linetype = "dotted", colour = "grey40") +
  scale_colour_brewer(palette = "Set1", name = NULL) +
  labs(
    title    = "Mean panel size by year and regime type (1990–2024)",
    subtitle = "Autocracies have persistently thinner panels; all regime types thin toward the leading edge.",
    x = "Year", y = "Mean coders per country-year"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

## 4. Analysis 1: Score Volatility

Country-indicator pairs with chronically thin panels produce more variable scores over
time. This is the most direct illustration of measurement unreliability: even if a thin
panel is unbiased on average, its scores jump around more from year to year as a function
of small-sample noise rather than real political change.

**Design**: one row per country × indicator. Outcome = SD of the annual raw panel mean
across 1990–2024 (minimum 5 years required). Indicator FEs absorb indicator-specific
baseline variance; region FEs absorb regional patterns in political volatility.
M3 adds an interaction with the country's mean polyarchy (time-invariant, not absorbed
by country FEs in a cross-sectional model) to test whether the volatility effect is
concentrated in autocracies. M4 splits the panel-size predictor pre/post 2013 to ask
whether the post-2013 era specifically drives the result.

```{r vol-data}
#| cache: true

volatility_data <- panel_full |>
  group_by(country_text_id, indicator) |>
  summarise(
    sd_raw_mean      = sd(raw_mean, na.rm = TRUE),
    mean_n_coders    = mean(n_coders, na.rm = TRUE),
    mean_n_pre2013   = mean(n_coders[year <= 2013], na.rm = TRUE),
    mean_n_post2013  = mean(n_coders[year >  2013], na.rm = TRUE),
    mean_polyarchy_c = first(mean_polyarchy_c),
    region_modal     = first(region_modal),
    n_years          = n_distinct(year),
    .groups          = "drop"
  ) |>
  filter(n_years >= 5, !is.na(sd_raw_mean))

cat("Country-indicator pairs:", nrow(volatility_data), "\n")
cat("Median years per pair:", median(volatility_data$n_years), "\n")
```

```{r vol-regressions}
#| cache: true

m1_vol <- feols(sd_raw_mean ~ mean_n_coders,
                data = volatility_data, cluster = ~country_text_id)

m2_vol <- feols(sd_raw_mean ~ mean_n_coders + n_years | indicator + region_modal,
                data = volatility_data, cluster = ~country_text_id)

m3_vol <- feols(sd_raw_mean ~ mean_n_coders * mean_polyarchy_c + n_years |
                  indicator + region_modal,
                data = volatility_data, cluster = ~country_text_id)

m4_vol <- feols(sd_raw_mean ~ mean_n_pre2013 + mean_n_post2013 + n_years |
                  indicator + region_modal,
                data = volatility_data, cluster = ~country_text_id)
```

```{r vol-table}
coef_map_vol <- c(
  "mean_n_coders"              = "Mean panel size",
  "mean_n_pre2013"             = "Mean panel size (1990–2013)",
  "mean_n_post2013"            = "Mean panel size (2014–2024)",
  "mean_polyarchy_c"           = "Mean polyarchy (centered)",
  "mean_n_coders:mean_polyarchy_c" = "Panel size × Mean polyarchy",
  "n_years"                    = "Years of coverage"
)

modelsummary(
  list(M1 = m1_vol, M2 = m2_vol, M3 = m3_vol, M4 = m4_vol),
  coef_map  = coef_map_vol,
  statistic = "({std.error})",
  stars     = c("*" = 0.05, "**" = 0.01, "***" = 0.001),
  gof_map   = list(
    list(raw = "nobs",      clean = "N",  fmt = 0),
    list(raw = "r.squared", clean = "R²", fmt = 3)
  ),
  notes  = "OLS. M2–M4 absorb indicator and region FEs. SEs clustered by country.",
  title  = "Score volatility (SD of annual raw panel mean, 1990–2024) ~ panel size",
  output = "gt"
)
```

```{r vol-scatter, fig.width = 9, fig.height = 5}
volatility_data |>
  mutate(
    regime_group = case_when(
      mean_polyarchy_c < -0.15 ~ "Autocracy",
      mean_polyarchy_c <  0.15 ~ "Hybrid",
      TRUE                     ~ "Democracy"
    ),
    regime_group = factor(regime_group, levels = c("Autocracy", "Hybrid", "Democracy"))
  ) |>
  ggplot(aes(x = mean_n_coders, y = sd_raw_mean, colour = regime_group)) +
  geom_point(alpha = 0.15, size = 0.7) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9) +
  scale_colour_manual(
    values = c(Autocracy = "firebrick", Hybrid = "goldenrod", Democracy = "steelblue"),
    name   = NULL
  ) +
  labs(
    x        = "Mean panel size (distinct coders per year, 1990–2024)",
    y        = "SD of annual raw panel mean",
    title    = "Larger panels produce more stable scores",
    subtitle = "Each point = one country × indicator. Effect strongest in autocratic contexts."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

## 5. Analysis 2: Level Bias (TWFE)

If thin panels produce biased scores, the mechanism is regional reference frame
compression: panels dominated by coders from authoritarian contexts inflate raw means.
This produces a systematic level shift — thin-panel autocracies score higher than
same-level peers — distinct from the variance increase in Analysis 1.

**Design**: standard TWFE, one row per country × indicator × year. Country FEs absorb
time-invariant country characteristics. Year FEs absorb global annual shocks (real
democratic backsliding, V-Dem methodology changes). Indicator FEs absorb
indicator-level baselines. Identification comes from within-country variation in panel
size over time — primarily the post-2013 attrition.

`mean_polyarchy_c` is time-invariant (country mean over 1990–2024), so its main effect
is absorbed by the country FE; only its interaction with `n_coders` is estimated. This
interaction tests whether the within-country panel-size effect depends on a country's
typical democracy level. The `n_coders:post2013` interaction tests whether the panel-size
effect strengthened after 2013, without dropping year FEs (the interaction
varies cross-sectionally within years and is not collinear with year FEs).

```{r lev-regressions}
#| cache: true

m1_lev <- feols(
  raw_mean ~ n_coders | country_text_id + indicator + year,
  data = panel_full, cluster = ~country_text_id
)

m2_lev <- feols(
  raw_mean ~ n_coders + n_coders:mean_polyarchy_c | country_text_id + indicator + year,
  data = panel_full, cluster = ~country_text_id
)

m3_lev <- feols(
  raw_mean ~ n_coders + n_coders:post2013 | country_text_id + indicator + year,
  data = panel_full, cluster = ~country_text_id
)

m4_lev <- feols(
  raw_mean ~ n_coders + n_coders:mean_polyarchy_c + n_coders:post2013 |
    country_text_id + indicator + year,
  data = panel_full, cluster = ~country_text_id
)
```

```{r lev-table}
coef_map_lev <- c(
  "n_coders"                = "Panel size",
  "n_coders:mean_polyarchy_c" = "Panel size × Mean polyarchy",
  "n_coders:post2013"       = "Panel size × Post-2013"
)

modelsummary(
  list(M1 = m1_lev, M2 = m2_lev, M3 = m3_lev, M4 = m4_lev),
  coef_map  = coef_map_lev,
  statistic = "({std.error})",
  stars     = c("*" = 0.05, "**" = 0.01, "***" = 0.001),
  gof_map   = list(
    list(raw = "nobs",       clean = "N",        fmt = 0),
    list(raw = "r.squared",  clean = "R²",        fmt = 3),
    list(raw = "r2.within",  clean = "Within R²", fmt = 3)
  ),
  notes  = "TWFE: country + indicator + year FEs absorbed. SEs clustered by country.",
  title  = "Level bias: raw_mean ~ panel size, 1990–2024",
  output = "gt"
)
```

```{r lev-interaction-plot, fig.width = 8, fig.height = 5}
# Effective slope of n_coders as a function of country mean polyarchy (from M2)
poly_seq   <- seq(-0.5, 0.5, by = 0.01)
slope_seq  <- coef(m2_lev)["n_coders"] +
              coef(m2_lev)["n_coders:mean_polyarchy_c"] * poly_seq
slope_lo   <- (coef(m2_lev)["n_coders"] - 1.96 * se(m2_lev)["n_coders"]) +
              (coef(m2_lev)["n_coders:mean_polyarchy_c"] -
               1.96 * se(m2_lev)["n_coders:mean_polyarchy_c"]) * poly_seq
slope_hi   <- (coef(m2_lev)["n_coders"] + 1.96 * se(m2_lev)["n_coders"]) +
              (coef(m2_lev)["n_coders:mean_polyarchy_c"] +
               1.96 * se(m2_lev)["n_coders:mean_polyarchy_c"]) * poly_seq

ggplot(
  data.frame(poly = poly_seq + poly_global_mean, slope = slope_seq,
             lo = slope_lo, hi = slope_hi),
  aes(x = poly, y = slope)
) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, fill = "firebrick") +
  geom_line(colour = "firebrick", linewidth = 1.2) +
  labs(
    x = "Country mean polyarchy (v2x_polyarchy, 1990–2024 average)",
    y = "Within-country effect of one additional coder on raw_mean",
    title = "Thin panels push autocratic scores toward the floor",
    subtitle = paste0(
      "More coders raise raw means in autocracies and lower them in democracies, ",
      "compressing scores toward the global center. Thinning has the opposite effect."
    )
  ) +
  theme_minimal(base_size = 12)
```

---

## 6. Analysis 3: Generalized Afghanistan (First Differences)

The level-bias result is consistent with a compositional confound: autocratic countries
may systematically attract less-calibrated coders regardless of panel size, producing an
omitted-variables bias even within TWFE. The first-difference design eliminates this by
identifying from *changes* in composition within the same country-indicator over time.

The direct claim: year-over-year changes in panel calibration quality (`Δ mean anchor MAD`)
and bridge coder fraction (`Δ bridge_frac`) predict year-over-year score changes, net of
indicator-level change rates (indicator FEs) and global annual shocks (year FEs). Country
FEs drop out through differencing. Only consecutive year pairs are used.

The regime interaction in M3 tests whether the composition effect is concentrated in
autocracies — the direct generalisation of the Afghanistan mechanism.

```{r fd-data}
#| cache: true

fd_data <- panel_full |>
  filter(!is.na(mean_anchor_mad)) |>
  arrange(country_text_id, indicator, year) |>
  group_by(country_text_id, indicator) |>
  mutate(
    d_raw_mean    = raw_mean        - lag(raw_mean),
    d_n_coders    = n_coders        - lag(n_coders),
    d_anchor_mad  = mean_anchor_mad - lag(mean_anchor_mad),
    d_bridge_frac = bridge_frac     - lag(bridge_frac),
    regime_lag    = lag(v2x_regime),
    consec        = (year - lag(year)) == 1L
  ) |>
  ungroup() |>
  filter(!is.na(d_raw_mean), consec)

cat("First-difference observations:", nrow(fd_data), "\n")
cat("Countries:", n_distinct(fd_data$country_text_id), "\n")
cat("Years covered:", range(fd_data$year), "\n")
```

```{r fd-regressions}
#| cache: true

m1_fd <- feols(
  d_raw_mean ~ d_n_coders | indicator + year,
  data = fd_data, cluster = ~country_text_id
)

m2_fd <- feols(
  d_raw_mean ~ d_n_coders + d_anchor_mad + d_bridge_frac | indicator + year,
  data = fd_data, cluster = ~country_text_id
)

m3_fd <- feols(
  d_raw_mean ~ d_n_coders   * regime_lag +
               d_anchor_mad  * regime_lag +
               d_bridge_frac * regime_lag | indicator + year,
  data = fd_data, cluster = ~country_text_id
)
```

```{r fd-table}
coef_map_fd <- c(
  "d_n_coders"                   = "Δ Panel size",
  "d_anchor_mad"                 = "Δ Mean anchor MAD",
  "d_bridge_frac"                = "Δ Bridge coder fraction",
  "regime_lag"                   = "Regime type t−1 (0–3)",
  "d_n_coders:regime_lag"        = "Δ Panel size × Regime",
  "d_anchor_mad:regime_lag"      = "Δ Anchor MAD × Regime",
  "d_bridge_frac:regime_lag"     = "Δ Bridge frac × Regime"
)

modelsummary(
  list(M1 = m1_fd, M2 = m2_fd, M3 = m3_fd),
  coef_map  = coef_map_fd,
  statistic = "({std.error})",
  stars     = c("*" = 0.05, "**" = 0.01, "***" = 0.001),
  gof_map   = list(
    list(raw = "nobs",       clean = "N",        fmt = 0),
    list(raw = "r.squared",  clean = "R²",        fmt = 3),
    list(raw = "r2.within",  clean = "Within R²", fmt = 3)
  ),
  notes  = paste0(
    "First differences. Country FEs removed by differencing. ",
    "Indicator + year FEs absorbed. Consecutive year pairs only. ",
    "SEs clustered by country."
  ),
  title  = "Generalized Afghanistan: composition changes predict score changes (1990–2024)",
  output = "gt"
)
```

```{r fd-coef-plot, fig.width = 9, fig.height = 5.5}
bind_rows(
  broom::tidy(m2_fd, conf.int = TRUE) |> mutate(model = "M2: Composition"),
  broom::tidy(m3_fd, conf.int = TRUE) |> mutate(model = "M3: × Regime")
) |>
  filter(term %in% names(coef_map_fd)) |>
  mutate(term = coef_map_fd[term],
         term = factor(term, levels = rev(coef_map_fd))) |>
  ggplot(aes(x = estimate, y = term, colour = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.25, linewidth = 0.6,
                 position = position_dodge(0.5)) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  scale_colour_manual(
    values = c("M2: Composition" = "steelblue", "M3: × Regime" = "firebrick"),
    name   = NULL
  ) +
  labs(
    x        = "Coefficient (95% CI, SEs clustered by country)",
    y        = NULL,
    title    = "Composition changes drive score changes — concentrated in autocracies",
    subtitle = "First-difference estimates. Indicator + year FEs. Consecutive year pairs only."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r fd-country-scan}
#| cache: true

# For each country: R² from regressing d_raw_mean on d_anchor_mad,
# pooled across indicators and years. Countries where R² is highest are
# those where composition changes most reliably predict score changes —
# generalisations of the Afghanistan pattern.
country_r2 <- fd_data |>
  filter(!is.na(d_anchor_mad)) |>
  group_by(country_text_id) |>
  filter(n() >= 20) |>
  group_modify(~ {
    m <- lm(d_raw_mean ~ d_anchor_mad, data = .x)
    tibble(
      r2          = summary(m)$r.squared,
      coef_mad    = coef(m)["d_anchor_mad"],
      n_obs       = nrow(.x),
      mean_n      = round(mean(.x$n_coders, na.rm = TRUE), 1),
      mean_regime = round(mean(.x$regime_lag, na.rm = TRUE), 2)
    )
  }) |>
  ungroup() |>
  arrange(desc(r2))

cat("Countries with ≥ 20 FD observations:", nrow(country_r2), "\n")
cat("Median R²:", round(median(country_r2$r2), 3), "\n")
cat("Share with positive coefficient:", round(mean(country_r2$coef_mad > 0), 2), "\n")
```

```{r fd-country-table}
country_r2 |>
  slice_head(n = 20) |>
  mutate(across(c(r2, coef_mad), ~ round(., 3))) |>
  select(country_text_id, r2, coef_mad, n_obs, mean_n, mean_regime) |>
  rename(
    Country             = country_text_id,
    `R²`                = r2,
    `Coef (Δ MAD)`      = coef_mad,
    `N obs`             = n_obs,
    `Mean panel size`   = mean_n,
    `Mean regime (0–3)` = mean_regime
  ) |>
  knitr::kable(
    caption = paste0(
      "Top 20 countries where Δ mean anchor MAD explains Δ raw_mean. ",
      "R² from bivariate regression of Δraw_mean ~ Δanchor_MAD, pooled across ",
      "all indicators and consecutive year pairs. ",
      "Positive coefficient = less-calibrated panel → higher scores (floor inflation). ",
      "Mean regime: 0 = closed autocracy, 3 = liberal democracy."
    )
  )
```

---

## 7. Analysis 4: Panel Size and IRT Posterior Uncertainty

The V-Dem codebook documents a `_nr` suffix (number of coders per country-variable-year)
and a `_sd` suffix (posterior standard deviation from the measurement model) for every
C-type indicator. These are available directly via `vdemdata::vdem`, so no coder-level
data is needed here. The codebook notes explicitly: *"We strongly advise against using
observations based on three or fewer coders."*

```{r uncertainty-data}
nr_cols <- paste0(analysis_indicators, "_nr")
sd_cols <- paste0(analysis_indicators, "_sd")

vdem_raw <- vdemdata::vdem |>
  select(country_text_id, year, any_of(c(nr_cols, sd_cols))) |>
  filter(year >= 1990, year <= 2024)

unc_long <- vdem_raw |>
  pivot_longer(
    cols          = -c(country_text_id, year),
    names_to      = c("indicator", ".value"),
    names_pattern = "^(.+)_(nr|sd)$"
  ) |>
  rename(n_coders = nr, posterior_sd = sd) |>
  filter(!is.na(n_coders), !is.na(posterior_sd), n_coders >= 1,
         indicator %in% analysis_indicators)

cat("Country-indicator-years:", nrow(unc_long), "\n")
cat("Panel size range:", paste(range(unc_long$n_coders), collapse = "–"), "\n")
cat("Posterior SD range:", paste(round(range(unc_long$posterior_sd), 3), collapse = "–"), "\n")
```

```{r uncertainty-binned-plot}
unc_binned <- unc_long |>
  mutate(n_bin = case_when(
    n_coders <= 3  ~ "1–3\n(V-Dem\nwarns)",
    n_coders <= 5  ~ "4–5",
    n_coders <= 8  ~ "6–8",
    n_coders <= 12 ~ "9–12",
    TRUE           ~ "13+"
  ),
  n_bin = factor(n_bin, levels = c("1–3\n(V-Dem\nwarns)", "4–5", "6–8", "9–12", "13+"))) |>
  group_by(n_bin) |>
  summarise(
    median_sd = median(posterior_sd),
    q25_sd    = quantile(posterior_sd, 0.25),
    q75_sd    = quantile(posterior_sd, 0.75),
    n         = n(),
    .groups   = "drop"
  )

ggplot(unc_binned, aes(x = n_bin, y = median_sd)) +
  geom_col(fill = "steelblue", alpha = 0.75, width = 0.6) +
  geom_errorbar(aes(ymin = q25_sd, ymax = q75_sd), width = 0.25) +
  labs(
    title    = "Fewer coders → wider IRT posterior intervals",
    subtitle = paste0("Median posterior SD by panel size bin. Error bars = IQR. Pooled across ", length(analysis_indicators), " scored indicators, 1990\u20132024."),
    x        = "Panel size (number of coders)",
    y        = "Posterior SD (V-Dem measurement model)"
  ) +
  theme_minimal(base_size = 12)
```

The bars above are anchored at zero, so most of the panel is empty space, and their
whiskers are the IQR — the spread of individual country-indicator-years within each bin,
not uncertainty about the bin's median. Those are different quantities and they differ in
width by more than an order of magnitude. The two panels below put each on its own
median-centred axis.

```{r uncertainty-dotwhisker, fig.width = 10, fig.height = 4.4}
library(patchwork)

unc_bin_iv <- unc_long |>
  mutate(n_bin = case_when(
    n_coders <= 3  ~ "1–3",
    n_coders <= 5  ~ "4–5",
    n_coders <= 8  ~ "6–8",
    n_coders <= 12 ~ "9–12",
    TRUE           ~ "13+"
  ),
  n_bin = factor(n_bin, levels = c("1–3", "4–5", "6–8", "9–12", "13+"))) |>
  group_by(n_bin) |>
  summarise(
    median_sd = median(posterior_sd),
    q25_sd    = quantile(posterior_sd, 0.25),
    q75_sd    = quantile(posterior_sd, 0.75),
    n         = n(),
    # Asymptotic SE of a median for a large sample: 1.253 * sd / sqrt(n).
    se_med    = 1.253 * sd(posterior_sd) / sqrt(n()),
    .groups   = "drop"
  ) |>
  mutate(ci_lo = median_sd - 1.96 * se_med,
         ci_hi = median_sd + 1.96 * se_med)

BLUE <- "#0072B2"   # Okabe-Ito blue, the repo's figure convention

p_ci <- ggplot(unc_bin_iv, aes(x = n_bin, y = median_sd)) +
  geom_linerange(aes(ymin = ci_lo, ymax = ci_hi), linewidth = 1.2, colour = BLUE) +
  geom_point(size = 3, colour = BLUE) +
  geom_text(aes(label = sprintf("%.3f", median_sd)),
            hjust = -0.4, size = 3.1, colour = "grey25") +
  scale_y_continuous(expand = expansion(mult = 0.13)) +
  labs(title = "Median, with 95% CI",
       subtitle = "Do the bins differ? Every interval is narrower than its own point.",
       x = NULL, y = "Posterior SD") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(), plot.title.position = "plot")

p_iqr <- ggplot(unc_bin_iv, aes(x = n_bin, y = median_sd)) +
  geom_linerange(aes(ymin = q25_sd, ymax = q75_sd),
                 linewidth = 1.2, colour = BLUE, alpha = 0.45) +
  geom_point(size = 3, colour = BLUE) +
  scale_y_continuous(expand = expansion(mult = 0.06)) +
  labs(title = "Median, with IQR",
       subtitle = "How much do cells vary? More than the bin medians span.",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(), plot.title.position = "plot")

(p_ci | p_iqr) +
  plot_annotation(
    title    = "Panel size shifts the median reliably; individual cells barely sort",
    subtitle = "Panel size (coders) on the x-axis. Pooled across ~150 working indicators, 1990–2024.",
    caption  = paste("Left: 95% CI on each median. Right: interquartile range of the underlying",
                     "country-indicator-years.\nNote the different y-axis ranges: the CIs are 14-58x",
                     "narrower than the IQRs, so a shared scale would hide them entirely."),
    theme    = theme(plot.title = element_text(size = 13, face = "bold"))
  )
```

```{r uncertainty-dotwhisker-table}
unc_bin_iv |>
  transmute(
    `Panel size` = n_bin, n,
    Median  = round(median_sd, 4),
    `CI low` = round(ci_lo, 4), `CI high` = round(ci_hi, 4),
    `CI width`  = round(ci_hi - ci_lo, 4),
    `IQR width` = round(q75_sd - q25_sd, 4),
    `IQR / CI`  = round((q75_sd - q25_sd) / (ci_hi - ci_lo))
  ) |>
  gt::gt() |>
  gt::tab_header(title = "Bin medians: estimation uncertainty vs cell-level spread")
```


```{r uncertainty-time-plot}
unc_year <- unc_long |>
  group_by(year) |>
  summarise(
    mean_sd  = mean(posterior_sd, na.rm = TRUE),
    mean_nr  = mean(n_coders,     na.rm = TRUE),
    .groups  = "drop"
  )

ggplot(unc_year, aes(x = year, y = mean_sd)) +
  geom_line(colour = "steelblue", linewidth = 0.9) +
  geom_smooth(method = "loess", span = 0.35, colour = "darkred", se = FALSE) +
  geom_vline(xintercept = 2013, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 2014, y = max(unc_year$mean_sd) * 0.98,
           label = "Original coding round ends", hjust = 0, size = 3.5, colour = "grey30") +
  labs(
    title    = "IRT uncertainty is highest in the years still accumulating coders",
    subtitle = "Mean posterior SD across all country-indicator-years, 1990–2024.",
    x        = NULL,
    y        = "Mean posterior SD"
  ) +
  theme_minimal(base_size = 12)
```

```{r uncertainty-regression}
m_unc1 <- feols(posterior_sd ~ n_coders,
                data = unc_long, cluster = ~country_text_id)
m_unc2 <- feols(posterior_sd ~ n_coders | indicator,
                data = unc_long, cluster = ~country_text_id)
m_unc3 <- feols(posterior_sd ~ n_coders | indicator + country_text_id,
                data = unc_long, cluster = ~country_text_id)
m_unc4 <- feols(posterior_sd ~ n_coders | indicator + country_text_id + year,
                data = unc_long, cluster = ~country_text_id)

modelsummary(
  list(M1 = m_unc1, M2 = m_unc2, M3 = m_unc3, M4 = m_unc4),
  stars      = TRUE,
  coef_rename = c(n_coders = "Panel size"),
  gof_map    = c("nobs", "r.squared", "r2.within"),
  notes      = "OLS. M2 absorbs indicator FE. M3 adds country FE. M4 adds year FE. SEs clustered by country.",
  output     = "gt",
  title      = "Panel size and IRT posterior uncertainty (posterior SD ~ panel size)"
)
```

### Marginal value of an added coder

The pooled coefficient above is a single slope, but the relationship is not a straight
line. The two figures below trace posterior SD at each integer panel size and the change
contributed by each additional coder. The first uses raw conditional means — what a data
user encounters. The second re-estimates the same quantities as panel-size dummies inside
the full fixed-effects specification, netting out the indicator, country and era
differences the raw means absorb. Both use the same estimator and the same
country-clustered SEs, so the only thing separating them is the fixed effects.

Restricted to 2005 onward. Before 2005 the contemporary panel does not operate: those
country-years are covered by the retrospective historical programme, where one or two
specialists coded a country's whole political history at once. Pooling the two puts ~80%
of all 1-3 coder cells into the figure's left end, where they read as solid anchors while
actually describing a different coding process.

```{r marginal-value-setup}
NRMAX   <- 15
MV_FROM <- 2005   # contemporary panel only; see note above

mv_dat <- unc_long |>
  filter(year >= MV_FROM, n_coders >= 1, n_coders <= NRMAX) |>
  mutate(nrf = factor(n_coders))

# Both curves come from the same estimator and the same country-clustered SEs, so the
# ONLY difference between the two figures is whether the fixed effects are absorbed.
# (An unclustered SE on a raw cell mean would be ~15x too narrow -- it treats every
# country-indicator-year as independent, which they plainly are not.)
mv_raw_fit <- feols(posterior_sd ~ 0 + nrf,
                    data = mv_dat, cluster = ~country_text_id)
mv_adj_fit <- feols(posterior_sd ~ i(nrf, ref = "5") | indicator + country_text_id + year,
                    data = mv_dat, cluster = ~country_text_id)

# Levels. The raw fit gives group means directly; the adjusted fit gives contrasts
# against n = 5, re-anchored onto the raw mean at n = 5 so both sit on one scale.
mv_anchor <- unname(coef(mv_raw_fit)[["nrf5"]])

mv_raw_lvl <- tibble(
  n_coders = as.integer(sub("^nrf", "", names(coef(mv_raw_fit)))),
  est      = unname(coef(mv_raw_fit)),
  se       = unname(sqrt(diag(vcov(mv_raw_fit))))
) |> arrange(n_coders)

mv_adj_cf <- coef(mv_adj_fit); mv_adj_V <- vcov(mv_adj_fit)
mv_adj_lv <- as.integer(sub(".*::", "", names(mv_adj_cf)))
mv_adj_lvl <- tibble(n_coders = c(mv_adj_lv, 5L),
                     est = c(unname(mv_adj_cf) + mv_anchor, mv_anchor),
                     se  = c(unname(sqrt(diag(mv_adj_V))), 0)) |>
  arrange(n_coders)

# Marginal change per added coder: a difference of two coefficients, so its SE needs the
# covariance between them, not a sum of squares.
mv_marginal <- function(cf, V, levels_vec, ref = NA_integer_) {
  idx <- setNames(seq_along(cf), levels_vec)
  bind_rows(lapply(2:NRMAX, function(k) {
    gk <- function(j) if (!is.na(ref) && j == ref) NULL else idx[[as.character(j)]]
    a <- gk(k); b <- gk(k - 1)
    d <- (if (is.null(a)) 0 else cf[a]) - (if (is.null(b)) 0 else cf[b])
    v <- (if (is.null(a)) 0 else V[a, a]) + (if (is.null(b)) 0 else V[b, b]) -
         2 * (if (is.null(a) || is.null(b)) 0 else V[a, b])
    tibble(n_coders = k, d = unname(d), se = sqrt(max(v, 0)))
  }))
}

mv_marg_raw <- mv_marginal(coef(mv_raw_fit), vcov(mv_raw_fit),
                           as.integer(sub("^nrf", "", names(coef(mv_raw_fit)))))
mv_marg_adj <- mv_marginal(mv_adj_cf, mv_adj_V, mv_adj_lv, ref = 5L)

MV_YL <- range(c(mv_raw_lvl$est - 1.96 * mv_raw_lvl$se, mv_raw_lvl$est + 1.96 * mv_raw_lvl$se,
                 mv_adj_lvl$est - 1.96 * mv_adj_lvl$se, mv_adj_lvl$est + 1.96 * mv_adj_lvl$se))
MV_YM <- range(c(mv_marg_raw$d - 1.96 * mv_marg_raw$se, mv_marg_raw$d + 1.96 * mv_marg_raw$se,
                 mv_marg_adj$d - 1.96 * mv_marg_adj$se, mv_marg_adj$d + 1.96 * mv_marg_adj$se))

# Sign convention is kept as-is: this is a change in SD, the same quantity and sign a
# regression coefficient would report. Plotting "reduction" instead would make the axis
# positive-is-better but would no longer be the number in the table. Labelled instead.
mv_figure <- function(lvl, marg, colour, tag) {
  bands <- list(
    annotate("rect", xmin = 0.5, xmax = 3.5, ymin = -Inf, ymax = Inf,
             fill = "grey70", alpha = 0.16),
    annotate("segment", x = 5, xend = 5, y = -Inf, yend = Inf,
             linetype = "22", colour = "grey45", linewidth = 0.4)
  )

  p1 <- ggplot(lvl, aes(n_coders, est)) + bands +
    geom_ribbon(aes(ymin = est - 1.96 * se, ymax = est + 1.96 * se),
                fill = colour, alpha = 0.30) +
    geom_line(colour = colour, linewidth = 0.5) +
    geom_point(colour = colour, size = 1.5) +
    scale_x_continuous(breaks = seq(1, NRMAX, 2)) +
    coord_cartesian(ylim = MV_YL) +
    labs(title = "Level — how uncertain is a panel of this size?",
         subtitle = "Lower is less uncertainty",
         x = "Panel size (coders)", y = "Posterior SD") +
    theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank(), plot.title.position = "plot")

  p2 <- ggplot(marg, aes(n_coders, d)) + bands +
    geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.5) +
    annotate("text", x = NRMAX, y = 0, hjust = 1, vjust = -0.8, size = 2.9,
             colour = "grey35", label = "zero = the added coder changed nothing") +
    geom_linerange(aes(ymin = d - 1.96 * se, ymax = d + 1.96 * se),
                   colour = colour, alpha = 0.55, linewidth = 0.9) +
    geom_point(colour = colour, size = 2.1) +
    scale_x_continuous(breaks = seq(2, NRMAX, 2)) +
    coord_cartesian(ylim = MV_YM) +
    labs(title = "Marginal — what did the next coder buy?",
         subtitle = "Change in SD from adding one coder. More negative = bought more.",
         x = "Panel size after adding (coders)", y = "Δ Posterior SD") +
    theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank(), plot.title.position = "plot")

  (p1 | p2) + plot_annotation(
    title = tag,
    subtitle = paste0("Contemporary panel only (", MV_FROM, "-2024). Intervals 95%, ",
                      "clustered by country.\nShaded = V-Dem's \"three or fewer\" warning ",
                      "zone; dashed line = its recommended floor of five."),
    theme = theme(plot.title = element_text(size = 13, face = "bold")))
}
```

#### Note: why the marginal panel's intervals are not read off the level panel

The marginal panel's point estimates are exactly the first differences of the level
panel's — a deterministic transformation carrying no new information. Its *intervals* are
a different matter, and they cannot be recovered by eye from the level ribbons.

The variance of a difference has three terms, not two:

$$
\operatorname{Var}(\beta_k - \beta_{k-1})
  = \operatorname{Var}(\beta_k) + \operatorname{Var}(\beta_{k-1})
  - 2\operatorname{Cov}(\beta_k,\, \beta_{k-1})
$$

In code this is the full variance-covariance matrix rather than just its diagonal —
`V[a, a] + V[b, b] - 2 * V[a, b]`, as in `mv_marginal()` above. Eyeballing two ribbons on
the level panel implicitly drops the covariance term, because a level plot shows each
estimate's own uncertainty and nothing about how the two move together.

Here the adjacent level estimates are strongly positively correlated — they are
coefficients from one regression, measured against the same reference category and sharing
the same absorbed country, indicator and year effects, so anything shifting the common
baseline moves them together. That shared component cancels in the difference, and the
covariance term credits the cancellation.

In prose: it is the difference between measuring where two knots on a swaying rope are,
and measuring the gap between them. Each position is uncertain because the rope moves, but
the gap is steady, because the knots sway together. Treating the positions as independent
would report the gap as far less certain than it is.

Worked example, the eighth coder:

| term | value |
|---|---|
| $\operatorname{Var}(\beta_8)$ | 1.492e-05 (SE 0.0039) |
| $\operatorname{Var}(\beta_7)$ | 1.188e-05 (SE 0.0034) |
| $\operatorname{Cov}(\beta_8, \beta_7)$ | 1.013e-05 (correlation 0.761) |
| naive SE, covariance dropped | 0.0052 |
| correct SE | **0.0026** |

The covariance is nearly as large as either variance, so subtracting twice it removes about
three-quarters of the total. Across the range the naive SE averages 1.6x the correct one,
and close to 2x for every panel size above six.

Two sanity checks on the sign and the mechanism. Positive covariance *shrinks* a
difference's SE; negative covariance would inflate it past the naive value. And at
n = 5 and n = 6 the two SEs coincide exactly, because n = 5 is the reference category —
$\beta_5$ is fixed at zero with no variance, so there is no covariance term to subtract.
Those are the only two rows where naive and correct agree, which is a useful check that
the code is indexing the right matrix entries.

```{r marginal-value-raw, fig.width = 10, fig.height = 4.6}
mv_figure(mv_raw_lvl, mv_marg_raw, "#0072B2",
          "Raw conditional means: what a data user encounters")
```

```{r marginal-value-adjusted, fig.width = 10, fig.height = 4.6}
mv_figure(mv_adj_lvl, mv_marg_adj, "#D55E00",
          "Adjusted: panel-size dummies within indicator, country and year fixed effects")
```

```{r marginal-value-table}
mv_n <- mv_dat |> count(n_coders, name = "n")

full_join(
  mv_marg_raw |> transmute(n_coders, `Raw Δ` = round(d, 4), `Raw SE` = round(se, 4)),
  mv_marg_adj |> transmute(n_coders, `Adjusted Δ` = round(d, 4), `Adjusted SE` = round(se, 4)),
  by = "n_coders"
) |>
  left_join(mv_n, by = "n_coders") |>
  rename(`Panel size` = n_coders, `N cells` = n) |>
  relocate(`N cells`, .after = `Panel size`) |>
  gt::gt() |>
  gt::tab_header(
    title = "Change in posterior SD per added coder",
    subtitle = paste("Contemporary panel only (2005-2024). Both columns use",
                     "country-clustered SEs and differ only in whether fixed effects are",
                     "absorbed. Negative = the added coder reduced uncertainty.",
                     "The adjusted series is anchored at n = 5.")
  )
```

#### The published thresholds against the marginal series

V-Dem publishes two thresholds, both in the v15 codebook's `_nr` entry at p. 35 (PDF
sheet 36):

> We strongly advise against using observations based on three or fewer coders.

> ...we suggest that users primarily base analyses on observations based on five or more
> coders.

The first is a usage instruction; the Cautionary Notes attribute it to internal analysis
that the published documents do not reproduce. The second is a recruitment quota the
methodology describes as a staffing target (p. 13, p. 16) and the codebook calls an
assumption. Analysis 14 §7 documents the provenance.

The marginal series above is the corresponding quantity in the data: what each additional
coder actually buys. Grouping it around the two published thresholds:

```{r threshold-gap}
mv_marg_adj |>
  mutate(zone = case_when(
    n_coders <= 3 ~ "2-3  (V-Dem: advise against)",
    n_coders == 4 ~ "4    (no guidance given)",
    n_coders <= 8 ~ "5-8  (at/above recommended floor)",
    TRUE          ~ "9-15 (well above floor)"
  )) |>
  group_by(zone) |>
  summarise(
    `Coder positions` = n(),
    `Mean Δ SD`   = round(mean(d), 4),
    `Largest Δ`   = round(min(d), 4),
    `Smallest Δ`  = round(max(d), 4),
    .groups = "drop"
  ) |>
  gt::gt() |>
  gt::tab_header(
    title = "What each added coder buys, grouped by V-Dem's published thresholds",
    subtitle = "Adjusted series, contemporary panel. More negative = larger reduction in posterior SD."
  )
```

Three features of the adjusted series, read off the table above.

**The low end is not monotone.** The second coder's contribution is null
(−0.011, CI −0.068 to +0.047), the third buys −0.042 (CI −0.077 to −0.007), and the
fourth is null again and slightly positive (+0.005, CI −0.006 to +0.016).

**Every coder from the fifth through the fourteenth excludes zero**, beginning at −0.028
(fifth) and −0.028 (sixth) and declining smoothly to −0.009 by the fourteenth. The
fifteenth is exactly zero (CI −0.008 to +0.008). Cumulatively, going from four coders to
nine buys 0.114; going from one to four buys 0.047.

**The series does not step down at the recommended floor.** The fifth and sixth coders buy
as much as any coder below them, and the decline toward zero runs from roughly the tenth
onward rather than from the fifth.

Two caveats before any of this is read as a dose-response. The n = 2 and n = 3 estimates
rest on 616 and 3,038 cells against 85,581 at n = 6, and those cells concentrate in a small
number of countries, so country-clustered intervals there are wide. More importantly the
left end is *selected*, not merely noisy: contemporary cells with four or fewer coders are
disproportionately leading-edge years and hard-to-recruit countries. The fixed effects
absorb country, indicator and year main effects but not their interaction, and "this
country in this year lost a coder" is exactly an interaction — so the 3–4–5 wobble, which
sits where N jumps from 3,038 to 19,463, is better read as composition than as a property
of panel size. Finally, the shape is a property of this outcome — posterior SD — not of
point estimates, which Analyses 1-3 found panel size does not move.

<!-- TODO (EJT): interpretation, and how it lands against the two published thresholds.
Note that neither threshold was derived from this quantity: the ≤3 warning is a usage
instruction whose stated empirical basis is unpublished, and five is a recruitment quota
the methodology describes as a staffing target (p. 13, p. 16). Analysis 14 §7 has the
provenance. -->


---

## 8. Conclusion

The four analyses tell a consistent but hierarchically ordered story about what panel
degradation does to V-Dem measurement.

**Analyses 1–3 are honest nulls.** Score volatility (Analysis 1) is driven almost
entirely by indicator-level differences — which indicators are inherently noisier — not
by panel size. The TWFE level-bias analysis (Analysis 2) finds a statistically
significant but small interaction: additional coders compress raw means toward the
global center, raising scores in autocracies and lowering them in democracies, with an
effect that is regime-conditional and robust but tiny in explained variance (within
R² = 0.002). The first-difference analysis (Analysis 3) is broadly null until regime
type is conditioned on, at which point composition changes — particularly bridge coder
entry — produce marginally significant score shifts in autocracies. None of these
results constitutes a crisis. Panel degradation does not produce dramatic or consistent
bias in V-Dem's published point estimates.

**Analysis 4 is where the consequence is measurable and direct.** V-Dem's own
posterior standard deviations — the uncertainty the measurement model places around
each country-indicator-year estimate — are strongly and robustly predicted by panel
size. Each additional coder reduces posterior SD by 0.012–0.013 units (p < 0.001),
and this effect survives indicator, country, and year fixed effects (within R² = 0.046
in the most saturated specification). The effect is structural: more observations give
the IRT model more information, producing tighter posteriors. This is not a finding
that could go the other way.

The time series makes the practical consequence visible. Mean posterior SD fell
continuously from 1990 to a minimum around 2008–2010 — the peak of the panel-size
pyramid — then rose again through the years the current release reports most recently.
By 2024, average IRT uncertainty sits approximately 0.05 SD units, or roughly 12%, above
the 2008–2010 minimum, and the regression coefficient almost exactly accounts for it: the
~4–5 coder difference in mean panel size implies a +0.05–0.06 increase at the estimated
−0.012 per coder.

What that rise is *not* is attrition. §3 and Analysis 14 establish that the active roster
has been roughly flat since 2020 and that each year arrives about equally thin before
filling in over subsequent releases. The recent years therefore carry wider posteriors
because their panels have not finished accumulating, not because coders left — which makes
the gap a permanent property of the leading edge rather than a trend, and one that the
most recent data, the data users most want, will always sit inside.

The argument for panel augmentation therefore rests on uncertainty, not bias. V-Dem
expert panels were built to provide the IRT model with sufficient information to
estimate latent democracy scores with reasonable precision. At the leading edge that
information has not yet arrived, in a way that is measurable and that recurs with every
release. The replacement experiment tests whether AI coders can supply effective panel
size where the human panel has not filled in — using V-Dem's own posterior SD as the
direct measure of whether the augmentation is working.

---

## Appendix: combined context figure (experiment)

An experiment in consolidation: the two context figures from §3 — panel size over time,
and panel size across countries — as one two-panel figure. Both measure the same quantity,
coders per country-indicator-year, so they share a y-axis; only the thing being varied
differs. Panel A varies time, panel B varies country.

The originals remain in §3. This is here to judge whether the combined version can replace
them.

```{r combined-context, fig.width = 13, fig.height = 5.6}
ctx_ylim <- range(country_var$mean_n)

# Regime comes from vdem directly rather than panel_full, so this panel uses the same
# 205-indicator analytic set as panel B. Warm-to-cool ordering follows the regime scale
# (closed autocracy -> liberal democracy); validated, worst normal-vision dE 15.6.
regime_ctx <- vdemdata::vdem |>
  select(country_text_id, year, v2x_regime,
         any_of(paste0(analysis_indicators, "_nr"))) |>
  filter(year >= 1990, year <= 2024, !is.na(v2x_regime)) |>
  pivot_longer(-c(country_text_id, year, v2x_regime),
               names_to = "indicator", values_to = "n_coders") |>
  filter(!is.na(n_coders)) |>
  mutate(regime = factor(v2x_regime, 0:3,
                         labels = c("Closed autocracy", "Electoral autocracy",
                                    "Electoral democracy", "Liberal democracy")))

regime_pal <- c("Closed autocracy"    = "#D55E00", "Electoral autocracy" = "#E69F00",
                "Electoral democracy" = "#56B4E9", "Liberal democracy"   = "#0072B2")

p_ctx_time <- regime_ctx |>
  group_by(year, regime) |>
  summarise(mean_n = mean(n_coders), .groups = "drop") |>
  ggplot(aes(year, mean_n, colour = regime)) +
  geom_line(linewidth = 0.85) +
  geom_vline(xintercept = c(2005, 2013), linetype = "dotted", colour = "grey45") +
  annotate("text", x = 2005, y = Inf, label = "2005", hjust = -0.15, vjust = 1.6,
           size = 2.9, colour = "grey40") +
  annotate("text", x = 2013, y = Inf, label = "2013", hjust = -0.15, vjust = 1.6,
           size = 2.9, colour = "grey40") +
  scale_colour_manual(values = regime_pal, name = NULL,
                      guide = guide_legend(nrow = 2)) +
  coord_cartesian(ylim = ctx_ylim) +
  labs(title = "A. Over time, by regime type",
       subtitle = paste("Autocracies run persistently thinner. Dotted lines: the",
                        "back-coding floor\nand the end of the original coding round."),
       x = "Year being rated", y = "Coders per country-indicator-year") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        plot.title.position = "plot")

# Labels are placed at explicit coordinates rather than by repel-with-nudges. The curve
# runs corner to corner, so the two clear regions are the wedge ABOVE it on the left and
# the wedge BELOW it on the right. Thin-country labels go in the first, thick-country in
# the second, each as an evenly spaced column with a drawn leader line. Longest labels sit
# furthest from the curve, since text extends horizontally into it.
ctx_lab <- country_var |>
  filter(rank <= 6 | rank > n() - 6) |>
  mutate(side = if_else(rank <= 6, "thin", "thick")) |>
  group_by(side) |>
  # within a side, longest text gets the position furthest from the curve
  arrange(desc(nchar(country_name)), .by_group = TRUE) |>
  mutate(
    slot  = row_number(),
    # Thick-country labels fan left as they RISE. The points sit together in the top-right
    # corner, so a label placed low and far left needs a long leader line that would cross
    # every label between it and its point. Anchoring the lowest label furthest right keeps
    # each leader line to the right of the labels below it, so none are crossed.
    lab_x = if_else(side == "thin", 20, 176 - (slot - 1) * 4.6),
    lab_y = if_else(side == "thin", 13.2 - (slot - 1) * 0.72,
                                     6.4 + (slot - 1) * 0.72),
    hj    = if_else(side == "thin", 0, 1)
  ) |>
  ungroup()

p_ctx_country <- ggplot(country_var, aes(rank, mean_n)) +
  geom_hline(yintercept = 5, linetype = "22", colour = "grey55") +
  annotate("text", x = 55, y = 5, label = "V-Dem's floor of five",
           hjust = 0, vjust = -0.7, size = 2.9, colour = "grey45") +
  geom_segment(data = ctx_lab,
               aes(x = rank, y = mean_n, xend = lab_x, yend = lab_y),
               colour = "grey70", linewidth = 0.22, inherit.aes = FALSE) +
  geom_point(aes(colour = lt5_bin), size = 1.6) +
  geom_text(data = ctx_lab, aes(x = lab_x, y = lab_y, label = country_name, hjust = hj),
            size = 2.7, colour = "grey20", inherit.aes = FALSE) +
  scale_colour_manual(values = lt5_pal, name = "Cells below five",
                      guide = guide_legend(nrow = 1)) +
  coord_cartesian(ylim = c(4.8, 15.8), xlim = c(-2, 184)) +
  labs(title = "B. Across countries",
       subtitle = paste(nrow(country_var),
                        "countries, 2005–2024, ranked. Labels: 6 thinnest and 6 thickest."),
       x = "Country (ranked)", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(), plot.title.position = "plot")

(p_ctx_time | p_ctx_country) + plot_layout(widths = c(1, 1.1))
```
