---
title: "Calibration Metrics: LOO MAE, Exact Match, and Compression Diagnostics"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
execute:
  warning: false
  message: false
params:
  n_boot: 500
  year: 2019
  min_coders: 2
---

Three prompt conditions × five models evaluated against 12 V-Dem indicators for
`r params$year`. Primary metric is leave-one-out MAE: the AI error against the panel mean
compared to the typical error of a randomly held-out human coder against the rest of their
panel. Secondary metrics are exact match rate and adjacent-category agreement (±1).
Signed deviation by democracy quintile is reported as a compression diagnostic.

Bootstrap CIs use `r params$n_boot` resamples at the country-year-indicator (CYI) level.
See `notes/evaluation-metrics.md` for full metric rationale.

---

## 1. Setup

```{r setup}
library(tidyverse)
library(rsample)
library(gt)
library(here)

indicators <- c(
  "v2clkill", "v2cltort", "v2mecenefm", "v2csreprss", "v2jupoatck",
  "v2mecenefi", "v2juhcind", "v2clacfree", "v2clslavef",
  "v2psoppaut", "v2excrptps", "v2pepwrsoc"
)

observability <- tribble(
  ~indicator,    ~tier,
  "v2clkill",    "High",
  "v2cltort",    "High",
  "v2mecenefm",  "High",
  "v2csreprss",  "High",
  "v2jupoatck",  "High",
  "v2mecenefi",  "Medium",
  "v2juhcind",   "Medium",
  "v2clacfree",  "Medium",
  "v2clslavef",  "Medium",
  "v2psoppaut",  "Medium",
  "v2excrptps",  "Medium",
  "v2pepwrsoc",  "Low"
)

model_labels <- c(
  "claude-sonnet"      = "Claude Sonnet 4.6",
  "llama-405b"         = "Llama 405B",
  "llama-70b"          = "Llama 70B",
  "llama-9b"           = "Llama 9B",
  "llama-70b-finetuned" = "Llama 70B (fine-tuned)"
)

condition_labels <- c(
  "codebook"   = "Codebook",
  "evidence"   = "Evidence",
  "anonymized" = "Anonymized"
)

data_dir  <- here("data", "processed")
runs_dir  <- here("data", "output", "runs")
```

---

## 2. Data loading

```{r load-data}
#| cache: true
#| cache.lazy: false

# AI ratings — all JSONL files in runs_dir
ai_raw <- list.files(runs_dir, pattern = "\\.jsonl$", full.names = TRUE) |>
  map(\(f) {
    readLines(f) |>
      keep(\(l) nzchar(trimws(l))) |>
      map(jsonlite::fromJSON) |>
      bind_rows()
  }) |>
  bind_rows()

# Human panel means — country_text_id, year, indicator, raw_mean, n_coders, theta_quintile
panel_means <- read_csv(file.path(data_dir, "panel_means.csv"), show_col_types = FALSE)

# Individual coder ratings — country_text_id, year, indicator, coder_id, rating
human_ratings <- read_csv(file.path(data_dir, "human_ratings.csv"), show_col_types = FALSE)
```

---

## 3. Metric construction

### Human LOO baseline

For each CYI with ≥ `r params$min_coders` coders, compute each coder's leave-one-out
error: how far their rating falls from the panel mean with them removed.

```{r human-loo}
#| cache: true

human_loo <- human_ratings |>
  filter(year == params$year, indicator %in% indicators) |>
  group_by(country_text_id, year, indicator) |>
  filter(n() >= params$min_coders) |>
  mutate(
    panel_sum  = sum(rating),
    n_coders   = n(),
    loo_mean   = (panel_sum - rating) / (n_coders - 1),
    loo_error  = abs(rating - loo_mean)
  ) |>
  summarise(
    human_loo_mae = mean(loo_error),
    panel_mean    = mean(rating),
    n_coders      = first(n_coders),
    .groups       = "drop"
  )
```

### AI metrics per CYI

```{r ai-metrics}
ai_metrics <- ai_raw |>
  filter(year == params$year, indicator %in% indicators) |>
  rename(country_text_id = country) |>
  inner_join(human_loo, by = c("country_text_id", "year", "indicator")) |>
  left_join(
    select(panel_means, country_text_id, year, indicator, theta_quintile),
    by = c("country_text_id", "year", "indicator")
  ) |>
  mutate(
    ai_error    = abs(rating - panel_mean),
    exact_match = as.integer(rating == round(panel_mean)),
    adjacent    = as.integer(abs(rating - round(panel_mean)) <= 1),
    signed_dev  = rating - panel_mean,
    model_label = recode(model_key, !!!model_labels),
    cond_label  = recode(condition, !!!condition_labels)
  )
```

---

## 4. Bootstrap CIs

Resample at the CYI level. For each draw, compute AI LOO MAE, human LOO MAE, and their
difference.

```{r bootstrap}
#| cache: true

loo_stat <- function(split) {
  d <- analysis(split)
  d |>
    group_by(model_key, condition) |>
    summarise(
      ai_mae    = mean(ai_error),
      human_mae = mean(human_loo_mae),
      diff      = mean(ai_error) - mean(human_loo_mae),
      .groups   = "drop"
    )
}

set.seed(42)

# One bootstrap per unique CYI (pool across models first, then resample CYIs)
cyi_pool <- ai_metrics |>
  distinct(country_text_id, year, indicator, human_loo_mae, panel_mean)

boots <- bootstraps(cyi_pool, times = params$n_boot, apparent = TRUE)

# Re-attach AI errors for each bootstrap draw
boot_results <- boots |>
  mutate(
    result = map(splits, \(s) {
      cyi_s <- analysis(s)
      ai_metrics |>
        semi_join(cyi_s, by = c("country_text_id", "year", "indicator")) |>
        group_by(model_key, condition) |>
        summarise(
          ai_mae    = mean(ai_error),
          human_mae = mean(human_loo_mae),
          diff      = mean(ai_error) - mean(human_loo_mae),
          .groups   = "drop"
        )
    })
  ) |>
  unnest(result)

boot_ci <- boot_results |>
  group_by(model_key, condition) |>
  summarise(
    ai_mae_mean  = mean(ai_mae[id == "Apparent"]),
    human_mae    = mean(human_mae[id == "Apparent"]),
    ci_lo        = quantile(diff, 0.025),
    ci_hi        = quantile(diff, 0.975),
    p_gt_zero    = mean(diff > 0),
    .groups      = "drop"
  )
```

---

## 5. Primary table: LOO MAE by model × indicator

```{r loo-table}
loo_by_indicator <- ai_metrics |>
  group_by(model_key, condition, indicator) |>
  summarise(ai_mae = mean(ai_error), .groups = "drop") |>
  left_join(
    ai_metrics |>
      group_by(indicator) |>
      summarise(human_mae = mean(human_loo_mae), .groups = "drop"),
    by = "indicator"
  )

# Aggregate column
loo_agg <- ai_metrics |>
  group_by(model_key, condition) |>
  summarise(ai_mae = mean(ai_error), .groups = "drop") |>
  mutate(indicator = "Overall")

loo_wide <- bind_rows(
    loo_by_indicator |> select(model_key, condition, indicator, ai_mae),
    loo_agg
  ) |>
  mutate(label = paste0(
    recode(model_key, !!!model_labels), " — ",
    recode(condition, !!!condition_labels)
  )) |>
  pivot_wider(names_from = indicator, values_from = ai_mae, id_cols = label)

# Human baseline row
human_row <- ai_metrics |>
  summarise(across(
    c(human_loo_mae),
    \(x) mean(x),
    .names = "Overall"
  )) |>
  mutate(label = "Human LOO baseline")

loo_wide |>
  gt() |>
  tab_header(
    title    = "LOO MAE by model, condition, and indicator",
    subtitle = glue::glue("Year: {params$year}  |  Bootstrap B = {params$n_boot}")
  ) |>
  fmt_number(columns = where(is.numeric), decimals = 2) |>
  tab_spanner(label = "High observability",
              columns = c(v2clkill, v2cltort, v2mecenefm, v2csreprss, v2jupoatck)) |>
  tab_spanner(label = "Medium observability",
              columns = c(v2mecenefi, v2juhcind, v2clacfree, v2clslavef,
                          v2psoppaut, v2excrptps)) |>
  tab_spanner(label = "Low observability", columns = v2pepwrsoc) |>
  tab_style(
    style = cell_fill(color = "#f0f0f0"),
    locations = cells_body(rows = str_detect(label, "fine-tuned"))
  )
```

---

## 6. Secondary metrics

```{r secondary-metrics}
ai_metrics |>
  group_by(model_key, condition) |>
  summarise(
    exact_match = mean(exact_match),
    adjacent    = mean(adjacent),
    .groups     = "drop"
  ) |>
  mutate(
    model_label = recode(model_key, !!!model_labels),
    cond_label  = recode(condition, !!!condition_labels)
  ) |>
  select(model_label, cond_label, exact_match, adjacent) |>
  gt() |>
  tab_header(title = "Exact match and adjacent-category agreement") |>
  fmt_percent(columns = c(exact_match, adjacent), decimals = 1) |>
  cols_label(
    model_label = "Model",
    cond_label  = "Condition",
    exact_match = "Exact match",
    adjacent    = "Within ±1"
  )
```

---

## 7. Compression diagnostic: signed deviation by quintile

The human baseline for signed deviation is ~0 by construction. Any systematic AI
deviation reveals directional bias: positive values in low quintiles indicate the AI
rates autocracies too generously; negative values in high quintiles indicate it rates
democracies too harshly.

```{r signed-deviation}
#| fig-width: 8
#| fig-height: 5

ai_metrics |>
  filter(!is.na(theta_quintile)) |>
  group_by(model_key, condition, theta_quintile) |>
  summarise(signed_dev = mean(signed_dev), .groups = "drop") |>
  mutate(
    model_label = recode(model_key, !!!model_labels),
    cond_label  = recode(condition, !!!condition_labels),
    series      = paste0(model_label, " — ", cond_label)
  ) |>
  ggplot(aes(x = theta_quintile, y = signed_dev, color = model_label,
             linetype = cond_label, group = series)) +
  geom_line() +
  geom_point(size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  scale_x_continuous(breaks = 1:5,
                     labels = c("Q1\n(most\nautocratic)", "Q2", "Q3", "Q4",
                                "Q5\n(most\ndemocratic)")) +
  labs(
    x        = "Democracy quintile (v2x_polyarchy)",
    y        = "Signed deviation (AI − panel mean)",
    color    = "Model",
    linetype = "Condition",
    title    = "Compression bias by regime type",
    subtitle = glue::glue("Year: {params$year}")
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```
