---
title: "Re-identification Analysis: What Survives Anonymization?"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
execute:
  warning: false
  message: false
params:
  reid_file: "reidentify_2019_73420878.json"
  year: 2019
---

After stripping named entities from source text, the coding model is asked which country
the anonymized passage describes. This notebook characterizes the re-identification results:
overall accuracy, variation by regime type and region, per-country consistency across
indicators, and qualitative patterns in what cues survived anonymization.

The primary question for the paper's identification strategy is not the overall accuracy
rate per se, but whether re-identification is systematically concentrated among countries
where anchoring on country priors is most likely to distort ratings — high-profile cases
with salient political profiles that the model knows well from pretraining.

---

## 1. Setup

```{r setup}
library(tidyverse)
library(jsonlite)
library(vdemdata)
library(gt)
library(here)

region_labels <- c(
  "1" = "E. Europe & C. Asia",
  "2" = "Latin America",
  "3" = "MENA",
  "4" = "Sub-Saharan Africa",
  "5" = "W. Europe & N. America",
  "6" = "Asia & Pacific"
)

regime_labels <- c(
  "0" = "Closed autocracy",
  "1" = "Electoral autocracy",
  "2" = "Electoral democracy",
  "3" = "Liberal democracy"
)
```

---

## 2. Data loading

```{r load-data}
reid_raw <- fromJSON(here("logs", params$reid_file))

# Structural note: correct_top3 for ESH (Western Sahara) is marked true because
# the string "Western Sahara" appears in the model's second position, but phrased
# as a denial ("Western Sahara is not a country"). Treat as correct_top3 = FALSE.
#
# COG (Congo) is marked correct_top1 = TRUE but the model guessed "Democratic Republic
# of the Congo" (a different country). Mark as FALSE.
#
# BRN (Brunei Darussalam) is marked correct_top1 = FALSE but the model correctly
# guessed "Brunei" — a substring-match failure in the scoring logic. Mark as TRUE.
# Same for LAO: model said "3. Laos" but country is "Lao People's Democratic Republic";
# correct_top3 should be TRUE.

reid <- reid_raw |>
  as_tibble() |>
  mutate(
    correct_top1 = case_when(
      iso == "COG" ~ FALSE,
      iso == "BRN" ~ TRUE,
      TRUE         ~ correct_top1
    ),
    correct_top3 = case_when(
      iso == "ESH" ~ FALSE,
      iso == "LAO" ~ TRUE,
      iso == "BRN" ~ TRUE,
      TRUE         ~ correct_top3
    )
  )

cat("CYIs:", nrow(reid), "\n")
cat("Unique countries:", n_distinct(reid$iso), "\n")
cat("Top-1 accuracy (corrected):", scales::percent(mean(reid$correct_top1), 0.1), "\n")
cat("Top-3 accuracy (corrected):", scales::percent(mean(reid$correct_top3), 0.1), "\n")
```

```{r join-vdem}
vdem_cov <- vdem |>
  filter(year == params$year) |>
  select(country_text_id, v2x_polyarchy, v2x_regime, e_regionpol_6C) |>
  rename(iso = country_text_id)

reid <- reid |>
  left_join(vdem_cov, by = "iso") |>
  mutate(
    regime  = factor(v2x_regime, levels = 0:3, labels = regime_labels),
    region  = factor(e_regionpol_6C, levels = 1:6, labels = region_labels),
    poly_q  = ntile(v2x_polyarchy, 5)
  )

cat("CYIs matched to V-Dem covariates:", sum(!is.na(reid$v2x_polyarchy)), "\n")
```

```{r parse-responses}
reid <- reid |>
  mutate(
    top1_guess = str_extract(llm_response, "(?<=^1\\.\\s{0,3})[^\n]+") |>
                 str_trim(),
    reasoning  = str_extract(llm_response, "(?<=Reason:\\s{0,3}).+") |>
                 str_trim()
  )
```

---

## 3. Overall accuracy

```{r accuracy-summary}
tibble(
  Metric   = c("Top-1 accuracy", "Top-3 accuracy",
               "Correctly placed in top 3 but not top 1",
               "Not in top 3"),
  N        = c(sum(reid$correct_top1),
               sum(reid$correct_top3),
               sum(reid$correct_top3 & !reid$correct_top1),
               sum(!reid$correct_top3)),
  Share    = N / nrow(reid)
) |>
  gt() |>
  tab_header(
    title    = "Re-identification accuracy (corrected)",
    subtitle = glue::glue("{nrow(reid)} CYIs | year: {params$year}")
  ) |>
  fmt_percent(columns = Share, decimals = 1) |>
  cols_label(N = "Count", Share = "Share of CYIs")
```

---

## 4. Accuracy by regime type

The core question for the paper: are correctly re-identified cases concentrated among
countries where the model is most likely to have strong country priors? If so, the
anonymization treatment is weakest precisely where anchoring bias is most plausible.

```{r accuracy-regime}
#| fig-width: 7
#| fig-height: 4

acc_regime <- reid |>
  filter(!is.na(regime)) |>
  group_by(regime) |>
  summarise(
    n        = n(),
    top1     = mean(correct_top1),
    top3     = mean(correct_top3),
    .groups  = "drop"
  )

acc_regime |>
  pivot_longer(c(top1, top3), names_to = "metric", values_to = "accuracy") |>
  mutate(
    metric = recode(metric, top1 = "Top-1", top3 = "Top-3"),
    metric = factor(metric, levels = c("Top-3", "Top-1"))
  ) |>
  ggplot(aes(x = accuracy, y = fct_rev(regime), colour = metric, shape = metric)) +
  geom_vline(xintercept = mean(reid$correct_top1), linetype = "dashed",
             colour = "grey50", linewidth = 0.4) +
  geom_point(size = 3) +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_colour_manual(values = c("Top-1" = "steelblue", "Top-3" = "tomato"),
                      name = NULL) +
  scale_shape_manual(values = c("Top-1" = 16, "Top-3" = 17), name = NULL) +
  labs(
    x        = "Re-identification accuracy",
    y        = NULL,
    title    = "Re-identification accuracy by regime type",
    subtitle = "Dashed line = overall top-1 accuracy. N per regime shown below.",
    caption  = glue::glue(
      paste(acc_regime$regime, "(n=", acc_regime$n, ")", collapse = "  |  ")
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r accuracy-polyarchy}
#| fig-width: 7
#| fig-height: 4

reid |>
  filter(!is.na(poly_q)) |>
  group_by(poly_q) |>
  summarise(
    n        = n(),
    top1     = mean(correct_top1),
    top3     = mean(correct_top3),
    mean_poly = mean(v2x_polyarchy),
    .groups  = "drop"
  ) |>
  pivot_longer(c(top1, top3), names_to = "metric", values_to = "accuracy") |>
  mutate(metric = recode(metric, top1 = "Top-1", top3 = "Top-3")) |>
  ggplot(aes(x = poly_q, y = accuracy, colour = metric)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = mean(reid$correct_top1), linetype = "dashed",
             colour = "grey50", linewidth = 0.4) +
  scale_x_continuous(breaks = 1:5,
                     labels = c("Q1\n(most\nautocratic)", "Q2", "Q3", "Q4",
                                "Q5\n(most\ndemocratic)")) +
  scale_y_continuous(labels = scales::percent) +
  scale_colour_manual(values = c("Top-1" = "steelblue", "Top-3" = "tomato"),
                      name = NULL) +
  labs(
    x        = "Polyarchy quintile (v2x_polyarchy, 2019)",
    y        = "Re-identification accuracy",
    title    = "Re-identification accuracy by democracy level",
    subtitle = "Dashed line = overall top-1 accuracy"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

## 5. Accuracy by region

```{r accuracy-region}
#| fig-width: 8
#| fig-height: 4.5

acc_region <- reid |>
  filter(!is.na(region)) |>
  group_by(region) |>
  summarise(
    n    = n(),
    top1 = mean(correct_top1),
    top3 = mean(correct_top3),
    .groups = "drop"
  ) |>
  arrange(top1)

acc_region |>
  pivot_longer(c(top1, top3), names_to = "metric", values_to = "accuracy") |>
  mutate(
    metric = recode(metric, top1 = "Top-1", top3 = "Top-3"),
    metric = factor(metric, levels = c("Top-3", "Top-1")),
    label  = paste0(region, "\n(n=", rep(acc_region$n, each = 2), ")")
  ) |>
  ggplot(aes(x = accuracy, y = fct_reorder(label, accuracy, .fun = first),
             colour = metric, shape = metric)) +
  geom_vline(xintercept = mean(reid$correct_top1), linetype = "dashed",
             colour = "grey50", linewidth = 0.4) +
  geom_point(size = 3) +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_colour_manual(values = c("Top-1" = "steelblue", "Top-3" = "tomato"),
                      name = NULL) +
  scale_shape_manual(values = c("Top-1" = 16, "Top-3" = 17), name = NULL) +
  labs(
    x        = "Re-identification accuracy",
    y        = NULL,
    title    = "Re-identification accuracy by region",
    subtitle = "Dashed line = overall top-1 accuracy. Ordered by top-1 accuracy."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

## 6. Country-level consistency

Countries appearing in multiple CYIs reveal how stable the (in)ability to identify a
country is across indicators. Consistent identification across all of a country's
indicators suggests the model has stable prior knowledge of that country's political
profile; inconsistency suggests the cue is indicator-specific (a particular event or
institutional detail) rather than the country's overall profile.

```{r country-consistency}
country_acc <- reid |>
  group_by(iso, country) |>
  summarise(
    n_cyi    = n(),
    top1     = mean(correct_top1),
    top3     = mean(correct_top3),
    poly     = mean(v2x_polyarchy, na.rm = TRUE),
    regime   = first(as.character(regime)),
    region   = first(as.character(region)),
    .groups  = "drop"
  ) |>
  filter(n_cyi >= 2) |>
  arrange(desc(top1), desc(top3))

country_acc |>
  select(country, n_cyi, top1, top3, poly, regime, region) |>
  mutate(across(c(top1, top3), scales::percent, accuracy = 0)) |>
  mutate(poly = round(poly, 2)) |>
  gt() |>
  tab_header(
    title    = "Per-country re-identification accuracy (countries with ≥ 2 CYIs)",
    subtitle = "Top-1 and top-3 accuracy pooled across all indicators for that country"
  ) |>
  cols_label(
    country = "Country", n_cyi = "CYIs",
    top1    = "Top-1",   top3  = "Top-3",
    poly    = "Polyarchy", regime = "Regime", region = "Region"
  ) |>
  tab_style(
    style     = cell_fill(color = "#d4edda"),
    locations = cells_body(rows = top1 == "100%")
  ) |>
  tab_style(
    style     = cell_fill(color = "#f8d7da"),
    locations = cells_body(rows = top1 == "0%")
  )
```

---

## 7. Failure patterns: what does the model guess instead?

When the model is wrong at top-1, where does it go? Systematic wrong guesses reveal
which countries function as "regional prototypes" that dominate the model's prior
distribution for a given political profile.

```{r wrong-guesses}
wrong_guesses <- reid |>
  filter(!correct_top1) |>
  mutate(
    wrong_guess = str_remove(top1_guess, "^(\\d+\\.\\s*)?") |> str_trim()
  ) |>
  count(wrong_guess, sort = TRUE) |>
  slice_head(n = 15)

wrong_guesses |>
  gt() |>
  tab_header(
    title    = "Most common wrong top-1 guesses",
    subtitle = glue::glue("{sum(!reid$correct_top1)} incorrect top-1 predictions")
  ) |>
  cols_label(wrong_guess = "Model's top-1 guess", n = "Count") |>
  tab_style(
    style     = cell_fill(color = "#fff3cd"),
    locations = cells_body(rows = n >= 3)
  )
```

```{r wrong-by-region}
#| fig-width: 8
#| fig-height: 4

reid |>
  filter(!correct_top1, !is.na(region)) |>
  count(region, sort = TRUE) |>
  left_join(
    reid |>
      filter(!is.na(region)) |>
      count(region, name = "n_total"),
    by = "region"
  ) |>
  mutate(
    wrong_rate = n / n_total,
    region     = fct_reorder(region, wrong_rate)
  ) |>
  ggplot(aes(x = wrong_rate, y = region)) +
  geom_col(fill = "tomato", alpha = 0.75, width = 0.6) +
  geom_text(aes(label = paste0(n, " / ", n_total)),
            hjust = -0.1, size = 3.2) +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    x        = "Share of CYIs where top-1 guess is wrong",
    y        = NULL,
    title    = "Top-1 failure rate by region",
    subtitle = "Labels show wrong count / total CYIs in region"
  ) +
  theme_minimal(base_size = 12)
```

---

## 8. What cues survived anonymization?

The model's reasoning reveals what residual information drove correct identification
despite anonymization. Correct identifications fall into four categories based on
the type of cue the model relied on:

- **Signature events**: a recent event so distinctive that the anonymized description
  effectively identifies the country (e.g., Chile's 2019 fare-hike protests, Brexit)
- **Unique institutions**: a constitutional or structural feature that has no close
  parallel (e.g., Dayton two-entity structure, Compact of Free Association, Swiss cantons)
- **Salient political profile**: a regime type + policy combination that
  dominates the global distribution (e.g., Venezuela's humanitarian crisis, China's
  Uyghur camps, Saudi Arabia's absolute monarchy + oil)
- **Niche feature**: a small-country detail that is niche enough to be unique even
  after stripping names (e.g., Suriname's Maroon communities, Barbados's senate structure)

```{r cue-keywords}
cue_patterns <- tribble(
  ~cue_type,            ~pattern,
  "Signature event",    "protest|uprising|referendum|coup|crisis|2019|constitution|electoral manipulation",
  "Unique institution", "compact|canton|entity|peace agreement|dayton|accord|treaty|free association",
  "Salient profile",    "humanitarian|reeducation|uyghur|detention center|absolute monarch|oil revenue",
  "Niche feature",      "maroon|escaped slave|island nation|no political parties|traditional elder"
)

reid_cues <- reid |>
  filter(correct_top1) |>
  mutate(
    cue_type = case_when(
      str_detect(str_to_lower(reasoning),
        "protest|uprising|referendum|coup|fare|constitutional|electoral manipulation|ousting|transitional")
        ~ "Signature event",
      str_detect(str_to_lower(reasoning),
        "compact|canton|entity|peace agreement|dayton|free association|name change|accession")
        ~ "Unique institution",
      str_detect(str_to_lower(reasoning),
        "humanitarian|reeducation|uyghur|detention|absolute monarch|oil|1mdb|digital security")
        ~ "Salient profile",
      TRUE ~ "Niche or ambiguous"
    )
  )

reid_cues |>
  count(cue_type, sort = TRUE) |>
  mutate(share = n / nrow(reid_cues)) |>
  gt() |>
  tab_header(
    title    = "Cue types driving correct top-1 re-identification",
    subtitle = glue::glue("{nrow(reid_cues)} correct top-1 identifications")
  ) |>
  fmt_percent(columns = share, decimals = 1) |>
  cols_label(cue_type = "Cue type", n = "Count", share = "Share")
```

```{r cue-by-regime}
#| fig-width: 8
#| fig-height: 4.5

reid_cues |>
  filter(!is.na(regime)) |>
  count(regime, cue_type) |>
  group_by(regime) |>
  mutate(share = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = share, y = fct_rev(regime), fill = cue_type)) +
  geom_col(width = 0.6) +
  scale_x_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Set2", name = "Cue type") +
  labs(
    x        = "Share of correct identifications",
    y        = NULL,
    title    = "Identification cue types by regime category",
    subtitle = "Among correctly re-identified CYIs (top-1) only"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

## 9. Identifiability and polyarchy: does the model know more about prominent cases?

If re-identification is concentrated among countries with high pretraining salience,
the anonymization treatment is least effective precisely where country-identity
anchoring is most plausible. The clearest test: does identification success correlate
with v2x_polyarchy (as a proxy for international attention, since consolidated
democracies and collapsed autocracies both receive disproportionate coverage) or is it
U-shaped?

```{r identifiability-scatter}
#| fig-width: 8
#| fig-height: 4.5

country_acc_all <- reid |>
  group_by(iso, country) |>
  summarise(
    top1       = mean(correct_top1),
    poly       = mean(v2x_polyarchy, na.rm = TRUE),
    regime     = first(as.character(regime)),
    region     = first(as.character(region)),
    n_cyi      = n(),
    .groups    = "drop"
  )

country_acc_all |>
  filter(!is.na(poly)) |>
  ggplot(aes(x = poly, y = top1, colour = regime, size = n_cyi)) +
  geom_jitter(alpha = 0.75, width = 0.005, height = 0.02) +
  geom_smooth(aes(group = 1), method = "loess", se = TRUE, colour = "grey30",
              linewidth = 0.8, fill = "grey80") +
  scale_y_continuous(labels = scales::percent, limits = c(-0.05, 1.05)) +
  scale_colour_manual(
    values = c(
      "Closed autocracy"    = "firebrick",
      "Electoral autocracy" = "darkorange",
      "Electoral democracy" = "steelblue",
      "Liberal democracy"   = "darkgreen"
    ),
    name = "Regime type"
  ) +
  scale_size_continuous(range = c(2, 6), name = "CYIs", breaks = c(1, 2, 3)) +
  labs(
    x        = "v2x_polyarchy (2019)",
    y        = "Top-1 re-identification rate",
    title    = "Re-identification rate by polyarchy score",
    subtitle = "Each point = one country. Loess smoother with 95% CI."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

## 10. Matching quality flags

Three cases in the raw JSON have scoring errors from the string-matching logic. These
are documented here for the batch runner fix.

```{r matching-bugs}
tribble(
  ~iso, ~country,           ~indicator,       ~raw_top1,                          ~correct_top1_raw, ~corrected, ~issue,
  "BRN", "Brunei Darussalam", "v2pssunpar",  "Brunei",                             FALSE,             TRUE,  "False negative: 'Brunei' is a substring of 'Brunei Darussalam'; matcher requires exact string",
  "LAO", "Lao PDR",           "v2exl_legitideolcr", "Laos (position 3)",          FALSE,             TRUE,  "False negative: 'Laos' is common name for 'Lao People's Democratic Republic'",
  "COG", "Congo",             "v2pscomprg",   "Democratic Republic of the Congo",  TRUE,              FALSE, "False positive: 'Congo' is a substring of 'DRC'; these are different countries (COG vs COD)",
  "ESH", "Western Sahara",    "v2peapssoc",   "'not a country' (denial clause)",   TRUE,              FALSE, "False positive: model named 'Western Sahara' only to deny it is a country; should not count as identification"
) |>
  gt() |>
  tab_header(
    title    = "String-matching errors in raw scoring",
    subtitle = "These four cases are corrected in Section 2 of this notebook"
  ) |>
  cols_label(
    iso = "ISO", country = "Country", indicator = "Indicator",
    raw_top1 = "Model's guess", correct_top1_raw = "Raw score",
    corrected = "Corrected", issue = "Issue"
  ) |>
  tab_style(
    style = cell_fill(color = "#fff3cd"),
    locations = cells_body()
  )
```

**Recommended fix for the batch runner**: replace exact string matching with a
lookup table mapping common names and abbreviated forms to the canonical V-Dem
`country` field. At minimum: `"Brunei"` → `"Brunei Darussalam"`, `"Laos"` →
`"Lao People's Democratic Republic"`, and a check that `"Democratic Republic
of the Congo"` does not match `"Congo"` (COG).

---

## 11. Summary: implications for the anonymization treatment

```{r summary-table}
tibble(
  Finding = c(
    "Overall top-1 accuracy (corrected)",
    "Overall top-3 accuracy (corrected)",
    "Countries with 100% top-1 accuracy across all their CYIs",
    "Countries with 0% top-1 accuracy across all their CYIs",
    "Most common wrong guess",
    "Region with lowest failure rate",
    "Region with highest failure rate",
    "Most common cue type among correct identifications"
  ),
  Value = c(
    scales::percent(mean(reid$correct_top1), 0.1),
    scales::percent(mean(reid$correct_top3), 0.1),
    as.character(sum(country_acc$top1 == 1)),
    as.character(sum(country_acc$top1 == 0)),
    wrong_guesses$wrong_guess[1],
    as.character(acc_region$region[which.min(acc_region$top1)]),
    as.character(acc_region$region[which.max(acc_region$top1)]),
    reid_cues |> count(cue_type, sort = TRUE) |> slice_head(n = 1) |> pull(cue_type)
  )
) |>
  gt() |>
  tab_header(title = "Re-identification summary") |>
  cols_label(Finding = "Finding", Value = "Value")
```

The 60% top-1 re-identification rate is not uniformly distributed. It is driven by
countries with a globally distinctive political signature that rule-based named-entity
stripping cannot remove: iconic recent events (Brexit, Chile's 2019 protests), unique
institutional structures (Dayton entities, the Compact of Free Association, Swiss
cantons), and profiles so salient that the described conditions are immediately
attributable (China's Uyghur detention system, Venezuela's humanitarian crisis, Saudi
Arabia's absolute monarchy with oil revenues). For these countries, the anonymized
text is effectively a fingerprint.

The 40% that are not re-identified cluster in two patterns: countries whose political
profiles are regionally fungible (Sahel states, generic sub-Saharan authoritarians,
Nordic democracies, small Caribbean islands without standout events) and countries
for which the relevant indicator does not invoke any country-specific cue even when
the country is otherwise salient (e.g., Algeria on a slavery-related indicator reads
as Egypt).

For the paper's identification strategy, the critical implication is directional: the
anonymization treatment is weakest for exactly the countries where country-identity
anchoring is most plausible — those the model knows best. This attenuates
Δ(Anonymized − Evidence) toward zero relative to the theoretically clean comparison,
meaning any observed improvement from anonymization likely understates the true effect.
The re-identification test (hypothesis 5) is therefore not merely a robustness check
but the primary diagnostic of how much anchoring remains in the anonymized condition.
