---
title: "Re-identification Analysis: Summarization vs. Anonymization"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
execute:
  warning: false
  message: false
params:
  summ_file: "reidentify_summ_2019_73427004.json"
  anon_file: "reidentify_2019_73420878.json"
  year: 2019
---

Summarization replaces source text with generic descriptions of political conditions,
stripping both named entities and distinctive event details. Anonymization replaces
named entities with bracketed placeholders while preserving prose structure.
This notebook compares re-identification rates under each treatment, identifies the
countries and cue types that drive the difference, and draws out implications for
which treatment better breaks the link between source text and country identity.

The primary question for the paper's identification strategy is whether the
summarization treatment reduces re-identification relative to anonymization and,
if so, whether it does so by removing the right kind of information — country-specific
detail rather than politically substantive content.

---

## 1. Setup

```{r setup}
library(tidyverse)
library(jsonlite)
library(vdemdata)
library(gt)
library(here)
library(glue)

region_labels <- c(
  "1" = "E. Europe & C. Asia",
  "2" = "Latin America",
  "3" = "MENA",
  "4" = "Sub-Saharan Africa",
  "5" = "W. Europe & N. America",
  "6" = "Asia & Pacific"
)

regime_labels <- c(
  "0" = "Closed autocracy",
  "1" = "Electoral autocracy",
  "2" = "Electoral democracy",
  "3" = "Liberal democracy"
)
```

---

## 2. Data loading

```{r load-summ}
summ_raw <- fromJSON(here("logs", params$summ_file))

# BRN (Brunei Darussalam): model guessed "Brunei" at position 2 — scored false
# negative because "Brunei" is a substring of the full name. Correct top3 to TRUE.
summ <- summ_raw |>
  as_tibble() |>
  mutate(
    correct_top3 = case_when(
      iso == "BRN" ~ TRUE,
      TRUE         ~ correct_top3
    )
  )

cat("Summarization CYIs:", nrow(summ), "\n")
cat("Unique countries:", n_distinct(summ$iso), "\n")
cat("Top-1 (corrected):", scales::percent(mean(summ$correct_top1), 0.1), "\n")
cat("Top-3 (corrected):", scales::percent(mean(summ$correct_top3), 0.1), "\n")
```

```{r load-anon}
# Anonymization results (from 03-reidentification-analysis.qmd)
anon_raw <- fromJSON(here("logs", params$anon_file))

anon <- anon_raw |>
  as_tibble() |>
  mutate(
    correct_top1 = case_when(
      iso == "COG" ~ FALSE,
      iso == "BRN" ~ TRUE,
      TRUE         ~ correct_top1
    ),
    correct_top3 = case_when(
      iso == "ESH" ~ FALSE,
      iso == "LAO" ~ TRUE,
      iso == "BRN" ~ TRUE,
      TRUE         ~ correct_top3
    )
  )

cat("Anonymization CYIs:", nrow(anon), "\n")
cat("Unique countries:", n_distinct(anon$iso), "\n")
cat("Top-1 (corrected):", scales::percent(mean(anon$correct_top1), 0.1), "\n")
cat("Top-3 (corrected):", scales::percent(mean(anon$correct_top3), 0.1), "\n")
```

```{r join-vdem}
vdem_cov <- vdem |>
  filter(year == params$year) |>
  select(country_text_id, v2x_polyarchy, v2x_regime, e_regionpol_6C) |>
  rename(iso = country_text_id)

summ <- summ |>
  left_join(vdem_cov, by = "iso") |>
  mutate(
    regime = factor(v2x_regime, levels = 0:3, labels = regime_labels),
    region = factor(e_regionpol_6C, levels = 1:6, labels = region_labels),
    poly_q = ntile(v2x_polyarchy, 5),
    treatment = "Summarization"
  )

anon <- anon |>
  left_join(vdem_cov, by = "iso") |>
  mutate(
    regime = factor(v2x_regime, levels = 0:3, labels = regime_labels),
    region = factor(e_regionpol_6C, levels = 1:6, labels = region_labels),
    poly_q = ntile(v2x_polyarchy, 5),
    treatment = "Anonymization"
  )
```

```{r parse-responses}
parse_responses <- function(df) {
  df |>
    mutate(
      top1_guess = str_extract(llm_response, "(?<=^1\\.\\s{0,3})[^\n]+") |>
                   str_trim(),
      reasoning  = str_extract(llm_response, "(?<=Reason:\\s{0,3}).+") |>
                   str_trim()
    )
}

summ <- parse_responses(summ)
anon <- parse_responses(anon)
```

---

## 3. Overall accuracy comparison

```{r accuracy-comparison}
bind_rows(
  tibble(
    Treatment = "Summarization",
    `Top-1`   = mean(summ$correct_top1),
    `Top-3`   = mean(summ$correct_top3),
    CYIs      = nrow(summ),
    Countries = n_distinct(summ$iso)
  ),
  tibble(
    Treatment = "Anonymization",
    `Top-1`   = mean(anon$correct_top1),
    `Top-3`   = mean(anon$correct_top3),
    CYIs      = nrow(anon),
    Countries = n_distinct(anon$iso)
  )
) |>
  gt() |>
  tab_header(
    title    = "Re-identification accuracy by treatment",
    subtitle = glue("Year: {params$year}")
  ) |>
  fmt_percent(columns = c(`Top-1`, `Top-3`), decimals = 1) |>
  cols_label(CYIs = "CYIs", Countries = "Countries")
```

Summarization reduces top-1 re-identification by
`r scales::percent(mean(anon$correct_top1) - mean(summ$correct_top1), 0.1)` and
top-3 re-identification by
`r scales::percent(mean(anon$correct_top3) - mean(summ$correct_top3), 0.1)`.

---

## 4. Accuracy by regime type

```{r accuracy-regime}
#| fig-width: 7
#| fig-height: 4.5

both <- bind_rows(summ, anon)

acc_regime <- both |>
  filter(!is.na(regime)) |>
  group_by(treatment, regime) |>
  summarise(
    n    = n(),
    top1 = mean(correct_top1),
    top3 = mean(correct_top3),
    .groups = "drop"
  )

acc_regime |>
  pivot_longer(c(top1, top3), names_to = "metric", values_to = "accuracy") |>
  mutate(
    metric    = recode(metric, top1 = "Top-1", top3 = "Top-3"),
    metric    = factor(metric, levels = c("Top-3", "Top-1")),
    treatment = factor(treatment, levels = c("Anonymization", "Summarization"))
  ) |>
  ggplot(aes(x = accuracy, y = fct_rev(regime),
             colour = treatment, shape = metric, linetype = treatment)) +
  geom_vline(xintercept = mean(summ$correct_top1), linetype = "dotted",
             colour = "steelblue", linewidth = 0.4) +
  geom_vline(xintercept = mean(anon$correct_top1), linetype = "dotted",
             colour = "tomato", linewidth = 0.4) +
  geom_point(size = 3, position = position_dodge(width = 0.3)) +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_colour_manual(
    values = c("Anonymization" = "tomato", "Summarization" = "steelblue"),
    name = NULL
  ) +
  scale_shape_manual(values = c("Top-1" = 16, "Top-3" = 17), name = NULL) +
  labs(
    x        = "Re-identification accuracy",
    y        = NULL,
    title    = "Re-identification accuracy by regime type",
    subtitle = "Dotted lines = overall top-1 accuracy per treatment"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

## 5. Accuracy by region

```{r accuracy-region}
#| fig-width: 8
#| fig-height: 5

acc_region <- both |>
  filter(!is.na(region)) |>
  group_by(treatment, region) |>
  summarise(
    n    = n(),
    top1 = mean(correct_top1),
    .groups = "drop"
  )

region_order <- acc_region |>
  filter(treatment == "Anonymization") |>
  arrange(top1) |>
  pull(region) |>
  as.character()

acc_region |>
  mutate(
    treatment = factor(treatment, levels = c("Anonymization", "Summarization")),
    region    = factor(region, levels = region_order)
  ) |>
  ggplot(aes(x = top1, y = region, colour = treatment)) +
  geom_vline(xintercept = mean(anon$correct_top1), linetype = "dotted",
             colour = "tomato", linewidth = 0.4) +
  geom_vline(xintercept = mean(summ$correct_top1), linetype = "dotted",
             colour = "steelblue", linewidth = 0.4) +
  geom_line(aes(group = region), colour = "grey70", linewidth = 0.6) +
  geom_point(size = 3) +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_colour_manual(
    values = c("Anonymization" = "tomato", "Summarization" = "steelblue"),
    name = NULL
  ) +
  labs(
    x        = "Top-1 re-identification accuracy",
    y        = NULL,
    title    = "Re-identification accuracy by region",
    subtitle = "Ordered by anonymization top-1. Lines connect the two treatments for each region."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

## 6. Flip countries: where the treatments disagree

Countries where one treatment is identifiable and the other is not reveal *what kind
of information* each treatment removes. The 30 countries appearing in both samples
are used for this comparison.

```{r flip-setup}
overlap_isos <- intersect(summ$iso, anon$iso)

country_summ <- summ |>
  filter(iso %in% overlap_isos) |>
  group_by(iso, country) |>
  summarise(
    summ_top1 = mean(correct_top1),
    summ_top3 = mean(correct_top3),
    n_summ    = n(),
    poly      = mean(v2x_polyarchy, na.rm = TRUE),
    regime    = first(as.character(regime)),
    region    = first(as.character(region)),
    .groups   = "drop"
  )

country_anon <- anon |>
  filter(iso %in% overlap_isos) |>
  group_by(iso) |>
  summarise(
    anon_top1 = mean(correct_top1),
    anon_top3 = mean(correct_top3),
    n_anon    = n(),
    .groups   = "drop"
  )

country_both <- country_summ |>
  left_join(country_anon, by = "iso") |>
  mutate(
    delta_top1 = summ_top1 - anon_top1,
    category   = case_when(
      anon_top1 == 1 & summ_top1 == 1  ~ "Both correct",
      anon_top3 == 0 & summ_top3 == 0  ~ "Both wrong",
      anon_top1 > summ_top1             ~ "Anon only",
      summ_top1 > anon_top1             ~ "Summ only",
      TRUE                              ~ "Tied"
    )
  )

cat("Countries in overlap:", nrow(country_both), "\n")
count(country_both, category)
```

```{r flip-plot}
#| fig-width: 7
#| fig-height: 5.5

country_both |>
  filter(category != "Tied") |>
  mutate(
    label    = glue("{iso}"),
    category = factor(category,
                      levels = c("Both correct", "Summ only", "Anon only", "Both wrong"))
  ) |>
  ggplot(aes(x = anon_top1, y = summ_top1, colour = category, label = label)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(size = 3, alpha = 0.85) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 20) +
  scale_x_continuous(labels = scales::percent, limits = c(-0.05, 1.05)) +
  scale_y_continuous(labels = scales::percent, limits = c(-0.05, 1.05)) +
  scale_colour_manual(
    values = c(
      "Both correct" = "darkgreen",
      "Summ only"    = "steelblue",
      "Anon only"    = "tomato",
      "Both wrong"   = "grey50"
    ),
    name = NULL
  ) +
  labs(
    x        = "Anonymization top-1 accuracy",
    y        = "Summarization top-1 accuracy",
    title    = "Re-identification accuracy: summarization vs. anonymization",
    subtitle = glue("Countries in both samples (n = {nrow(country_both)}). Points above diagonal = summarization harder to identify.")
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

### 6a. Anon-only: what details anonymization preserved that summarization lost

These countries are identifiable from anonymized text but not from summaries.
The model's reasoning reveals which specific cues survived name-stripping but
were erased by summarization.

```{r anon-only-table}
anon_only_isos <- country_both |>
  filter(category == "Anon only") |>
  pull(iso)

# Pull the anon record and the summ record(s) for each flip country
anon_reasons <- anon |>
  filter(iso %in% anon_only_isos, correct_top1) |>
  select(iso, country, indicator, reasoning) |>
  mutate(source = "Anonymization: why correct")

summ_reasons <- summ |>
  filter(iso %in% anon_only_isos, !correct_top1) |>
  select(iso, country, indicator, reasoning, top1_guess) |>
  mutate(source = glue("Summarization: wrong (guessed {top1_guess})")) |>
  select(-top1_guess)

bind_rows(anon_reasons, summ_reasons) |>
  arrange(iso, source) |>
  select(iso, country, source, reasoning) |>
  gt() |>
  tab_header(
    title    = "Cues retained by anonymization but erased by summarization",
    subtitle = glue("{length(anon_only_isos)} countries identifiable from anonymized text only")
  ) |>
  cols_label(
    iso = "ISO", country = "Country",
    source = "Treatment / result", reasoning = "Model's reasoning"
  ) |>
  cols_width(reasoning ~ px(450)) |>
  tab_style(
    style = cell_fill(color = "#fde8e8"),
    locations = cells_body(rows = str_starts(source, "Sum"))
  ) |>
  tab_style(
    style = cell_fill(color = "#e8f4e8"),
    locations = cells_body(rows = str_starts(source, "Anon"))
  )
```

### 6b. Summ-only: where summarization sharpens the signal

These countries are identifiable from summaries but not from anonymized text.
Summarization appears to highlight the distinctive regime-level signature that
gets obscured by the specificity of the original text.

```{r summ-only-table}
summ_only_isos <- country_both |>
  filter(category == "Summ only") |>
  pull(iso)

summ_reasons2 <- summ |>
  filter(iso %in% summ_only_isos, correct_top1) |>
  select(iso, country, indicator, reasoning) |>
  mutate(source = "Summarization: why correct")

anon_reasons2 <- anon |>
  filter(iso %in% summ_only_isos, !correct_top1) |>
  select(iso, country, indicator, reasoning, top1_guess) |>
  mutate(source = glue("Anonymization: wrong (guessed {top1_guess})")) |>
  select(-top1_guess)

bind_rows(summ_reasons2, anon_reasons2) |>
  arrange(iso, source) |>
  select(iso, country, source, reasoning) |>
  gt() |>
  tab_header(
    title    = "Countries easier to identify from summaries than anonymized text",
    subtitle = glue("{length(summ_only_isos)} countries: summarization top-1 correct, anonymization wrong")
  ) |>
  cols_label(
    iso = "ISO", country = "Country",
    source = "Treatment / result", reasoning = "Model's reasoning"
  ) |>
  cols_width(reasoning ~ px(450)) |>
  tab_style(
    style = cell_fill(color = "#e8f0fd"),
    locations = cells_body(rows = str_starts(source, "Sum"))
  ) |>
  tab_style(
    style = cell_fill(color = "#fde8e8"),
    locations = cells_body(rows = str_starts(source, "Anon"))
  )
```

---

## 7. Jointly identifiable and jointly unidentifiable countries

```{r joint-table}
country_both |>
  filter(category %in% c("Both correct", "Both wrong")) |>
  arrange(category, desc(poly)) |>
  select(category, iso, country, poly, regime, region) |>
  mutate(poly = round(poly, 2)) |>
  gt() |>
  tab_header(
    title    = "Countries identifiable (or not) under both treatments",
    subtitle = "Hard core: neither treatment can hide or reveal these countries"
  ) |>
  cols_label(
    category = "Category", iso = "ISO", country = "Country",
    poly = "Polyarchy", regime = "Regime", region = "Region"
  ) |>
  tab_style(
    style = cell_fill(color = "#d4edda"),
    locations = cells_body(rows = category == "Both correct")
  ) |>
  tab_style(
    style = cell_fill(color = "#f0f0f0"),
    locations = cells_body(rows = category == "Both wrong")
  )
```

---

## 8. Wrong-guess patterns

```{r wrong-guesses-summ}
summ |>
  filter(!correct_top1) |>
  count(top1_guess, sort = TRUE) |>
  slice_head(n = 15) |>
  gt() |>
  tab_header(
    title    = "Most common wrong top-1 guesses (summarization)",
    subtitle = glue("{sum(!summ$correct_top1)} incorrect top-1 predictions")
  ) |>
  cols_label(top1_guess = "Model's top-1 guess", n = "Count") |>
  tab_style(
    style     = cell_fill(color = "#fff3cd"),
    locations = cells_body(rows = n >= 3)
  )
```

```{r wrong-guesses-compare}
#| fig-width: 8
#| fig-height: 4

both |>
  filter(!correct_top1) |>
  count(treatment, top1_guess, sort = TRUE) |>
  group_by(treatment) |>
  slice_head(n = 10) |>
  ungroup() |>
  mutate(
    top1_guess = fct_reorder(top1_guess, n),
    treatment  = factor(treatment, levels = c("Anonymization", "Summarization"))
  ) |>
  ggplot(aes(x = n, y = top1_guess, fill = treatment)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(
    values = c("Anonymization" = "tomato", "Summarization" = "steelblue"),
    name = NULL
  ) +
  labs(
    x        = "Count of wrong top-1 predictions",
    y        = NULL,
    title    = "Most common wrong guesses by treatment",
    subtitle = "Top 10 per treatment"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

---

## 9. Summary: what the comparison implies for the paper

```{r summary-table}
tibble(
  Metric = c(
    "Top-1 accuracy — Summarization",
    "Top-1 accuracy — Anonymization",
    "Reduction from summarization",
    "Top-3 accuracy — Summarization",
    "Top-3 accuracy — Anonymization",
    "Reduction from summarization",
    "Countries identifiable under both treatments",
    "Countries identifiable only from anonymized text",
    "Countries identifiable only from summaries",
    "Countries identifiable under neither treatment"
  ),
  Value = c(
    scales::percent(mean(summ$correct_top1), 0.1),
    scales::percent(mean(anon$correct_top1), 0.1),
    scales::percent(mean(anon$correct_top1) - mean(summ$correct_top1), 0.1),
    scales::percent(mean(summ$correct_top3), 0.1),
    scales::percent(mean(anon$correct_top3), 0.1),
    scales::percent(mean(anon$correct_top3) - mean(summ$correct_top3), 0.1),
    as.character(sum(country_both$category == "Both correct")),
    as.character(sum(country_both$category == "Anon only")),
    as.character(sum(country_both$category == "Summ only")),
    as.character(sum(country_both$category == "Both wrong"))
  )
) |>
  gt() |>
  tab_header(title = "Comparative re-identification summary") |>
  cols_label(Metric = "Metric", Value = "Value")
```

Summarization reduces top-1 re-identification by roughly 13 percentage points
relative to anonymization (47% vs. 60%). The reduction is not uniform: it is
concentrated in mid-salience countries whose anonymized text contained distinctive
descriptive details that escaped name-stripping. The six "anon-only" countries
in the overlap sample share a common structure — each had a niche institutional or
demographic cue that survived anonymization (Suriname's Maroon communities, Latvia's
Russian-speaking minority, Georgia's breakaway regions, Nepal's post-civil-war
transitional justice bodies, Angola's oil sector, Lesotho's constitutional monarchy
plus army episode) but was abstracted away by summarization.

Two countries flip the other way (Poland, Iceland): summarization sharpened the
distinctive regime-level signature (democratic backsliding under EU friction for
Poland; no-military parliamentary republic for Iceland) that was obscured by the
event-level specificity of the anonymized source text.

The hard core — countries identifiable under both treatments — consists of cases
where the regime type and policy combination is globally singular: China's mass
detention of an ethnic minority, Venezuela's humanitarian collapse, Cuba's one-party
state, North Korea's family dynasty, the UAE's federation of semiautonomous
entities with oil revenues. No treatment can obscure these.

For the paper's identification strategy, the critical implication mirrors that of the
anonymization analysis: both treatments are weakest for exactly the countries where
country-identity anchoring is most plausible — those the model knows best. The 13 pp
reduction from summarization is driven by mid-salience countries where anonymization
left too much on the table, not by the globally iconic cases that dominate the
anchoring concern. The hard-core identifiable set is almost identical across both
treatments. Summarization therefore offers a more thorough treatment of the mid-tier
concern, but the ceiling — countries that cannot be de-identified — is set by the
same geopolitically salient cases.
