---
title: "Human MAE Floor: Rounding Floor and Leave-One-Out Baseline"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
execute:
  warning: false
  message: false
params:
  n_boot: 500
  years: !expr c(2019, 2023, 2024)
  min_coders: 2
---

AI MAE against the panel mean has no natural zero. Even a hypothetical oracle coder
constrained to integer ratings cannot always reproduce a fractional panel mean exactly,
and the panel mean itself is a noisy estimate from a small panel. This notebook computes
both floors from human data alone — no AI output required, so it can be finalized now:

1. **Rounding floor** — the error a hypothetical oracle rater would post if it always
   guessed `round(panel_mean)` exactly. A property of the panel means alone.
2. **Human leave-one-out (LOO) MAE** — how far a real human coder's rating falls from the
   panel mean of everyone else in their panel. Folds in panel-size noise as well as the
   rounding constraint, since humans code the same integer scale. This is the benchmark
   used in the agreement test and the intended reference line for the primary results
   figure.

Restricted to the 205 evaluable indicators (`config/indicator_sections.yaml` minus
`v2exl_legitideolcr`, which has no panel-mean data in any year — see
`preregistration/indicator-selection.md`). See `notes/loo-mae-computation.md` and
`preregistration/evaluation-metrics.md` for full construction and rationale.

---

## 1. Setup

```{r setup}
library(tidyverse)
library(rsample)
library(gt)
library(here)
library(yaml)

config <- read_yaml(here("config", "indicator_sections.yaml"))
eval_indicators <- names(config)
length(eval_indicators)

data_dir <- here("data", "processed")
```

---

## 2. Data loading

```{r load-data}
#| cache: true
#| cache.lazy: false

panel_means <- read_csv(file.path(data_dir, "panel_means.csv"), show_col_types = FALSE) |>
  filter(indicator %in% eval_indicators, year %in% params$years)

human_ratings <- read_csv(file.path(data_dir, "human_ratings.csv"), show_col_types = FALSE) |>
  filter(indicator %in% eval_indicators, year %in% params$years)
```

### Sanity check: 206 config indicators → 205 evaluable

```{r indicator-check}
missing_indicators <- setdiff(eval_indicators, unique(panel_means$indicator))
missing_indicators

stopifnot(identical(missing_indicators, "v2exl_legitideolcr"))
length(unique(panel_means$indicator))
```

---

## 3. Rounding floor

```{r rounding-floor}
rounding_floor <- panel_means |>
  mutate(rounding_error = abs(round(raw_mean) - raw_mean))

rounding_floor_summary <- rounding_floor |>
  group_by(year) |>
  summarise(
    mean_floor   = mean(rounding_error),
    median_floor = median(rounding_error),
    n_cyi        = n(),
    .groups      = "drop"
  )

rounding_floor_summary
```

### Bootstrap CI (CYI-level resampling, per year)

```{r rounding-floor-boot}
set.seed(42)

boot_mean <- function(df, col, times) {
  bootstraps(df, times = times, apparent = TRUE) |>
    mutate(stat = map_dbl(splits, \(s) mean(analysis(s)[[col]]))) |>
    summarise(
      mean_est = mean(stat[id == "Apparent"]),
      ci_lo    = quantile(stat, 0.025),
      ci_hi    = quantile(stat, 0.975)
    )
}

floor_boot <- rounding_floor |>
  group_by(year) |>
  group_modify(~ boot_mean(.x, "rounding_error", params$n_boot)) |>
  ungroup()

floor_boot
```

---

## 4. Human leave-one-out (LOO) MAE

For each CYI with at least `r params$min_coders` coders, compute each coder's
leave-one-out error — how far their rating falls from the panel mean with them removed —
using the O(n) shortcut (`(panel_sum - rating) / (n_coders - 1)`) rather than recomputing
the mean n times. See `notes/loo-mae-computation.md`.

```{r human-loo}
#| cache: true

human_loo <- human_ratings |>
  group_by(country_text_id, year, indicator) |>
  filter(n() >= params$min_coders) |>
  mutate(
    panel_sum = sum(rating),
    n_coders  = n(),
    loo_mean  = (panel_sum - rating) / (n_coders - 1),
    loo_error = abs(rating - loo_mean)
  ) |>
  summarise(
    human_loo_mae = mean(loo_error),
    n_coders      = first(n_coders),
    .groups       = "drop"
  )

human_loo_summary <- human_loo |>
  group_by(year) |>
  summarise(
    mean_loo_mae   = mean(human_loo_mae),
    median_loo_mae = median(human_loo_mae),
    n_cyi          = n(),
    .groups        = "drop"
  )

human_loo_summary
```

### Bootstrap CI (CYI-level resampling, per year)

```{r human-loo-boot}
set.seed(42)

loo_boot <- human_loo |>
  group_by(year) |>
  group_modify(~ boot_mean(.x, "human_loo_mae", params$n_boot)) |>
  ungroup()

loo_boot
```

---

## 5. Combined summary

```{r combined-summary}
combined <- floor_boot |>
  rename(rounding_floor = mean_est, floor_ci_lo = ci_lo, floor_ci_hi = ci_hi) |>
  left_join(
    loo_boot |> rename(human_loo_mae = mean_est, loo_ci_lo = ci_lo, loo_ci_hi = ci_hi),
    by = "year"
  ) |>
  left_join(rounding_floor_summary |> select(year, floor_n_cyi = n_cyi), by = "year") |>
  left_join(human_loo_summary |> select(year, loo_n_cyi = n_cyi), by = "year")

combined |>
  gt() |>
  tab_header(
    title    = "Human MAE floor: rounding floor vs. leave-one-out MAE",
    subtitle = glue::glue("205-indicator evaluable universe; bootstrap B = {params$n_boot}")
  ) |>
  fmt_number(columns = where(is.numeric), decimals = 3) |>
  fmt_integer(columns = c(floor_n_cyi, loo_n_cyi)) |>
  cols_label(
    year           = "Year",
    rounding_floor = "Rounding floor",
    floor_ci_lo    = "Floor 95% CI (lo)",
    floor_ci_hi    = "Floor 95% CI (hi)",
    human_loo_mae  = "Human LOO MAE",
    loo_ci_lo      = "LOO 95% CI (lo)",
    loo_ci_hi      = "LOO 95% CI (hi)",
    floor_n_cyi    = "N (CYI, floor)",
    loo_n_cyi      = "N (CYI, LOO ≥2 coders)"
  )
```

These numbers are the two reference points for the metrics write-up:

- The **rounding floor** is the theoretical minimum achievable by any integer-constrained
  rater — it belongs in `preregistration/evaluation-metrics.md` and
  `docs/experimental-design.md` as the answer to "why can't AI MAE be read against zero."
- The **human LOO MAE** (2019 row) is the intended reference line for Figure 1 and the
  benchmark for the registered agreement-test hypothesis (R10, `preregistration/key-hypotheses.md`).
  The 2023 row is the confirmatory replication value for the same test.

Both require no AI output and are therefore finalized as of this notebook — they will not
change when inference results land.
