---
title: "2019 bootstrap metrics — compute & prototype figures"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
params:
  n_boot: 2000
  year: 2019
  min_coders: 2
  runs_subdir: ""
---

This is the **2019 compute + prototype document**. Everything here is built on 2019 data;
the 2023/2024 holdout work is a separate document, so the selected model is never validated
on the data that selected it. The figure/table plan lives in `paper/figures-and-tables.md`.

This document owns the expensive bootstrap and serializes its results (§9) to
`data/derived/bootstrap_2019.rds`. The paper deliverables — `paper/results-2019.qmd`
(main-text figures + descriptive tables) and `paper/appendix-2019.qmd` (the formal prereg
tables) — load that bundle and never rerun the bootstrap. The figures below are working
prototypes kept here as a lab notebook; the clean re-homed versions live in `paper/`.

Primary metric is **AI MAE against the panel mean**, read against the human leave-one-out
(LOO) MAE — the typical error of a randomly held-out human coder against the rest of
their panel. Bootstrap CIs use `r params$n_boot` **country-clustered** resamples — countries
are drawn with replacement, each carrying its whole indicator block — because a country's
indicators are correlated and resampling them independently understates uncertainty.

---

## 1. Setup

```{r setup}
library(tidyverse)
library(rsample)
library(gt)
library(here)
library(glue)
library(patchwork)   # install.packages("patchwork") if missing — used for the 2-panel figures

# `params` is injected by knitr from the YAML header only at render time. Running chunks
# interactively there is no `params`, so define the same defaults when it's absent. At
# render the guard is skipped (knitr has already set params), so the file works both ways.
if (!exists("params")) {
  params <- list(n_boot = 2000, year = 2019, min_coders = 2, runs_subdir = "")
}

# Resolve the panel-member project root robustly, whatever the session's working dir is.
# First walk UP to a .git (wd inside the project); if that fails because the session is
# rooted in the parent workspace (v-dem-coding, which holds panel-member as a subdir),
# step DOWN into panel-member. Beats here()'s cached root, which pins to wherever `here`
# first loaded and is the usual cause of "dir.exists(runs_dir) is not TRUE".
find_panel_member_root <- function() {
  up <- tryCatch(rprojroot::find_root(rprojroot::is_git_root), error = function(e) NA_character_)
  if (!is.na(up)) return(up)
  down <- file.path(getwd(), "panel-member")
  if (dir.exists(file.path(down, ".git"))) return(down)
  stop("Could not locate the panel-member project root from working dir: ", getwd())
}
proj_root <- find_panel_member_root()

# Shared vocabulary + paired-delta machinery (model_labels, family_pal, canon_col,
# ft_diag, cell_series(), paired_delta()). Single source of truth for both this
# compute doc and the paper deliverables.
source(file.path(proj_root, "helpers", "bootstrap_helpers.R"))

data_dir    <- file.path(proj_root, "data", "processed")
runs_dir    <- file.path(proj_root, "data", "output", "runs", params$runs_subdir)
derived_dir <- file.path(proj_root, "data", "derived")
dir.create(derived_dir, showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Data loading

All JSONL files in `runs_dir` (one per model × condition). Each record carries
`country`, `year`, `indicator`, `condition`, `model_key`, `rating`.

```{r load-data}
#| cache: true
#| cache.lazy: false

if (!dir.exists(runs_dir)) {
  stop("runs dir not found:\n  ", runs_dir,
       "\n  (working dir: ", getwd(), ", project root: ", proj_root, ")",
       "\nDrop the 19 JSONL files in <project>/data/output/runs/ and confirm the root above.")
}
run_files <- list.files(runs_dir, pattern = "\\.jsonl$", full.names = TRUE)
if (length(run_files) == 0) stop("No .jsonl files in ", runs_dir)

# Records carry nested fields (sources, section_keys, tokens) and large text blobs
# (justification, raw_response); a naive fromJSON |> bind_rows() trips on the ragged nested
# arrays. Pull only the six scalar columns we need via stream_in, which reads NDJSON
# column-wise and tolerates the nested fields we then drop. Base runs are logged under the
# production vLLM key `llama-70b-local`; normalize the `-local` suffix back to `llama-70b`.
needed_cols <- c("country", "year", "indicator", "condition", "model_key", "rating")

read_run <- function(f) {
  con <- file(f, "r")
  on.exit(close(con))
  jsonlite::stream_in(con, verbose = FALSE) |>
    as_tibble() |>
    select(all_of(needed_cols))
}

ai_raw <- run_files |>
  map(read_run) |>
  bind_rows() |>
  mutate(model_key = str_remove(model_key, "-local$"))

# Human panel means — country_text_id, year, indicator, raw_mean, n_coders, theta_quintile
panel_means <- read_csv(file.path(data_dir, "panel_means.csv"), show_col_types = FALSE)

# Individual coder ratings — country_text_id, iso3, year, indicator, coder_id, rating
human_ratings <- read_csv(file.path(data_dir, "human_ratings.csv"), show_col_types = FALSE)
```

**Load diagnostic** — confirm all expected model × condition cells are present before
trusting anything downstream. The 2019 grid should show 19 cells: base (7 = 4 primary +
3 zero-shot ablation) and 12 FT (3 variants × 4 conditions).

```{r load-diagnostic}
ai_raw |>
  filter(year == params$year) |>
  count(model_key, condition) |>
  mutate(model = recode(model_key, !!!model_labels)) |>
  arrange(model_key, condition) |>
  gt() |>
  tab_header(
    title    = glue("Cells present in runs/ for {params$year}"),
    subtitle = glue("{n_distinct(paste(ai_raw$model_key, ai_raw$condition))} model×condition cells loaded")
  )
```

---

## 3. Metric construction

### Human LOO baseline (per CYI)

For each CYI with ≥ `r params$min_coders` coders, each coder's leave-one-out error is how
far their rating falls from the panel mean computed with them removed.

```{r human-loo}
#| cache: true

human_loo <- human_ratings |>
  filter(year == params$year) |>
  group_by(country_text_id, year, indicator) |>
  filter(n() >= params$min_coders) |>
  mutate(
    panel_sum = sum(rating),
    n_coders  = n(),
    loo_mean  = (panel_sum - rating) / (n_coders - 1),
    loo_error = abs(rating - loo_mean)
  ) |>
  summarise(
    human_loo_mae = mean(loo_error),
    n_coders      = first(n_coders),
    .groups       = "drop"
  )
```

### AI metrics (per CYI × model × condition)

AI error is against the canonical panel mean (`raw_mean`). Restricted to CYIs that carry
a human LOO floor, so AI and human are compared on the same cases.

```{r ai-metrics}
ai_metrics <- ai_raw |>
  filter(year == params$year) |>
  rename(country_text_id = country) |>
  inner_join(
    select(panel_means, country_text_id, year, indicator, raw_mean),
    by = c("country_text_id", "year", "indicator")
  ) |>
  inner_join(
    select(human_loo, country_text_id, year, indicator, human_loo_mae),
    by = c("country_text_id", "year", "indicator")
  ) |>
  mutate(
    ai_error = abs(rating - raw_mean), #calling ai_error here because not weighted yet
    col      = canon_col(condition),
    family   = recode(model_key, !!!model_family)
  )
```

---

## 4. Bootstrap CIs (grid-wide, reused by every figure)

Resample **countries** with replacement; for each draw compute AI MAE per cell and the human
LOO MAE, plus their difference. Countries are the independent unit — a country's ~200
indicators are correlated, so resampling country-year-indicators (CYIs) independently, as the
prereg specified, treats correlated draws as independent and yields anti-conservative
(too-narrow) level intervals. A country drawn *k* times carries its whole indicator block *k*
times — the per-country weight `w` (from the shared `country_boot_weights()` helper),
broadcast to all of that country's rows via an `inner_join`, so the aggregation is otherwise
unchanged. This block is deliberately general — every figure reads its cells and CIs off
`boot_ci` / `boot_results`. See `notes/bootstrap-clustering-figures-check.qmd` for the derivation.

```{r bootstrap}
#| cache: true
#| cache.lazy: false

# One row per unique CYI (shared human floor); the country universe is resampled below.
cyi_pool <- ai_metrics |>
  distinct(country_text_id, year, indicator, human_loo_mae)

# Shared country-clustered draw step: a (countries x n_boot+1) weight matrix, seed 42,
# column "Apparent" = the all-ones point-estimate draw. The persistence rail (§4b) reuses
# this same `W`, so the naive rail and the cell MAEs share one country universe.
W <- country_boot_weights(cyi_pool$country_text_id, params$n_boot, seed = 42)

boot_results <- map_dfr(colnames(W), \(draw) {
  # Per-country weight for this draw, broadcast to every one of the country's CYIs. `w > 0`
  # drops undrawn countries; a country drawn k times is weighted k (bootstrap multiplicity).
  country_w <- tibble(country_text_id = rownames(W), w = W[, draw]) |>
    filter(w > 0)
  ai_metrics |>
    inner_join(country_w, by = "country_text_id") |>
    group_by(model_key, condition) |>
    summarise(
      ai_mae    = weighted.mean(ai_error, w),
      human_mae = weighted.mean(human_loo_mae, w),
      diff      = weighted.mean(ai_error, w) - weighted.mean(human_loo_mae, w),
      .groups   = "drop"
    ) |>
    mutate(id = draw, .before = 1)
})

# Per-cell AI MAE point estimate (apparent draw) + percentile CI on the AI–human diff.
boot_ci <- boot_results |>
  group_by(model_key, condition) |>
  summarise(
    # Percentile CIs from the resamples (id != "Apparent"). Compute these FIRST, while
    # `ai_mae`/`diff` still refer to the full bootstrap columns — the point-estimate lines
    # below reuse those names, and any reference after the reassignment would collapse to a
    # scalar (the degenerate-CI bug).
    ai_lo     = quantile(ai_mae[id != "Apparent"], 0.025),
    ai_hi     = quantile(ai_mae[id != "Apparent"], 0.975),
    diff_lo   = quantile(diff[id != "Apparent"], 0.025),
    diff_hi   = quantile(diff[id != "Apparent"], 0.975),
    p_gt_zero = mean(diff[id != "Apparent"] > 0),
    # Point estimates from the apparent (full-sample) draw — defined last.
    ai_mae    = mean(ai_mae[id == "Apparent"]),
    diff      = mean(diff[id == "Apparent"]),
    .groups   = "drop"
  )

# Human LOO reference line: one value per bootstrap draw (constant across cells).
human_ref <- boot_results |>
  group_by(id) |>
  summarise(human_mae = first(human_mae), .groups = "drop") |>
  summarise(
    lo        = quantile(human_mae[id != "Apparent"], 0.025),
    hi        = quantile(human_mae[id != "Apparent"], 0.975),
    human_mae = mean(human_mae[id == "Apparent"])
  )
```

---

## 4b. Reference rails (persistence, SESOI) + paired-delta helper

Two things every figure below reuses. **Persistence** = the 2018 panel mean carried to 2019
and *rounded to an integer*, so it is an apples-to-apples "naive coder" against the
integer-emitting AI (~98% of 2019 CYIs have a 2018 value; the rail is computed over those).
**SESOI** = half the 2019 rounding floor (the irreducible integer-vs-fractional-mean error a
perfect rater still absorbs); any delta whose CI sits inside ±SESOI is "not substantively
meaningful" per the prereg.

```{r reference-rails}
#| cache: true
#| cache.lazy: false

# Prior-year copy-forward (year-1 → year), over the same >=2-coder CYI pool used everywhere
# else. Parametrized by params$year so the naive rail follows the analysis year (2018→2019,
# 2022→2023, …); the `mean_2018` column name is kept for continuity with downstream code.
pm18 <- panel_means |>
  filter(year == params$year - 1) |>
  select(country_text_id, indicator, mean_2018 = raw_mean)

persist_cyi <- cyi_pool |>
  left_join(
    panel_means |> filter(year == params$year) |>
      select(country_text_id, indicator, raw_mean),
    by = c("country_text_id", "indicator")
  ) |>
  left_join(pm18, by = c("country_text_id", "indicator")) |>
  mutate(persist_err = abs(round(mean_2018) - raw_mean))

# Rounding floor and SESOI (half the floor), on the same pool.
rounding_floor <- mean(abs(round(persist_cyi$raw_mean) - persist_cyi$raw_mean))
sesoi <- rounding_floor / 2

# Persistence rail: MAE on the same pool under the SAME country-clustered draws (reuse `W`),
# so the naive rail and the cell MAEs share one country universe. Weight each CYI by its
# country's draw count (broadcast by country_text_id); drop undrawn/NA rows per draw.
pv <- vapply(colnames(W), \(draw) {
  wt <- W[persist_cyi$country_text_id, draw]
  ok <- wt > 0 & !is.na(persist_cyi$persist_err)
  weighted.mean(persist_cyi$persist_err[ok], wt[ok])
}, numeric(1))
persist_ref <- tibble(
  lo          = quantile(pv[names(pv) != "Apparent"], 0.025),
  hi          = quantile(pv[names(pv) != "Apparent"], 0.975),
  persist_mae = pv[["Apparent"]]
)

glue("rounding floor = {round(rounding_floor,3)} · SESOI (±) = {round(sesoi,3)} · ",
     "persistence MAE = {round(persist_ref$persist_mae,3)} ",
     "[{round(persist_ref$lo,3)}, {round(persist_ref$hi,3)}]")
```

```{r delta-helper}
# `ft_diag`, `cell_series()` and `paired_delta()` now live in R/bootstrap_helpers.R
# (sourced in the setup chunk) so the paper docs share the same pairing logic. Their
# defaults read `boot_results` and `sesoi` from this environment, both defined above.
```

---

## 5. Figure 1 — Synthetic coder performance (the MAE landscape)

One panel, all seven cells on a common AI-MAE axis: the base condition ladder (Cb→Ev→An→Su)
and the fine-tuned diagonal, against the **Naive model** (persistence copy-forward) and
**human LOO** rails. Two dynamics read straight off it — the *centrifugal* base ladder (adding
evidence and stripping identity pushes the base model rightward, away from consensus) and the
*centripetal* fine-tuned cluster (training pulls the models onto the human LOO line).

```{r}
#| label: fig-landscape
#| fig-width: 8
#| fig-height: 4.8

land <- bind_rows(
  boot_ci |>
    filter(model_key == "llama-70b",
           condition %in% c("codebook", "evidence", "anonymized", "summarized")) |>
    mutate(family = "Base",
           row = recode(condition, codebook = "Codebook", evidence = "Raw Text",
                        anonymized = "Anonymized", summarized = "Summarized")),
  boot_ci |>
    semi_join(ft_diag, by = c("model_key", "condition")) |>
    mutate(family = "Fine-tuned",
           row = recode(model_key,
             "llama-70b-ft-raw"  = "FT · Raw Text",
             "llama-70b-ft-anon" = "FT · Anonymized",
             "llama-70b-ft-summ" = "FT · Summarized"))
) |>
  mutate(
    family = factor(family, levels = c("Base", "Fine-tuned")),
    # first level plots at the bottom: FT block below, base ladder above
    row = factor(row, levels = c("FT · Summarized", "FT · Anonymized", "FT · Raw Text",
                                 "Summarized", "Anonymized", "Raw Text", "Codebook"))
  )

ggplot(land, aes(ai_mae, row, color = family)) +
  # ±SESOI of the human LOO level: the band within which a synthetic cell's error is not
  # substantively distinguishable from the human coder's — same grey-band idiom SESOI uses in
  # Figs 2/2b/3. Reference lines and bands are capped just above the top row (y = 7.55) so the
  # two labels can sit in the clean strip above the plotting region (below the legend), with no
  # line or shading crossing the text. No CI band around human LOO — the dashed line alone marks it.
  annotate("rect", xmin = human_ref$human_mae - sesoi, xmax = human_ref$human_mae + sesoi,
           ymin = -Inf, ymax = 7.3, fill = "grey85", alpha = 0.55) +
  annotate("segment", x = persist_ref$persist_mae, xend = persist_ref$persist_mae,
           y = -Inf, yend = 7.3, linetype = "dashed", color = "grey40") +
  annotate("segment", x = human_ref$human_mae, xend = human_ref$human_mae,
           y = -Inf, yend = 7.3, linetype = "longdash", color = "grey40") +
  annotate("text", x = persist_ref$persist_mae, y = 6.8, label = "Naive model",
           color = "grey40", size = 2.9, hjust = -0.07, vjust = 0.5) +
  annotate("text", x = human_ref$human_mae, y = 7.5, label = "±SESOI of human LOO MAE",
           color = "grey40", size = 2.9, hjust = 0.5, vjust = 0.5) +
  # faint separator between the base block (top 4) and the FT block (bottom 3)
  geom_hline(yintercept = 3.5, color = "grey85", linewidth = 0.4) +
  geom_errorbar(aes(xmin = ai_lo, xmax = ai_hi), orientation = "y",
                width = 0.20, linewidth = 0.7) +
  geom_point(size = 2) +
  scale_color_manual(values = family_pal, name = NULL) +
  scale_y_discrete() +   # default (non-stretching) expansion; keeps the axis discrete
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.08), add = c(0.02, 0))) +   # small left room so the Naive line isn't flush against the edge
  coord_cartesian(clip = "off") +
  labs(x = "AI Mean Absolute Error", y = NULL,
       title = "Figure 1 — Synthetic coder performance on V-Dem ratings",
       subtitle = "2019 · mean absolute error relative to panel mean (lower = closer to consensus)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", legend.justification = "left",
        panel.grid.major.y = element_blank(), plot.title.position = "plot",
        plot.title = element_text(face = "bold"))
```

---

## 6. Figure 2 — Base model: reading the evidence or reading the country

The evidence-*centrifugal* dynamic, organized around the two anchors. **Top group** = each
condition vs. the **Codebook** prior (does adding text / de-identification move the base model
off its prior?). **Bottom group** = the de-identified conditions vs. full **Evidence** (does
removing identity hurt with the text held fixed?). The story: adding text barely moves it
(Ev−Cb inside SESOI); stripping *identity* is what pushes it out (An−Cb, Su−Cb, An−Ev beyond
SESOI). B4 (Su−An) is dropped to the appendix.

```{r}
#| label: fig-base-deltas
#| fig-width: 8
#| fig-height: 4.0

base_deltas <- bind_rows(
  paired_delta("llama-70b", "evidence",   "llama-70b", "codebook", "B1 · Ev−Cb"),
  paired_delta("llama-70b", "anonymized", "llama-70b", "codebook", "B2 · An−Cb"),
  paired_delta("llama-70b", "summarized", "llama-70b", "codebook", "Su−Cb (descr.)"),
  paired_delta("llama-70b", "anonymized", "llama-70b", "evidence", "B3 · An−Ev"),
  paired_delta("llama-70b", "summarized", "llama-70b", "evidence", "Su−Ev (descr.)")
) |>
  mutate(label = factor(label, levels = rev(c(
    "B1 · Ev−Cb", "B2 · An−Cb", "Su−Cb (descr.)", "B3 · An−Ev", "Su−Ev (descr.)"))))

ggplot(base_deltas, aes(est, label)) +
  annotate("rect", xmin = -sesoi, xmax = sesoi, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = 0.55) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  # separator between the "vs Codebook" group (top 3) and the "vs Evidence" group (bottom 2)
  geom_hline(yintercept = 2.5, color = "grey85", linewidth = 0.4) +
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y",
                width = 0.15, linewidth = 0.7, color = "#0072B2") +
  geom_point(size = 3, color = "#0072B2") +
  scale_x_continuous(expand = expansion(mult = c(0.06, 0.06))) +
  labs(x = "ΔMAE · right = higher error · grey band = ±SESOI", y = NULL,
       title = "Figure 2 — Base model: reading the evidence or reading the country",
       subtitle = glue("2019 · top: vs Codebook · bottom: vs Evidence · paired · SESOI ±{round(sesoi, 3)}")) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(), plot.title.position = "plot",
        plot.title = element_text(face = "bold"))
```

### B-series readout

```{r fig2-readout}
base_deltas |>
  transmute(
    Contrast = label,
    ΔMAE     = est,
    `95% CI` = glue("[{round(lo, 3)}, {round(hi, 3)}]"),
    `P(Δ>0)` = p_gt,
    `Beyond SESOI?` = if_else(sesoi_out, "yes", "no")
  ) |>
  gt() |>
  tab_header(
    title    = "Base-model identification contrasts",
    subtitle = glue("Paired CYI bootstrap · SESOI = ±{round(sesoi, 3)}")
  ) |>
  fmt_number(columns = c(ΔMAE, `P(Δ>0)`), decimals = 3)
```

---

## 6b. Figure 2b — Everything against the base:Cb prior (SESOI ruler)

A single ruler that collapses the Fig 2 / Fig 3-B idea: **base:Cb is the reference (0)** and
every other thing — the base de-identified conditions, the fine-tuned diagonal, and the human
coder — is its own paired ΔMAE from that pure-prior floor, read against ±SESOI. Reading top to
bottom: the base conditions fan *right* as identity is stripped (An, Su past SESOI), the FT
models sit near the prior, and the human coder marks where a real held-out rater lands relative
to the same anchor.

```{r}
#| label: fig-vs-prior-all
#| fig-width: 8
#| fig-height: 4.6

vs_cb_all <- bind_rows(
  paired_delta("llama-70b", "evidence",   "llama-70b", "codebook", "Ev") |> mutate(kind = "Base"),
  paired_delta("llama-70b", "anonymized", "llama-70b", "codebook", "An") |> mutate(kind = "Base"),
  paired_delta("llama-70b", "summarized", "llama-70b", "codebook", "Su") |> mutate(kind = "Base"),
  paired_delta("llama-70b-ft-raw",  "evidence-zeroshot",   "llama-70b", "codebook", "FT-raw")  |> mutate(kind = "Fine-tuned"),
  paired_delta("llama-70b-ft-anon", "anonymized-zeroshot", "llama-70b", "codebook", "FT-anon") |> mutate(kind = "Fine-tuned"),
  paired_delta("llama-70b-ft-summ", "summarized-zeroshot", "llama-70b", "codebook", "FT-summ") |> mutate(kind = "Fine-tuned")
)

# human LOO − base:Cb, paired per draw off boot_results (human_mae = the LOO for that draw).
loo_draws <- boot_results |>
  filter(model_key == "llama-70b", condition == "codebook") |>
  transmute(id, delta = human_mae - ai_mae)
loo_vs_cb <- tibble(
  label = "human LOO", kind = "Human coder",
  est   = loo_draws$delta[loo_draws$id == "Apparent"],
  lo    = quantile(loo_draws$delta[loo_draws$id != "Apparent"], 0.025),
  hi    = quantile(loo_draws$delta[loo_draws$id != "Apparent"], 0.975),
  p_gt  = mean(loo_draws$delta[loo_draws$id != "Apparent"] > 0),
  sesoi_out = abs(loo_draws$delta[loo_draws$id == "Apparent"]) > sesoi
)

vs_cb_all <- bind_rows(vs_cb_all, loo_vs_cb) |>
  mutate(
    kind  = factor(kind, levels = c("Base", "Fine-tuned", "Human coder")),
    # first level plots at the bottom: human at the very bottom, FT block, then base ladder on top
    label = factor(label, levels = rev(c("Ev", "An", "Su",
                                         "FT-raw", "FT-anon", "FT-summ", "human LOO")))
  )

ggplot(vs_cb_all, aes(est, label, color = kind)) +
  annotate("rect", xmin = -sesoi, xmax = sesoi, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = 0.55) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  # separators: base ladder (top 3) | FT block (middle 3) | human coder (bottom 1)
  geom_hline(yintercept = 4.5, color = "grey85", linewidth = 0.4) +
  geom_hline(yintercept = 1.5, color = "grey85", linewidth = 0.4) +
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y", width = 0.18, linewidth = 0.7) +
  geom_point(size = 3) +
  scale_color_manual(values = c(family_pal, "Human coder" = "grey25"), name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0.04, 0.06))) +
  labs(x = "ΔMAE vs. base:Cb · right = more divergent than the pure prior · grey band = ±SESOI",
       y = NULL,
       title = "Figure 2b — Everything against the base:Cb prior",
       subtitle = glue("2019 · base conditions, FT diagonal & human coder vs base:Cb · paired · SESOI ±{round(sesoi, 3)}")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", legend.justification = "left",
        panel.grid.major.y = element_blank(), plot.title.position = "plot",
        plot.title = element_text(face = "bold"))
```

```{r fig2b-readout}
vs_cb_all |>
  arrange(desc(as.integer(label))) |>
  transmute(
    Cell     = label,
    Family   = kind,
    ΔMAE     = est,
    `95% CI` = glue("[{round(lo, 3)}, {round(hi, 3)}]"),
    `P(Δ>0)` = p_gt,
    `Beyond SESOI?` = if_else(sesoi_out, "yes", "no")
  ) |>
  gt() |>
  tab_header(
    title    = "Divergence from base:Cb",
    subtitle = glue("Paired CYI bootstrap · SESOI = ±{round(sesoi, 3)}")
  ) |>
  fmt_number(columns = c(ΔMAE, `P(Δ>0)`), decimals = 3)
```

**Reading these as preliminary.** Both figures are descriptive, single-year (2019) estimates on
a metric — static MAE against the panel mean — that rewards central tendency: the naive
copy-forward sits closest of all, and the pool is dominated by country-years that barely move
between waves, where no independent rater can beat a copy-forward. The question of whether the
synthetic coders earn their keep on the country-years that actually *move* — where a persistence
copier is structurally blind and independent reading should pay — is the one **R2** decides, with
a pre-registered movement flag rather than an endogenous split. The 2019 movement-bin split that
motivates this (exploratory only) is the |Δ| table in
`notes/stage1-preliminary-observations.md` ("the reversal is a stable-cell artifact").

**The resampling unit.** CIs throughout are from a **country-clustered** bootstrap: countries
are resampled with replacement, each carrying its whole ~200-indicator block. This is the honest
independent unit — a country's indicators are correlated, so the prereg's independent CYI resample
treated them as more independent than they are and produced anti-conservative (too-narrow) level
intervals. The switch leaves point estimates and cell ordering untouched; every SESOI verdict
across Figures 1–3 is preserved, with only two non-registered adjacent-pair distinctions (An vs.
Su, Su vs. FT-summ) losing statistical resolution. The full CYI-vs-clustered comparison is worked
through in `notes/bootstrap-clustering-figures-check.qmd`.

---

## 7. Figure 3 — Divergence from the reciting prior (base:Cb)

Two candidates, both measuring how far each thing sits from the pure-prior floor (base:Cb = 0),
paired within-draw, with the SESOI band. **3-A** puts the base de-identified conditions and the
FT diagonal on one ruler: de-identifying the *base* model flings it past the prior (An, Su beyond
SESOI), while the fine-tuned models stay close to it. **3-B** replaces the base conditions with
the **human coder** as a benchmark point — the FT models bracket the human coder's own divergence
from the prior (FT-raw ≈ human, FT-anon below, FT-summ above). *(The human point landing near the
SESOI edge is a numerical coincidence — three unrelated quantities — not a result; do not narrate
it.)*

### Figure 3-A — base conditions and FT models vs. base:Cb

```{r}
#| label: fig-vs-prior
#| fig-width: 8
#| fig-height: 4.2

vs_cb <- bind_rows(
  paired_delta("llama-70b", "evidence",   "llama-70b", "codebook", "Ev")      |> mutate(kind = "Base"),
  paired_delta("llama-70b", "anonymized", "llama-70b", "codebook", "An")      |> mutate(kind = "Base"),
  paired_delta("llama-70b", "summarized", "llama-70b", "codebook", "Su")      |> mutate(kind = "Base"),
  paired_delta("llama-70b-ft-raw",  "evidence-zeroshot",   "llama-70b", "codebook", "FT-raw")  |> mutate(kind = "Fine-tuned"),
  paired_delta("llama-70b-ft-anon", "anonymized-zeroshot", "llama-70b", "codebook", "FT-anon") |> mutate(kind = "Fine-tuned"),
  paired_delta("llama-70b-ft-summ", "summarized-zeroshot", "llama-70b", "codebook", "FT-summ") |> mutate(kind = "Fine-tuned")
) |>
  mutate(
    kind  = factor(kind, levels = c("Base", "Fine-tuned")),
    label = factor(label, levels = rev(c("Ev", "An", "Su", "FT-raw", "FT-anon", "FT-summ")))
  )

ggplot(vs_cb, aes(est, label, color = kind)) +
  annotate("rect", xmin = -sesoi, xmax = sesoi, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = 0.55) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = 3.5, color = "grey85", linewidth = 0.4) +
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y", width = 0.18, linewidth = 0.7) +
  geom_point(size = 3) +
  scale_color_manual(values = family_pal, name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0.04, 0.06))) +
  labs(x = "ΔMAE vs. base:Cb · right = more divergent than the pure prior · grey band = ±SESOI",
       y = NULL,
       title = "Figure 3-A — Divergence from the reciting prior",
       subtitle = glue("2019 · base de-id conditions (top) & FT diagonal (bottom) vs base:Cb · SESOI ±{round(sesoi, 3)}")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", legend.justification = "left",
        panel.grid.major.y = element_blank(), plot.title.position = "plot",
        plot.title = element_text(face = "bold"))
```

### Figure 3-B — FT models vs. the human coder, both anchored to base:Cb

```{r}
#| label: fig-vs-prior-human
#| fig-width: 8
#| fig-height: 3.6

ft_vs_cb <- bind_rows(
  paired_delta("llama-70b-ft-raw",  "evidence-zeroshot",   "llama-70b", "codebook", "FT-raw")  |> mutate(kind = "Fine-tuned"),
  paired_delta("llama-70b-ft-anon", "anonymized-zeroshot", "llama-70b", "codebook", "FT-anon") |> mutate(kind = "Fine-tuned"),
  paired_delta("llama-70b-ft-summ", "summarized-zeroshot", "llama-70b", "codebook", "FT-summ") |> mutate(kind = "Fine-tuned")
)

# human LOO − base:Cb, paired per draw off boot_results (human_mae = the LOO for that draw).
loo_draws <- boot_results |>
  filter(model_key == "llama-70b", condition == "codebook") |>
  transmute(id, delta = human_mae - ai_mae)
loo_vs_cb <- tibble(
  label = "human LOO", kind = "Human coder",
  est   = loo_draws$delta[loo_draws$id == "Apparent"],
  lo    = quantile(loo_draws$delta[loo_draws$id != "Apparent"], 0.025),
  hi    = quantile(loo_draws$delta[loo_draws$id != "Apparent"], 0.975),
  p_gt  = mean(loo_draws$delta[loo_draws$id != "Apparent"] > 0),
  sesoi_out = abs(loo_draws$delta[loo_draws$id == "Apparent"]) > sesoi
)

vs_cb2 <- bind_rows(ft_vs_cb, loo_vs_cb) |>
  mutate(label = factor(label, levels = c("FT-anon", "FT-raw", "human LOO", "FT-summ")))

ggplot(vs_cb2, aes(est, label, color = kind)) +
  annotate("rect", xmin = -sesoi, xmax = sesoi, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = 0.55) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y", width = 0.16, linewidth = 0.7) +
  geom_point(size = 3.2) +
  scale_color_manual(values = c("Fine-tuned" = "#E69F00", "Human coder" = "grey25"), name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0.04, 0.06))) +
  labs(x = "ΔMAE vs. base:Cb · right = more divergent than the pure prior · grey band = ±SESOI",
       y = NULL,
       title = "Figure 3-B — Fine-tuned models vs. the human coder",
       subtitle = glue("2019 · distance from base:Cb · FT-raw ≈ human · FT-anon below · FT-summ above · SESOI ±{round(sesoi, 3)}")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", legend.justification = "left",
        panel.grid.major.y = element_blank(), plot.title.position = "plot",
        plot.title = element_text(face = "bold"))
```

---

## 8. Appendix figures (moved out of the main text)

Secondary tests kept for the appendix and discussed in the main-text prose: **F1**
(weight- vs. prompt-calibration), the **selection-rule readout**, and the **A3 anchoring probe**
(B4 · Su−An can be added from `paired_delta` if wanted). All read off the same `boot_ci` /
`paired_delta` layer.

### Appendix — F1: fine-tuning vs. few-shot (matched conditions)

```{r}
#| label: fig-f1
#| fig-width: 8
#| fig-height: 3.0

f1 <- bind_rows(
  paired_delta("llama-70b-ft-raw",  "evidence-zeroshot",   "llama-70b", "evidence",   "Raw Text"),
  paired_delta("llama-70b-ft-anon", "anonymized-zeroshot", "llama-70b", "anonymized", "Anonymized"),
  paired_delta("llama-70b-ft-summ", "summarized-zeroshot", "llama-70b", "summarized", "Summarized")
) |>
  mutate(label = factor(label, levels = rev(c("Raw Text", "Anonymized", "Summarized"))))

ggplot(f1, aes(est, label)) +
  annotate("rect", xmin = -sesoi, xmax = sesoi, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = 0.55) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y",
                width = 0.15, linewidth = 0.7, color = "#E69F00") +
  geom_point(size = 3, color = "#E69F00") +
  scale_x_continuous(expand = expansion(mult = c(0.08, 0.08))) +
  labs(x = "ΔMAE (FT − base) · left = fine-tuning reduces error · grey band = ±SESOI", y = NULL,
       title = "Appendix — F1: fine-tuning vs. few-shot (matched)",
       subtitle = glue("2019 · registered F1 · SESOI ±{round(sesoi, 3)}")) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(), plot.title.position = "plot")
```

```{r f1-readout}
f1 |>
  transmute(Condition = label, `ΔMAE (FT − base)` = est,
            `95% CI` = glue("[{round(lo, 3)}, {round(hi, 3)}]"),
            `P(Δ>0)` = p_gt,
            `Beyond SESOI?` = if_else(sesoi_out, "yes", "no")) |>
  gt() |>
  tab_header(title = "F1 — fine-tuning vs. few-shot, matched conditions",
             subtitle = glue("Negative = fine-tuning reduces error · SESOI ±{round(sesoi, 3)}")) |>
  fmt_number(columns = c(`ΔMAE (FT − base)`, `P(Δ>0)`), decimals = 3)
```

### Appendix — selection-rule readout

```{r selection-readout}
bind_rows(
  tibble(model_key = "llama-70b", condition = "evidence", row = "Base : Ev"),
  transmute(ft_diag, model_key, condition, row)
) |>
  inner_join(boot_ci, by = c("model_key", "condition")) |>
  transmute(Cell = row, `AI MAE` = ai_mae,
            `95% CI` = glue("[{round(ai_lo, 3)}, {round(ai_hi, 3)}]"),
            `Δ vs human` = diff, `P(Δ>0)` = p_gt_zero) |>
  arrange(`AI MAE`) |>
  gt() |>
  tab_header(title = "Diagonal selection readout",
             subtitle = "Lowest AI MAE in its own condition = carried model (ties → base)") |>
  fmt_number(columns = c(`AI MAE`, `Δ vs human`, `P(Δ>0)`), decimals = 3)
```

### Appendix — A3 anchoring probe (codebook only)

All four models on the **codebook** condition only. A fine-tuned model that matches the base:Cb
anchor with no evidence to read has baked a country→rating shortcut into its weights; higher
codebook MAE = *less* reciting (the desirable direction). Dotted line = base:Cb anchor.

```{r}
#| label: fig-ft-anchor
#| fig-width: 8
#| fig-height: 3.6
anchor <- boot_ci |>
  filter(model_key == "llama-70b", condition == "codebook") |>
  pull(ai_mae)

fig_anchor <- boot_ci |>
  filter(condition == "codebook") |>
  mutate(model_label = recode(model_key, !!!model_labels),
         family      = recode(model_key, !!!model_family))

ggplot(fig_anchor, aes(ai_mae, fct_reorder(model_label, -ai_mae), color = family)) +
  annotate("rect", xmin = human_ref$lo, xmax = human_ref$hi,
           ymin = -Inf, ymax = Inf, fill = "grey70", alpha = 0.25) +
  geom_vline(xintercept = human_ref$human_mae, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = anchor, linetype = "dotted", color = "#0072B2") +
  annotate("text", x = human_ref$human_mae, y = Inf, label = "human LOO",
           color = "grey40", size = 3, hjust = -0.05, vjust = 1.5) +
  geom_errorbar(aes(xmin = ai_lo, xmax = ai_hi), orientation = "y",
                width = 0.18, linewidth = 0.7) +
  geom_point(size = 3) +
  scale_color_manual(values = family_pal, name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.12))) +
  coord_cartesian(clip = "off") +
  labs(x = "AI MAE vs. panel mean, codebook only", y = NULL,
       title = "Appendix — A3 anchoring probe (codebook only)",
       subtitle = glue("2019 · dotted line = base:Cb anchor ({round(anchor, 3)})")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", legend.justification = "left",
        panel.grid.major.y = element_blank(), plot.title.position = "plot")
```

---

## 9. Movement-bin descriptive table (exploratory)

The 2018→2019 |Δ| split that motivates making R2 central: on the same ≥2-coder CYI pool used
everywhere else, split cells by how far the panel mean actually moved between waves, and read
persistence, base:Cb, and base:Ev MAE within each bin. **Exploratory / hypothesis-generating
only** — the bin edge is the realized change, so persistence's per-bin MAE is mechanical (it is
just the bin's mean |Δ|). The confirmatory movement test is R2 on 2023 with the pre-fixed ERT
flag; do not retrofit this as a confirmatory 2019 result. Serialized into the bundle so the
paper's descriptive tables read it without recomputing.

```{r movement-bins}
# Continuous 2018→2019 move per CYI, restricted to cells with a 2018 predecessor.
abs_delta_cyi <- persist_cyi |>
  filter(!is.na(mean_2018)) |>
  transmute(country_text_id, year, indicator,
            persist_err,                         # integer-fair: |round(2018) − 2019|
            abs_delta = abs(raw_mean - mean_2018))

# Base model's per-CYI error on the two identity-present conditions.
base_err_wide <- ai_metrics |>
  filter(model_key == "llama-70b", condition %in% c("codebook", "evidence")) |>
  select(country_text_id, year, indicator, condition, ai_error) |>
  pivot_wider(names_from = condition, values_from = ai_error)

movement_bins <- abs_delta_cyi |>
  inner_join(base_err_wide, by = c("country_text_id", "year", "indicator")) |>
  mutate(bin = cut(abs_delta, breaks = c(0, 0.25, 0.5, 1, 2, Inf), right = FALSE,
                   labels = c("[0.00, 0.25)", "[0.25, 0.50)", "[0.50, 1.00)",
                              "[1.00, 2.00)", "[2.00, +)"))) |>
  group_by(bin) |>
  summarise(
    n            = n(),
    n_ev         = sum(!is.na(evidence)),  # evidence covers fewer cells than codebook (needs a source doc)
    persist_cont = mean(abs_delta),        # = bin mean |Δ| (matches the notes' persist column)
    persist_round= mean(persist_err),      # integer-fair persistence vs the integer-emitting AI
    # codebook covers the whole pool; evidence is averaged over its available cells (n_ev),
    # so the two columns are each over their own coverage, matching the notes' framing.
    base_cb_mae  = mean(codebook, na.rm = TRUE),
    base_ev_mae  = mean(evidence, na.rm = TRUE),
    .groups      = "drop"
  )

movement_bins |>
  gt() |>
  tab_header(title = "2018→2019 movement bins (≥2-coder pool)",
             subtitle = "Exploratory. persist_cont = bin mean |Δ|; persist_round is integer-fair vs AI") |>
  fmt_number(columns = c(persist_cont, persist_round, base_cb_mae, base_ev_mae), decimals = 3)
```

---

## 10. Serialize the bundle for the paper documents

Everything the paper deliverables need, written to one gitignored `.rds` (`*.rds` is ignored
repo-wide). `paper/results-2019.qmd` and `paper/appendix-2019.qmd` load this and never rerun
the bootstrap; the static vocabulary and `paired_delta()` come from `R/bootstrap_helpers.R`,
which those docs source directly.

```{r serialize-bundle}
bundle_path <- file.path(derived_dir, glue("bootstrap_{params$year}.rds"))

saveRDS(
  list(
    boot_ci        = boot_ci,         # per-cell AI MAE point + CIs
    boot_results   = boot_results,    # per-draw per-cell — required by paired_delta()
    human_ref      = human_ref,       # human LOO reference line + CI
    persist_ref    = persist_ref,     # persistence (naive) reference line + CI
    sesoi          = sesoi,           # smallest effect size of interest (± band)
    rounding_floor = rounding_floor,  # integer-vs-fractional-mean floor
    movement_bins  = movement_bins    # exploratory 2018→2019 |Δ| split
  ),
  bundle_path
)

glue("bundle written: {bundle_path} ({round(file.size(bundle_path)/1024)} KB)")
```
