---
title: "2019 regime-transition split (R2) — compute & formal tests"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: true
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
params:
  n_boot: 2000
  year: 2019
---

Formal **R2 regime-transition** analysis (hypothesis R2 in
`preregistration/preregistration-draft.qmd`; strategy: `notes/alternative-test-strategy.md`;
measurement design: `preregistration/mechanism-test-design.md`). R2 asks whether the **evidence
gain** — the paired MAE change from the codebook condition to a text condition — is *more
negative* (evidence helps more) on country-years undergoing a regime transition than on stable
ones. The registered mechanism: the model's frozen country prior is stale exactly where the
country is moving, so the text describing current conditions carries more marginal information
there.

**This document is 2019 only, and this is a within-year detection test, not a staleness test.**
With a one-year gap after the 2016–2018 fine-tuning window (and a base model whose pretraining
runs through Dec 2023), no prior has had time to go *stale* for 2019. What a 2019 transition CYI
still tests is local: in a transition year the panel mean has moved off the country's baseline,
and the question is whether the model reads the text to catch that this year is anomalous for the
country rather than defaulting to its reputation prior. The multi-year design (accumulated
transitions in the 2023 pool — lag vs. holdout) is deferred and noted at the end; **do not build
2023 machinery here.**

Three movement instruments, a ladder of the same test (see `notes/stage1-preliminary-observations.md`):

- **Endogenous within-CYI |Δ| (background only).** Bin CYIs by |Δ| of *the scored indicator's own*
  2018→2019 panel mean. Circular — the bin edge is the realized change in the very quantity MAE is
  scored against — so it is reported as background, not a test. This reproduces the Python table in
  the stage-1 note in R.
- **ERT episode flag (registered binary treatment).** V-Dem Episodes of Regime Transformation
  (v15): a country-year is transition-adjacent if it sits in a democratization or autocratization
  episode. External to the scored indicator, so non-circular.
- **|Δv2x_polyarchy| (continuous moderator).** Year-over-year change in the *aggregate*
  polyarchy index — one level removed from the scored indicator, for statistical resolution.

The R2-style quantity is the **difference-in-differences**: is the evidence gain (base Δ(Ev−Cb),
Δ(An−Cb), Δ(Su−Cb), and each FT model's diagonal − own codebook) more negative in transition CYIs
than in stable CYIs? R2 is registered on 2023; on 2019 these are exploratory, additional analyses
in the same spirit, run on both the base conditions and the FT diagonal.

**Country-clustered resampling.** The ERT flag is assigned at the country-year level, so all ~205
indicators of a transition country share it; resampling CYIs independently would treat them as
independent and *understate* uncertainty on the split — most severely for the movement instruments,
whose entire signal is between-country. This document therefore resamples **countries** throughout
— the §5 registered/DiD engine, the Figure 2 ribbon, the Table 5 level slopes, and the §7 horse
race alike — consistent with the country-clustered switch now used in `06`/`07`/`08`. The registered
confirmatory R2 lives on 2023 (a separate document); these 2019 analyses are exploratory, where
country clustering is the honest default.

## 1. Setup

```{r setup}
library(tidyverse)
library(rsample)
library(gt)
library(glue)

if (!exists("params")) params <- list(n_boot = 2000, year = 2019)

find_panel_member_root <- function() {
  up <- tryCatch(rprojroot::find_root(rprojroot::is_git_root), error = function(e) NA_character_)
  if (!is.na(up)) return(up)
  down <- file.path(getwd(), "panel-member")
  if (dir.exists(file.path(down, ".git"))) return(down)
  stop("Could not locate the panel-member project root from working dir: ", getwd())
}
proj_root   <- find_panel_member_root()
source(file.path(proj_root, "helpers", "bootstrap_helpers.R"))
data_dir    <- file.path(proj_root, "data", "processed")
runs_dir    <- file.path(proj_root, "data", "output", "runs")
derived_dir <- file.path(proj_root, "data", "derived")

# SESOI is fixed by the main analysis; carry it over so every document bands on the identical
# value. The R2 quantities (evidence gains, DiDs) are on the rating-point axis, so ±SESOI is a
# valid "substantively negligible" band here.
sesoi <- readRDS(file.path(derived_dir, "bootstrap_2019.rds"))$sesoi
```

## 2. Data loading

Three sources: the JSONL runs, the panel means (also carries the 2018 predecessor for
persistence), and the ERT v15 country-year file (episode flags + `v2x_polyarchy` for both the
binary and continuous instruments). No coder-level ratings — the registered R2 is a panel-mean
MAE test, so this runs on the full confirmatory pool (no coder floor), like `08`.

```{r load-data}
#| cache: true
#| cache.lazy: false
needed_cols <- c("country", "year", "indicator", "condition", "model_key", "rating")
read_run <- function(f) {
  con <- file(f, "r"); on.exit(close(con))
  jsonlite::stream_in(con, verbose = FALSE) |> as_tibble() |> select(all_of(needed_cols))
}
ai_raw <- list.files(runs_dir, pattern = "\\.jsonl$", full.names = TRUE) |>
  map(read_run) |> bind_rows() |> mutate(model_key = str_remove(model_key, "-local$"))

panel_means <- read_csv(file.path(data_dir, "panel_means.csv"), show_col_types = FALSE)

# ERT v15 (symlink → shared/vdem-data/ert_v15.csv). The leading unnamed row-index column reads
# in as `...1`; drop it. Keep only what the two instruments need.
ert <- read_csv(file.path(data_dir, "ert.csv"), show_col_types = FALSE) |>
  select(country_text_id, year, v2x_polyarchy,
         dem_ep, aut_ep, dem_ep_start_year, aut_ep_start_year)
```

## 3. Metric construction

### The transition instruments (country-year level)

```{r instruments}
# Binary ERT flag for the evaluation year: in a democratization OR autocratization episode.
# `onset` (episode start year) is the sharper variant reported alongside as robustness.
ert_flag <- ert |>
  filter(year == params$year) |>
  transmute(country_text_id,
            in_ep = (dem_ep == 1) | (aut_ep == 1),
            onset = (dem_ep == 1 & year == dem_ep_start_year) |
                    (aut_ep == 1 & year == aut_ep_start_year)) |>
  mutate(across(c(in_ep, onset), \(x) replace_na(x, FALSE)))

# Continuous moderator: |Δ v2x_polyarchy| from the prior year to the evaluation year, taken from
# the ERT file itself (one value per country-year, shared across that country's indicators).
dpoly <- ert |>
  filter(year %in% c(params$year - 1L, params$year)) |>
  select(country_text_id, year, v2x_polyarchy) |>
  pivot_wider(names_from = year, values_from = v2x_polyarchy,
              names_prefix = "p") |>
  transmute(country_text_id,
            dpoly = abs(.data[[paste0("p", params$year)]] -
                        .data[[paste0("p", params$year - 1L)]]))
```

### Base-model per-CYI errors (wide) + persistence

```{r base-errors}
# The seven Figure-1 cells — base ladder (few-shot Cb/Ev/An/Su) + the FT diagonal (each FT
# variant on the input it was built for, zero-shot) — plus each FT model's own codebook baseline
# (ft_*_cb), needed to form the FT evidence gains (diagonal − own codebook) that parallel the
# base Ev/An/Su − Cb gains. Short labels double as column names.
cell_map <- tibble::tribble(
  ~lab,          ~model_key,           ~condition,
  "cb",          "llama-70b",          "codebook",
  "ev",          "llama-70b",          "evidence",
  "an",          "llama-70b",          "anonymized",
  "su",          "llama-70b",          "summarized",
  "ft_raw",      "llama-70b-ft-raw",   "evidence-zeroshot",
  "ft_anon",     "llama-70b-ft-anon",  "anonymized-zeroshot",
  "ft_summ",     "llama-70b-ft-summ",  "summarized-zeroshot",
  "ft_raw_cb",   "llama-70b-ft-raw",   "codebook",
  "ft_anon_cb",  "llama-70b-ft-anon",  "codebook",
  "ft_summ_cb",  "llama-70b-ft-summ",  "codebook")

pm_year <- panel_means |> filter(year == params$year) |>
  select(country_text_id, indicator, raw_mean)

# 2018 predecessor mean → integer-fair persistence (round the carry-forward, since the AI emits
# integers), same construction as 06 §9 / the stage-1 note.
pm_prev <- panel_means |> filter(year == params$year - 1L) |>
  select(country_text_id, indicator, mean_prev = raw_mean)

# Per-CYI absolute error for all seven cells (full confirmatory pool, no coder floor).
cell_wide <- ai_raw |>
  filter(year == params$year) |>
  rename(country_text_id = country) |>
  inner_join(cell_map, by = c("model_key", "condition")) |>
  inner_join(pm_year, by = c("country_text_id", "indicator")) |>
  mutate(err = abs(rating - raw_mean)) |>
  select(country_text_id, indicator, lab, err) |>
  pivot_wider(names_from = lab, values_from = err)

# Human leave-one-out MAE per CYI (≥2-coder cells; NA for singletons) — the held-out-coder
# reference row for Table 2, same LOO construction as 06.
human_ratings <- read_csv(file.path(data_dir, "human_ratings.csv"), show_col_types = FALSE)
human_loo <- human_ratings |>
  filter(year == params$year) |>
  group_by(country_text_id, indicator) |>
  filter(n() >= 2) |>
  summarise(hloo = mean(abs(rating - (sum(rating) - rating) / (n() - 1))), .groups = "drop")

# One row per CYI: the seven cell errors, the base evidence-gain deltas, the transition flags,
# the continuous moderator, and integer-fair persistence error.
cyi <- cell_wide |>
  left_join(pm_year, by = c("country_text_id", "indicator")) |>
  left_join(pm_prev, by = c("country_text_id", "indicator")) |>
  left_join(human_loo, by = c("country_text_id", "indicator")) |>
  left_join(ert_flag, by = "country_text_id") |>
  left_join(dpoly,    by = "country_text_id") |>
  mutate(
    d_ev = ev - cb, d_an = an - cb, d_su = su - cb,   # base evidence gains (paired, per CYI)
    # FT evidence gains: each FT model's diagonal condition minus its own codebook baseline.
    d_ftraw = ft_raw - ft_raw_cb, d_ftanon = ft_anon - ft_anon_cb, d_ftsumm = ft_summ - ft_summ_cb,
    persist_err = abs(round(mean_prev) - raw_mean),
    abs_delta   = abs(raw_mean - mean_prev),          # endogenous within-CYI move (background)
    in_ep = replace_na(in_ep, FALSE),
    onset = replace_na(onset, FALSE)
  ) |>
  mutate(.rowid = row_number())   # stable resample index (tibbles drop rownames)

glue("pool: {nrow(cyi)} CYIs · in-episode {sum(cyi$in_ep)} ({round(100*mean(cyi$in_ep),1)}%) · ",
     "onset {sum(cyi$onset)} · with 2018 predecessor {sum(!is.na(cyi$mean_prev))} · ",
     "with dpoly {sum(!is.na(cyi$dpoly))}")
```

## 4. Table 1 — Background: endogenous within-CYI |Δ| bins

*Reproduces the stage-1 note's movement table in R.* Bin CYIs by how far **the scored
indicator's own** panel mean moved 2018→2019, and read persistence and base MAE within each bin.
**Background only, not a test:** the bin edge is the realized change in the quantity MAE scores
against, so persistence's per-bin MAE is mechanical (≈ the bin's mean |Δ|). This is exactly the
circularity the ERT and |Δpolyarchy| instruments below avoid.

```{r tbl-endogenous}
endog <- cyi |>
  filter(!is.na(mean_prev)) |>
  mutate(bin = cut(abs_delta, breaks = c(0, 0.25, 0.5, 1, 2, Inf), right = FALSE,
                   labels = c("[0.00, 0.25)", "[0.25, 0.50)", "[0.50, 1.00)",
                              "[1.00, 2.00)", "[2.00, +)"))) |>
  group_by(bin) |>
  summarise(
    n       = n(),
    persist = mean(abs_delta),                       # = bin mean |Δ| (mechanical)
    across(c(cb, ev, an, su, ft_raw, ft_anon, ft_summ), \(x) mean(x, na.rm = TRUE)),
    .groups = "drop"
  )

endog |>
  gt() |>
  fmt_number(columns = c(persist, cb, ev, an, su, ft_raw, ft_anon, ft_summ), decimals = 3) |>
  cols_label(cb = "Base:Cb", ev = "Base:Ev", an = "Base:An", su = "Base:Su",
             ft_raw = "FT-raw", ft_anon = "FT-anon", ft_summ = "FT-summ") |>
  tab_header(title = "Table 1 — Endogenous within-CYI |Δ| bins, all cells (background, not a test)",
             subtitle = "MAE per cell within each movement bin · persist = bin mean |Δ| (mechanical)")
```

## 5. Bootstrap engine (transition instruments)

Seed-42 country-clustered resample (via the shared `country_boot_weights()` helper), like
`06`/`07`/`08`. Per draw the engine
returns the evidence gains (transition vs stable) under the ERT flag — for the **base** conditions
(Ev/An/Su − Cb) and the **FT diagonal** (each FT model's diagonal − own codebook), both the
in-episode primary and the onset-year cut — the weighted-OLS slope of each gain on |Δpolyarchy|,
and the MAE **level of all seven cells** (plus persistence and the human LOO reference) split by
stable / in-episode / onset. The difference-in-differences (transition − stable) is formed from
these in §6.

```{r bootstrap}
#| cache: true
#| cache.lazy: false

# Precompute per-CYI vectors (indexed by pool row) so each draw is a handful of weighted sums.
d_ev <- cyi$d_ev; d_an <- cyi$d_an; d_su <- cyi$d_su
d_ftraw <- cyi$d_ftraw; d_ftanon <- cyi$d_ftanon; d_ftsumm <- cyi$d_ftsumm
in_ep <- cyi$in_ep; onset <- cyi$onset
dpoly_v <- cyi$dpoly; persist_v <- cyi$persist_err; hloo_v <- cyi$hloo
# The seven cell error vectors, for the MAE-level tables (Table 2) split by transition/onset.
cellmat <- list(cb = cyi$cb, ev = cyi$ev, an = cyi$an, su = cyi$su,
                ft_raw = cyi$ft_raw, ft_anon = cyi$ft_anon, ft_summ = cyi$ft_summ)

# weighted group means of `val` over `grp` (logical: TRUE = transition), on the non-NA mask
gmean <- function(val, grp, w) {
  m  <- !is.na(val)
  wt <- w[m]; v <- val[m]; g <- grp[m]
  num <- tapply(wt * v, g, sum); den <- tapply(wt, g, sum)
  st  <- num / den
  c(tr = unname(st["TRUE"]), st = unname(st["FALSE"]))
}
# weighted OLS slope of y on x (closed form), non-NA on both
wslope <- function(y, x, w) {
  m <- !is.na(y) & !is.na(x)
  wt <- w[m]; yy <- y[m]; xx <- x[m]; W <- sum(wt)
  mx <- sum(wt * xx) / W; my <- sum(wt * yy) / W
  sum(wt * (xx - mx) * (yy - my)) / sum(wt * (xx - mx)^2)
}

compute_draw <- function(w) {
  g_ev <- gmean(d_ev, in_ep, w); g_an <- gmean(d_an, in_ep, w); g_su <- gmean(d_su, in_ep, w)
  o_ev <- gmean(d_ev, onset, w); o_an <- gmean(d_an, onset, w); o_su <- gmean(d_su, onset, w)
  g_fr <- gmean(d_ftraw, in_ep, w); g_fa <- gmean(d_ftanon, in_ep, w); g_fs <- gmean(d_ftsumm, in_ep, w)
  o_fr <- gmean(d_ftraw, onset, w); o_fa <- gmean(d_ftanon, onset, w); o_fs <- gmean(d_ftsumm, onset, w)
  l_ps <- gmean(persist_v, in_ep, w); o_ps <- gmean(persist_v, onset, w)
  l_hl <- gmean(hloo_v, in_ep, w);    o_hl <- gmean(hloo_v, onset, w)
  # MAE level per cell, split by in-episode (tr/st) and onset (tr = onset MAE). L_<cell>_<grp>.
  lev <- do.call(c, lapply(names(cellmat), function(nm) {
    gi <- gmean(cellmat[[nm]], in_ep, w); go <- gmean(cellmat[[nm]], onset, w)
    setNames(c(gi[["tr"]], gi[["st"]], go[["tr"]]), paste0("L_", nm, c("_tr", "_st", "_on")))
  }))
  # `[[ ]]` strips the "tr"/"st" names so c(...) keeps the outer names verbatim.
  c(ev_tr = g_ev[["tr"]], ev_st = g_ev[["st"]], an_tr = g_an[["tr"]], an_st = g_an[["st"]],
    su_tr = g_su[["tr"]], su_st = g_su[["st"]],
    ev_tr_on = o_ev[["tr"]], ev_st_on = o_ev[["st"]], an_tr_on = o_an[["tr"]], an_st_on = o_an[["st"]],
    su_tr_on = o_su[["tr"]], su_st_on = o_su[["st"]],
    fr_tr = g_fr[["tr"]], fr_st = g_fr[["st"]], fa_tr = g_fa[["tr"]], fa_st = g_fa[["st"]],
    fs_tr = g_fs[["tr"]], fs_st = g_fs[["st"]],
    fr_tr_on = o_fr[["tr"]], fr_st_on = o_fr[["st"]], fa_tr_on = o_fa[["tr"]], fa_st_on = o_fa[["st"]],
    fs_tr_on = o_fs[["tr"]], fs_st_on = o_fs[["st"]],
    ps_tr = l_ps[["tr"]], ps_st = l_ps[["st"]], ps_on = o_ps[["tr"]],
    hl_tr = l_hl[["tr"]], hl_st = l_hl[["st"]], hl_on = o_hl[["tr"]],
    lev,
    slope_ev = wslope(d_ev, dpoly_v, w),
    slope_an = wslope(d_an, dpoly_v, w),
    slope_su = wslope(d_su, dpoly_v, w),
    slope_ftraw = wslope(d_ftraw, dpoly_v, w),
    slope_ftanon = wslope(d_ftanon, dpoly_v, w),
    slope_ftsumm = wslope(d_ftsumm, dpoly_v, w))
}

# Country-clustered draw step (shared helper), same seed/scheme as 06/07/08. Broadcast each draw's
# per-country weight to that country's CYIs; cyi is ordered by `.rowid` (= row_number), so indexing
# W's rows by country yields a weight vector aligned to the pool rows compute_draw() expects.
W <- country_boot_weights(cyi$country_text_id, params$n_boot, seed = 42)
draws <- map(colnames(W), function(draw) {
  w <- unname(W[, draw][cyi$country_text_id])
  compute_draw(w)
})
app <- which(colnames(W) == "Apparent")   # apparent (full-sample) draw = point estimate
M   <- do.call(rbind, draws)
```

## 6. Aggregate to point estimates + 95% CIs

```{r aggregate}
boot_rows <- setdiff(seq_len(nrow(M)), app)
ci <- function(col) {
  v <- M[, col]
  # unname() the quantiles — otherwise the "2.5%"/"97.5%" tags mangle the lo/hi names
  # and downstream lookups by name return NA. na.rm because under country resampling a draw can
  # contain zero onset countries (only 8 exist) → NaN onset means; those draws drop from the CI.
  c(est = unname(v[app]), lo = unname(quantile(v[boot_rows], 0.025, na.rm = TRUE)),
    hi = unname(quantile(v[boot_rows], 0.975, na.rm = TRUE)), p_gt = mean(v[boot_rows] > 0, na.rm = TRUE))
}
# Difference-in-differences formed per draw (transition − stable), then quantiled — paired, so the
# CI is on the DiD itself, not the overlap of two group bars. na.rm as in ci() (onset draws).
did <- function(tr_col, st_col, label) {
  d <- M[, tr_col] - M[, st_col]
  tibble(Contrast = label, tr = M[app, tr_col], st = M[app, st_col],
         DiD = d[app], lo = quantile(d[boot_rows], 0.025, na.rm = TRUE),
         hi = quantile(d[boot_rows], 0.975, na.rm = TRUE),
         p_gt = mean(d[boot_rows] > 0, na.rm = TRUE),
         sesoi_out = abs(d[app]) > sesoi)
}
```

### Table 2 — ERT split, MAE levels (external-flag analog of Table 1)

The non-circular version of Table 1: split by the ERT episode flag instead of the endogenous
|Δ|. `n` and coverage are fixed; MAE levels carry bootstrap CIs.

```{r tbl-levels}
n_tr <- sum(cyi$in_ep); n_st <- sum(!cyi$in_ep); n_on <- sum(cyi$onset)
cifmt <- function(x) glue("[{round(x['lo'], 3)}, {round(x['hi'], 3)}]")
lvl <- function(prefix, label) {
  a <- ci(paste0(prefix, "_st")); b <- ci(paste0(prefix, "_tr")); d <- ci(paste0(prefix, "_on"))
  tibble(Quantity = label,
         Stable = a["est"],       `Stable CI` = cifmt(a),
         `In-episode` = b["est"], `In-ep CI` = cifmt(b),
         Onset = d["est"],        `Onset CI` = cifmt(d))
}
bind_rows(
  lvl("ps", "persistence (integer-fair)"),
  lvl("hl", "human LOO (held-out coder)"),
  lvl("L_cb", "Base:Cb"), lvl("L_ev", "Base:Ev"), lvl("L_an", "Base:An"), lvl("L_su", "Base:Su"),
  lvl("L_ft_raw", "FT-raw"), lvl("L_ft_anon", "FT-anon"), lvl("L_ft_summ", "FT-summ")
) |>
  gt() |>
  fmt_number(columns = c(Stable, `In-episode`, Onset), decimals = 3) |>
  tab_header(
    title = "Table 2 — ERT episode split: MAE levels, all cells",
    subtitle = glue("2019 · stable n={n_st}, in-episode n={n_tr}, onset n={n_on} (onset ⊂ in-episode) · ",
                    "country-clustered bootstrap — onset spans only 8 countries, so its CI is wide/indicative"))
```

### Figure 1 — MAE landscape by transition subgroup

Table 2 as a picture, in the Figure-1 idiom (`06`): the seven cells on a common AI-MAE axis
against the **human LOO** rail, faceted by subgroup so each panel carries its own rail and ±SESOI
band. Reading left panel → right panel, the human LOO line and the whole cell cluster slide
*right* together — coders themselves sit farther from consensus in transition years (disagreement
during change) — while FT-raw stays pinned to the human line throughout. Most cells fall *within*
±SESOI of the human line: on error *magnitude* they are not substantively distinguishable from a
held-out coder (only base Anonymized/Summarized clear the band). The separation between cells
lives in error *shape* and *direction*, which is what the horse-race tables (§7) measure — this
figure is the level-view background. (Onset spans only 8 countries, so its error bars are wide
under the country-clustered bootstrap — read them as indicative.)

```{r}
#| label: fig-landscape-subgroups
#| fig-width: 9.5
#| fig-height: 4.6
cells_tbl <- tibble(
  pref   = c("L_cb", "L_ev", "L_an", "L_su", "L_ft_raw", "L_ft_anon", "L_ft_summ"),
  row    = c("Codebook", "Raw Text", "Anonymized", "Summarized",
             "FT · Raw Text", "FT · Anonymized", "FT · Summarized"),
  family = c("Base", "Base", "Base", "Base", "Fine-tuned", "Fine-tuned", "Fine-tuned"))
getci <- function(col) { z <- ci(col); tibble(est = z["est"], lo = z["lo"], hi = z["hi"]) }

sub_lv <- c("Stable", "In-episode", "Onset")
land <- tidyr::crossing(cells_tbl, suf = c("st", "tr", "on")) |>
  mutate(cidf = map(paste0(pref, "_", suf), getci)) |> unnest(cidf) |>
  mutate(sub = factor(recode(suf, st = "Stable", tr = "In-episode", on = "Onset"), levels = sub_lv),
         row = factor(row, levels = c("FT · Summarized", "FT · Anonymized", "FT · Raw Text",
                                      "Summarized", "Anonymized", "Raw Text", "Codebook")),
         family = factor(family, levels = c("Base", "Fine-tuned")))

refs <- tibble(sub = factor(sub_lv, levels = sub_lv),
               human   = c(ci("hl_st")["est"], ci("hl_tr")["est"], ci("hl_on")["est"]),
               persist = c(ci("ps_st")["est"], ci("ps_tr")["est"], ci("ps_on")["est"]))

ggplot(land, aes(est, row, color = family)) +
  geom_rect(data = refs, inherit.aes = FALSE, ymin = -Inf, ymax = Inf,
            aes(xmin = human - sesoi, xmax = human + sesoi), fill = "grey85", alpha = 0.55) +
  geom_vline(data = refs, aes(xintercept = human), linetype = "longdash", color = "grey40") +
  geom_hline(yintercept = 3.5, color = "grey85", linewidth = 0.4) +   # base | FT separator
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y", width = 0.22, linewidth = 0.6) +
  geom_point(size = 2) +
  facet_wrap(~ sub, nrow = 1) +
  scale_color_manual(values = family_pal, name = NULL) +
  labs(x = "AI Mean Absolute Error", y = NULL,
       title = "Figure 1 — Coder performance by regime-transition subgroup",
       subtitle = glue("2019 · MAE vs panel mean · long-dash = human LOO · ",
                       "grey band = ±SESOI ({round(sesoi,3)}) around human LOO")) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", legend.justification = "left",
        panel.grid.major.y = element_blank(), plot.title.position = "plot",
        plot.title = element_text(face = "bold"))
```

### Table 3 — Evidence-gain difference-in-differences (2019, exploratory)

An exploratory 2019 analysis in the spirit of R2, which is registered on 2023; on 2019 it is an
additional, hypothesis-generating companion. For each model's **evidence gain** — the base
Ev/An/Su − Cb, and each FT model's diagonal condition minus its **own** codebook baseline — the
mean paired Δ in transition vs. stable CYIs, and the **DiD = transition − stable**. In R2's logic a
negative DiD means evidence helps more where the country is moving. In-episode is the primary cut;
onset-year is the sharper one.

```{r tbl-did}
did_tbl <- bind_rows(
  did("ev_tr", "ev_st", "Base Ev−Cb · in-episode"),
  did("an_tr", "an_st", "Base An−Cb · in-episode"),
  did("su_tr", "su_st", "Base Su−Cb · in-episode"),
  did("fr_tr", "fr_st", "FT-raw · in-episode"),
  did("fa_tr", "fa_st", "FT-anon · in-episode"),
  did("fs_tr", "fs_st", "FT-summ · in-episode"),
  did("ev_tr_on", "ev_st_on", "Base Ev−Cb · onset-year"),
  did("an_tr_on", "an_st_on", "Base An−Cb · onset-year"),
  did("su_tr_on", "su_st_on", "Base Su−Cb · onset-year"),
  did("fr_tr_on", "fr_st_on", "FT-raw · onset-year"),
  did("fa_tr_on", "fa_st_on", "FT-anon · onset-year"),
  did("fs_tr_on", "fs_st_on", "FT-summ · onset-year")
)

did_tbl |>
  transmute(Contrast,
            `Δ transition` = tr, `Δ stable` = st, DiD,
            `95% CI (DiD)` = glue("[{round(lo,3)}, {round(hi,3)}]"),
            `P(DiD>0)` = p_gt,
            `Beyond SESOI?` = if_else(sesoi_out, "yes", "no")) |>
  gt() |>
  fmt_number(columns = c(`Δ transition`, `Δ stable`, DiD, `P(DiD>0)`), decimals = 3) |>
  tab_header(
    title = "Table 3 — Evidence-gain difference-in-differences, base + FT (2019, exploratory)",
    subtitle = glue("In the spirit of R2 (registered on 2023) · DiD<0 = evidence helps more in ",
                    "transition CYIs · SESOI ±{round(sesoi,3)}"))
```

**Reading the DiD in the reversed regime.** Because $d = \text{MAE(condition)} - \text{MAE(codebook)}$
is *positive* on 2019 — evidence raises error over the bare prior, the identification reversal — a
negative DiD does **not** mean evidence helps; it means the penalty is *smaller* in transition than
in stable. Base Ev−Cb's −0.018 decomposes cleanly against Table 2: codebook error rises +0.035
(stable → in-episode) while evidence rises only +0.017, so the Ev−Cb gap narrows from 0.075 to
0.057. Mechanistically this is R2's stale-prior logic — the naked prior (codebook) degrades faster
where the country moves, so reading evidence is comparatively less costly there — but it is not the
"evidence grounds the rating" result the hypothesis imagined: evidence still raises error
everywhere, the effect is sub-SESOI, and it is confined to the identity-present Ev condition (An−Cb
and Su−Cb are null). The large *shared* component — everyone's error rises in transition, human LOO
most of all (+0.070) — is what the DiD differences away; that common rise, not the small residual,
is the headline of Figure 1. Most DiD rows do not clear zero: the binary split is underpowered,
and the continuous moderator below — though higher-resolution — is no more decisive once the
bootstrap resamples countries, because its slope is carried by only the handful of country-years
that actually moved.

### Table 4 — Continuous moderator: evidence gain vs |Δpolyarchy| (2019, exploratory)

The higher-resolution companion to Table 3, on the same 2019 exploratory footing. Weighted-OLS
slope of each per-CYI evidence gain (base and FT) on |Δv2x_polyarchy|. A **negative** slope means
the evidence gain grows more negative (evidence helps more) as the country's aggregate movement
increases — the continuous analog of a negative DiD. Slope is on ΔMAE per unit polyarchy change
(dimensionless axis change), so no SESOI band applies.

```{r tbl-slope}
bind_rows(
  as_tibble_row(ci("slope_ev")) |> mutate(Gain = "Base Ev−Cb"),
  as_tibble_row(ci("slope_an")) |> mutate(Gain = "Base An−Cb"),
  as_tibble_row(ci("slope_su")) |> mutate(Gain = "Base Su−Cb"),
  as_tibble_row(ci("slope_ftraw"))  |> mutate(Gain = "FT-raw"),
  as_tibble_row(ci("slope_ftanon")) |> mutate(Gain = "FT-anon"),
  as_tibble_row(ci("slope_ftsumm")) |> mutate(Gain = "FT-summ")
) |>
  transmute(Gain, Slope = est, `95% CI` = glue("[{round(lo,3)}, {round(hi,3)}]"),
            `P(slope>0)` = p_gt) |>
  gt() |>
  fmt_number(columns = c(Slope, `P(slope>0)`), decimals = 3) |>
  tab_header(
    title = "Table 4 — Evidence gain vs |Δpolyarchy|, base + FT (2019, exploratory)",
    subtitle = "slope<0 = evidence helps more as aggregate movement grows")
```

**Reading the slope, and why it does not survive clustering.** The continuous moderator regresses
the per-CYI gain $d$ on movement magnitude rather than collapsing it to a bin, so each country-year
enters weighted by how far it actually moved. Read as *point estimates*, the pattern is the one R2
predicts: a negative slope means the positive evidence penalty *erodes* as the country moves more,
and Base Ev−Cb's −0.563 slope reaches $d = 0$ near |Δpolyarchy| ≈ 0.14 and turns negative beyond it
(Figure 2) — on the few country-years that moved most, reading evidence stops raising error and
begins to lower it, the R2-predicted payoff realized only at the tail of movement. **But the slopes
do not survive honest uncertainty.** Because the moderator is constant within country, the slope is
identified entirely by the ~dozen country-years that actually moved, so the country-clustered
bootstrap widens these intervals ~3–4× and **Base Ev, Base Su, and FT-summ all cross zero** (only
FT-anon still, barely, excludes it; FT-raw is flat either way, its calibration baked into the
weights). The point estimates keep their old order — base conditions slope down, Ev steepest,
FT-summ positive — but in 2019 the instrument cannot resolve any of them from zero. Two further
cautions reinforce this: the footprint is tiny even at the point estimate (≈ −0.01 at the 3rd
quartile of movement), and 2019 has no genuinely stale prior for movement to act on (§intro). These
remain exploratory 2019 companions; the movement payoff is properly a 2023 question.

### Figure 2 — Evidence penalty vs. movement (the slope as a shape)

`07`'s "slope as a shape" idiom applied to the moderator: the per-CYI evidence gain against
|Δpolyarchy|, with a dashed zero line. Above the line, evidence raises error (the penalty); below
it, evidence lowers error. Points are 20 equal-count movement bins (dense near zero, where almost
every country sits); the coloured lines are the weighted fits with a country-year bootstrap ribbon,
extended across the movement range so the **Base Ev−Cb** line's descent *through zero* is visible
(the high-movement region is sparse, so the ribbon widens there). **FT-summ** is drawn as the
contrast — its penalty rises with movement — and the other four gains are greyed.

```{r}
#| label: fig-polyarchy-crossing
#| cache: true
#| cache.lazy: false
#| fig-width: 8
#| fig-height: 4.8
xv <- cyi$dpoly
gains <- list("Base Ev" = cyi$d_ev, "Base An" = cyi$d_an, "Base Su" = cyi$d_su,
              "FT-raw" = cyi$d_ftraw, "FT-anon" = cyi$d_ftanon, "FT-summ" = cyi$d_ftsumm)

# weighted OLS intercept + slope of a gain on movement
wfit <- function(y, w) {
  m <- !is.na(y) & !is.na(xv); wt <- w[m]; yy <- y[m]; xx <- xv[m]
  W <- sum(wt); mx <- sum(wt * xx) / W; my <- sum(wt * yy) / W
  b <- sum(wt * (xx - mx) * (yy - my)) / sum(wt * (xx - mx)^2)
  c(a = my - b * mx, b = b)
}

# axis: reach a bit past the Base Ev zero-crossing so it is visible
f_ev  <- wfit(gains[["Base Ev"]], rep(1, length(xv)))
cross <- as.numeric(-f_ev["a"] / f_ev["b"])
xmax  <- min(max(xv, na.rm = TRUE), cross * 1.4)
xg    <- seq(0, xmax, length.out = 80)

line_df <- bind_rows(lapply(names(gains), function(nm) {
  f <- wfit(gains[[nm]], rep(1, length(xv))); tibble(cell = nm, x = xg, yhat = f[["a"]] + f[["b"]] * xg)
}))

hl <- c("Base Ev", "FT-summ")

# 20 equal-count movement bins for the empirical points (highlighted cells). ntile() rather than
# quantile-cut because |Δpolyarchy| is heavily tied near zero, which collapses quantile breaks.
nbn <- 20L
binf <- dplyr::ntile(xv, nbn)
xbin <- tapply(xv, binf, mean, na.rm = TRUE)
bin_df <- bind_rows(lapply(hl, function(nm) {
  tibble(cell = nm, x = as.numeric(xbin),
         ybar = as.numeric(tapply(gains[[nm]], binf, mean, na.rm = TRUE)))
})) |> filter(x <= xmax)

# country-clustered bootstrap ribbon for the highlighted fitted lines (the moderator is
# country-level, so resample whole countries and broadcast each draw's weight to the country's CYIs)
Wf <- country_boot_weights(cyi$country_text_id, 1000L, seed = 42)
ribbon <- bind_rows(lapply(hl, function(nm) {
  P <- vapply(colnames(Wf)[-1], function(draw) {
    w <- unname(Wf[, draw][cyi$country_text_id]); f <- wfit(gains[[nm]], w); f[["a"]] + f[["b"]] * xg
  }, numeric(length(xg)))
  tibble(cell = nm, x = xg, lo = apply(P, 1, quantile, 0.025, na.rm = TRUE),
         hi = apply(P, 1, quantile, 0.975, na.rm = TRUE))
}))

pal2 <- c("Base Ev" = "#0072B2", "FT-summ" = "#E69F00")
ggplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_line(data = filter(line_df, !cell %in% hl), aes(x, yhat, group = cell),
            color = "grey80", linewidth = 0.5) +
  geom_ribbon(data = ribbon, aes(x, ymin = lo, ymax = hi, fill = cell), alpha = 0.18) +
  geom_line(data = filter(line_df, cell %in% hl), aes(x, yhat, color = cell), linewidth = 1.1) +
  geom_point(data = bin_df, aes(x, ybar, color = cell), size = 1.5, alpha = 0.65) +
  scale_color_manual(values = pal2, name = NULL, aesthetics = c("color", "fill")) +
  coord_cartesian(xlim = c(0, xmax)) +
  labs(x = "|Δ v2x_polyarchy|  (2018→2019 movement)",
       y = "evidence gain  d = MAE(condition) − MAE(codebook)",
       title = "Figure 2 — Evidence penalty vs. movement: the slope as a shape",
       subtitle = glue("above 0 = evidence raises error · below 0 = evidence lowers error · ",
                       "points = 20 equal-count bins · grey = other gains · high-movement tail is sparse")) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", legend.justification = "left",
        panel.grid.minor = element_blank(), plot.title.position = "plot",
        plot.title = element_text(face = "bold"))
```

### Table 5 — Level slopes: how fast each predictor degrades with movement

The gain slope in Table 4 is a *difference* of two level slopes — slope(Ev−Cb) = slope(MAE_Ev) −
slope(MAE_Cb) — so it has no direct human analog (a coder has no codebook-vs-evidence contrast).
But the two *pieces* do: the **level slope** of a cell's MAE on |Δpolyarchy| — how fast its error
grows as the country moves — and the human LOO MAE has exactly that same level slope. So the
**human LOO slope is the natural baseline** here, the continuous version of "human error rose most
in the transition bin." Each predictor is read against it: a slope above the human line degrades
*faster* than a held-out coder as the country moves, below it *slower*.

The engine is a self-contained country-clustered bootstrap, like the rest of this document; it
reuses the cached `cyi`, so it does not touch the registered or horse-race bootstraps.

```{r level-slopes}
#| cache: true
#| cache.lazy: false
lvl_vars <- list(
  "human LOO"   = cyi$hloo,      "persistence" = cyi$persist_err,
  "Base:Cb" = cyi$cb, "Base:Ev" = cyi$ev, "Base:An" = cyi$an, "Base:Su" = cyi$su,
  "FT-raw"  = cyi$ft_raw, "FT-anon" = cyi$ft_anon, "FT-summ" = cyi$ft_summ)
xpoly <- cyi$dpoly
cci   <- match(cyi$country_text_id, sort(unique(cyi$country_text_id))); n_cc <- max(cci)

wsl <- function(y, w) {                      # weighted slope of y on xpoly
  m <- !is.na(y) & !is.na(xpoly); wt <- w[m]; yy <- y[m]; xx <- xpoly[m]
  W <- sum(wt); mx <- sum(wt * xx) / W; my <- sum(wt * yy) / W
  sum(wt * (xx - mx) * (yy - my)) / sum(wt * (xx - mx)^2)
}
set.seed(42); Bc <- 1000L
lvl_est  <- sapply(lvl_vars, \(y) wsl(y, rep(1, nrow(cyi))))
lvl_boot <- sapply(seq_len(Bc), function(b) {
  w <- tabulate(sample.int(n_cc, n_cc, replace = TRUE), n_cc)[cci]
  sapply(lvl_vars, \(y) wsl(y, w))
})
lvl_slopes <- tibble(
  var = names(lvl_vars), est = lvl_est,
  lo = apply(lvl_boot, 1, quantile, 0.025), hi = apply(lvl_boot, 1, quantile, 0.975)) |>
  mutate(family = case_when(var == "human LOO" ~ "Human", var == "persistence" ~ "Naive",
                            str_starts(var, "FT") ~ "Fine-tuned", TRUE ~ "Base"))
human_slope <- lvl_slopes$est[lvl_slopes$var == "human LOO"]

lvl_slopes |>
  transmute(Predictor = var, `Level slope` = est,
            `95% CI` = glue("[{round(lo, 3)}, {round(hi, 3)}]"),
            `vs. human` = case_when(var == "human LOO" ~ "—",
                                    est > human_slope ~ "steeper (degrades faster)",
                                    TRUE ~ "flatter (more robust)")) |>
  gt() |>
  fmt_number(`Level slope`, decimals = 3) |>
  tab_header(title = "Table 5 — Level slopes: MAE growth with movement",
             subtitle = glue("weighted slope of each predictor's MAE on |Δpolyarchy| · country-clustered · ",
                             "human LOO baseline = {round(human_slope, 3)}"))
```

The reading: **persistence (+0.45) and Codebook (+0.43) degrade faster than a human coder (+0.33)** —
the past-leaning predictors (naive copy-forward, frozen prior) go stale faster than a person does.
**Base:Ev (−0.14) is the one movement-robust cell** — flatter than the human, its error essentially
unmoved by regime change (this is the numerator of the negative Ev−Cb gain slope). **FT-raw (+0.36)
sits right on the human line** — another sense in which it degrades like a held-out coder. *Caveat:*
these level slopes are noisy — almost every CI crosses zero (human LOO included), because the slope
is carried by the handful of high-movement countries and the clustered bootstrap is honest about
that. Differencing Ev − Cb (Table 4) cancels the shared "everyone degrades under movement"
component, so the gain *level* is sharper than either raw level slope — but it does **not** rescue
the gain *slope* on |Δpolyarchy|, which is still a between-country quantity carried by the same few
movers and, under clustering, widens and mostly crosses zero just like the levels here.

### Figure 3 — Degradation with movement, against the human baseline

**This is Table 5 as a picture** (and the continuous analog of Figure 1): each predictor's level
slope (how fast its error grows with movement) as a point with its country-clustered 95% CI,
ordered slowest → fastest, with the **human LOO slope drawn as the reference line**. Points to the right of the line degrade faster
than a coder; to the left, slower. The wide intervals are shown deliberately — they are the honest
uncertainty, and the figure should be read for the *ordering* (naive/prior fastest, evidence
slowest, FT-raw ≈ human), not for precise magnitudes.

```{r}
#| label: fig-level-slopes
#| fig-width: 8
#| fig-height: 4.4
ord <- lvl_slopes |> arrange(est) |> pull(var)
lvl_slopes |>
  mutate(var = factor(var, levels = ord),
         family = factor(family, levels = c("Naive", "Base", "Fine-tuned", "Human"))) |>
  ggplot(aes(est, var, color = family)) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey65") +
  geom_vline(xintercept = human_slope, linetype = "longdash", color = "grey30") +
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y", width = 0.28, linewidth = 0.6) +
  geom_point(size = 2.4) +
  scale_color_manual(values = c("Naive" = "grey55", "Base" = "#0072B2",
                                "Fine-tuned" = "#E69F00", "Human" = "grey25"), name = NULL) +
  labs(x = "level slope: ΔMAE per unit |Δpolyarchy|  (right = degrades faster with movement)",
       y = NULL,
       title = "Figure 3 (Table 5 as a picture) — How fast each predictor degrades with movement",
       subtitle = glue("2019 · slope of MAE on |Δpolyarchy| · long-dash = human LOO baseline ",
                       "({round(human_slope, 3)}) · wide CIs (sparse movers, country-clustered)")) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", legend.justification = "left",
        panel.grid.major.y = element_blank(), plot.title.position = "plot",
        plot.title = element_text(face = "bold"))
```

## 7. Human-likeness horse race (exploratory)

The tables above ask R2's oracle-shaped question (does evidence lower MAE where the country
moves). This section asks the **debater** question the reframe cares about: conditional on
transition, **which model behaves most like a held-out human coder?** Two human-referenced
diagnostics from `07`/`08`, recomputed within transition subgroups:

- **W1 to human** (`07`): earth-mover distance between a cell's error distribution `|AI − mean|`
  and the human per-coder LOO error distribution. Smaller = more human-shaped. The adjudicator.
- **Signed shift** (`08`): mean `AI − mean` (direction kept). Humans are pinned at ~0, so any
  departure is directional bias. Split by episode direction (democratization vs. autocratization),
  since a direction-specific lag — if present — would self-cancel when pooled.

**Country-clustered bootstrap.** The ERT flag is a country-level treatment and the subgroups are
small in *countries* (dem 18, aut 41), so the effective *n* is countries, not CYIs. This section
resamples **countries** with replacement (seed 42), as everywhere in this document — essential here
because the subgroups are so small in countries. Reference is the human LOO distribution / zero; onset subgroups are carried in
the bundle as descriptive only (too few countries for a CI).

### Data: human LOO errors + seven-cell AI errors (≥2-coder pool)

```{r horse-data}
#| cache: true
#| cache.lazy: false
human_ratings <- read_csv(file.path(data_dir, "human_ratings.csv"), show_col_types = FALSE)

# Per-coder leave-one-out error, restricted to ≥2-coder CYIs (the human floor). These CYIs are
# the eligible pool for the whole horse race, so AI and human distributions live on the same cells.
he <- human_ratings |>
  filter(year == params$year) |>
  group_by(country_text_id, indicator) |>
  filter(n() >= 2) |>
  mutate(e = abs(rating - (sum(rating) - rating) / (n() - 1))) |>
  ungroup() |>
  select(country_text_id, indicator, e)
elig <- distinct(he, country_text_id, indicator)

# Full episode/direction flags at the country level (2019).
ert_sub <- ert |>
  filter(year == params$year) |>
  transmute(country_text_id,
            dem = replace_na(dem_ep == 1, FALSE), aut = replace_na(aut_ep == 1, FALSE),
            onset_dem = replace_na(dem_ep == 1 & year == dem_ep_start_year, FALSE),
            onset_aut = replace_na(aut_ep == 1 & year == aut_ep_start_year, FALSE)) |>
  mutate(in_ep = dem | aut, onset = onset_dem | onset_aut)

fig_cells <- c("llama-70b|codebook", "llama-70b|evidence", "llama-70b|anonymized",
               "llama-70b|summarized", "llama-70b-ft-raw|evidence-zeroshot",
               "llama-70b-ft-anon|anonymized-zeroshot", "llama-70b-ft-summ|summarized-zeroshot")
cell_lbl <- c("Base:Cb", "Base:Ev", "Base:An", "Base:Su", "FT-raw", "FT-anon", "FT-summ")

pm_y <- panel_means |> filter(year == params$year) |> select(country_text_id, indicator, raw_mean)
ai7 <- ai_raw |>
  filter(year == params$year) |>
  rename(country_text_id = country) |>
  mutate(cell = paste(model_key, condition, sep = "|")) |>
  filter(cell %in% fig_cells) |>
  inner_join(pm_y, by = c("country_text_id", "indicator")) |>
  semi_join(elig, by = c("country_text_id", "indicator")) |>
  mutate(a = abs(rating - raw_mean), s = rating - raw_mean,
         cell = factor(cell, levels = fig_cells)) |>
  left_join(ert_sub, by = "country_text_id") |>
  mutate(across(c(dem, aut, in_ep, onset, onset_dem, onset_aut), \(x) replace_na(x, FALSE)))
he <- he |> semi_join(elig, by = c("country_text_id", "indicator")) |>
  left_join(ert_sub, by = "country_text_id") |>
  mutate(across(c(dem, aut, in_ep, onset, onset_dem, onset_aut), \(x) replace_na(x, FALSE)))

# Shared country index (the resample unit).
countries <- sort(unique(c(ai7$country_text_id, he$country_text_id)))
ai7$ci <- match(ai7$country_text_id, countries); he$ci <- match(he$country_text_id, countries)
n_ctry <- length(countries)
glue("horse-race pool: AI {nrow(ai7)} rows · human {nrow(he)} rows · {n_ctry} countries")
```

### Bootstrap engine (W1 + signed shift, by subgroup)

```{r horse-bootstrap}
#| cache: true
#| cache.lazy: false
# W1 on a fixed grid (0.01 rating points; coarser than 07's 0.002 to keep the clustered bootstrap
# affordable — the cell *ordering*, which is what the horse race reads, is grid-insensitive).
hstep <- 0.01; hnb <- as.integer(ceiling(4 / hstep))
ai7$bin <- pmin(pmax(as.integer(round(ai7$a / hstep)), 1L), hnb)
he$bin  <- pmin(pmax(as.integer(round(he$e  / hstep)), 1L), hnb)
cell_i  <- as.integer(ai7$cell)
scat <- function(tab, len) { v <- numeric(len); v[as.integer(rownames(tab))] <- tab; v }

# Subgroup row masks (AI side and human side).
subs <- list(
  stable = list(a = !ai7$in_ep, h = !he$in_ep), in_ep = list(a = ai7$in_ep, h = he$in_ep),
  in_dem = list(a = ai7$dem,    h = he$dem),    in_aut = list(a = ai7$aut,  h = he$aut),
  onset  = list(a = ai7$onset,  h = he$onset),
  on_dem = list(a = ai7$onset_dem, h = he$onset_dem), on_aut = list(a = ai7$onset_aut, h = he$onset_aut))

w1_sub <- function(wc, ma, mh) {
  wa <- wc[ai7$ci[ma]]; wh <- wc[he$ci[mh]]
  H  <- cumsum(scat(rowsum(wh, he$bin[mh], reorder = FALSE), hnb)); H <- H / H[hnb]
  g  <- (cell_i[ma] - 1L) * hnb + ai7$bin[ma]
  AH <- matrix(scat(rowsum(wa, g, reorder = FALSE), hnb * 7L), nrow = hnb)
  apply(AH, 2, function(col) { A <- cumsum(col); sum(abs(A / A[hnb] - H)) * hstep })
}
sh_sub <- function(wc, ma) {
  wa <- wc[ai7$ci[ma]]
  scat(rowsum(wa * ai7$s[ma], cell_i[ma], reorder = FALSE), 7L) /
    scat(rowsum(wa, cell_i[ma], reorder = FALSE), 7L)
}
compute_h <- function(wc) {
  w1 <- do.call(c, lapply(names(subs), \(s)
    setNames(w1_sub(wc, subs[[s]]$a, subs[[s]]$h), paste("W1", s, cell_lbl, sep = "|"))))
  sh <- do.call(c, lapply(c("stable", "in_dem", "in_aut", "on_dem", "on_aut"), \(s)
    setNames(sh_sub(wc, subs[[s]]$a), paste("SH", s, cell_lbl, sep = "|"))))
  c(w1, sh)
}

set.seed(42)
hdraws <- vector("list", params$n_boot + 1L)
hdraws[[1]] <- compute_h(rep(1, n_ctry))                       # apparent (all weights 1)
for (b in seq_len(params$n_boot)) {
  hdraws[[b + 1L]] <- compute_h(tabulate(sample.int(n_ctry, n_ctry, replace = TRUE), nbins = n_ctry))
}
Mh <- do.call(rbind, hdraws); apph <- 1L; hbr <- seq.int(2L, params$n_boot + 1L)
```

### Aggregate + tables

```{r horse-agg}
h_agg <- tibble(key = colnames(Mh), est = Mh[apph, ],
                lo = apply(Mh[hbr, ], 2, quantile, 0.025, na.rm = TRUE),
                hi = apply(Mh[hbr, ], 2, quantile, 0.975, na.rm = TRUE)) |>
  separate(key, c("stat", "sub", "cell"), sep = "\\|") |>
  mutate(cell = factor(cell, levels = cell_lbl))

cellfmt <- function(est, lo, hi) glue("{sprintf('%.3f', est)} [{sprintf('%.3f', lo)}, {sprintf('%.3f', hi)}]")
```

**W1 to the human error distribution — closest to human wins.** All seven cells; the powered
subgroups (dem 18 / aut 41 countries; dem CIs are wide).

```{r tbl-w1}
h_agg |>
  filter(stat == "W1", sub %in% c("stable", "in_ep", "in_dem", "in_aut", "onset")) |>
  mutate(val = cellfmt(est, lo, hi),
         sub = factor(sub, levels = c("stable", "in_ep", "in_dem", "in_aut", "onset"),
                      labels = c("Stable", "In-episode", "In-ep · dem", "In-ep · aut", "Onset"))) |>
  select(cell, sub, val) |> pivot_wider(names_from = sub, values_from = val) |>
  arrange(cell) |> rename(Cell = cell) |>
  gt() |>
  tab_header(title = "Table 6 — W1 to human LOO error distribution (country-clustered)",
             subtitle = "smaller = more human-shaped · FT-raw / FT-summ closest throughout · onset = 8 countries (descriptive)")
```

Table 6 as a picture: each panel is a transition subgroup, cells sorted most-human-shaped (top) to
least (bottom). Read left = closer to a real coder (0 = identical error distribution). The two FT
diagonal cells hug the left edge in *every* subgroup — the "FT-raw is the most human-shaped cell"
result survives conditioning on transition — while the bare Codebook prior stays farthest right.
CIs widen sharply left → right (Stable is well-powered; the dem/aut splits and Onset have few
countries), so read the *ordering*, not the small-subgroup gaps.

```{r}
#| label: fig-w1-subgroups
#| fig-width: 11
#| fig-height: 4.2
sub_lv6  <- c("stable", "in_ep", "in_dem", "in_aut", "onset")
sub_lab6 <- c(stable = "Stable", in_ep = "In-episode", in_dem = "In-ep · dem",
              in_aut = "In-ep · aut", onset = "Onset (8 ctry, descriptive)")
w1_fig <- h_agg |> filter(stat == "W1", sub %in% sub_lv6) |>
  mutate(sub = factor(sub_lab6[sub], levels = sub_lab6[sub_lv6]),
         family = factor(if_else(str_starts(as.character(cell), "FT"), "Fine-tuned", "Base"),
                         levels = c("Base", "Fine-tuned")))
# order cells by their in-episode W1 (most human-shaped at top)
ord6 <- w1_fig |> filter(sub == sub_lab6[["in_ep"]]) |> arrange(desc(est)) |> pull(cell) |> as.character()
w1_fig <- w1_fig |> mutate(cell = factor(as.character(cell), levels = ord6))

ggplot(w1_fig, aes(est, cell, color = family)) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey55") +
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y", width = 0.25, linewidth = 0.6) +
  geom_point(size = 2.2) +
  facet_wrap(~ sub, nrow = 1) +
  scale_color_manual(values = family_pal, name = NULL) +
  labs(x = "W1 to human LOO error distribution  (0 = identical to a coder · left = more human-shaped)",
       y = NULL,
       title = "Figure 4 (Table 6 as a picture) — how human-shaped each cell is, by transition subgroup",
       subtitle = "country-clustered 95% CI · smaller = closer to a held-out human coder · FT-raw/FT-summ hug the left throughout") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", legend.justification = "left",
        panel.grid.major.y = element_blank(), plot.title.position = "plot",
        plot.title = element_text(face = "bold"))
```

**Signed shift — directional bias, split by episode direction.** Human reference ≈ 0; a uniform
negative value is the harshness/level-shift effect (`08`), not a transition-specific lag.

```{r tbl-shift}
h_agg |>
  filter(stat == "SH", sub %in% c("in_dem", "in_aut", "stable", "on_dem", "on_aut")) |>
  mutate(val = cellfmt(est, lo, hi),
         sub = factor(sub, levels = c("in_dem", "in_aut", "stable", "on_dem", "on_aut"),
                      labels = c("In-ep · dem", "In-ep · aut", "Stable",
                                 "Onset · dem", "Onset · aut"))) |>
  select(cell, sub, val) |> pivot_wider(names_from = sub, values_from = val) |>
  arrange(cell) |> rename(Cell = cell) |>
  gt() |>
  tab_header(title = "Table 7 — Signed shift (AI − panel mean), by episode direction",
             subtitle = "≈0 = directionally human · FT-raw nearest zero throughout · negative = harsher than panel · onset descriptive (dem = 2 countries, aut = 6)")
```

Table 7 as a picture: signed shift per cell, faceted by episode direction, against the human/oracle
target (dashed 0) and the ±SESOI band. Read distance from 0 = directional bias; negative = harsher
than the panel. FT-raw sits nearest zero throughout, and — the load-bearing point — **no cell flips
sign between the dem and aut panels**: a genuine direction-specific stale-prior lag would push
democratizers and autocratizers *opposite* ways, but here every cell leans the same direction in
both, so this is a uniform harshness/level-shift (`08`), not a transition lag. (FT-anon is the lone
positive/generous cell; onset panels are 2–6 countries, descriptive only.)

```{r}
#| label: fig-shift-subgroups
#| fig-width: 11
#| fig-height: 4.2
sub_lv7  <- c("stable", "in_dem", "in_aut", "on_dem", "on_aut")
sub_lab7 <- c(stable = "Stable", in_dem = "In-ep · dem", in_aut = "In-ep · aut",
              on_dem = "Onset · dem (2 ctry)", on_aut = "Onset · aut (6 ctry)")
sh_fig <- h_agg |> filter(stat == "SH", sub %in% sub_lv7) |>
  mutate(sub = factor(sub_lab7[sub], levels = sub_lab7[sub_lv7]),
         family = factor(if_else(str_starts(as.character(cell), "FT"), "Fine-tuned", "Base"),
                         levels = c("Base", "Fine-tuned")))
# order cells by |stable shift| — nearest human-zero at top
ord7 <- sh_fig |> filter(sub == sub_lab7[["stable"]]) |> arrange(desc(abs(est))) |> pull(cell) |> as.character()
sh_fig <- sh_fig |> mutate(cell = factor(as.character(cell), levels = ord7))

ggplot(sh_fig, aes(est, cell, color = family)) +
  annotate("rect", xmin = -sesoi, xmax = sesoi, ymin = -Inf, ymax = Inf, fill = "grey85", alpha = 0.55) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y", width = 0.25, linewidth = 0.6) +
  geom_point(size = 2.2) +
  facet_wrap(~ sub, nrow = 1) +
  scale_color_manual(values = family_pal, name = NULL) +
  labs(x = "signed shift (AI − panel mean)  ·  0 = human/oracle target  ·  negative = harsher than panel",
       y = NULL,
       title = "Figure 5 (Table 7 as a picture) — directional bias by episode direction",
       subtitle = glue("dashed 0 = human target · grey band = ±SESOI ({round(sesoi,3)}) · ",
                       "dem & aut lean the same way = uniform harshness, not a direction-specific lag")) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", legend.justification = "left",
        panel.grid.major.y = element_blank(), plot.title.position = "plot",
        plot.title = element_text(face = "bold"))
```

## 8. Reading (prototype)

Working readout; the paper narrative and the visualizations we discussed go in later sections of
this same document. Three things to read across the tables:

- **Table 1 vs Table 2** — the same "does evidence help where the country moved" question, first
  with the circular within-CYI |Δ| (background) and then with the external ERT flag. If the ERT
  split reproduces the direction of the endogenous split without the mechanical persistence
  column, the effect is not an artifact of binning on the target.
- **Table 3 (DiD)** — the exploratory 2019 companion to R2, now across base and FT: a negative DiD
  means evidence helps more in transition CYIs. R2 itself is registered on 2023, so these 2019
  values are hypothesis-generating; a null here says nothing about the 2023 result.
- **Table 4 (slope)** — the higher-resolution version of the same, base and FT; agreement in sign
  with Table 3 (theoretically-meaningful binary + continuous moderator) is the cross-check.
- **Table 5 / Figure 3 (level slopes)** — the level decomposition against the human LOO baseline:
  the past-leaning predictors (persistence, codebook) degrade faster than a coder under movement,
  evidence is uniquely flat, and FT-raw ≈ the human rate. Noisy (wide CIs) — read the ordering, not
  the magnitudes.
- **Tables 6–7 (horse race)** — the debater diagnostic. Table 6: FT-raw and FT-summ carry the
  smallest W1 to the human error distribution in every subgroup, so the "FT-raw is the most
  human-shaped cell" result survives conditioning on transition — the diagnostic value that further
  supports carrying FT-raw forward as the human-behavior emulator. Table 7: the signed shift is a
  uniform harshness/level-shift (`08`), *not* a direction-specific stale-prior lag (democratizers
  and autocratizers both run negative, not opposite); its use here is confirming FT-raw sits nearest
  human-zero directionally throughout, consistent with the W1 verdict.

## 9. Serialize the bundle

```{r serialize}
saveRDS(
  list(endog = endog,                 # Table 1 (background)
       did_tbl = did_tbl,             # Table 3 (DiD, base + FT)
       lvl_slopes = lvl_slopes, human_slope = human_slope,   # Table 5 / Figure 3 (level slopes)
       M = M, app = app,              # registered per-draw matrix (country-clustered) + apparent index
       # Horse race (§7): tidy point+CI table (all subgroups incl. onset descriptive), the raw
       # per-draw country-clustered matrix for later figures, and the cell vocabulary.
       h_agg = h_agg, Mh = Mh, apph = apph, cell_lbl = cell_lbl,
       n_tr = n_tr, n_st = n_st, sesoi = sesoi, n_boot = params$n_boot),
  file.path(derived_dir, "regimetrans_2019.rds")
)
glue("regimetrans_2019.rds written · B = {params$n_boot} · SESOI ±{round(sesoi,3)} · ",
     "transition n = {n_tr}, stable n = {n_st}")
```

---

## Note — 2023 design, deferred

Out of scope for this 2019 document; recorded here to revisit. R2's actual mechanism (a *stale*
frozen prior) only bites once real time has passed, so the substantive R2 lives in the 2023 pool.
Open question: whether to treat post-2018 transitions as a single accumulated holdout subset
(preferred if the year span is sufficient) rather than a **lag analysis** (evidence gain / human-
likeness as a function of years-since-transition). Also unresolved for 2023: the base carried
model's pretraining runs through Dec 2023, so its prior is not truly stale even there — the clean
temporal probe is R3's 2024 post-cutoff year, and the FT models (fine-tuned on ≤2018 signal) are
where the staleness contrast is sharpest. Fix the binary flag and the |Δpolyarchy| threshold
before running the 2023 tests, per the prereg.
