---
title: "2019 name-swap (R4) — compute & prototype figures"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: true
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
params:
  n_boot: 2000
  year: 2019
  min_coders: 2
---

> **SCAFFOLD — not yet run.** No name-swap JSONL exists in `data/output/runs/` yet; this document
> is written to the known output schema so the analysis can start the moment the runs land. On the
> first real render, check the **load diagnostic (§2)** before trusting anything downstream: cell
> counts per arm/model, that `arm` splits cleanly on `source == named`, and the double-join
> coverage (every swapped CYI carries both a `source_mean` and a `named_mean`).

Exploratory **name-swap** analysis (hypothesis **R4** in `preregistration/preregistration-draft.qmd`;
design: `notes/name-swap-design.md`). R4 asks whether a swapped rating tracks the **described
conditions** (the source country's summarized text) or the **injected name** (a different,
within-region country). Registered R4 is on 2023/the carried model; **this document is the
exploratory 2019 prototype**, in the 06–09 mould — the correct-name arm already gives the
cue-shift baseline, so both arms are new inference here (see the design note).

Two models, both on the **summarized** substrate: **base** (`summarized`, few-shot kept) and
**FT-summ** (`summarized-zeroshot`, calibration in the adapter weights). FT-raw/FT-anon are dropped
(off-diagonal on this substrate — see the design note §Arms and models).

**Three name conditions over the same source text.** Each is the *source's* summarized text with a
different name in the framing line:

- **swapped** — framed with σ(source)'s name. Carries the registered tracking metric and the
  cue-shift treatment.
- **correct** — framed with the source's own name. The name-visible baseline the cue-shift
  differences against.
- **no-name** — the existing name-hidden summarized runs ("the focal country"). Reused, no new
  inference; lets the cue-shift decompose into "any name vs. none" and "wrong name vs. right name".

**Two axes, both headline.** (1) *Tracking* — does the rating land closer to the source or the
named panel mean (the registered R4 quantity). (2) *Coder axis* — is the rating's distance to the
**source** panel within a held-out human coder's own distance (**human LOO**)? The tracking metric
is a *difference of two distances*, so the human-like elevation that reverses the level hypotheses
(06–09) largely cancels; the LOO rail is carried on every level view so the result can be read in
coder-terms, not only oracle-terms. **Cue-shift** (Metric 2) never touches a panel mean at all.

**Country-clustered resampling on the *source* country** (seed 42, `country_boot_weights`), like
06–09 — a source carries its whole swapped indicator block.

## 1. Setup

```{r setup}
library(tidyverse)
library(rsample)
library(gt)
library(glue)
library(patchwork)

if (!exists("params")) params <- list(n_boot = 2000, year = 2019, min_coders = 2)

find_panel_member_root <- function() {
  up <- tryCatch(rprojroot::find_root(rprojroot::is_git_root), error = function(e) NA_character_)
  if (!is.na(up)) return(up)
  down <- file.path(getwd(), "panel-member")
  if (dir.exists(file.path(down, ".git"))) return(down)
  stop("Could not locate the panel-member project root from working dir: ", getwd())
}
proj_root   <- find_panel_member_root()
source(file.path(proj_root, "R", "bootstrap_helpers.R"))
data_dir    <- file.path(proj_root, "data", "processed")
runs_dir    <- file.path(proj_root, "data", "output", "runs")
derived_dir <- file.path(proj_root, "data", "derived")
dir.create(derived_dir, showWarnings = FALSE, recursive = TRUE)

# SESOI is fixed by the main analysis; carry it over so every document bands on the identical
# value. Tracking and cue-shift are on the rating-point axis, so ±SESOI is a valid band here.
sesoi <- readRDS(file.path(derived_dir, "bootstrap_2019.rds"))$sesoi

# The two name-swap cells, keyed by model_key. Base keeps its few-shot summarized condition;
# FT-summ runs zero-shot. Both read the same summarized substrate.
ns_models <- tibble::tribble(
  ~model_key,            ~condition,             ~label,
  "llama-70b",           "summarized",           "Base",
  "llama-70b-ft-summ",   "summarized-zeroshot",  "FT-summ")
```

## 2. Data loading

Name-swap runs (both arms) join to `panel_means` **twice** — on `source` and on `named` — so each
swapped rating carries both benchmarks. The **no-name control** is the existing summarized runs,
matched to a source country on `country == source`. The **human LOO** floor is built on the source
cells (≥ `r params$min_coders` coders). ERT flags feed the §8 transition split.

```{r load-data}
#| cache: true
#| cache.lazy: false

if (!dir.exists(runs_dir)) stop("runs dir not found: ", runs_dir)
all_jsonl <- list.files(runs_dir, pattern = "\\.jsonl$", full.names = TRUE)

ns_files    <- all_jsonl[grepl("^nameswap_", basename(all_jsonl))]
plain_files <- setdiff(all_jsonl, ns_files)
if (length(ns_files) == 0)
  stop("No name-swap JSONL in ", runs_dir, " — run slurm/run_nameswap_*.sh first.")

# Name-swap records carry `source`/`named`; plain summarized runs do not — hence two readers.
read_ns <- function(f) {
  con <- file(f, "r"); on.exit(close(con))
  jsonlite::stream_in(con, verbose = FALSE) |> as_tibble() |>
    select(source, named, year, indicator, condition, model_key, rating)
}
read_plain <- function(f) {
  con <- file(f, "r"); on.exit(close(con))
  jsonlite::stream_in(con, verbose = FALSE) |> as_tibble() |>
    select(country, year, indicator, condition, model_key, rating)
}

# Both arms. Arm is derived from the data (correct: source == named; swapped: source != named),
# so it does not depend on the filename. `-local` is normalized back to the config key, as in 06.
ns_raw <- ns_files |>
  map(read_ns) |> bind_rows() |>
  mutate(model_key = str_remove(model_key, "-local$"),
         arm = if_else(source == named, "correct", "swapped")) |>
  filter(year == params$year)

# No-name control: the two summarized cells from the ordinary runs.
control_raw <- plain_files |>
  map(read_plain) |> bind_rows() |>
  mutate(model_key = str_remove(model_key, "-local$")) |>
  filter(year == params$year,
         model_key %in% ns_models$model_key,
         (model_key == "llama-70b"         & condition == "summarized") |
         (model_key == "llama-70b-ft-summ" & condition == "summarized-zeroshot")) |>
  transmute(model_key, source = country, indicator, rating_nn = rating)

panel_means   <- read_csv(file.path(data_dir, "panel_means.csv"),   show_col_types = FALSE)
human_ratings <- read_csv(file.path(data_dir, "human_ratings.csv"), show_col_types = FALSE)
ert <- read_csv(file.path(data_dir, "ert.csv"), show_col_types = FALSE) |>
  select(country_text_id, year, dem_ep, aut_ep)

pm_year <- panel_means |> filter(year == params$year) |>
  select(country_text_id, indicator, raw_mean)

# Human LOO per source cell (≥ min_coders), same construction as 06/09.
human_loo <- human_ratings |>
  filter(year == params$year) |>
  group_by(country_text_id, indicator) |>
  filter(n() >= params$min_coders) |>
  summarise(human_loo = mean(abs(rating - (sum(rating) - rating) / (n() - 1))),
            .groups = "drop")
```

**Load diagnostic** — confirm both arms and both models are present, and that `arm` split cleanly.

```{r load-diagnostic}
ns_raw |>
  count(model_key, arm, condition) |>
  mutate(model = recode(model_key, !!!model_labels)) |>
  arrange(model_key, arm) |>
  gt() |>
  tab_header(title = glue("Name-swap cells present for {params$year}"),
             subtitle = glue("{n_distinct(ns_raw$source)} source countries · ",
                             "no-name control rows: {nrow(control_raw)}"))
```

## 3. Metric construction (per CYI × model)

One row per (model, source, indicator): the swapped rating and both means, the correct-name and
no-name ratings on the same source cell, and the derived quantities.

```{r metrics}
swapped <- ns_raw |>
  filter(arm == "swapped") |>
  transmute(model_key, source, named, indicator, rating_sw = rating) |>
  inner_join(rename(pm_year, source = country_text_id, source_mean = raw_mean),
             by = c("source", "indicator")) |>
  inner_join(rename(pm_year, named = country_text_id, named_mean = raw_mean),
             by = c("named", "indicator"))

correct <- ns_raw |>
  filter(arm == "correct") |>
  transmute(model_key, source, indicator, rating_cor = rating)

ns <- swapped |>
  left_join(correct,     by = c("model_key", "source", "indicator")) |>
  left_join(control_raw, by = c("model_key", "source", "indicator")) |>
  left_join(rename(human_loo, source = country_text_id), by = c("source", "indicator")) |>
  mutate(
    # Metric 1 — tracking: positive = closer to source (content) than named (name).
    track       = abs(rating_sw - named_mean) - abs(rating_sw - source_mean),
    gap         = abs(source_mean - named_mean),         # conditioning variable
    signed_gap  = named_mean - source_mean,              # anchoring x-var (Metric 2)
    # Level errors (carry the LOO rail):
    src_err_sw  = abs(rating_sw  - source_mean),
    src_err_cor = abs(rating_cor - source_mean),
    named_err_sw= abs(rating_sw  - named_mean),
    # Metric 2 — cue-shift: the within-text effect of the name (swapped − correct name).
    cue         = rating_sw - rating_cor,
    flip        = as.integer(abs(rating_sw - rating_cor) >= 1)
  )

glue("per-CYI pool: {nrow(ns)} rows · ",
     "with correct-name arm {sum(!is.na(ns$rating_cor))} · ",
     "with no-name control {sum(!is.na(ns$rating_nn))} · ",
     "with human LOO {sum(!is.na(ns$human_loo))} · ",
     "median source–named gap {round(median(ns$gap, na.rm = TRUE), 3)}")
```

```{r gap-distribution}
#| fig-width: 7
#| fig-height: 3
ns |>
  ggplot(aes(gap)) +
  geom_histogram(bins = 40, fill = "#0072B2", alpha = 0.8) +
  labs(x = "source–named panel-mean gap  |source_mean − named_mean|", y = "CYIs",
       title = "Distribution of the source–named gap",
       subtitle = "Near-zero pairs carry no tracking signal; Metric 1 is read conditioned on this") +
  theme_minimal(base_size = 11) +
  theme(plot.title.position = "plot", plot.title = element_text(face = "bold"))
```

## 4. Bootstrap engine (country-clustered on source)

Seed-42 country-clustered resample via the shared helper. Per draw × model: the tracking mean, the
level MAEs (swapped/correct vs source, swapped vs named), the human LOO level, the cue-shift flip
rate and signed mean, and two slopes — tracking on the gap (the gap-conditioning) and cue-shift on
the signed gap (the anchoring slope). All weighted means use `na.rm` so cells missing the
correct-name arm or a LOO floor drop from that statistic only.

```{r bootstrap}
#| cache: true
#| cache.lazy: false

# weighted OLS slope of y on x (closed form), non-NA on both — as in 09.
wslope <- function(y, x, w) {
  m <- !is.na(y) & !is.na(x); wt <- w[m]; yy <- y[m]; xx <- x[m]
  W <- sum(wt); mx <- sum(wt * xx) / W; my <- sum(wt * yy) / W
  sum(wt * (xx - mx) * (yy - my)) / sum(wt * (xx - mx)^2)
}

W <- country_boot_weights(ns$source, params$n_boot, seed = 42)

boot <- map_dfr(colnames(W), \(draw) {
  cw <- tibble(source = rownames(W), w = W[, draw]) |> filter(w > 0)
  ns |>
    inner_join(cw, by = "source") |>
    group_by(model_key) |>
    summarise(
      track_mae    = weighted.mean(track,        w, na.rm = TRUE),
      src_sw_mae   = weighted.mean(src_err_sw,   w, na.rm = TRUE),
      src_cor_mae  = weighted.mean(src_err_cor,  w, na.rm = TRUE),
      named_sw_mae = weighted.mean(named_err_sw, w, na.rm = TRUE),
      human_mae    = weighted.mean(human_loo,    w, na.rm = TRUE),
      flip_rate    = weighted.mean(flip,         w, na.rm = TRUE),
      cue_mean     = weighted.mean(cue,          w, na.rm = TRUE),
      slope_gap    = wslope(track, gap,        w),   # tracking vs gap
      slope_anchor = wslope(cue,   signed_gap, w),   # cue-shift vs (named − source)
      .groups = "drop") |>
    mutate(id = draw)
})

# est / lo / hi / P(>0) per model × metric.
boot_ci <- boot |>
  pivot_longer(-c(id, model_key), names_to = "metric", values_to = "value") |>
  group_by(model_key, metric) |>
  summarise(
    est  = value[id == "Apparent"],
    lo   = quantile(value[id != "Apparent"], 0.025, na.rm = TRUE),
    hi   = quantile(value[id != "Apparent"], 0.975, na.rm = TRUE),
    p_gt = mean(value[id != "Apparent"] > 0, na.rm = TRUE),
    .groups = "drop") |>
  mutate(model = recode(model_key, !!!model_labels))

getm <- function(m) filter(boot_ci, metric == m)   # convenience accessor for figures
```

## 5. Metric 1 — Tracking (registered R4 quantity)

Positive = the swapped rating sits closer to the **source** (content) than the **named** country;
negative = it anchors to the name. Paired within case (the delta is formed per CYI before
resampling), country-clustered on source, with the ±SESOI band.

```{r}
#| label: fig-tracking
#| fig-width: 8
#| fig-height: 2.8
getm("track_mae") |>
  mutate(model = factor(model, levels = c("FT-summ", "Base"))) |>
  ggplot(aes(est, model)) +
  annotate("rect", xmin = -sesoi, xmax = sesoi, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = 0.55) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y", width = 0.15,
                linewidth = 0.7, color = "#0072B2") +
  geom_point(size = 3, color = "#0072B2") +
  labs(x = "tracking Δ = |r − named| − |r − source| · right = tracks source (content) · grey = ±SESOI",
       y = NULL,
       title = "Figure 1 — Name-swap tracking: content vs. name",
       subtitle = glue("2019 · positive = reads the text, not the name · paired · SESOI ±{round(sesoi, 3)}")) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(), plot.title.position = "plot",
        plot.title = element_text(face = "bold"))
```

```{r tracking-readout}
getm("track_mae") |>
  transmute(Model = model, `Tracking Δ` = est,
            `95% CI` = glue("[{round(lo, 3)}, {round(hi, 3)}]"),
            `P(Δ>0)` = p_gt,
            `Beyond SESOI?` = if_else(abs(est) > sesoi, "yes", "no")) |>
  gt() |>
  tab_header(title = "Metric 1 — tracking (R4 quantity, 2019 exploratory)",
             subtitle = glue("Positive = tracks source · paired CYI bootstrap · SESOI ±{round(sesoi, 3)}")) |>
  fmt_number(columns = c(`Tracking Δ`, `P(Δ>0)`), decimals = 3)
```

### Gap-conditioned tracking (the slope as a shape)

Near-equal pairs are mechanically ~0 and dilute the headline, so the design reports tracking
**conditioned on the source–named gap**. Two views: the bootstrapped **slope** of tracking on the
gap (should be positive — more signal where the two countries differ), and a descriptive gap-tercile
table (point estimates).

```{r tracking-slope}
getm("slope_gap") |>
  transmute(Model = model, `Slope (track on gap)` = est,
            `95% CI` = glue("[{round(lo, 3)}, {round(hi, 3)}]"), `P(slope>0)` = p_gt) |>
  gt() |>
  tab_header(title = "Tracking vs. source–named gap",
             subtitle = "slope>0 = tracking grows where the countries differ more") |>
  fmt_number(columns = c(`Slope (track on gap)`, `P(slope>0)`), decimals = 3)
```

```{r tracking-tercile}
ns |>
  filter(!is.na(gap)) |>
  group_by(model_key) |>
  mutate(gap_tercile = ntile(gap, 3)) |>
  group_by(model_key, gap_tercile) |>
  summarise(n = n(), track = mean(track, na.rm = TRUE),
            gap_mid = median(gap), .groups = "drop") |>
  mutate(model = recode(model_key, !!!model_labels),
         tercile = recode(gap_tercile, `1` = "low gap", `2` = "mid", `3` = "high gap")) |>
  select(model, tercile, n, `median gap` = gap_mid, `mean tracking Δ` = track) |>
  gt() |>
  tab_header(title = "Tracking by source–named gap tercile (descriptive)",
             subtitle = "Point estimates only — the CI on the trend is the bootstrapped slope above") |>
  fmt_number(columns = c(`median gap`, `mean tracking Δ`), decimals = 3)
```

## 6. Axis 2 — Coder axis: source-directed error vs. human LOO

The reframe. Even under a swapped name, is the rating's distance to the **source** panel within a
held-out human coder's own distance? Level MAEs on a common axis against the **human LOO** rail
(±SESOI band), for both arms: swapped-vs-source and correct-vs-source (the coder-distance whether or
not the name was swapped), with swapped-vs-named for contrast.

```{r}
#| label: fig-coder-axis
#| fig-width: 8.5
#| fig-height: 3.8
coder <- bind_rows(
  getm("src_sw_mae")   |> mutate(row = "swapped vs source"),
  getm("src_cor_mae")  |> mutate(row = "correct vs source"),
  getm("named_sw_mae") |> mutate(row = "swapped vs named")
) |>
  mutate(row = factor(row, levels = c("swapped vs named", "correct vs source", "swapped vs source")))

human_rail <- getm("human_mae") |> select(model, human_mae = est)

ggplot(coder, aes(est, row, color = model)) +
  geom_vline(data = human_rail, aes(xintercept = human_mae, color = model),
             linetype = "longdash", alpha = 0.7) +
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y",
                width = 0.2, linewidth = 0.7,
                position = position_dodge(width = 0.4)) +
  geom_point(size = 2.6, position = position_dodge(width = 0.4)) +
  scale_color_manual(values = c("Base" = "#0072B2", "FT-summ" = "#E69F00"), name = NULL) +
  labs(x = "MAE (rating vs. the stated benchmark) · long-dash = that model's human LOO",
       y = NULL,
       title = "Figure 2 — Coder axis: is the source-directed error within human range?",
       subtitle = glue("2019 · dashed = human LOO per model · lower = closer · SESOI ±{round(sesoi, 3)}")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", legend.justification = "left",
        panel.grid.major.y = element_blank(), plot.title.position = "plot",
        plot.title = element_text(face = "bold"))
```

```{r coder-readout}
bind_rows(
  getm("src_sw_mae")   |> mutate(Quantity = "swapped vs source"),
  getm("src_cor_mae")  |> mutate(Quantity = "correct vs source"),
  getm("named_sw_mae") |> mutate(Quantity = "swapped vs named"),
  getm("human_mae")    |> mutate(Quantity = "human LOO (source cells)")
) |>
  transmute(Model = model, Quantity, MAE = est,
            `95% CI` = glue("[{round(lo, 3)}, {round(hi, 3)}]")) |>
  arrange(Model, Quantity) |>
  gt() |>
  tab_header(title = "Axis 2 — level MAEs against the human LOO floor",
             subtitle = "Coder-terms reading: is the swapped rating within LOO of the source panel?") |>
  fmt_number(columns = MAE, decimals = 3)
```

## 7. Metric 2 — Cue-shift (exploratory companion)

The pure within-text name effect: `rating(swapped) − rating(correct)` on the *same* source text — no
panel mean, so immune to the oracle/coder ambiguity. Integer outputs make it coarse: a difference of
two integers, reported as a **flip rate** (|Δ| ≥ 1) and a signed mean.

```{r cueshift-readout}
bind_rows(
  getm("flip_rate") |> mutate(Quantity = "flip rate  P(|cue| ≥ 1)"),
  getm("cue_mean")  |> mutate(Quantity = "signed cue-shift  mean(swapped − correct)")
) |>
  transmute(Model = model, Quantity, est,
            `95% CI` = glue("[{round(lo, 3)}, {round(hi, 3)}]")) |>
  gt() |>
  tab_header(title = "Metric 2 — cue-shift (2019 exploratory)",
             subtitle = "flip rate = how often the name flips the integer; signed = net direction") |>
  fmt_number(columns = est, decimals = 3)
```

### Anchoring — does a higher-rated name pull the rating up?

Signed cue-shift against the signed gap `named_mean − source_mean`. A **positive slope** = swapping
in a more-democratic name pushes the rating up: the anchoring mechanism, stated directly.

```{r anchoring-slope}
getm("slope_anchor") |>
  transmute(Model = model, `Anchoring slope` = est,
            `95% CI` = glue("[{round(lo, 3)}, {round(hi, 3)}]"), `P(slope>0)` = p_gt) |>
  gt() |>
  tab_header(title = "Cue-shift vs. (named − source) panel-mean gap",
             subtitle = "slope>0 = a higher-rated injected name pulls the rating up (anchoring)") |>
  fmt_number(columns = c(`Anchoring slope`, `P(slope>0)`), decimals = 3)
```

```{r}
#| label: fig-anchoring
#| cache: true
#| cache.lazy: false
#| fig-width: 8
#| fig-height: 4.4
# 20 equal-count bins of the signed gap (per model), with the weighted fit + country-clustered
# ribbon — the 09-Fig-2 "slope as a shape" idiom.
Wf <- country_boot_weights(ns$source, 1000L, seed = 42)
fit_anchor <- function(mk) {
  d   <- filter(ns, model_key == mk, !is.na(cue), !is.na(signed_gap))
  lab <- recode(mk, !!!model_labels)
  xg  <- seq(quantile(d$signed_gap, 0.02), quantile(d$signed_gap, 0.98), length.out = 60)
  wfit <- function(w) {
    Wt <- sum(w); mx <- sum(w * d$signed_gap) / Wt; my <- sum(w * d$cue) / Wt
    b  <- sum(w * (d$signed_gap - mx) * (d$cue - my)) / sum(w * (d$signed_gap - mx)^2)
    c(a = my - b * mx, b = b)
  }
  f0 <- wfit(rep(1, nrow(d)))
  P  <- vapply(colnames(Wf)[-1], function(dr) {
    w <- unname(Wf[, dr][d$source]); f <- wfit(w); f[["a"]] + f[["b"]] * xg
  }, numeric(length(xg)))
  bins <- ntile(d$signed_gap, 20)
  list(
    line = tibble(model = lab, x = xg, yhat = f0[["a"]] + f0[["b"]] * xg,
                  lo = apply(P, 1, quantile, 0.025, na.rm = TRUE),
                  hi = apply(P, 1, quantile, 0.975, na.rm = TRUE)),
    pts  = tibble(model = lab, x = as.numeric(tapply(d$signed_gap, bins, mean)),
                  y = as.numeric(tapply(d$cue, bins, mean)))
  )
}
fits    <- map(ns_models$model_key, fit_anchor)
line_df <- bind_rows(map(fits, "line"))
pts_df  <- bind_rows(map(fits, "pts"))

ggplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "grey65") +
  geom_ribbon(data = line_df, aes(x, ymin = lo, ymax = hi, fill = model), alpha = 0.15) +
  geom_line(data = line_df, aes(x, yhat, color = model), linewidth = 1.1) +
  geom_point(data = pts_df, aes(x, y, color = model), size = 1.5, alpha = 0.65) +
  scale_color_manual(values = c("Base" = "#0072B2", "FT-summ" = "#E69F00"),
                     aesthetics = c("color", "fill"), name = NULL) +
  labs(x = "named − source panel mean  (higher = swapped-in name is more democratic)",
       y = "cue-shift  (swapped − correct rating)",
       title = "Figure 3 — Anchoring: does the injected name pull the rating toward its reputation?",
       subtitle = "above 0 & rising = a more-democratic name raises the rating · points = 20 equal-count bins") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", legend.justification = "left",
        plot.title.position = "plot", plot.title = element_text(face = "bold"))
```

### Three-point ladder (none → correct → swapped)

The no-name control decomposes the name effect: **none → correct** = the effect of adding the
*true* name (should be small if the model reads content); **correct → swapped** = the cue-shift (the
effect of the name being *wrong*). Both signed, per model.

```{r ladder}
ns |>
  group_by(model_key) |>
  summarise(
    `add true name (correct − none)` = mean(rating_cor - rating_nn, na.rm = TRUE),
    `swap name (swapped − correct)`  = mean(rating_sw  - rating_cor, na.rm = TRUE),
    n_none = sum(!is.na(rating_nn)), .groups = "drop") |>
  mutate(model = recode(model_key, !!!model_labels)) |>
  select(model, `add true name (correct − none)`, `swap name (swapped − correct)`, n_none) |>
  gt() |>
  tab_header(title = "Name-effect ladder (descriptive point estimates)",
             subtitle = "none = existing name-hidden summarized runs · signed rating change") |>
  fmt_number(columns = 2:3, decimals = 3)
```

## 8. Transition-adjacency split (secondary, descriptive)

Per the design note, transition is a **split of the result, not a pairing criterion** — the
source–named gap concentrates in transition-adjacent countries, so tracking signal does too. A
single descriptive table (09 showed the 2019 movement signal is underpowered under clustering, so no
DiD engine here): mean tracking Δ for source countries in an ERT episode vs. stable.

```{r transition-split}
ert_flag <- ert |>
  filter(year == params$year) |>
  transmute(source = country_text_id,
            in_ep = replace_na((dem_ep == 1) | (aut_ep == 1), FALSE))

ns |>
  left_join(ert_flag, by = "source") |>
  mutate(in_ep = replace_na(in_ep, FALSE),
         grp = if_else(in_ep, "transition-adjacent", "stable")) |>
  group_by(model_key, grp) |>
  summarise(n = n(), `mean tracking Δ` = mean(track, na.rm = TRUE),
            `median gap` = median(gap, na.rm = TRUE), .groups = "drop") |>
  mutate(model = recode(model_key, !!!model_labels)) |>
  select(model, grp, n, `median gap`, `mean tracking Δ`) |>
  gt() |>
  tab_header(title = "Table — tracking by transition-adjacency (2019, descriptive)",
             subtitle = "ERT episode flag on the source country · point estimates (see 09 on 2019 power)") |>
  fmt_number(columns = c(`median gap`, `mean tracking Δ`), decimals = 3)
```

## 9. Serialize the bundle for the paper documents

```{r serialize-bundle}
bundle_path <- file.path(derived_dir, "nameswap_2019.rds")
saveRDS(
  list(
    boot_ci = boot_ci,   # per-model × metric: est/lo/hi/p_gt (tracking, level MAEs, cue-shift, slopes)
    ns      = ns,        # per-CYI table (for any paper-side re-cut; carries both means + all arms)
    sesoi   = sesoi
  ),
  bundle_path
)
glue("bundle written: {bundle_path} ({round(file.size(bundle_path)/1024)} KB)")
```
