---
title: "2019 secondary & robustness metrics — RMSE, exact-match/adjacent, QWK, Krippendorff α"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
params:
  n_boot: 2000
  year: 2019
  # Confusion-matrix panels: base:Cb, base:Ev (reading-vs-reciting), and the carried cell.
  # Carried model is finalized at 2019 unblinding; FT-raw:Ev is the working proxy (the cell
  # the signed-deviation doc also treats as protagonist). Override when the rule is applied.
  carried_key: "llama-70b-ft-raw"
  carried_cond: "evidence-zeroshot"
---

Plan and rationale: `notes/secondary-and-robustness-metrics-roadmap.md`. This document re-scores
the identical 2019 ratings `06` scores, under a different loss (squared error) and different
agreement metrics (exact-match, adjacent-category, quadratic weighted kappa, Krippendorff's α).
Nothing here is confirmatory: RMSE is a robustness re-run of `06`'s registered contrasts, and the
agreement metrics are secondary/descriptive summaries. The identification story stays entirely on
AI MAE. See the roadmap's integrity note for why adding these on 2019 is clean.

**Reference object.** Exact-match, adjacent, QWK, and the confusion matrix are scored against
`round(panel_mean)` — the same reference object `06`'s MAE uses, collapsed to one gold class — so
the secondary table reads as "same reference, different loss." The per-coder / panel-distribution
framing has its homes in `07` (distributional match) and §Check-4 (α), not here.

**Mode/mean caveat, carried throughout.** Greedy decoding emits the *mode*, and
`round(panel_mean) ≠ panel mode` in general, so exact-match is slightly misaligned with what the
decoder actually optimizes. High exact-match is a raw hit rate on the rounded rating, **not**
"closer to consensus" (that is MAE's job). The FT models are mode-targeted and should look best on
exact-match while "losing" mean-MAE; that split is itself the evidence for the mode/mean story in
`notes/mode-vs-mean-decode-and-act1-next-steps.md`.

The paper-facing narrative lives in the deliverable; this is the working readout.

## 1. Setup

```{r setup}
library(tidyverse)
library(gt)
library(glue)

if (!exists("params")) {
  params <- list(n_boot = 2000, year = 2019,
                 carried_key = "llama-70b-ft-raw", carried_cond = "evidence-zeroshot")
}

# Same root finder as 06/08/11, so the doc knits from any working dir in the workspace.
find_panel_member_root <- function() {
  up <- tryCatch(rprojroot::find_root(rprojroot::is_git_root), error = function(e) NA_character_)
  if (!is.na(up)) return(up)
  down <- file.path(getwd(), "panel-member")
  if (dir.exists(file.path(down, ".git"))) return(down)
  stop("Could not locate the panel-member project root from working dir: ", getwd())
}
proj_root   <- find_panel_member_root()
source(file.path(proj_root, "R", "bootstrap_helpers.R"))  # model_labels, col_labels, canon_col, country_boot_weights, ...
data_dir    <- file.path(proj_root, "data", "processed")
runs_dir    <- file.path(proj_root, "data", "output", "runs")
derived_dir <- file.path(proj_root, "data", "derived")

# SESOI + rounding floor are fixed by 06; carry them over so every doc bands on one value.
.b06           <- readRDS(file.path(derived_dir, "bootstrap_2019.rds"))
sesoi          <- .b06$sesoi
rounding_floor <- 2 * sesoi
# The coded indicators are ordinal but on HETEROGENEOUS scales: most are 0..4, but 12 run 0..5
# and others 0..1/0..2/0..3. 0..5 is the widest in the AI-coded set (verified: no AI rating ever
# exceeds its indicator's human scale), so K = 6 classes covers every rating. The class grid is
# global and pooled across scales, exactly as 06/07/08 pool all indicators for MAE — see the
# mixed-scale caveat under Checks 2/3.
K              <- 6L   # 0..5 -> 6 classes (widest ordinal scale in the coded set)

# A valid rating is an integer on the 0..K-1 scale; anything else (NA, negative, non-integer, or
# above 5) is dropped before scoring so it cannot corrupt a confusion table or agreement metric.
# In practice nothing legitimate is dropped: no model emits a rating above its indicator's scale.
valid_rating <- function(x) !is.na(x) & x == round(x) & x >= 0 & x <= (K - 1)

# α (Check 4) needs the `irr` package; the section self-gates so the doc knits without it.
have_irr <- requireNamespace("irr", quietly = TRUE)
```

## 2. Data loading

The three sources `06` uses: the JSONL runs, the panel means (the `raw_mean` reference), and the
coder-level ratings (for the human LOO reference lines). Loaded once, cached.

```{r load-data}
#| cache: true
#| cache.lazy: false
needed_cols <- c("country", "year", "indicator", "condition", "model_key", "rating")
read_run <- function(f) {
  con <- file(f, "r"); on.exit(close(con))
  jsonlite::stream_in(con, verbose = FALSE) |> as_tibble() |> select(all_of(needed_cols))
}
ai_raw <- list.files(runs_dir, pattern = "\\.jsonl$", full.names = TRUE) |>
  map(read_run) |> bind_rows() |> mutate(model_key = str_remove(model_key, "-local$"))

panel_means   <- read_csv(file.path(data_dir, "panel_means.csv"),   show_col_types = FALSE)
human_ratings <- read_csv(file.path(data_dir, "human_ratings.csv"), show_col_types = FALSE)
```

## 3. Metric construction

### The 16-cell grid and the gold class

Rows are the 4 models × 4 conditions. Base runs its four conditions few-shot; each FT adapter runs
codebook plus the three zero-shot text conditions, which `canon_col()` collapses to the same
Cb/Ev/An/Su columns. The gold class is `round(panel_mean)` (round-half-up, matching `08`'s
`floor(x + 0.5)`), and the AI class is its integer rating.

```{r grid}
grid_cells <- bind_rows(
  tibble(model_key = "llama-70b",
         condition = c("codebook", "evidence", "anonymized", "summarized")),
  tidyr::crossing(
    model_key = c("llama-70b-ft-raw", "llama-70b-ft-anon", "llama-70b-ft-summ"),
    condition = c("codebook", "evidence-zeroshot", "anonymized-zeroshot", "summarized-zeroshot"))
) |>
  mutate(canon = canon_col(condition),
         cell  = paste(model_key, condition, sep = "|"),
         cell_lbl = paste0(recode(model_key, !!!model_labels), ":", recode(canon, !!!col_labels)))

cell_levels <- grid_cells$cell
cell_lab_of <- setNames(grid_cells$cell_lbl, grid_cells$cell)

# The 7 "core" cells featured across 06-11: the base ladder (few-shot) + the FT diagonal (each FT
# adapter on the input it was built for). Level figures (Checks 2/3) and the α table focus on these;
# the other 9 off-diagonal cells go to the appendix figures at the bottom.
core_cells <- c("llama-70b|codebook", "llama-70b|evidence", "llama-70b|anonymized",
                "llama-70b|summarized", "llama-70b-ft-raw|evidence-zeroshot",
                "llama-70b-ft-anon|anonymized-zeroshot", "llama-70b-ft-summ|summarized-zeroshot")
core_lbl  <- unname(cell_lab_of[core_cells])
other_lbl <- setdiff(unname(cell_lab_of[cell_levels]), core_lbl)

pm_year <- panel_means |>
  filter(year == params$year) |>
  transmute(country_text_id, indicator, raw_mean, gold = as.integer(floor(raw_mean + 0.5)))

# The indicators the AI actually coded (the grid runs), used to restrict the human reference to
# the SAME indicator set. human_ratings.csv also carries human-only indicators on other scales
# (e.g. 0..13, 0..100 interval variables) that the AI never coded; without this restriction the
# human LOO reference would be contaminated by indicators absent from the AI comparison.
ai_indicators <- ai_raw |>
  filter(year == params$year) |>
  inner_join(grid_cells |> select(model_key, condition), by = c("model_key", "condition")) |>
  distinct(indicator) |> pull(indicator)

# Per-CYI AI metrics on the 16 cells, restricted (like 06) to CYIs carrying a human LOO floor,
# so the AI and human references live on the identical case set.
```

### Human LOO reference (per coder-cell)

Each held-out coder's rating scored against the rounded *reduced* mean (the panel mean recomputed
with them removed). This is the LOO analogue of every secondary metric: exact-match, adjacent, and
squared error per coder-cell, plus the (gold, rating) pairs QWK/α read. Restricted to CYIs with
≥ 2 coders (an LOO floor needs one to hold out and ≥ 1 to average).

```{r human-loo}
#| cache: true
human_loo <- human_ratings |>
  filter(year == params$year, valid_rating(rating), indicator %in% ai_indicators) |>
  group_by(country_text_id, indicator) |>
  filter(n() >= 2) |>
  mutate(loo_mean  = (sum(rating) - rating) / (n() - 1),
         loo_gold  = as.integer(floor(loo_mean + 0.5)),
         h_se      = (rating - loo_mean)^2,
         h_exact   = as.integer(rating == loo_gold),
         h_adj     = as.integer(abs(rating - loo_gold) <= 1)) |>
  ungroup()

# The case set: CYIs with a human LOO floor. AI metrics are inner-joined to it.
loo_cyi <- human_loo |> distinct(country_text_id, indicator)
```

### AI per-CYI metrics on the LOO case set

```{r ai-metrics}
ai_joined <- ai_raw |>
  filter(year == params$year) |>
  rename(country_text_id = country) |>
  inner_join(grid_cells |> select(model_key, condition, canon, cell), by = c("model_key", "condition")) |>
  inner_join(pm_year,  by = c("country_text_id", "indicator")) |>
  inner_join(loo_cyi,  by = c("country_text_id", "indicator"))

# A rater that emits an unparseable / off-scale value must not silently corrupt the confusion
# table or the agreement metrics. Drop ratings not integer-valued in 0..K-1 (incl. NA), and
# report how many per cell so the loss is visible rather than silent (valid_rating from setup).
dropped <- ai_joined |> filter(!valid_rating(rating)) |> count(cell, name = "n_dropped")

ai_metrics <- ai_joined |>
  filter(valid_rating(rating)) |>
  mutate(rating = as.integer(rating),
         cell   = factor(cell, levels = cell_levels),
         se     = (rating - raw_mean)^2,       # squared error (Check 1)
         exact  = as.integer(rating == gold),  # 0/1 vs rounded mean (Check 2)
         adj    = as.integer(abs(rating - gold) <= 1))

cat("AI rows:", nrow(ai_metrics), "· cells present:",
    nlevels(droplevels(ai_metrics$cell)), "of", length(cell_levels), "\n")
cat("Dropped off-scale/NA ratings:", sum(dropped$n_dropped), "\n")
if (nrow(dropped) > 0) print(dropped)
cat("LOO case CYIs:", nrow(loo_cyi), "· coder-cells:", nrow(human_loo), "\n")
```

### Shared country-clustered weights

One weight matrix over the country universe (seed 42), reused by every metric below so all CIs
share one resampling scheme — the same `country_boot_weights()` primitive `06`/`08` use. A cell (or
the human reference) broadcasts each draw's per-country weight onto its own rows.

```{r weights}
#| cache: true
countries_all <- sort(unique(c(ai_metrics$country_text_id, human_loo$country_text_id)))
W <- country_boot_weights(countries_all, params$n_boot, seed = 42)
draw_id <- colnames(W)                 # col 1 = "Apparent"
app     <- which(draw_id == "Apparent")
boot    <- setdiff(seq_along(draw_id), app)

# Weighted-mean-per-draw of a per-row value, grouped into `ncell` blocks, via rowsum (as in 08).
# Returns an (ndraw x ncell) matrix. `grp` is a 1..ncell integer per row; `cid` the country id.
wmean_by_cell <- function(val, grp, cid, ncell) {
  Wc <- W[cid, , drop = FALSE]                              # (nrow x ndraw)
  num <- rowsum(Wc * val, grp, reorder = FALSE)             # (ncell x ndraw)
  den <- rowsum(Wc,       grp, reorder = FALSE)
  # rowsum keeps only present groups; scatter back to full 1..ncell then transpose.
  present <- as.integer(rownames(num))
  M <- matrix(NA_real_, ncell, length(draw_id))
  M[present, ] <- num / den
  t(M)
}
```

---

## Check 1 — RMSE robustness on the registered contrasts

Re-score 2019 under squared error and re-run `06`'s registered paired contrasts, asking
verdict-by-verdict whether each SESOI call survives the L1→L2 switch. RMSE-vs-mean is a genuine
robustness check on the L1-vs-L2 choice (per-coder RMSE would add nothing — squared error to each
coder equals panel-mean MSE plus a per-cell constant, so every contrast is identical; see roadmap).

The contrast machinery mirrors `06`'s `paired_delta`, but on `ai_rmse` rather than `ai_mae`, so
this rebuilds a per-draw metric table from the raw squared errors rather than reading `06`'s
`ai_mae` bundle (which stores L1 only). Resampling is country-clustered, seed 42 — identical to the
MAE contrasts, so the two are comparable.

**SESOI caveat.** SESOI (± `r round(sesoi, 3)`) was calibrated on the L1 axis as half the 2019
rounding floor. RMSE deltas are on the root-mean-square axis, a different scale, so applying the
same numeric band is a deliberate common-yardstick choice (per the roadmap's "same SESOI verdict
rule"), not an exact re-derivation. Read the "survives" column as "does the L2 delta clear the same
±SESOI ruler," and weight the sign/ordering as much as the strict verdict.

```{r rmse-boot}
#| cache: true
#| cache.lazy: false
# Per-draw RMSE per cell = sqrt(weighted-mean squared error). Reuse the block aggregator on `se`,
# then sqrt. Cells map to 1..ncell by factor level.
ncell   <- length(cell_levels)
grp_ai  <- as.integer(ai_metrics$cell)
mse_mat <- wmean_by_cell(ai_metrics$se, grp_ai, ai_metrics$country_text_id, ncell)  # (ndraw x ncell)
rmse_mat <- sqrt(mse_mat)
colnames(rmse_mat) <- cell_levels

# Generic paired within-draw delta on any (ndraw x cell) metric matrix, against ±SESOI —
# the L2 analogue of R/bootstrap_helpers.R::paired_delta, parameterized on the metric.
paired_metric_delta <- function(M, a_cell, b_cell, label, sesoi_val = sesoi) {
  if (!all(c(a_cell, b_cell) %in% colnames(M)))
    return(tibble(label = label, est = NA, lo = NA, hi = NA, p_gt = NA, sesoi_out = NA))
  d <- M[, a_cell] - M[, b_cell]
  tibble(label = label,
         est  = d[app], lo = quantile(d[boot], 0.025), hi = quantile(d[boot], 0.975),
         p_gt = mean(d[boot] > 0), sesoi_out = abs(d[app]) > sesoi_val)
}

# The registered contrasts, exactly as 06 states them (cell keys = model_key|condition).
B  <- "llama-70b|codebook"; EV <- "llama-70b|evidence"
AN <- "llama-70b|anonymized"; SU <- "llama-70b|summarized"
rmse_contrasts <- bind_rows(
  # B-series (Figure 2): identity-stripping pushes error out; reading evidence does not.
  paired_metric_delta(rmse_mat, EV, B,  "B1 · Ev−Cb"),
  paired_metric_delta(rmse_mat, AN, B,  "B2 · An−Cb"),
  paired_metric_delta(rmse_mat, SU, B,  "Su−Cb (descr.)"),
  paired_metric_delta(rmse_mat, AN, EV, "B3 · An−Ev"),
  paired_metric_delta(rmse_mat, SU, EV, "Su−Ev (descr.)"),
  # F1 (appendix): fine-tuning vs. few-shot on matched conditions.
  paired_metric_delta(rmse_mat, "llama-70b-ft-raw|evidence-zeroshot",   EV, "F1 · Raw Text"),
  paired_metric_delta(rmse_mat, "llama-70b-ft-anon|anonymized-zeroshot", AN, "F1 · Anonymized"),
  paired_metric_delta(rmse_mat, "llama-70b-ft-summ|summarized-zeroshot", SU, "F1 · Summarized")
)
```

The MAE verdicts to compare against, read from `06`'s serialized `paired_delta` layer, so the
"survives" column is L1-verdict vs L2-verdict on the same rows.

```{r rmse-vs-mae}
# Recompute the L1 verdicts from 06's boot_results with the shared helper, so both columns use
# the identical SESOI rule. (list2env exposes boot_results/sesoi for paired_delta's defaults.)
list2env(.b06[c("boot_results", "sesoi")], environment())
mae_contrasts <- bind_rows(
  paired_delta("llama-70b", "evidence",   "llama-70b", "codebook",   "B1 · Ev−Cb"),
  paired_delta("llama-70b", "anonymized", "llama-70b", "codebook",   "B2 · An−Cb"),
  paired_delta("llama-70b", "summarized", "llama-70b", "codebook",   "Su−Cb (descr.)"),
  paired_delta("llama-70b", "anonymized", "llama-70b", "evidence",   "B3 · An−Ev"),
  paired_delta("llama-70b", "summarized", "llama-70b", "evidence",   "Su−Ev (descr.)"),
  paired_delta("llama-70b-ft-raw",  "evidence-zeroshot",   "llama-70b", "evidence",   "F1 · Raw Text"),
  paired_delta("llama-70b-ft-anon", "anonymized-zeroshot", "llama-70b", "anonymized", "F1 · Anonymized"),
  paired_delta("llama-70b-ft-summ", "summarized-zeroshot", "llama-70b", "summarized", "F1 · Summarized")
) |> transmute(label, mae_est = est, mae_sesoi_out = sesoi_out)

rmse_tbl <- rmse_contrasts |>
  left_join(mae_contrasts, by = "label") |>
  mutate(survives = if_else(sesoi_out == mae_sesoi_out, "yes", "FLIPPED"))

# Shared renderer for the L1-vs-L2 survival table, called once per contrast group so the B-series
# and F1 tables stay separate (mirroring 06's split readouts). Strips the "F1 · " label prefix.
survival_gt <- function(df, title) {
  df |>
    transmute(Contrast = str_remove(label, "^F1 · "),
              `ΔMAE` = round(mae_est, 3), `MAE > SESOI` = if_else(mae_sesoi_out, "yes", "no"),
              `ΔRMSE` = round(est, 3), `RMSE 95% CI` = glue("[{round(lo,3)}, {round(hi,3)}]"),
              `RMSE > SESOI` = if_else(sesoi_out, "yes", "no"),
              `Survives L1→L2` = survives) |>
    gt() |>
    tab_header(title = title,
               subtitle = glue("Paired country-clustered bootstrap · B = {params$n_boot} · ",
                               "SESOI ±{round(sesoi, 3)} (L1-calibrated, common ruler)"))
}
```

### RMSE robustness — base model (mirror of `06` Figure 2)

The base-model B-series re-scored under squared error, in the same layout as `06`'s Figure 2:
**top group** = each condition vs. the Codebook prior; **bottom group** = the de-identified
conditions vs. full Evidence. The question is only whether each SESOI verdict survives L1→L2.

```{r rmse-bseries-tbl}
survival_gt(filter(rmse_tbl, !str_starts(label, "F1")),
            "RMSE robustness — base-model identification contrasts (B-series)")
```

```{r rmse-fig-bseries}
#| fig-height: 4.0
rmse_b <- rmse_contrasts |>
  filter(!str_starts(label, "F1")) |>
  mutate(label = factor(label, levels = rev(c(
    "B1 · Ev−Cb", "B2 · An−Cb", "Su−Cb (descr.)", "B3 · An−Ev", "Su−Ev (descr.)"))))

ggplot(rmse_b, aes(est, label)) +
  annotate("rect", xmin = -sesoi, xmax = sesoi, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = 0.55) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  # separator between the "vs Codebook" group (top 3) and the "vs Evidence" group (bottom 2)
  geom_hline(yintercept = 2.5, color = "grey85", linewidth = 0.4) +
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y",
                width = 0.15, linewidth = 0.7, color = "#0072B2") +
  geom_point(size = 3, color = "#0072B2") +
  scale_x_continuous(expand = expansion(mult = c(0.06, 0.06))) +
  labs(x = "ΔRMSE · right = higher error · grey band = ±SESOI", y = NULL,
       title = "RMSE robustness — base model: reading the evidence or reading the country",
       subtitle = glue("2019 · top: vs Codebook · bottom: vs Evidence · paired · SESOI ±{round(sesoi, 3)}")) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(), plot.title.position = "plot",
        plot.title = element_text(face = "bold"))
```

### RMSE robustness (appendix) — F1: fine-tuning vs. few-shot

The registered F1 contrast (each FT adapter minus the matched-condition base cell) under squared
error, kept as a separate figure exactly as `06` does — different anchor (FT − matched base),
different narrative. Left = fine-tuning reduces error.

```{r rmse-f1-tbl}
survival_gt(filter(rmse_tbl, str_starts(label, "F1")),
            "RMSE robustness — F1: fine-tuning vs. few-shot (matched)")
```

```{r rmse-fig-f1}
#| fig-height: 3.2
rmse_f1 <- rmse_contrasts |>
  filter(str_starts(label, "F1")) |>
  mutate(label = factor(str_remove(label, "^F1 · "),
                        levels = rev(c("Raw Text", "Anonymized", "Summarized"))))

ggplot(rmse_f1, aes(est, label)) +
  annotate("rect", xmin = -sesoi, xmax = sesoi, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = 0.55) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y",
                width = 0.15, linewidth = 0.7, color = "#E69F00") +
  geom_point(size = 3, color = "#E69F00") +
  scale_x_continuous(expand = expansion(mult = c(0.08, 0.08))) +
  labs(x = "ΔRMSE (FT − base) · left = fine-tuning reduces error · grey band = ±SESOI", y = NULL,
       title = "RMSE robustness (appendix) — F1: fine-tuning vs. few-shot (matched)",
       subtitle = glue("2019 · registered F1 under L2 · SESOI ±{round(sesoi, 3)}")) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(), plot.title.position = "plot")
```

*Note.* The R-series contrasts (R2 in `09`, R5 in `11`) are difference-in-differences, not simple
paired cell deltas, so their L2 robustness belongs in those docs if wanted, not here. This check
covers the `06` contrasts (B-series and F1) — the registered paired-MAE deltas.

---

## Check 2 — Exact-match and adjacent-category agreement (levels)

Per-cell 0/1 metrics against `round(panel_mean)`, so they support the same country-clustered
weighting as MAE. Exact-match averaged to the rounded mean is the raw hit rate; adjacent is the
tri-diagonal band (within one category). High adjacent + low exact is the expected signature of an
integer rater against a fractional mean (close but rarely exact). A human LOO reference row is
included, exactly as MAE carries one.

```{r exact-adj-levels}
#| fig-height: 4.6
ex_mat  <- wmean_by_cell(ai_metrics$exact, grp_ai, ai_metrics$country_text_id, ncell)
adj_mat <- wmean_by_cell(ai_metrics$adj,   grp_ai, ai_metrics$country_text_id, ncell)
colnames(ex_mat) <- colnames(adj_mat) <- cell_levels

level_ci <- function(M, labels = colnames(M)) {
  tibble(cell = labels, est = M[app, ],
         lo = apply(M[boot, , drop = FALSE], 2, quantile, 0.025, na.rm = TRUE),
         hi = apply(M[boot, , drop = FALSE], 2, quantile, 0.975, na.rm = TRUE))
}
ex_lvl  <- level_ci(ex_mat)  |> mutate(metric = "Exact")
adj_lvl <- level_ci(adj_mat) |> mutate(metric = "Adjacent")

# Human LOO reference: one row, per-coder exact/adjacent, country-clustered on the coder-cells.
h_grp <- rep(1L, nrow(human_loo))
h_ex  <- wmean_by_cell(human_loo$h_exact, h_grp, human_loo$country_text_id, 1L)[, 1]
h_adj <- wmean_by_cell(human_loo$h_adj,   h_grp, human_loo$country_text_id, 1L)[, 1]
human_ref_row <- tibble(
  cell = "Human (LOO)",
  ex_est = h_ex[app], ex_lo = quantile(h_ex[boot], 0.025), ex_hi = quantile(h_ex[boot], 0.975),
  adj_est = h_adj[app], adj_lo = quantile(h_adj[boot], 0.025), adj_hi = quantile(h_adj[boot], 0.975))

exact_adj_tbl <- ex_lvl |>
  transmute(Cell = cell_lab_of[cell],
            `Exact` = est, `exact 95% CI` = glue("[{scales::percent(lo,0.1)}, {scales::percent(hi,0.1)}]")) |>
  left_join(adj_lvl |> transmute(Cell = cell_lab_of[cell],
            `Adjacent` = est, `adj 95% CI` = glue("[{scales::percent(lo,0.1)}, {scales::percent(hi,0.1)}]")),
            by = "Cell") |>
  bind_rows(human_ref_row |> transmute(Cell = cell,
            Exact = ex_est, `exact 95% CI` = glue("[{scales::percent(ex_lo,0.1)}, {scales::percent(ex_hi,0.1)}]"),
            Adjacent = adj_est, `adj 95% CI` = glue("[{scales::percent(adj_lo,0.1)}, {scales::percent(adj_hi,0.1)}]")))

# Coefficient plot (mirrors 07/08): 16 cells as points+CIs coloured Base vs Fine-tuned, faceted by
# metric, with the human LOO level as a dashed reference line and its 95% CI as a grey band. The
# gt-source `exact_adj_tbl` above is kept for the serialized bundle / paper.
ea_long <- bind_rows(
  ex_lvl  |> transmute(cell_lbl = cell_lab_of[cell], metric = "Exact match",    est, lo, hi),
  adj_lvl |> transmute(cell_lbl = cell_lab_of[cell], metric = "Adjacent (±1)", est, lo, hi)
) |>
  mutate(family = factor(if_else(str_starts(cell_lbl, "FT"), "Fine-tuned", "Base"),
                         levels = c("Base", "Fine-tuned")),
         metric = factor(metric, levels = c("Exact match", "Adjacent (±1)")))

# Shared y-order across both facets: grouped by family (Base at top, FT below, as in 06-08),
# ranked by exact-match within each group. `arrange(desc(fam), est)` yields bottom-to-top order:
# FT (ascending) then Base (ascending), so Base sits above FT with the best cell atop each group.
cell_order  <- ex_lvl |>
  mutate(l = cell_lab_of[cell], fam = if_else(str_starts(l, "FT"), "Fine-tuned", "Base")) |>
  arrange(desc(fam), est) |> pull(l)
ea_long     <- ea_long |> mutate(cell_lbl = factor(cell_lbl, levels = cell_order))
human_ref_ea <- bind_rows(
  tibble(metric = "Exact match",    est = human_ref_row$ex_est,  lo = human_ref_row$ex_lo,  hi = human_ref_row$ex_hi),
  tibble(metric = "Adjacent (±1)", est = human_ref_row$adj_est, lo = human_ref_row$adj_lo, hi = human_ref_row$adj_hi)
) |> mutate(metric = factor(metric, levels = c("Exact match", "Adjacent (±1)")))

# Reusable coefficient plot over a subset of cell labels (main = core cells; appendix = the rest),
# keeping the shared exact-match ranking within the subset.
ea_coef_plot <- function(labs, subtitle) {
  ea_long |>
    filter(cell_lbl %in% labs) |>
    mutate(cell_lbl = factor(as.character(cell_lbl), levels = intersect(cell_order, labs))) |>
    ggplot(aes(est, cell_lbl, color = family)) +
    geom_rect(data = human_ref_ea, inherit.aes = FALSE,
              aes(xmin = lo, xmax = hi, ymin = -Inf, ymax = Inf), fill = "grey85", alpha = 0.5) +
    geom_vline(data = human_ref_ea, aes(xintercept = est), linetype = "dashed", color = "grey40") +
    geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y", width = 0.3, linewidth = 0.6) +
    geom_point(size = 2.6) +
    facet_wrap(~ metric, scales = "free_x") +
    scale_color_manual(values = family_pal, name = NULL) +
    scale_x_continuous(labels = scales::percent, expand = expansion(mult = c(0.06, 0.06))) +
    labs(x = "agreement vs round(panel_mean) · dashed = human LOO (grey band = its 95% CI)", y = NULL,
         title = "Exact-match and adjacent-category agreement (levels)", subtitle = subtitle) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "top", legend.justification = "left",
          panel.grid.minor = element_blank(), panel.spacing.x = grid::unit(1.4, "lines"),
          plot.title.position = "plot", plot.title = element_text(face = "bold"))
}

ea_coef_plot(core_lbl, glue("2019 · core cells (base ladder + FT diagonal) · country-clustered 95% CI · ",
                            "B = {params$n_boot} · Base over FT, ranked within"))
```

*Mode caveat.* `round(panel_mean)` is not the panel mode, so even this exact-match is slightly
misaligned with what greedy emits; do not narrate exact-match as consensus-fidelity. The FT models
being mode-targeted should show here.

*Mixed-scale caveat.* Exact and adjacent pool indicators of different ordinal widths (0..1 up to
0..5). Both are scale-agnostic per cell, but the *baseline rate* differs by width: on a 0..1 binary
indicator adjacent-agreement (`|rating − gold| ≤ 1`) is always 1, and exact-match is a coin-flip
baseline, so short-scale indicators inflate both. The pooled numbers are read as a within-column
contrast across models/conditions (the pool is fixed), not as absolute agreement rates.

---

## Check 3 — Quadratic weighted kappa (levels)

QWK is chance-corrected ordinal agreement between the AI class and `round(panel_mean)`, the one
scalar that gives no credit for marginal / central-tendency matching (its chance term is computed
from both marginals), which is the direct rebuttal to "MAE rewards central tendency." It is a
**sample-level** metric (defined over a *set* of cells), so there is no per-cell value and no
within-cell pairing: reported as a level per model × condition with a country-clustered bootstrap
CI, compared by CI overlap. It re-imposes a single gold label, so it is a secondary agreement
number, never a target.

QWK is computed from a weighted K×K contingency table (K = 6, spanning the 0..5 union of scales;
rows = gold, cols = AI), so the country-clustered weights enter directly as table mass — no row
expansion needed. Verified against `psych::cohen.kappa` at the point estimate below.

**Mixed-scale caveat.** Ratings are pooled across indicators of different ordinal widths (0..1 up
to 0..5), the same pool 06 uses for MAE. QWK's quadratic weights are normalized by `(K-1)^2 = 25`,
so a one-step disagreement on a narrow (e.g. 0..1) indicator contributes less weighted disagreement
than the same step on a 0..5 indicator, and the pooled scalar is dominated by the wider scales.
This is inherent to a single pooled kappa (as raw-rating-point MAE is), not a defect of this
computation; a per-indicator-then-averaged QWK is the more principled refinement if ever wanted.

```{r qwk-helper}
# Quadratic-weighted kappa from a K×K count matrix (weighted counts allowed). Weights
# w_ij = (i-j)^2/(K-1)^2; kappa = 1 - sum(w*O)/sum(w*E), E = outer(row, col)/N under independence.
qwk_from_counts <- function(N) {
  Wt <- outer(0:(K-1), 0:(K-1), \(i, j) (i - j)^2) / (K - 1)^2
  tot <- sum(N); if (tot <= 0) return(NA_real_)
  O <- N / tot
  E <- outer(rowSums(O), colSums(O))
  denom <- sum(Wt * E)
  if (denom == 0) return(NA_real_)
  1 - sum(Wt * O) / denom
}

# Per-draw QWK for a cell: aggregate country weights into the K*K (gold,AI) bins via rowsum, then
# reshape each draw's column to K×K. `gold`,`pred` are 0..K-1; `cid` the country ids.
qwk_draws <- function(gold, pred, cid) {
  keep <- valid_rating(gold) & valid_rating(pred)   # defensive: only in-scale integer pairs
  gold <- gold[keep]; pred <- pred[keep]; cid <- cid[keep]
  bin <- gold * K + pred + 1L                       # 1..K*K, row-major (gold, pred)
  Wc  <- W[cid, , drop = FALSE]                      # (n x ndraw)
  cnt <- matrix(0, K * K, length(draw_id))
  agg <- rowsum(Wc, bin, reorder = FALSE)
  cnt[as.integer(rownames(agg)), ] <- agg           # scatter to all K*K bins
  apply(cnt, 2, \(col) qwk_from_counts(matrix(col, K, K, byrow = TRUE)))
}
```

```{r qwk-levels}
#| cache: true
#| cache.lazy: false
qwk_cells <- ai_metrics |>
  group_by(cell) |>
  group_map(~ {
    v <- qwk_draws(.x$gold, .x$rating, .x$country_text_id)
    tibble(cell = as.character(.y$cell[1]),
           est = v[app], lo = quantile(v[boot], 0.025, na.rm = TRUE),
           hi = quantile(v[boot], 0.975, na.rm = TRUE))
  }) |>
  bind_rows()

# Human LOO QWK reference: (loo_gold, coder rating) pairs, same machinery.
h_qwk <- qwk_draws(human_loo$loo_gold, human_loo$rating, human_loo$country_text_id)
qwk_human <- tibble(cell = "Human (LOO)", est = h_qwk[app],
                    lo = quantile(h_qwk[boot], 0.025, na.rm = TRUE),
                    hi = quantile(h_qwk[boot], 0.975, na.rm = TRUE))

# Point-estimate cross-check against psych on one cell (apparent draw = unweighted counts).
qwk_check <- {
  d <- ai_metrics |> filter(cell == cell_levels[1])
  ck <- suppressWarnings(psych::cohen.kappa(cbind(d$gold, d$rating))$weighted.kappa)
  mine <- qwk_from_counts(table(factor(d$gold, 0:(K-1)), factor(d$rating, 0:(K-1))))
  glue("cross-check {cell_lab_of[cell_levels[1]]}: mine = {round(mine,4)} · psych = {round(ck,4)}")
}

```

```{r qwk-fig}
#| fig-height: 4.2
# Reusable coefficient plot (mirrors 07): cells as points+CIs coloured Base vs Fine-tuned, human LOO
# QWK as a dashed reference line + 95% CI band. Main = core cells; appendix = the rest. Ranked by QWK.
qwk_all <- qwk_cells |>
  transmute(cell_lbl = cell_lab_of[cell], est, lo, hi) |>
  mutate(family = factor(if_else(str_starts(cell_lbl, "FT"), "Fine-tuned", "Base"),
                         levels = c("Base", "Fine-tuned")))
# Grouped by family (Base at top, FT below, as in 06-08), ranked by QWK within each group.
qwk_order <- qwk_all |> arrange(desc(family), est) |> pull(cell_lbl)   # bottom-to-top: FT then Base

qwk_coef_plot <- function(labs, subtitle) {
  qwk_all |>
    filter(cell_lbl %in% labs) |>
    mutate(cell_lbl = factor(cell_lbl, levels = intersect(qwk_order, labs))) |>
    ggplot(aes(est, cell_lbl, color = family)) +
    annotate("rect", xmin = qwk_human$lo, xmax = qwk_human$hi, ymin = -Inf, ymax = Inf,
             fill = "grey85", alpha = 0.5) +
    geom_vline(xintercept = qwk_human$est, linetype = "dashed", color = "grey40") +
    geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y", width = 0.3, linewidth = 0.6) +
    geom_point(size = 2.8) +
    scale_color_manual(values = family_pal, name = NULL) +
    labs(x = "quadratic weighted kappa vs round(panel_mean) · dashed = human LOO (grey band = 95% CI)",
         y = NULL, title = "Quadratic weighted kappa (levels)", subtitle = subtitle, caption = qwk_check) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "top", legend.justification = "left",
          panel.grid.minor = element_blank(), plot.title.position = "plot",
          plot.title = element_text(face = "bold"))
}

qwk_coef_plot(core_lbl, glue("2019 · core cells (base ladder + FT diagonal) · sample-level · ",
                             "country-clustered 95% CI · B = {params$n_boot} · Base over FT, ranked within"))
```

---

## Check 4 — Krippendorff's α with the AI as an added coder (exploratory)

Reliability-framed, and native to the multi-coder structure (no single gold label). Compute panel
reliability on humans only, then with the AI inserted as one added coder, and ask whether
reliability holds. This never collapses the panel; it is the reliability cousin of the D2
augmentation test and points forward to the confirmatory D2 on 2023. Ordinal α (ordinal difference
on the 0..5 union of scales). Reported for the same **7 core cells** as the other figures (base
ladder + FT diagonal), since the added AI coder is that cell's ratings.

```{r alpha}
#| eval: !expr have_irr
#| cache: true
#| cache.lazy: false
# Reliability data matrix: raters x subjects, NA where a rater did not score a subject. Coders are
# not consistent across CYIs, so humans are slotted by arbitrary within-CYI order (α's coincidence
# matrix is invariant to slot identity). The humans-only panel is built once; each core cell then
# enters as one extra coder row (its own AI ratings, on the subjects that cell scored).
alpha_base <- human_loo |>
  transmute(country_text_id, indicator, rating) |>
  group_by(country_text_id, indicator) |>
  mutate(slot = row_number()) |>
  ungroup()
alpha_maxslot <- max(alpha_base$slot)
alpha_subj    <- alpha_base |> distinct(country_text_id, indicator) |> mutate(subj = row_number())
alpha_base    <- alpha_base |> left_join(alpha_subj, by = c("country_text_id", "indicator"))

Hm <- matrix(NA_real_, alpha_maxslot, nrow(alpha_subj))
Hm[cbind(alpha_base$slot, alpha_base$subj)] <- alpha_base$rating
a_humans <- irr::kripp.alpha(Hm, method = "ordinal")$value   # humans-only (fixed across cells)

alpha_for_cell <- function(cell_key) {
  ai_cell <- ai_metrics |> filter(cell == cell_key) |>
    distinct(country_text_id, indicator, ai = rating)
  ai_j <- alpha_subj |> inner_join(ai_cell, by = c("country_text_id", "indicator"))
  Ha <- rbind(Hm, NA_real_)
  Ha[alpha_maxslot + 1, ai_j$subj] <- ai_j$ai
  a_ha <- irr::kripp.alpha(Ha, method = "ordinal")$value
  tibble(Cell = cell_lab_of[cell_key],
         `α humans only` = a_humans, `α humans + AI` = a_ha, `Δα` = a_ha - a_humans,
         n_added = nrow(ai_j))
}

alpha_tbl <- map_dfr(core_cells, alpha_for_cell) |>
  mutate(family = if_else(str_starts(Cell, "FT"), "Fine-tuned", "Base"))
alpha_tbl |>
  select(-family) |>
  gt() |>
  fmt_number(columns = c(`α humans only`, `α humans + AI`, `Δα`), decimals = 3) |>
  tab_header(title = "Krippendorff's α — humans only vs humans + AI (core cells, exploratory)",
             subtitle = glue("Ordinal α · humans-only α = {round(a_humans, 3)} on {nrow(alpha_subj)} CYIs · ",
                             "Δα > 0 = adding that cell's AI as a coder raises panel reliability")) |>
  tab_style(style = cell_fill(color = "#fdf1dd"),
            locations = cells_body(rows = str_starts(Cell, "FT")))
```

```{r alpha-skip}
#| eval: !expr '!have_irr'
cat("Check 4 (Krippendorff's α) skipped: package `irr` is not installed.\n",
    "Install it (install.packages(\"irr\")) and re-knit to populate this section.\n")
```

---

## Check 5 — Scale-normalized MAE robustness on the primary metric

Companion to Check 1. The coded indicators do not share a single rating scale. Some are dichotomous
(scored 0 or 1), most run 0 to 4, and a handful run 0 to 5. `06`'s primary MAE pools raw rating-point
errors across all of them, but a one-point error means very different things depending on the scale:
on a 0–1 dichotomy it is the entire range, whereas on a 0–5 scale it is one step out of five. Because
the pool is dominated by the 0–4 indicators, the headline MAE is effectively tilted toward the longer
scales rather than being an equal-weighted average of per-indicator skill.

The scales present in the 2019 coded set, and how many indicators sit on each:

| Indicator scale | Range (the divisor) | # indicators |
|:---|:---:|:---:|
| 0–1 (dichotomous) | 1 | 9 |
| 0–2 | 2 | 19 |
| 0–3 | 3 | 22 |
| 0–4 | 4 | 143 |
| 0–5 | 5 | 12 |

This check re-scores the identical 2019 ratings under a scale-normalized loss: each error is divided
by its indicator's range, so every indicator lands on a common 0-to-1 footing. It then re-runs `06`'s
landscape (Figure 1) and base-model B-series (Figure 2), asking whether the identification story holds
once a dichotomy and a 0–5 scale count equally.

**Why the primary metric is still defensible (this is a robustness check, not a replacement).** The
scale weighting applies equally to every cell being compared. The base ladder, the FT diagonal, and
the human LOO reference are all scored on the same pooled country-year-indicators with the same
scales, so every paired contrast is on matched scales on both sides and the weighting largely cancels
in the deltas. Raw-point MAE also stays the interpretable, pre-registered unit ("rating points off
consensus"). The purpose here is to confirm the picture is not an artifact of the longer scales
dominating, not to re-anchor the paper on a normalized number.

**Normalized SESOI.** Check 1's RMSE lives on a squared-error axis where the rounding floor is not
cleanly defined, which is why it borrowed the raw SESOI as a common ruler. Normalized MAE stays on the
absolute-error axis, so the floor is well defined per indicator as
`|round(panel_mean) − panel_mean| / range`, and `nSESOI` is half its pool mean: a properly normalized
band rather than a borrowed one.

**How the scale is measured.** Each indicator's range is the largest human rating it ever receives
(the minimum is 0 for every ordinal V-Dem C-variable). This recovers exactly the 12 known 0–5
indicators and the 0–1 dichotomies. Reading the range off the codebook instead would be marginally
more principled but gives the same answer here.

```{r nmae-setup}
#| cache: true
#| cache.lazy: false
# Indicator scale width: observed max rating across all human years for each AI-coded indicator.
width_tbl <- human_ratings |>
  filter(indicator %in% ai_indicators, valid_rating(rating)) |>
  group_by(indicator) |>
  summarise(width = max(rating), .groups = "drop") |>
  filter(width > 0)                                  # no all-zero indicator in the AI set

# Per-row scale-normalized absolute error: |rating - panel_mean| / width. Same fractional reference
# (raw_mean) and same LOO case set as 06's primary MAE — only the per-indicator rescaling is new.
ai_nmae <- ai_metrics |>
  inner_join(width_tbl, by = "indicator") |>
  mutate(cell = factor(cell, levels = cell_levels),
         nerr = abs(rating - raw_mean) / width)

# Normalized SESOI: the integer-vs-fractional-mean floor on the same 0..1-per-indicator footing,
# halved. Computed over the CYI pool (one row per CYI), matching 06's rounding-floor construction.
nfloor <- ai_nmae |>
  distinct(country_text_id, indicator, raw_mean, width) |>
  summarise(f = mean(abs(round(raw_mean) - raw_mean) / width)) |>
  pull(f)
nsesoi <- nfloor / 2

# Human LOO rail (normalized): |rating - reduced_mean| / width, country-clustered single group.
human_nmae <- human_loo |>
  inner_join(width_tbl, by = "indicator") |>
  mutate(h_nerr = abs(rating - loo_mean) / width)
h_nv <- wmean_by_cell(human_nmae$h_nerr, rep(1L, nrow(human_nmae)),
                      human_nmae$country_text_id, 1L)[, 1]
human_nmae_ref <- tibble(est = h_nv[app],
                         lo = quantile(h_nv[boot], 0.025), hi = quantile(h_nv[boot], 0.975))

# Persistence ("Naive") rail (normalized): |round(2018 mean) - 2019 mean| / width, on the LOO pool.
pm18 <- panel_means |> filter(year == 2018) |>
  select(country_text_id, indicator, mean_2018 = raw_mean)
persist_nmae <- loo_cyi |>
  inner_join(select(pm_year, country_text_id, indicator, raw_mean),
             by = c("country_text_id", "indicator")) |>
  inner_join(width_tbl, by = "indicator") |>
  left_join(pm18, by = c("country_text_id", "indicator")) |>
  filter(!is.na(mean_2018)) |>
  mutate(p_nerr = abs(round(mean_2018) - raw_mean) / width)
p_nv <- wmean_by_cell(persist_nmae$p_nerr, rep(1L, nrow(persist_nmae)),
                      persist_nmae$country_text_id, 1L)[, 1]
persist_nmae_ref <- tibble(est = p_nv[app],
                           lo = quantile(p_nv[boot], 0.025), hi = quantile(p_nv[boot], 0.975))

# Per-cell normalized MAE per draw, via the shared block aggregator; level CIs via Check 2's helper.
nmae_mat <- wmean_by_cell(ai_nmae$nerr, as.integer(ai_nmae$cell),
                          ai_nmae$country_text_id, ncell)
colnames(nmae_mat) <- cell_levels
nmae_lvl <- level_ci(nmae_mat)

glue("nSESOI (±) = {round(nsesoi, 3)} · normalized rounding floor = {round(nfloor, 3)} · ",
     "human LOO nMAE = {round(human_nmae_ref$est, 3)} · persistence nMAE = {round(persist_nmae_ref$est, 3)}")
```

### Check 5 · levels — the normalized-MAE landscape (mirror of `06` Figure 1)

The 7 core cells on one normalized-MAE axis — the base ladder (Cb→Ev→An→Su) on top, the FT diagonal
below — against the **Naive model** (persistence) and **human LOO** rails, with the ±nSESOI band on
the human line. Same layout as `06`'s Figure 1; the question is whether the same two dynamics (base
ladder fanning right as identity is stripped, FT cluster pulling onto the human line) hold once every
indicator counts equally.

```{r nmae-fig-levels}
#| fig-width: 8
#| fig-height: 4.8
land_n <- nmae_lvl |>
  filter(cell %in% core_cells) |>
  mutate(lbl    = cell_lab_of[cell],
         family = factor(if_else(str_starts(lbl, "FT"), "Fine-tuned", "Base"),
                         levels = c("Base", "Fine-tuned")),
         # first level plots at the bottom: FT block below, base ladder above (as in 06 Fig 1)
         row = factor(lbl, levels = c("FT-summ:Su", "FT-anon:An", "FT-raw:Ev",
                                      "Base:Su", "Base:An", "Base:Ev", "Base:Cb")))

ggplot(land_n, aes(est, row, color = family)) +
  annotate("rect", xmin = human_nmae_ref$est - nsesoi, xmax = human_nmae_ref$est + nsesoi,
           ymin = -Inf, ymax = 7.3, fill = "grey85", alpha = 0.55) +
  annotate("segment", x = persist_nmae_ref$est, xend = persist_nmae_ref$est,
           y = -Inf, yend = 7.3, linetype = "dashed", color = "grey40") +
  annotate("segment", x = human_nmae_ref$est, xend = human_nmae_ref$est,
           y = -Inf, yend = 7.3, linetype = "longdash", color = "grey40") +
  annotate("text", x = persist_nmae_ref$est, y = 6.8, label = "Naive model",
           color = "grey40", size = 2.9, hjust = -0.07, vjust = 0.5) +
  annotate("text", x = human_nmae_ref$est, y = 7.5, label = "±nSESOI of human LOO nMAE",
           color = "grey40", size = 2.9, hjust = 0.5, vjust = 0.5) +
  geom_hline(yintercept = 3.5, color = "grey85", linewidth = 0.4) +
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y", width = 0.20, linewidth = 0.7) +
  geom_point(size = 2) +
  scale_color_manual(values = family_pal, name = NULL) +
  scale_y_discrete() +   # keep y discrete so the numeric rail labels (y = 6.8/7.3/7.5) place correctly
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.08), add = c(0.005, 0))) +
  coord_cartesian(clip = "off") +
  labs(x = "Scale-normalized MAE  (|error| / scale range)", y = NULL,
       title = "Check 5 — Scale-normalized synthetic-coder performance",
       subtitle = glue("2019 · core cells · every indicator on a common 0-1 footing (lower = closer to consensus) · nSESOI ±{round(nsesoi, 3)}")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", legend.justification = "left",
        panel.grid.major.y = element_blank(), plot.title.position = "plot",
        plot.title = element_text(face = "bold"))
```

### Check 5 · deltas — base-model B-series under normalized loss (mirror of `06` Figure 2)

The base-model identification contrasts re-scored under normalized MAE, in `06` Figure 2's layout:
**top group** = each condition vs. the Codebook prior; **bottom group** = the de-identified
conditions vs. full Evidence. The survival table reads each metric against its own properly-scaled
band — raw ΔMAE vs ±SESOI beside normalized ΔnMAE vs ±nSESOI — so "survives" means the SESOI verdict
is unchanged by neutralizing scale width.

```{r nmae-contrasts}
Bc  <- "llama-70b|codebook"; EVc <- "llama-70b|evidence"
ANc <- "llama-70b|anonymized"; SUc <- "llama-70b|summarized"
nmae_contrasts <- bind_rows(
  paired_metric_delta(nmae_mat, EVc, Bc,  "B1 · Ev−Cb",     nsesoi),
  paired_metric_delta(nmae_mat, ANc, Bc,  "B2 · An−Cb",     nsesoi),
  paired_metric_delta(nmae_mat, SUc, Bc,  "Su−Cb (descr.)", nsesoi),
  paired_metric_delta(nmae_mat, ANc, EVc, "B3 · An−Ev",     nsesoi),
  paired_metric_delta(nmae_mat, SUc, EVc, "Su−Ev (descr.)", nsesoi)
)

# Raw-MAE B-series verdicts (reuse Check 1's mae_contrasts, computed from 06's boot_results).
nmae_tbl <- nmae_contrasts |>
  left_join(select(mae_contrasts, label, mae_est, mae_sesoi_out), by = "label") |>
  mutate(survives = if_else(sesoi_out == mae_sesoi_out, "yes", "FLIPPED"))

nmae_tbl |>
  transmute(Contrast = label,
            `ΔMAE (raw)` = round(mae_est, 3),
            `raw > SESOI` = if_else(mae_sesoi_out, "yes", "no"),
            `ΔnMAE` = round(est, 3),
            `nMAE 95% CI` = glue("[{round(lo, 3)}, {round(hi, 3)}]"),
            `nMAE > nSESOI` = if_else(sesoi_out, "yes", "no"),
            `Survives` = survives) |>
  gt() |>
  tab_header(title = "Scale-normalized MAE robustness — base-model B-series",
             subtitle = glue("Raw MAE vs ±SESOI ({round(sesoi, 3)}) beside normalized MAE vs ±nSESOI ",
                             "({round(nsesoi, 3)}) · paired country-clustered bootstrap · B = {params$n_boot}"))
```

```{r nmae-fig-deltas}
#| fig-height: 4.0
nmae_b <- nmae_contrasts |>
  mutate(label = factor(label, levels = rev(c(
    "B1 · Ev−Cb", "B2 · An−Cb", "Su−Cb (descr.)", "B3 · An−Ev", "Su−Ev (descr.)"))))

ggplot(nmae_b, aes(est, label)) +
  annotate("rect", xmin = -nsesoi, xmax = nsesoi, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = 0.55) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  # separator between the "vs Codebook" group (top 3) and the "vs Evidence" group (bottom 2)
  geom_hline(yintercept = 2.5, color = "grey85", linewidth = 0.4) +
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y",
                width = 0.15, linewidth = 0.7, color = "#0072B2") +
  geom_point(size = 3, color = "#0072B2") +
  scale_x_continuous(expand = expansion(mult = c(0.06, 0.06))) +
  labs(x = "ΔnMAE · right = higher normalized error · grey band = ±nSESOI", y = NULL,
       title = "Check 5 — Scale-normalized base model: reading the evidence or reading the country",
       subtitle = glue("2019 · top: vs Codebook · bottom: vs Evidence · paired · nSESOI ±{round(nsesoi, 3)}")) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(), plot.title.position = "plot",
        plot.title = element_text(face = "bold"))
```

---

## Confusion matrix (display, not a scalar)

A 6×6 matrix (`round(panel_mean)` × AI class, classes 0..5) shows the error *structure* the scalars hide:
diagonal mass = exact-match, tri-diagonal band = adjacent, above/below-diagonal asymmetry = the
signed-deviation story as a matrix. Shown for the reading-vs-reciting shift (base:Cb → base:Ev)
plus the carried cell (`r glue("{recode(params$carried_key, !!!model_labels)}:{recode(canon_col(params$carried_cond), !!!col_labels)}")`,
finalized at unblinding). Not all sixteen.

```{r confusion}
#| fig-height: 3.8
cm_cells <- c("llama-70b|codebook", "llama-70b|evidence",
              paste(params$carried_key, params$carried_cond, sep = "|"))
cm_df <- ai_metrics |>
  filter(cell %in% cm_cells) |>
  mutate(cell = factor(cell, levels = cm_cells)) |>
  count(cell, gold, rating, name = "n") |>
  group_by(cell, gold) |>
  mutate(row_share = n / sum(n)) |>
  ungroup()

ggplot(cm_df, aes(rating, gold, fill = row_share)) +
  geom_tile(color = "grey90") +
  geom_text(aes(label = n), size = 2.8) +
  facet_wrap(~ cell, labeller = as_labeller(cm_cells |> setNames(cell_lab_of[cm_cells]) |> {\(x) setNames(names(x), x)}())) +
  scale_fill_gradient(low = "white", high = "#0072B2", labels = scales::percent, name = "row share") +
  scale_x_continuous(breaks = 0:5) + scale_y_continuous(breaks = 0:5) +
  coord_equal() +
  labs(x = "AI rating", y = "round(panel mean)",
       title = "Confusion structure: gold class × AI class",
       subtitle = "Cell counts; fill = share within the gold row (diagonal = exact, band = adjacent)") +
  theme_minimal(base_size = 11) + theme(panel.grid.minor = element_blank())
```

*Facet labels* fall back to the raw cell key if the labeller lookup misses; the panels are
base:Cb, base:Ev, and the carried cell, left to right.

---

## Appendix — off-diagonal cells (levels)

The nine cells outside the core set (each base condition is already in the core figure; these are
the FT adapters run *off* their diagonal — e.g. FT-summ on codebook, FT-raw on anonymized). Same
metrics, same human LOO reference, same Base/Fine-tuned colouring as the core figures above.

```{r exact-adj-appendix}
#| fig-height: 4.6
ea_coef_plot(other_lbl, glue("2019 · off-diagonal cells · country-clustered 95% CI · ",
                             "B = {params$n_boot} · ranked by exact-match"))
```

```{r qwk-appendix}
#| fig-height: 4.0
qwk_coef_plot(other_lbl, glue("2019 · off-diagonal cells · sample-level · ",
                              "country-clustered 95% CI · B = {params$n_boot} · ranked by QWK"))
```

---

## Serialize the bundle

Parallel to `06`/`08`, so `paper/` reads it without recomputing.

```{r serialize}
saveRDS(
  list(
    rmse_tbl      = rmse_tbl,          # Check 1: L1 vs L2 verdicts
    rmse_contrasts = rmse_contrasts,
    exact_adj_tbl = exact_adj_tbl,     # Check 2 levels
    ex_lvl = ex_lvl, adj_lvl = adj_lvl, human_ref_row = human_ref_row,
    qwk_cells = qwk_cells, qwk_human = qwk_human,   # Check 3
    alpha_tbl = if (have_irr) alpha_tbl else NULL,  # Check 4 (may be NULL)
    nmae_lvl = nmae_lvl, nmae_contrasts = nmae_contrasts, nmae_tbl = nmae_tbl,  # Check 5
    human_nmae_ref = human_nmae_ref, persist_nmae_ref = persist_nmae_ref, nsesoi = nsesoi,
    confusion = cm_df,
    cell_lab_of = cell_lab_of, sesoi = sesoi, K = K, n_boot = params$n_boot),
  file.path(derived_dir, "secondary_2019.rds")
)
glue("secondary_2019.rds written · B = {params$n_boot} · SESOI ±{round(sesoi,3)} · ",
     "α section: {if (have_irr) 'computed' else 'skipped (install irr)'}")
```

## Summary

```{r summary}
tibble(
  Check = c("1 · RMSE robustness", "2 · Exact / adjacent", "3 · QWK", "4 · Krippendorff α",
            "5 · Scale-normalized MAE", "Confusion matrix"),
  Reference = c("squared error, paired B/F1 contrasts vs ±SESOI",
                "round(panel_mean), per-cell levels + human LOO",
                "round(panel_mean), sample-level levels + human LOO",
                "no gold label; humans vs humans+AI, base model",
                "|error|/width, landscape + B-series vs ±nSESOI",
                "round(panel_mean) × AI, base:Cb / base:Ev / carried"),
  Role = c("robustness on registered contrasts", "secondary descriptive", "secondary descriptive",
           "exploratory reliability (→ D2)", "robustness on primary metric", "display"),
  Status = c(if (any(rmse_tbl$survives == "FLIPPED")) "built · some verdicts flip" else "built · all verdicts survive",
             "built", "built", if (have_irr) "built" else "skipped (install irr)",
             if (any(nmae_tbl$survives == "FLIPPED")) "built · some verdicts flip" else "built · all verdicts survive",
             "built")
) |>
  gt() |>
  tab_header(title = glue("Secondary & robustness metrics — {params$year} status"),
             subtitle = "All exploratory / robustness; the identification story stays on AI MAE")
```
