---
title: "2019 expectation (mean) readout — sensitivity analysis"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
params:
  n_boot: 2000
  year: 2019
  min_coders: 2
  exp_subdir: "expectation"
---

This is the **expectation-readout sensitivity analysis** pre-committed in
`notes/expectation-readout-precommitment.md`. It recomputes the entire 2019 result set under
the expectation readout `Ê[y|X] = Σ y·p(y|X)` (the distribution's **mean**) and reports it
beside the pre-registered greedy readout (the **mode**), using the identical
country-clustered bootstrap.

It reads the logprob-capturing re-run only — the JSONL files that carry the `rating_dist`
field — from `data/output/runs/<exp_subdir>/`. Those files are a separate inference pass
(same adapters, same prompts, `temperature = 0`) whose sole difference from the confirmatory
runs is that the rating-token distribution is retained rather than discarded. The frozen
greedy files in `data/output/runs/` are never opened here.

**Integrity note.** Everything below is exploratory and pre-specified. The greedy readout
stays the confirmatory result; nothing here is relabelled confirmatory. Interpretive prose is
left as `TODO (EJT)` markers — this document computes; the paper narrates.

---

## 1. Setup

```{r setup}
library(tidyverse)
library(gt)
library(glue)

if (!exists("params")) {
  params <- list(n_boot = 2000, year = 2019, min_coders = 2, exp_subdir = "expectation")
}

# Same robust root resolution as analysis/06 (walk up to .git, else step down into
# panel-member); avoids here()'s cached-root pin.
find_panel_member_root <- function() {
  up <- tryCatch(rprojroot::find_root(rprojroot::is_git_root), error = function(e) NA_character_)
  if (!is.na(up)) return(up)
  down <- file.path(getwd(), "panel-member")
  if (dir.exists(file.path(down, ".git"))) return(down)
  stop("Could not locate the panel-member project root from working dir: ", getwd())
}
proj_root <- find_panel_member_root()

# Shared vocabulary + paired-delta machinery (model_labels, family_pal, canon_col, ft_diag,
# country_boot_weights, cell_series(), paired_delta()). Same source of truth as doc 06.
source(file.path(proj_root, "helpers", "bootstrap_helpers.R"))

data_dir    <- file.path(proj_root, "data", "processed")
exp_dir     <- file.path(proj_root, "data", "output", "runs", params$exp_subdir)
derived_dir <- file.path(proj_root, "data", "derived")
dir.create(derived_dir, showWarnings = FALSE, recursive = TRUE)
```

---

## 2. Data loading

Load the expectation re-run JSONLs. Each record carries the same scalar fields as the greedy
runs **plus** `rating_dist`: the model's probability over the integer ratings at the
rating-digit token, index `i` == p(rating `i`), raw (un-renormalized) `exp(logprob)` with `0`
for any rating absent from the captured top-k, or `null` on a capture miss.

```{r load-exp}
#| cache: true
#| cache.lazy: false

if (!dir.exists(exp_dir)) {
  stop("expectation runs dir not found:\n  ", exp_dir,
       "\nProduce it with the logprob-capturing re-run, e.g.:\n",
       "  OUTPUT_DIR=data/output/runs/expectation YEAR=2019 CONDITION=codebook sbatch slurm/run_coding_llama70b.sh\n",
       "  OUTPUT_DIR=data/output/runs/expectation VARIANT=raw sbatch slurm/run_inference_finetuned.sh\n",
       "then pull the archive locally into that directory.")
}
exp_files <- list.files(exp_dir, pattern = "\\.jsonl$", full.names = TRUE)
if (length(exp_files) == 0) stop("No .jsonl files in ", exp_dir)

needed_cols <- c("country", "year", "indicator", "condition", "model_key", "rating", "rating_dist")

read_exp <- function(f) {
  con <- file(f, "r")
  on.exit(close(con))
  # simplifyMatrix = FALSE keeps rating_dist a list-column even when every dist has the same
  # length (else jsonlite collapses equal-length arrays to a matrix column).
  df <- jsonlite::stream_in(con, simplifyMatrix = FALSE, verbose = FALSE) |> as_tibble()
  if (!"rating_dist" %in% names(df)) {
    stop("No 'rating_dist' in ", basename(f),
         " — these look like greedy files. Point exp_subdir at the logprob re-run.")
  }
  # Normalize to a list-column of numeric vectors regardless of how jsonlite simplified it.
  d <- df[["rating_dist"]]
  df[["rating_dist"]] <- if (is.matrix(d)) asplit(d, 1) else as.list(d)
  select(df, all_of(needed_cols))
}

exp_raw <- exp_files |>
  map(read_exp) |>
  bind_rows() |>
  mutate(model_key = str_remove(model_key, "-local$"))

panel_means   <- read_csv(file.path(data_dir, "panel_means.csv"), show_col_types = FALSE)
human_ratings <- read_csv(file.path(data_dir, "human_ratings.csv"), show_col_types = FALSE)
```

**Load diagnostic** — the expectation re-run captured **16 cells**: base (4 primary conditions)
and 12 FT (3 variants × 4 conditions). The 3 base `*-zeroshot` cells are optional (precommitment)
and were not run, so the A1/A2 grid contrasts in §9 are skipped until they are.

```{r load-diagnostic}
exp_raw |>
  filter(year == params$year) |>
  count(model_key, condition) |>
  mutate(model = recode(model_key, !!!model_labels)) |>
  arrange(model_key, condition) |>
  gt() |>
  tab_header(
    title    = glue("Expectation-run cells present for {params$year}"),
    subtitle = glue("{n_distinct(paste(exp_raw$model_key, exp_raw$condition))} model×condition cells loaded")
  )
```

---

## 3. Metric construction

### Human LOO baseline (per CYI)

Identical to doc 06: each coder's leave-one-out error against the panel mean computed with
them removed, averaged per CYI over panels of ≥ `r params$min_coders` coders.

```{r human-loo}
#| cache: true
human_loo <- human_ratings |>
  filter(year == params$year) |>
  group_by(country_text_id, year, indicator) |>
  filter(n() >= params$min_coders) |>
  mutate(
    panel_sum = sum(rating),
    n_coders  = n(),
    loo_mean  = (panel_sum - rating) / (n_coders - 1),
    loo_error = abs(rating - loo_mean)
  ) |>
  summarise(human_loo_mae = mean(loo_error), n_coders = first(n_coders), .groups = "drop")
```

### Greedy + expectation error (per CYI × model × condition)

`Ê = Σ_i i · p_i / Σ_i p_i` renormalized over the captured rating tokens. `greedy_error` uses
the emitted `rating` (the mode); `exp_error` uses `Ê` (the mean). Both are absolute error
against the same canonical panel mean, restricted to CYIs carrying a human LOO floor so AI and
human compare on identical cases — again matching doc 06.

```{r metrics}
#| cache: true
#| cache.lazy: false

exp_rating_of <- function(d) {
  d <- suppressWarnings(as.numeric(unlist(d)))
  if (length(d) == 0 || all(is.na(d)) || sum(d, na.rm = TRUE) <= 0) return(NA_real_)
  sum((seq_along(d) - 1) * d, na.rm = TRUE) / sum(d, na.rm = TRUE)
}

cell_metrics_all <- exp_raw |>
  filter(year == params$year) |>
  rename(country_text_id = country) |>
  inner_join(select(panel_means, country_text_id, year, indicator, raw_mean),
             by = c("country_text_id", "year", "indicator")) |>
  inner_join(select(human_loo, country_text_id, year, indicator, human_loo_mae),
             by = c("country_text_id", "year", "indicator")) |>
  mutate(
    exp_rating   = map_dbl(rating_dist, exp_rating_of),
    greedy_error = abs(rating - raw_mean),
    exp_error    = abs(exp_rating - raw_mean),
    col          = canon_col(condition),
    family       = recode(model_key, !!!model_family)
  )

# Capture-miss accounting: keep only CYIs with a valid expectation, so greedy and expectation
# are paired on exactly the same rows (a fair within-draw difference). Report what is dropped.
miss_tbl <- cell_metrics_all |>
  group_by(model_key, condition) |>
  summarise(n = n(), n_missing = sum(is.na(exp_rating)),
            pct_missing = round(100 * n_missing / n, 3), .groups = "drop")

cell_metrics <- filter(cell_metrics_all, !is.na(exp_rating))

miss_tbl |>
  arrange(desc(pct_missing)) |>
  gt() |>
  tab_header(title = "Rating-token capture misses per cell",
             subtitle = "Dropped from the paired analysis; expect ~0% if capture is clean")
```

---

## 4. Reference lines (loaded from the greedy bundle)

SESOI, the rounding floor, and the human-LOO / persistence rails are properties of the panel
and the pool, not of the readout, so they are reused from the confirmatory bundle rather than
recomputed. If the bundle is absent, render doc 06 first.

```{r rails}
bundle_path <- file.path(derived_dir, glue("bootstrap_{params$year}.rds"))
if (!file.exists(bundle_path)) {
  stop("greedy bundle not found: ", bundle_path, "\nRender analysis/06-bootstrap-metrics-2019.qmd first.")
}
greedy_bundle <- readRDS(bundle_path)
sesoi          <- greedy_bundle$sesoi
rounding_floor <- greedy_bundle$rounding_floor
human_ref      <- greedy_bundle$human_ref
persist_ref    <- greedy_bundle$persist_ref
greedy_boot_ci <- greedy_bundle$boot_ci   # confirmatory greedy MAE per cell (for cross-check)

glue("SESOI (±) = {round(sesoi,3)} · rounding floor = {round(rounding_floor,3)} · ",
     "human LOO = {round(human_ref$human_mae,3)} · persistence = {round(persist_ref$persist_mae,3)}")
```

---

## 5. Bootstrap (country-clustered, seed 42 — matches doc 06)

One draw step shared by greedy and expectation, so their difference is paired within draw.
Countries are resampled with replacement, each carrying its whole indicator block; a country
drawn `k` times is weighted `k`. `boot_both` carries both MAEs per cell per draw.

```{r bootstrap}
#| cache: true
#| cache.lazy: false

cyi_pool <- distinct(cell_metrics, country_text_id, year, indicator)
W <- country_boot_weights(cyi_pool$country_text_id, params$n_boot, seed = 42)

boot_both <- map_dfr(colnames(W), \(draw) {
  country_w <- tibble(country_text_id = rownames(W), w = W[, draw]) |> filter(w > 0)
  cell_metrics |>
    inner_join(country_w, by = "country_text_id") |>
    group_by(model_key, condition) |>
    summarise(
      greedy_mae = weighted.mean(greedy_error, w),
      exp_mae    = weighted.mean(exp_error, w),
      .groups    = "drop"
    ) |>
    mutate(id = draw, .before = 1)
})

# Two views of boot_both shaped like doc 06's `boot_results` so paired_delta() works unchanged
# on either readout. Greedy here is recomputed on the captured pool (fair apples-to-apples with
# expectation); §6 checks it against the confirmatory bundle.
greedy_boot_results <- transmute(boot_both, id, model_key, condition, ai_mae = greedy_mae)
exp_boot_results    <- transmute(boot_both, id, model_key, condition, ai_mae = exp_mae)

boot_ci_of <- function(results) {
  results |>
    group_by(model_key, condition) |>
    summarise(
      ai_lo  = quantile(ai_mae[id != "Apparent"], 0.025),
      ai_hi  = quantile(ai_mae[id != "Apparent"], 0.975),
      ai_mae = mean(ai_mae[id == "Apparent"]),
      .groups = "drop"
    )
}
greedy_ci <- boot_ci_of(greedy_boot_results)
exp_ci    <- boot_ci_of(exp_boot_results)
```

---

## 6. Reproducibility check: recomputed greedy vs the confirmatory bundle

`temperature = 0` means the re-run's emitted rating should match the archived greedy rating
cell for cell, so the greedy MAE recomputed here (on the captured pool) should track the
confirmatory bundle. Large gaps flag a pool mismatch or a nondeterministic decode, not an
expectation result.

```{r repro-check}
greedy_ci |>
  select(model_key, condition, ai_mae_rerun = ai_mae) |>
  inner_join(select(greedy_boot_ci, model_key, condition, ai_mae_bundle = ai_mae),
             by = c("model_key", "condition")) |>
  mutate(abs_diff = abs(ai_mae_rerun - ai_mae_bundle)) |>
  arrange(desc(abs_diff)) |>
  mutate(model = recode(model_key, !!!model_labels)) |>
  gt() |>
  tab_header(title = "Greedy MAE: expectation re-run vs confirmatory bundle",
             subtitle = "abs_diff should be ~0 (pool differs only by dropped capture misses)") |>
  fmt_number(columns = c(ai_mae_rerun, ai_mae_bundle, abs_diff), decimals = 4)
```

---

## 7. Expectation vs greedy MAE — the three-number landscape

Per cell: greedy (mode) point, expectation (mean) point, against the floor / human-LOO /
persistence rails. `TODO (EJT)`: narrative — does expectation drop toward the floor (readout
artifact) or stall short (real center-shift)?

```{r}
#| label: fig-exp-vs-greedy
#| fig-width: 8
#| fig-height: 5.2

grid_cells <- bind_rows(
  filter(greedy_ci, model_key == "llama-70b",
         condition %in% c("codebook","evidence","anonymized","summarized")) |>
    mutate(readout = "Greedy (mode)"),
  filter(exp_ci, model_key == "llama-70b",
         condition %in% c("codebook","evidence","anonymized","summarized")) |>
    mutate(readout = "Expectation (mean)"),
  semi_join(greedy_ci, ft_diag, by = c("model_key","condition")) |> mutate(readout = "Greedy (mode)"),
  semi_join(exp_ci,    ft_diag, by = c("model_key","condition")) |> mutate(readout = "Expectation (mean)")
) |>
  mutate(
    row = case_when(
      model_key == "llama-70b" ~ recode(condition, codebook="Codebook", evidence="Raw Text",
                                        anonymized="Anonymized", summarized="Summarized"),
      TRUE ~ recode(model_key, "llama-70b-ft-raw"="FT · Raw Text",
                    "llama-70b-ft-anon"="FT · Anonymized", "llama-70b-ft-summ"="FT · Summarized")
    ),
    row = factor(row, levels = c("FT · Summarized","FT · Anonymized","FT · Raw Text",
                                 "Summarized","Anonymized","Raw Text","Codebook")),
    readout = factor(readout, levels = c("Greedy (mode)", "Expectation (mean)"))
  )

ggplot(grid_cells, aes(ai_mae, row, color = readout)) +
  annotate("segment", x = rounding_floor, xend = rounding_floor, y = -Inf, yend = Inf,
           linetype = "dotted", color = "grey40") +
  annotate("segment", x = persist_ref$persist_mae, xend = persist_ref$persist_mae,
           y = -Inf, yend = Inf, linetype = "dashed", color = "grey40") +
  annotate("segment", x = human_ref$human_mae, xend = human_ref$human_mae,
           y = -Inf, yend = Inf, linetype = "longdash", color = "grey40") +
  geom_hline(yintercept = 3.5, color = "grey85", linewidth = 0.4) +
  geom_errorbar(aes(xmin = ai_lo, xmax = ai_hi), orientation = "y",
                width = 0.2, linewidth = 0.6, position = position_dodge(width = 0.5)) +
  geom_point(size = 2.2, position = position_dodge(width = 0.5)) +
  scale_color_manual(values = c("Greedy (mode)" = "#0072B2",
                                "Expectation (mean)" = "#D55E00"), name = NULL) +
  labs(x = "AI Mean Absolute Error vs panel mean",
       y = NULL,
       title = "Expectation (mean) vs greedy (mode) readout",
       subtitle = glue("2019 · dotted = floor · dashed = persistence · long-dash = human LOO")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", legend.justification = "left",
        panel.grid.major.y = element_blank(), plot.title.position = "plot",
        plot.title = element_text(face = "bold"))
```

---

## 8. Prediction 1 — asymmetric readout gain (FT gain ≫ base gain)

Per-cell **expectation gain** = greedy MAE − expectation MAE (how much switching readout cuts
error). The memo counts P1 supported iff the mean gain across FT cells exceeds the mean gain
across base cells by **more than SESOI** *and* the bootstrap CI on that difference excludes
zero. Paired within draw off `boot_both`.

```{r p1-gain}
gain_draws <- boot_both |>
  mutate(family = if_else(model_key == "llama-70b", "Base", "Fine-tuned"),
         gain   = greedy_mae - exp_mae)

# Headline: all base cells (7) vs all FT cells (12), unweighted cell means per draw.
by_group <- gain_draws |>
  group_by(id, family) |>
  summarise(mean_gain = mean(gain), .groups = "drop") |>
  pivot_wider(names_from = family, values_from = mean_gain) |>
  mutate(gap = `Fine-tuned` - Base)

ci <- function(x, draws = by_group) c(
  est = x[draws$id == "Apparent"],
  lo  = unname(quantile(x[draws$id != "Apparent"], 0.025)),
  hi  = unname(quantile(x[draws$id != "Apparent"], 0.975))
)
b <- ci(by_group$Base); f <- ci(by_group$`Fine-tuned`); g <- ci(by_group$gap)
p_gap_gt0 <- mean(by_group$gap[by_group$id != "Apparent"] > 0)
p1_supported <- (g["est"] > sesoi) && (g["lo"] > 0)

tibble(
  Quantity = c("Base mean gain", "FT mean gain", "FT − Base gap"),
  Estimate = c(b["est"], f["est"], g["est"]),
  `95% CI` = c(glue("[{round(b['lo'],3)}, {round(b['hi'],3)}]"),
               glue("[{round(f['lo'],3)}, {round(f['hi'],3)}]"),
               glue("[{round(g['lo'],3)}, {round(g['hi'],3)}]"))
) |>
  gt() |>
  tab_header(
    title    = "Prediction 1 — expectation gain, base vs fine-tuned",
    subtitle = glue("P(gap>0) = {round(p_gap_gt0,3)} · SESOI = {round(sesoi,3)} · ",
                    "gap > SESOI & CI excludes 0 → {ifelse(p1_supported,'SUPPORTED','not supported')}")
  ) |>
  fmt_number(columns = Estimate, decimals = 3)
```

**Cell-set robustness.** The memo says "fine-tuned cells" / "base cells" without pinning which
of the 7 base and 12 FT cells enter the means. `TODO (EJT)`: confirm the headline (all-base vs
all-FT). Two alternative reductions below — leave-off decision is yours, not the readout's.

```{r p1-variants}
gap_gap <- function(df) {
  bg <- df |> group_by(id, family) |> summarise(m = mean(gain), .groups = "drop") |>
    pivot_wider(names_from = family, values_from = m) |> mutate(gap = `Fine-tuned` - Base)
  tibble(
    gap_est = bg$gap[bg$id == "Apparent"],
    gap_lo  = quantile(bg$gap[bg$id != "Apparent"], 0.025),
    gap_hi  = quantile(bg$gap[bg$id != "Apparent"], 0.975),
    p_gt0   = mean(bg$gap[bg$id != "Apparent"] > 0)
  )
}

variants <- bind_rows(
  gap_gap(gain_draws) |> mutate(spec = "All base (7) vs all FT (12)"),
  gain_draws |>
    filter((model_key == "llama-70b" & condition %in% c("codebook","evidence","anonymized","summarized")) |
           (family == "Fine-tuned")) |>
    gap_gap() |> mutate(spec = "Primary-4 base vs all FT (12)"),
  gain_draws |>
    semi_join(bind_rows(
      tibble(model_key = "llama-70b", condition = c("codebook","evidence","anonymized","summarized")),
      ft_diag |> select(model_key, condition)
    ), by = c("model_key","condition")) |>
    gap_gap() |> mutate(spec = "Primary-4 base vs FT diagonal (3)")
) |>
  transmute(Spec = spec, `Gap (FT−Base)` = gap_est,
            `95% CI` = glue("[{round(gap_lo,3)}, {round(gap_hi,3)}]"),
            `P(gap>0)` = p_gt0,
            `> SESOI & CI>0` = if_else(gap_est > sesoi & gap_lo > 0, "yes", "no"))

variants |> gt() |>
  tab_header(title = "Prediction 1 — cell-set sensitivity",
             subtitle = glue("SESOI = {round(sesoi,3)}")) |>
  fmt_number(columns = c(`Gap (FT−Base)`, `P(gap>0)`), decimals = 3)
```

---

## 9. Prediction 2 — do the condition contrasts survive the switch?

Every registered and reported contrast, recomputed as a paired expectation-based delta and set
beside its greedy value on the **same captured pool** (so the comparison is like-for-like, not
against the confirmatory bundle). "Survives" = same side of the SESOI verdict under both
readouts. `paired_delta()` is reused unchanged, pointed at each readout's `boot_results`.
Contrasts whose endpoints include a cell absent from the captured pool (the A1/A2 grid and the
F1' primes, which need the optional base `*-zeroshot` expectation cells) are skipped and listed.

```{r p2-contrasts}
contrast_specs <- tribble(
  ~a_mk,               ~a_cond,               ~b_mk,               ~b_cond,               ~label,               ~grp,
  "llama-70b",         "evidence",            "llama-70b",         "codebook",            "B1 · Ev−Cb",         "B",
  "llama-70b",         "anonymized",          "llama-70b",         "codebook",            "B2 · An−Cb",         "B",
  "llama-70b",         "anonymized",          "llama-70b",         "evidence",            "B3 · An−Ev",         "B",
  "llama-70b",         "summarized",          "llama-70b",         "anonymized",          "B4 · Su−An",         "B",
  "llama-70b-ft-raw",  "evidence-zeroshot",   "llama-70b",         "evidence",            "F1 · Raw",           "F1",
  "llama-70b-ft-anon", "anonymized-zeroshot", "llama-70b",         "anonymized",          "F1 · Anon",          "F1",
  "llama-70b-ft-summ", "summarized-zeroshot", "llama-70b",         "summarized",          "F1 · Summ",          "F1",
  "llama-70b-ft-raw",  "evidence-zeroshot",   "llama-70b",         "evidence-zeroshot",   "F1' · Raw (pm)",     "F1pm",
  "llama-70b-ft-anon", "anonymized-zeroshot", "llama-70b",         "anonymized-zeroshot", "F1' · Anon (pm)",    "F1pm",
  "llama-70b-ft-summ", "summarized-zeroshot", "llama-70b",         "summarized-zeroshot", "F1' · Summ (pm)",    "F1pm",
  "llama-70b-ft-anon", "evidence-zeroshot",   "llama-70b-ft-raw",  "evidence-zeroshot",   "F2 · anon−raw",      "F2",
  "llama-70b-ft-summ", "evidence-zeroshot",   "llama-70b-ft-anon", "evidence-zeroshot",   "F2 · summ−anon",     "F2",
  "llama-70b",         "evidence-zeroshot",   "llama-70b",         "codebook",            "A1 · Ev-zs−Cb",      "A1",
  "llama-70b",         "anonymized-zeroshot", "llama-70b",         "codebook",            "A1 · An-zs−Cb",      "A1",
  "llama-70b",         "summarized-zeroshot", "llama-70b",         "codebook",            "A1 · Su-zs−Cb",      "A1",
  "llama-70b",         "evidence",            "llama-70b",         "evidence-zeroshot",   "A2 · Ev gap",        "A2",
  "llama-70b",         "anonymized",          "llama-70b",         "anonymized-zeroshot", "A2 · An gap",        "A2",
  "llama-70b",         "summarized",          "llama-70b",         "summarized-zeroshot", "A2 · Su gap",        "A2",
  "llama-70b-ft-anon", "codebook",            "llama-70b-ft-raw",  "codebook",            "A3 · anon−raw",      "A3",
  "llama-70b-ft-summ", "codebook",            "llama-70b-ft-anon", "codebook",            "A3 · summ−anon",     "A3",
  "llama-70b-ft-anon", "evidence-zeroshot",   "llama-70b",         "evidence",            "A5 · FT-anon−Base",  "A5",
  "llama-70b-ft-summ", "evidence-zeroshot",   "llama-70b",         "evidence",            "A5 · FT-summ−Base",  "A5"
)

# The captured pool covers the 4 base primary conditions + 12 FT cells; the 3 base *-zeroshot
# expectation cells are optional (precommitment) and were not run. Any contrast whose endpoints
# include a cell absent from the pool (groups A1, A2, and the F1' "(pm)" primes, which need base
# *-zeroshot) is not computable here — filter those out and report them rather than erroring.
avail_cells <- distinct(greedy_boot_results, model_key, condition)
has_cell <- function(mk, cond) any(avail_cells$model_key == mk & avail_cells$condition == cond)

contrast_specs <- contrast_specs |>
  mutate(computable = pmap_lgl(list(a_mk, a_cond, b_mk, b_cond),
    \(a_mk, a_cond, b_mk, b_cond) has_cell(a_mk, a_cond) && has_cell(b_mk, b_cond)))
specs_run  <- filter(contrast_specs, computable) |> select(-computable)
specs_skip <- filter(contrast_specs, !computable)

run_contrasts <- function(results) {
  pmap_dfr(specs_run, function(a_mk, a_cond, b_mk, b_cond, label, grp) {
    paired_delta(a_mk, a_cond, b_mk, b_cond, label, results = results, sesoi_val = sesoi) |>
      mutate(grp = grp)
  })
}

if (nrow(specs_skip) > 0) {
  cat("Skipped", nrow(specs_skip), "contrast(s) needing the optional base *-zeroshot expectation\n",
      "cells (not run): ", paste(specs_skip$label, collapse = ", "), "\n",
      "Run the 3 base zero-shot expectation cells and re-pull to populate A1/A2/F1' rows.\n", sep = "")
}

greedy_contr <- run_contrasts(greedy_boot_results)
exp_contr    <- run_contrasts(exp_boot_results)

survival <- greedy_contr |>
  select(label, grp, g_est = est, g_lo = lo, g_hi = hi, g_out = sesoi_out) |>
  inner_join(select(exp_contr, label, e_est = est, e_lo = lo, e_hi = hi, e_out = sesoi_out),
             by = "label") |>
  mutate(
    sign_flip = sign(g_est) != sign(e_est),
    survives  = (g_out == e_out) & !sign_flip
  )

survival |>
  transmute(
    Contrast = label, Group = grp,
    `Greedy Δ`      = g_est,
    `Greedy CI`     = glue("[{round(g_lo,3)}, {round(g_hi,3)}]"),
    `Greedy >SESOI` = if_else(g_out, "yes", "no"),
    `Exp Δ`         = e_est,
    `Exp CI`        = glue("[{round(e_lo,3)}, {round(e_hi,3)}]"),
    `Exp >SESOI`    = if_else(e_out, "yes", "no"),
    Survives        = if_else(survives, "yes", "NO")
  ) |>
  gt(groupname_col = "Group") |>
  tab_header(title = "Prediction 2 — contrast survival, greedy vs expectation",
             subtitle = glue("Survives = same SESOI verdict & no sign flip · SESOI = {round(sesoi,3)}")) |>
  fmt_number(columns = c(`Greedy Δ`, `Exp Δ`), decimals = 3)
```

```{r p2-summary}
glue("Prediction 2: {sum(survival$survives)}/{nrow(survival)} contrasts survive the switch",
     " · {sum(!survival$survives)} change verdict",
     if (any(!survival$survives)) glue(" ({paste(survival$label[!survival$survives], collapse=', ')})") else "")
```

---

## 10. Per-cell expectation gain (full grid)

```{r gain-table}
gain_ci <- boot_both |>
  mutate(gain = greedy_mae - exp_mae) |>
  group_by(model_key, condition) |>
  # CIs first: the row-wise `gain` must be read before the scalar point estimate below reuses the
  # name (dplyr exposes earlier-created columns to later args, which would shadow the vector).
  summarise(
    greedy_mae = mean(greedy_mae[id == "Apparent"]),
    exp_mae    = mean(exp_mae[id == "Apparent"]),
    gain_lo    = quantile(gain[id != "Apparent"], 0.025),
    gain_hi    = quantile(gain[id != "Apparent"], 0.975),
    gain       = mean(gain[id == "Apparent"]),
    .groups    = "drop"
  ) |>
  mutate(model = recode(model_key, !!!model_labels), col = canon_col(condition)) |>
  arrange(model_key, condition)

gain_ci |>
  transmute(Model = model, Condition = condition,
            `Greedy MAE` = greedy_mae, `Exp MAE` = exp_mae, Gain = gain,
            `95% CI` = glue("[{round(gain_lo,3)}, {round(gain_hi,3)}]")) |>
  gt() |>
  tab_header(title = "Per-cell expectation gain (greedy MAE − expectation MAE)",
             subtitle = glue("2019 · floor = {round(rounding_floor,3)} · SESOI = {round(sesoi,3)}")) |>
  fmt_number(columns = c(`Greedy MAE`, `Exp MAE`, Gain), decimals = 3)
```

---

## 11. Serialize the expectation bundle

```{r serialize}
exp_bundle_path <- file.path(derived_dir, glue("expectation_{params$year}.rds"))
saveRDS(
  list(
    exp_ci              = exp_ci,               # per-cell expectation MAE + CIs
    greedy_ci_rerun     = greedy_ci,            # greedy MAE recomputed on captured pool
    boot_both           = boot_both,            # per-draw per-cell greedy + exp MAE
    exp_boot_results    = exp_boot_results,     # paired_delta-ready (expectation)
    greedy_boot_results = greedy_boot_results,  # paired_delta-ready (greedy, captured pool)
    gain_ci             = gain_ci,              # per-cell gain
    p2_survival         = survival,             # contrast survival table
    miss_tbl            = miss_tbl,             # capture-miss accounting
    sesoi               = sesoi,
    rounding_floor      = rounding_floor
  ),
  exp_bundle_path
)
glue("bundle written: {exp_bundle_path} ({round(file.size(exp_bundle_path)/1024)} KB)")
```
