---
title: "Coder Workforce Across Releases: Separating Attrition from Back-Coding"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
execute:
  warning: false
  message: false
---

Analysis 01 measures panel size within a single release. This analysis measures the
same quantity across seven successive releases, which separates two things a single
release combines: how many coders were working at a given moment, and how many coders
have since back-coded into that moment.

The distinction matters because V-Dem's `year` is the year being *rated*, not the year
the rating was *produced*. A coder recruited in 2022 back-codes their country to 2005,
so within one release their rows are indistinguishable from a founding coder's. This
makes departures observable — an inactive coder's last rated year freezes and stays
frozen — while arrivals are invisible, since every new recruit enters at 2005 alongside
the originals. A single release therefore produces a declining profile toward recent
years whether or not recruitment has changed.

Successive releases break the tie. A published release is frozen: only coders on the
roster during that coding round can appear in its most recent year, because no one has
had the opportunity to back-code into a file already printed. Reading each release at
its own leading edge gives one back-coding-free measurement of workforce size per
release. Reading a *fixed* historical year across releases gives the complementary
quantity — the rate at which later hires accumulate into settled years.

Releases V10 (published 2020) through V16 (published 2026) are mirrored in
`shared/vdem-data/vdem-release-archive/` from the tagged `vdemdata` package. V10 is the
earliest tagged release, so the series reaches 2020 and not 2013; earlier releases would
have to be requested from V-Dem directly. Country and indicator sets are held constant
across releases (202 countries, 257 indicators) so the series does not confound workforce
change with scope expansion.

---

## 1. Setup

```{r setup}
library(tidyverse)
library(knitr)

series <- readRDS(here::here("data", "derived", "workforce_series.rds"))
edge     <- series$edge
backfill <- series$backfill

OKABE_BLUE <- "#0072B2"   # as reported by the current release
OKABE_VERM <- "#D55E00"   # as measured at the time
```

## 2. Workforce at each release's leading edge

Each release is read at fixed offsets from its own most recent year. Offset 0 is the
leading edge itself, which is still accumulating ratings at publication; offsets 1 and 2
are progressively more settled. The comparison across releases is like-for-like because
the same offset carries the same degree of incompleteness in every release.

```{r edge-table}
edge |>
  mutate(offset = paste0("edge_minus_", offset)) |>
  pivot_wider(id_cols = c(release, pub_year, edge_year),
              names_from = offset, values_from = mean_nr) |>
  mutate(across(starts_with("edge_minus_"), ~round(.x, 2))) |>
  kable(caption = "Mean coders per country-indicator cell at each release's leading edge")
```

## 3. Back-coding accumulation into settled years

The same historical year, read from each successive release. Increments are the coders
added to an already-settled year between one release and the next.

```{r backfill-table}
backfill |>
  filter(year %in% c(2010, 2013, 2015, 2017)) |>
  mutate(mean_nr = round(mean_nr, 2), increment = round(increment, 3)) |>
  pivot_wider(id_cols = c(release, pub_year), names_from = year,
              values_from = c(mean_nr, increment)) |>
  kable(caption = "Coder count for fixed historical years, as reported by each release")
```

```{r backfill-rate}
backfill |>
  filter(year %in% 2010:2017, !is.na(increment)) |>
  group_by(release, pub_year) |>
  summarise(mean_increment = round(mean(increment), 3), .groups = "drop") |>
  arrange(pub_year) |>
  kable(caption = "Per-release back-coding increment, averaged over settled years 2010-2017")
```

## 4. Does a flat edge plus steady back-coding reproduce the profile?

If workforce size is roughly constant and each release adds a fixed increment to every
prior year, then a year's count in the current release should be its edge value plus the
increment times the number of releases since. Fitting that on V16's profile recovers both
parameters: the intercept estimates workforce at the edge, the slope estimates the
per-release back-coding rate.

```{r model}
v16_profile <- backfill |>
  filter(release == "V16", year >= 2005) |>
  mutate(age_releases = 2025 - year)

fit <- lm(mean_nr ~ age_releases, data = v16_profile)

tibble(
  term      = c("intercept (edge workforce)", "slope (back-coding per release)"),
  estimate  = round(unname(coef(fit)), 3),
  std_error = round(summary(fit)$coefficients[, 2], 4)
) |>
  kable(caption = sprintf("Linear fit to V16's within-release profile (R-squared = %.4f)",
                          summary(fit)$r.squared))
```

```{r model-fit}
v16_profile |>
  mutate(predicted = round(predict(fit), 2), observed = round(mean_nr, 2),
         residual = round(observed - predicted, 2)) |>
  filter(year %in% c(2005, 2010, 2013, 2016, 2019, 2022, 2023)) |>
  select(year, age_releases, predicted, observed, residual) |>
  kable(caption = "Predicted vs observed, V16 profile")
```

## 5. The two measurements plotted together

Both series are in the same units — mean coders per country-indicator cell — and both are
indexed by the historical year being rated. The blue line is what V16 reports today for
each year. The vermillion points are what each release reported for its own edge year at
the time it was published. The vertical gap between them at a given year is back-coding
accumulated since.

```{r fig-workforce, fig.width = 9, fig.height = 5.2, dev = "png", dpi = 200}
as_reported <- backfill |>
  filter(release == "V16", year >= 2005) |>
  transmute(year, mean_nr, series = "As reported by V16 (2026)")

at_the_time <- edge |>
  filter(offset == 0) |>
  transmute(year = edge_year, mean_nr, series = "Measured at the time (each release's edge)")

plot_dat <- bind_rows(as_reported, at_the_time)
pal <- c("As reported by V16 (2026)" = OKABE_BLUE,
         "Measured at the time (each release's edge)" = OKABE_VERM)

ggplot(plot_dat, aes(year, mean_nr, color = series)) +
  geom_line(data = as_reported, linewidth = 1) +
  geom_line(data = at_the_time, linewidth = 1, linetype = "22") +
  geom_point(data = at_the_time, size = 2.4) +
  geom_text(
    data = at_the_time |> filter(year %in% c(2019, 2025)),
    aes(label = sprintf("%.1f", mean_nr)), vjust = 1.9, size = 3.4, show.legend = FALSE
  ) +
  geom_text(
    data = as_reported |> filter(year %in% c(2005, 2013)),
    aes(label = sprintf("%.1f", mean_nr)), vjust = -1.1, size = 3.4, show.legend = FALSE
  ) +
  scale_color_manual(values = pal, name = NULL) +
  scale_x_continuous(breaks = seq(2005, 2025, 5)) +
  labs(
    title    = "Coders per country-year: as reported today vs measured at the time",
    subtitle = "Constant set of 202 countries and 257 indicators across releases V10-V16",
    x = "Year being rated", y = "Coders per country-indicator cell"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(),
        plot.title.position = "plot")
```

## 6. Reach of the series

The archive reaches back to V10 because that is the earliest tag carrying a bundled
dataset. The edge readings therefore cover rating-years 2019 through 2025. Extending the
"measured at the time" series to 2013 — the inflection point in Analysis 01 — requires the
v3 release, which is not distributed through the package and would have to be obtained
from V-Dem directly.

```{r reach}
edge |>
  filter(offset == 0) |>
  summarise(
    releases       = n(),
    first_release  = first(release),
    first_edge_yr  = min(edge_year),
    last_release   = last(release),
    last_edge_yr   = max(edge_year),
    edge_min       = round(min(mean_nr), 2),
    edge_max       = round(max(mean_nr), 2)
  ) |>
  kable(caption = "Coverage of the cross-release workforce series")
```

---

## 7. Turnover and threshold shortfall

Mean panel size is a poor summary of panel health. It averages the tail away, and the
tail is where the measurement model loses precision: a mean of 6 is compatible with a
distribution whose bottom quarter sits under the floor V-Dem tells users to work above.
The codebook names two thresholds directly.

> We strongly advise against using observations based on three or fewer coders.

> By providing the number of Country Experts for each country-variable-year/date, we
> suggest that users primarily base analyses on observations based on five or more
> coders.

Both appear in the v15 codebook on p. 35 (PDF sheet 36), in the entry for the `_nr`
variable. The first is repeated in the Cautionary Notes at p. 26 (sheet 27), which is
also issued standalone as `cautionary_notes.pdf`.

The two thresholds do not have the same standing, and it is worth keeping them apart.

**Three or fewer is a usage instruction with a stated empirical basis.** The Cautionary
Notes preface it with "From analysis, we have found that this at times can result in
significant changes in point estimates, most likely as a consequence of self-selected
attrition of Country Experts" (p. 26). No analysis is shown or cited.

**Five is a recruitment quota that is also passed on as user guidance.** The methodology
describes it as a staffing target rather than a property of the estimates:

> we endeavor to find a minimum of five Country Experts to code each country-year for
> every indicator

> The Program Managers at the V-Dem Institute (University of Gothenburg) issue invitations
> until the quota of five Country Experts per survey. We usually recruit 7 or more experts
> to be able to replace those who fail to begin or complete the survey on time.

(methodology p. 13 and p. 16; PDF sheets 14 and 17). The codebook calls it an assumption
the methodology "is based on" rather than a finding. The one substantive rationale offered
anywhere is outlier dilution — "wayward experts ... are unlikely to have a strong influence
on the point estimates that result from the measurement model's aggregation across five or
more experts" (methodology p. 25, sheet 26) — asserted without a derivation or simulation.

Four falls in a gap: not warned against, below the recommended floor, and given no
guidance either way. Analysis 01's marginal-value section reports where the data puts the
break.

Not checked: the Pemstein et al. measurement-model working paper is not among the PDFs in
`initial-exploration/vdem-docs/`, so a threshold justification may exist there.

A share crossing one of V-Dem's own lines is interpretable in a way a mean is not. This
section reports two such series, both built by `helpers/build_coder_attrition.R`.

```{r attrition-setup}
attr_dat <- readRDS(here::here("data", "derived", "coder_attrition.rds"))
turnover <- attr_dat$turnover
sf_year  <- attr_dat$shortfall_year
sf_edge  <- attr_dat$shortfall_edge
```

### Turnover

A coder's exit is inferred, not recorded: it is the last rating-year they ever produced.
Because the archive is additive and contemporary recruits back-code to 2005, a coder
present in year Y and absent from every later year has genuinely stopped contributing at
the leading edge. The final row is not an exit cohort — those coders are still active.

```{r turnover-table}
turnover |>
  filter(year >= 2013) |>
  transmute(Year = year, `Coders stopping` = coders_stopping,
            `% of pool` = pct_of_pool,
            Note = ifelse(still_active, "still active, not an exit cohort", "")) |>
  kable(caption = sprintf("Annual coder exits against a contemporary pool of %d (v15)",
                          attr_dat$meta$contemporary_pool))
```

```{r fig-turnover, fig.width = 8, fig.height = 3.6, dev = "png", dpi = 200}
turnover |>
  filter(year >= 2013, !still_active) |>
  ggplot(aes(year, coders_stopping)) +
  geom_segment(aes(xend = year, yend = 0), colour = OKABE_BLUE, linewidth = 0.8) +
  geom_point(colour = OKABE_BLUE, size = 2.4) +
  geom_hline(yintercept = median(turnover$coders_stopping[turnover$year %in% 2014:2023]),
             linetype = "22", colour = "grey45") +
  annotate("text", x = 2013.2, y = 175, hjust = 0, vjust = -0.5, size = 3.2,
           colour = "grey35", label = "median 163.5/yr") +
  scale_x_continuous(breaks = seq(2013, 2023, 2)) +
  labs(title = "Coders whose last rating-year is Y",
       subtitle = "v15 coder-level file. 2024 omitted: those coders are still active.",
       x = NULL, y = "Coders") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), plot.title.position = "plot")
```

Exits run at roughly 120–220 a year against a standing pool of about 4,000 — near 4%
annual turnover. Section 2's flat leading-edge series is what tells us these are being
replaced; turnover alone cannot distinguish replacement from decline.

### Threshold shortfall

The same two readings as Section 5. The blue line is what V16 reports today for each
year; the vermillion points are what each release reported at its own leading edge, before
back-coding had filled anything in.

```{r fig-shortfall, fig.width = 9, fig.height = 4.8, dev = "png", dpi = 200}
sf_plot <- bind_rows(
  sf_year |> transmute(year, pct = pct_lt_5, series = "As reported by V16 (2026)"),
  sf_edge |> transmute(year = edge_year, pct = pct_lt_5,
                       series = "Measured at the time (each release's edge)")
)

ggplot(sf_plot, aes(year, pct, colour = series)) +
  geom_line(data = ~filter(.x, series == "As reported by V16 (2026)"), linewidth = 1) +
  geom_line(data = ~filter(.x, series != "As reported by V16 (2026)"),
            linewidth = 1, linetype = "22") +
  geom_point(data = ~filter(.x, series != "As reported by V16 (2026)"), size = 2.4) +
  scale_colour_manual(values = c("As reported by V16 (2026)" = OKABE_BLUE,
                                 "Measured at the time (each release's edge)" = OKABE_VERM),
                      name = NULL) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(title = "Share of country-indicator cells below V-Dem's recommended floor of five coders",
       subtitle = sprintf("All %d indicators carrying an _nr column. Constant country set for the edge series.",
                          attr_dat$meta$n_indicators),
       x = "Year being rated", y = "Cells with fewer than 5 coders") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        plot.title.position = "plot")
```

```{r shortfall-tables}
sf_year |>
  filter(year %in% c(2005, 2010, 2015, 2019:2024)) |>
  transmute(Year = year, Cells = cells, `Mean n` = mean_n, `Median n` = median_n,
            `% <= 3` = pct_le_3, `% < 5` = pct_lt_5) |>
  kable(caption = "Within V16: shortfall by year, as the current release reports it")

sf_edge |>
  transmute(Release = release, `Published` = pub_year, `Edge year` = edge_year,
            Cells = cells, `Mean n` = mean_n, `% <= 3` = pct_le_3, `% < 5` = pct_lt_5) |>
  kable(caption = "At each release's own leading edge: shortfall before any back-coding")
```

Two readings sit side by side here. Within V16, the shortfall is near 7% for settled
years and rises to 21% at 2024 — the back-coding gradient of Section 5, expressed as a
share rather than a mean. At the leading edge, where no back-coding has occurred, the
shortfall has run between 24% and 42% in every release since V10, with V16 the highest
since V10.

Note the indicator set: these figures use every indicator carrying an `_nr` column. An
analysis restricted to a narrower working set will produce different levels, and
`build_coder_attrition()` takes an `indicators` argument for that purpose.
