---
title: "Paper figures & tables — draft gallery"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

**Reminder**: Yihui has released a fig.note option for `knitr`. https://github.com/yihui/knitr/pull/2453. Install development version and try it when you build the PDF.

```{r setup}
library(tidyverse)
library(gt)
library(glue)

proj_root <- rprojroot::find_root(rprojroot::is_git_root)
source(file.path(proj_root, "helpers", "bootstrap_helpers.R"))
source(file.path(proj_root, "helpers", "paper_figures.R"))
derived <- file.path(proj_root, "data", "derived")

# Panel-size motivation (Figures 1-2, Appendix A1-A2): how thin V-Dem's panels are, and what
# an added coder buys in the project's own posterior SD. One bundle serves all four figures.
# Guarded like the rest; build with:
#   Rscript helpers/build_coder_attrition.R   # prerequisite: reads the release archive
#   Rscript helpers/build_panelsize.R
ps23 <- file.path(derived, "panelsize.rds")
have_panelsize <- file.exists(ps23)
if (have_panelsize) panelsize <- readRDS(ps23)

boot_bundle <- readRDS(file.path(derived, "bootstrap_2019.rds"))
exp_bundle  <- readRDS(file.path(derived, "expectation_2019.rds"))
dm_slope    <- readRDS(file.path(derived, "distmatch_slope_2019.rds"))  # Fig 4 (Test-3 slope)

# 2023 holdout bundles (appendix replication of Figs 3 & 4). Guarded so the doc still renders if
# they haven't been built yet:
#   Rscript helpers/build_bundles.R    --year 2023 --runs-subdir 2023 --exp-subdir 2023
#   Rscript helpers/build_distmatch.R  --year 2023 --runs-subdir 2023
b23 <- file.path(derived, "bootstrap_2023.rds")
e23 <- file.path(derived, "expectation_2023.rds")
have_2023   <- file.exists(b23) && file.exists(e23)
if (have_2023) {
  boot_bundle_2023 <- readRDS(b23)
  exp_bundle_2023  <- readRDS(e23)
}
ds23 <- file.path(derived, "distmatch_slope_2023.rds")
have_2023_slope <- file.exists(ds23)
if (have_2023_slope) dm_slope_2023 <- readRDS(ds23)

# 2024 FH-only holdout (loose out-of-sample replication of Figure 3; see
# notes/plan-2024-holdout.md — private notes repo). Base ladder (4 conditions) + FT-raw diagonal
# restricted to {codebook, evidence-zeroshot} — FT-raw on anonymized/summarized text is
# appendix-only (2026-09-04 decision), matching the Fine-tuned block's 2-row design here. Guarded
# so the doc still renders if not yet built:
#   Rscript helpers/build_bundles.R --year 2024 --runs-subdir 2024-fhonly --exp-subdir 2024-fhonly
b24 <- file.path(derived, "bootstrap_2024.rds")
e24 <- file.path(derived, "expectation_2024.rds")
have_2024 <- file.exists(b24) && file.exists(e24)
if (have_2024) {
  boot_bundle_2024 <- readRDS(b24)
  exp_bundle_2024  <- readRDS(e24)
}

sd_signed <- readRDS(file.path(derived, "signeddev_xmodel_2019.rds"))  # Fig 5 (signed deviation)
sds23 <- file.path(derived, "signeddev_xmodel_2023.rds")
have_2023_signed <- file.exists(sds23)
if (have_2023_signed) sd_signed_2023 <- readRDS(sds23)

# Re-identification rates (Fig 6; Question 5). Build with: Rscript helpers/build_reidrates.R
rr23 <- file.path(derived, "reidrates_2023.rds")
have_reidrates <- file.exists(rr23)
if (have_reidrates) reidrates_bundle_2023 <- readRDS(rr23)

# Movement illustration (Appendix A13). Build with: Rscript helpers/build_movement.R --year 2023 --baseline-year 2018
mv23 <- file.path(derived, "movement_2023.rds")
have_movement <- file.exists(mv23)
if (have_movement) {
  movement_bundle_2023 <- readRDS(mv23)
  country_names <- read_csv(file.path(proj_root, "data", "processed", "ert.csv"),
                            show_col_types = FALSE) |>
    distinct(country_text_id, country_name)
}

# Movement/Prominence effects on the difficulty-tracking slope, by condition (Appendix A14).
# Superseded
# the single-panel Identity x Compression interaction version (2026-09-06) -- two separate,
# simpler models (movement-only, reid-only, no interaction term), each run across all four base
# conditions + the trimmed Codebook/Raw Text Fine-Tuned pair. Build with:
#   Rscript helpers/build_slope_by_condition.R --moderator movement
#   Rscript helpers/build_slope_by_condition.R --moderator reid
sbc_movement_path <- file.path(derived, "slopebycondition_movement_2023.rds")
sbc_reid_path     <- file.path(derived, "slopebycondition_reid_2023.rds")
have_sbc_movement <- file.exists(sbc_movement_path)
have_sbc_reid     <- file.exists(sbc_reid_path)
if (have_sbc_movement) sbc_movement <- readRDS(sbc_movement_path)
if (have_sbc_reid)     sbc_reid     <- readRDS(sbc_reid_path)

# Movement/Prominence effects on signed deviation, by condition (Appendix A15). Build with:
#   Rscript helpers/build_signeddev_by_condition.R --moderator movement
#   Rscript helpers/build_signeddev_by_condition.R --moderator reid
sdbc_movement_path <- file.path(derived, "signeddevbycondition_movement_2023.rds")
sdbc_reid_path     <- file.path(derived, "signeddevbycondition_reid_2023.rds")
have_sdbc_movement <- file.exists(sdbc_movement_path)
have_sdbc_reid     <- file.exists(sdbc_reid_path)
if (have_sdbc_movement) sdbc_movement <- readRDS(sdbc_movement_path)
if (have_sdbc_reid)     sdbc_reid     <- readRDS(sdbc_reid_path)

# Name-swap, base + fine-tuned, all three families (Fig 7, companion to Section 6, and the A11
# salience tests). Build with:
#   Rscript helpers/build_nameswap_channels.R
# Supersedes the position/win-rate pair (build_nameswap_rescaled.R) and the registered raw-track
# metric (build_priorreliance.R --panel B), neither of which is rendered any more; both are
# retained in notes/mockups/nameswap-rescaled.* with their rationale.
nsc23 <- file.path(derived, "nameswapchannels_2023.rds")
have_nameswap_channels <- file.exists(nsc23)
if (have_nameswap_channels) nameswap_channels_2023 <- readRDS(nsc23)

# Compression and de-identification (Fig 8) -- the two successive steps from raw source text to
# the Summarized condition, each a paired contrast on all three outcomes: MAE, signed deviation,
# and the difficulty-tracking slope. These bundles also still carry the Full Text contrast that
# the retired identity-effect appendix rendered (fig_identity_effect_combined), which nothing calls now.
# Build with:
#   Rscript helpers/build_identity_effect.R
#   Rscript helpers/build_identity_effect_signeddev.R
#   Rscript helpers/build_identity_effect_slope.R
identityeffect_path <- file.path(derived, "identityeffect_2023.rds")
have_identityeffect <- file.exists(identityeffect_path)
if (have_identityeffect) identityeffect_bundle <- readRDS(identityeffect_path)

identityeffect_sd_path <- file.path(derived, "identityeffect_signeddev_2023.rds")
have_identityeffect_sd <- file.exists(identityeffect_sd_path)
if (have_identityeffect_sd) identityeffect_sd_bundle <- readRDS(identityeffect_sd_path)

identityeffect_slope_path <- file.path(derived, "identityeffect_slope_2023.rds")
have_identityeffect_slope <- file.exists(identityeffect_slope_path)
if (have_identityeffect_slope) identityeffect_slope_bundle <- readRDS(identityeffect_slope_path)

# Synthetic-swap deployment, MEAN-SHIFT version. No longer rendered: Figure 9 now reports
# verdict changes (build_fliprate.R below) instead. Retained because the mean-shift
# estimator is (signed deviation)/(n+1), i.e. Figure 5 rescaled, and the numbers are still
# quotable in prose; fig_augmentation_combined() and fig_degradation_combined() still work.
# Build with:
#   Rscript helpers/build_augmentation.R
#   Rscript helpers/build_degradation.R
#   Rscript helpers/build_doseresponse_augmentation.R
#   Rscript helpers/build_doseresponse_degradation.R
augmentation_path <- file.path(derived, "augmentation_2023.rds")
degradation_path  <- file.path(derived, "degradation_2023.rds")
dr_aug_path       <- file.path(derived, "doseresponse_augmentation_2023.rds")
dr_deg_path       <- file.path(derived, "doseresponse_degradation_2023.rds")
have_augmentation <- file.exists(augmentation_path)
have_degradation  <- file.exists(degradation_path)
have_dr_aug       <- file.exists(dr_aug_path)
have_dr_deg       <- file.exists(dr_deg_path)
if (have_augmentation) augmentation_bundle_2023 <- readRDS(augmentation_path)
if (have_degradation)  degradation_bundle_2023  <- readRDS(degradation_path)
if (have_dr_aug)        dr_augmentation_2023    <- readRDS(dr_aug_path)
if (have_dr_deg)        dr_degradation_2023     <- readRDS(dr_deg_path)

# Verdict changes and directional push (Figure 9, four panels). A verdict is
# round(panel mean); the benchmark is real held-out-coder churn, measured rather than assumed.
# Two mechanics on one constructed-thinness design (base panel of 4 drawn from source panels
# with >=8 coders). Build with:
#   Rscript helpers/build_fliprate.R --mechanic augmentation
#   Rscript helpers/build_fliprate.R --mechanic degradation
fra23 <- file.path(derived, "fliprate_augmentation_2023.rds")
frd23 <- file.path(derived, "fliprate_degradation_2023.rds")
have_fliprate <- file.exists(fra23) && file.exists(frd23)
if (have_fliprate) {
  fliprate_augmentation_2023 <- readRDS(fra23)
  fliprate_degradation_2023  <- readRDS(frd23)
}

# Task framing vs. evidence body (Appendix A12): the three summarized arms measured from raw
# evidence, and the task-framing leg split by whether the model re-identified the country.
# Build with:
#   Rscript helpers/build_summarization_arms.R
#   Rscript helpers/build_reid_split.R
sa23 <- file.path(derived, "summarizationarms_2023.rds")
rs23 <- file.path(derived, "reidsplit_2023.rds")
have_summarization_arms <- file.exists(sa23)
have_reid_split         <- file.exists(rs23)
if (have_summarization_arms) summarization_arms_2023 <- readRDS(sa23)
if (have_reid_split)         reid_split_2023         <- readRDS(rs23)

# Decoding-temperature ladder (Appendix A9): what a SAMPLED rating would do to all three
# outcomes, computed in closed form from the stored rating distribution -- no new inference.
# Build with: Rscript helpers/build_temperature_ladder.R
tl23 <- file.path(derived, "temperature_ladder_2023.rds")
have_temperature_ladder <- file.exists(tl23)
if (have_temperature_ladder) temperature_ladder_2023 <- readRDS(tl23)
```

Broader question: how does information help an AI become a synthetic coder?

A synthetic coder can be judged against two different targets, and the same manipulation can move
the two in opposite directions. Under the first, the model is a **consensus estimator**: what
matters is the panel's aggregate judgment, and the model is scored by how close it comes to it.
Under the second, the model is a **panel member**: what matters is whether it behaves like one more
expert seated on the panel, which is a question about the distribution of its ratings rather than
about any single distance. Applied work on machine coding almost always adopts the first, because a
single agreed label is usually all that exists. V-Dem publishes every coder's individual rating on
every case, and that is what makes the second target measurable here.

The two targets separate because a human expert does not sit on the panel mean. A held-out V-Dem
coder is on average about 0.71 rating points away from the mean of the colleagues rating the same
case (0.73 in 2023), and that gap is a property of the task rather than a shortcoming of the coder:
these judgments are genuinely contestable, and one opinion carries the full spread that averaging
across a panel removes. A synthetic coder that succeeds as a panel member will therefore not
converge on the consensus. It will depart from it about as often as a human does, and — the part
this paper turns on — it will depart in the same *shape*: more error on the cases the human panel
itself found hard, and no standing tendency to rate above or below the panel.

That reference distance is a coordinate, not a bullseye. A merely noisy rater can sit 0.71 from the
mean while resembling a coder in no other respect. Distance from the panel mean locates a model;
the difficulty-tracking slope and the signed deviation are what say whether it arrived there the way
a human does.

We evaluate against the panel mean first because it is the field's default criterion for machine
coders. On that criterion the expectation is the field's own: supplying evidence should move the
model toward the panel mean; de-identifying that evidence should move it closer still, on the
reasoning that a model which cannot tell which country it is reading about cannot fall back on a
stored impression of that country, or on a V-Dem score it may have encountered in
training; and fine-tuning should amplify both.[^finetune-note] Registered alongside that
expectation is a second reading — that richer evidence leads the model to weigh the case
differently from the panel rather than to ignore it — along with the tests that decide between them:
the name-swap, the signed deviation, and the difficulty-tracking slope.

[^finetune-note]: (Drop this in if we decide to relegate the FT-summ and FT-anon models to the
appendix) We also tested models that were finetuned on anonymized and summarized evidence. The
results suggested that... We explore these results further in the online appendix.

## 1. How thin are V-Dem's panels, and when?

The paper proposes filling a panel seat with a synthetic coder. This section establishes
where the seats are actually empty.

Panel A places the contemporary era against V-Dem's whole history, in the units V-Dem's own
guidance is written in: the share of country-indicator-years falling below its recommended
floor of five coders. Panel B reads the contemporary era twice — as each year was first
published, when no back-coding had yet been possible, and as the current release reports
that same year now.

The second reading is what separates two things a single release combines. V-Dem's `year` is
the year being *rated*, not the year a rating was *produced*, and contemporary recruits
back-code to 2005, so any one release slopes downward toward recent years whether or not the
workforce changed. Reading each year at first publication removes that: a published release
is frozen, so only coders on the roster during that coding round can appear in its most
recent year. See `analysis/14-coder-workforce-cross-release.qmd` for the identification
argument and `helpers/build_panelsize.R` for the build.

<!-- TODO (EJT): interpretation. The vermillion bars carry no trend across cohorts (23.1 to
40.9%, entering); the blue bars ramp monotonically with release cycles elapsed (3.1, 4.2,
6.3, 8.5, 12.1, 18.3). The 2019 cohort has had six cycles and sits at 3.1%, still 2.2x the
settled level of 1.4%; no cohort in the observed window has finished filling. -->

```{r}
#| label: fig-panelsize-history
#| fig-cap: "Panel size across V-Dem's history, and how the contemporary era reaches its own floor. (A) Share of country-indicator-years at each panel size, by era; bold figures are the share below five. Oranges fall below V-Dem's recommended floor, blues at or above it. 205 scored indicators; the 19th-century column rests on the narrower set of indicators carrying pre-1900 data. (B) The same years read twice: as first published (vermillion), before any back-coding was possible, and as V16 reports them now (blue). The number below each pair is release cycles elapsed; the dashed line is the settled level of 1.4%, the share below five in fully filled years (2005-2015). The 2025 bars are equal because that year has had no cycles yet. Because V-Dem's year is the year being rated and contemporary recruits back-code to 2005, the within-release decline toward recent years is fill-in that has not arrived rather than coders leaving; the leading-edge series is flat. Appendix A1 gives the country and regime cuts."
#| fig-width: 12
#| fig-height: 5.4
#| eval: !expr have_panelsize
fig_panelsize_history(panelsize)
```

## 2. What does an added coder buy?

Figure 1 is a count of empty seats; it says nothing about what filling one is worth. This
section prices a coder in V-Dem's own uncertainty units — the posterior standard deviation
the measurement model publishes alongside every C-type indicator.

Panel-size dummies are estimated inside indicator, country and year fixed effects, so each
comparison holds the case constant rather than comparing thin countries to thick ones. Both
panels come from one fit and one country-clustered covariance matrix: Panel B's point
estimates are exactly Panel A's first differences, but its intervals are not readable off
Panel A's ribbons, because the variance of a difference carries a covariance term and
adjacent coefficients here correlate about 0.76.

The window is 2005 onward. Before 2005 the contemporary panel does not operate — those
country-years belong to the retrospective historical programme, where one or two specialists
coded a country's whole political history at once — and pooling the two would put roughly
80% of all 1-3 coder cells into the left end of the figure.

<!-- TODO (EJT): interpretation. Points the current estimates will support: the fourth
coder's contribution is null (+0.005, CI -0.006 to +0.016) and the fifteenth's is zero, but
every coder from the fifth through the fourteenth excludes zero; going from four coders to
nine buys 0.114, against 0.047 for going from one to four, so most of the available
precision is purchased above the floor rather than below it. Two cautions: V-Dem's floor of
five is a recruitment quota (methodology pp. 13, 16), not a finding derived from posterior
SD, so this figure is a check on it rather than a restatement of it; and the 3-4-5
non-monotonicity sits exactly where N jumps from 3,038 to 19,463 cells, in a range where
thin contemporary cells are disproportionately leading-edge years and hard-to-recruit
countries, so it is better read as composition than as dose-response. -->

```{r}
#| label: fig-coder-value
#| fig-cap: "What an added coder buys, in V-Dem's own uncertainty units. Panel-size dummies within indicator, country and year fixed effects, so each comparison holds the case constant; contemporary panel only (2005-2024), 95% intervals clustered by country. (A) Posterior SD at each panel size; lower is less uncertainty. (B) The change contributed by each added coder. These are exactly Panel A's first differences, but the intervals cannot be read off Panel A's ribbons: the variance of a difference is V[k,k] + V[k-1,k-1] - 2V[k,k-1], and adjacent coefficients correlate about 0.76, so dropping the covariance term inflates the SE by roughly 1.6x on average. Shaded = V-Dem's 'three or fewer' warning zone; dashed rule = its recommended floor of five. Neither threshold was derived from posterior SD -- the first is a usage instruction whose stated empirical basis is unpublished, the second a recruitment quota the methodology describes as a staffing target. The raw (unadjusted) companion is available from the same bundle via `fig_coder_value(panelsize, series = \"raw\")`."
#| fig-width: 10
#| fig-height: 4.6
#| eval: !expr have_panelsize
fig_coder_value(panelsize)
```

## 3. Synthetic-coder performance

The first of the three axes the paper tracks: how far each model and condition sits from the panel
mean. Under the consensus-estimator reading a short distance is the goal, and naive carry-forward is
the bar to clear. Under the panel-member reading this axis only *locates* a model — the human
reference marks where one expert coder sits relative to the mean of the others, so a synthetic coder
landing near that line is emitting one committed opinion rather than averaging toward the middle.
Neither reading makes a short distance good on its own, which is why Figures 4 and 5 follow.

The base rows carry the study's first result: distance from the panel mean grows as evidence is
added, and grows further as identity is stripped from it. That pattern is consistent with two
readings — the model failing to use the evidence, or using it and weighting the case differently
from the panel — and this axis cannot separate them. Figures 4 and 5 are what do.

```{r}
#| label: fig-crossmodel
#| fig-cap: "Synthetic coder performance across model families (Llama 70B / Qwen 72B / Gemma 27B). Base conditions (top, all four text inputs) and fine-tuned raw-text adapters (bottom, Codebook and Raw Text only — the matched-format check on Anonymized/Summarized is Appendix A7). Color = model, with country-clustered 95% CIs. The base rows show the greedy (mode) readout only — greedy and mean differ by at most 0.045 MAE there (Qwen; ≤0.017 for Llama and Gemma), well inside the SESOI band and with the same condition ordering under either readout; the fine-tuned rows show both greedy (filled) and expectation/mean (open), where the two readouts diverge by 0.16–0.30 MAE. Dashed line = naive persistence; long-dash = human reference; grey band = ±SESOI."
#| fig-width: 8
#| fig-height: 5.2
fig_crossmodel_landscape(exp_bundle, boot_bundle, band = "human_loo", base_readout = "greedy",
                         ft_conds = c("codebook", "evidence-zeroshot"))
```

We will highlight here how the B hypotheses were either null or confirmed the alternative hypotheses. The most surprising finding is that evidence appears to push the models further away from consensus. It is not beyond the SESOI, but it is striking that evidence appears to increase MAE of the models rather than reduce it. The models are paying attention to the evidence but it is making them disagree as much as or more than individual human coders rather than pushing them towards the consensus view.  

The F hypotheses were supported if we look at the expectation readout rather than greedy decoding. But the greedy decoding results are suggestive of something we did not anticipate which is that greedy mode induces the model to better emulate the human coder. However we cannot assess how human-like the AI is by looking at MAE alone. But rather we have to look at the distribution of the AI codings relative to human coders. (transition to results in QMD 07 and 08)

Difficulty slope, attitude, exaggeration gap analysis suggest that the finetuned model is the most human-like and, further, that it is the weights more than the information in the text that does the heavy lifting. The 2x2 suggests that the fine-tuned raw evidence model can perform as well or better without the evidence in the prompt. 

For QMD 09, part I may be underpowered and does not use stale priors. Need to repeat on 2023. The Part II horse race suggests that of the seven models, FT-Raw Text performs the best in terms of mimicking human coders. This is an underpowered version of QMD 07s distributional analysis.

Both readouts come off the same trained weights; they differ only in what is asked of the learned
distribution over ratings.

- **Greedy (mode) readout** → a panel member. One committed ordinal judgment, the rating the
  learned coder distribution puts most weight on.
- **Expectation (mean) readout** → a consensus estimate. The probability-weighted mean of that
  same distribution, i.e. an average over a synthetic panel of one.

The size of the gap between them is informative in itself: a model that learned a point estimate has
mode ≈ mean, so the two readouts nearly coincide (base, ≤ 0.045 MAE apart). The fine-tuned models
separate by 0.16–0.30, which is what a learned distribution looks like. The mean readout is reported
as a diagnostic of what fine-tuning taught, not as a deployment recommendation — naive carry-forward
is closer to the panel mean than it is.

## 4. Difficulty tracking

A companion to Figure 3 on the same rows and greedy readout. **Case difficulty** `h_c` is the
within-cell human leave-one-out MAE — how much the real coders disagreed on that country-year-
indicator (the per-case version of Figure 3's human reference line). The **difficulty-tracking
slope** is the weighted OLS slope of a cell's AI error `a_c` on `h_c`. Because `h_c` is defined
as the human error, the human reference lies on the 45° line (slope 1) by construction; a slope
below 1 is flatter than that reference.

Panel A draws the relationship for two cells against the human reference (curves = mean error per
difficulty bin, with country-clustered 95% CI bands), with every cell shown as a faint grey trace
behind them. Panel B is the slope for every cell, with country-clustered 95% CIs, against the
human reference at 1 (slope 1 = tracks difficulty like a human, flatter = leans on a prior).

```{r}
#| label: fig-crossmodel-slope
#| fig-cap: "Difficulty tracking (greedy readout). (A) Mean AI error vs case difficulty h_c for a flat and a steep exemplar, against the human reference (which lies on the 45° line because h_c is defined as human error), every cell a faint grey trace; 20 equal-count difficulty bins, country-clustered 95% CI bands. (B) Weighted OLS slope of AI error on h_c for every cell, country-clustered 95% CIs, long-dash = human reference (slope 1). Base block covers all four inputs; the FT-raw block shows Codebook and Raw Text only (the matched-format check on Anonymized/Summarized is Appendix A7)."
#| fig-width: 12
#| fig-height: 5.4
fig_crossmodel_slope_2panel(dm_slope, ft_conds = c("codebook", "evidence-zeroshot"))
```

<!-- TODO (EJT): interpretation of the difficulty-tracking pattern. Robustness: slopes on the
>=5-coder difficulty pool preserve the cross-model ordering and the FT-vs-base gap (see
build_distmatch.R min_coders arg); the >=2-coder pool is used here for consistency with Fig 3. -->

## 5. Directional bias — signed deviation

<!-- STOPGAP PROSE (EJT to rewrite): -->
The third axis keeps the *direction* MAE discards: `s_c = AI rating − panel mean`. Zero is the
human panel-member benchmark, since a real coder's signed deviations average to zero by
construction. Panel A traces `s_c` across V-Dem's Regimes of the World for three exemplar cells —
the best-calibrated cell, the one with the steepest regime tilt, and the harshest (most negative
average `s_c`) — with every cell drawn faintly behind them, against the dotted zero line and the
dashed V-Dem IRT `_ord` target on the same axis. Panel B is the average `s_c` per cell: negative
means the model rates below the panel on average (harsh), positive means above (generous). The
democracy-quintile cut of the same measure is in Appendix A5.

```{r}
#| label: fig-crossmodel-signeddev-regime
#| fig-cap: "Directional bias by regime type (greedy readout). (A) Signed deviation (AI − panel mean) across V-Dem's Regimes of the World for three exemplar cells — best-calibrated, steepest regime tilt, harshest average deviation — with every cell shown as a faint grey trace; dotted 0 = human panel-member benchmark, dashed = V-Dem IRT `_ord` target on the same axis, grey band = ±SESOI, country-clustered 95% CI whiskers. (B) Average signed deviation per cell (0 = panel-member benchmark; negative = rates below the panel on average), country-clustered 95% CIs, grey band = ±SESOI. Base block covers all four inputs; the FT-raw block shows Codebook and Raw Text only (the matched-format check on Anonymized/Summarized is Appendix A7)."
#| fig-width: 12
#| fig-height: 5.4
fig_crossmodel_signeddev_2panel_regime(sd_signed, ft_conds = c("codebook", "evidence-zeroshot"))
```

<!-- TODO (EJT): interpretation. Llama and Qwen rate near-zero under Codebook; every base cell
with de-identified text is strongly negative (rates below the panel), worst under
Anonymized/Summarized. Gemma's base Codebook is itself strongly negative (-0.25). Fine-tuning
shrinks the large negative deviations but overshoots into mild generosity for Qwen and Gemma
on Codebook. -->

## 6. Re-identification

Even shown only de-identified evidence, how often does the model name the true country anyway?
**Exact-match re-identification rate** by model family (Llama 70B / Qwen 72B / Gemma 27B), by weights
(Base vs. Fine-Tuned raw-text adapter — the only fine-tuned condition in the study now), and by
de-identification condition (Anonymized vs. Summarized), 2023. Bars ordered left to right within
each condition by descending Base × Anonymized rate. This is a descriptive result in its own
right — the rates are far above chance regardless of condition or model (chance ≈ 1/176 countries
≈ 0.6%; not drawn as a reference line since it would sit indistinguishably close to zero at this
scale) — and one of the ingredients the prominence-crossing figures in Appendix A14-A15 build
on: a country has to be
identifiable before the model can lean on an identity-based prior about it. See
`notes/mockups/reid-rates-fig-concept.md` for the full design rationale and a data-integrity fix
(duplicated rows from a job-submission race) this bundle already reflects.

```{r}
#| label: fig-reid-rates
#| fig-cap: "Exact-match re-identification rate by model family, weights, and de-identification condition (2023). Plain bars, no CI (a descriptive proportion, not a paired contrast). Chance level ≈0.6%, not drawn."
#| fig-width: 9
#| fig-height: 5
#| eval: !expr have_reidrates
fig_reid_rates(reidrates_bundle_2023)
```

## 7. Name-swap: does the rating follow the evidence or the label?

Section 6 asks whether the model can still *recognize* a country from de-identified evidence. This
section asks the complementary question directly: hold the real evidence text fixed, swap in a
different country's name, and see whether the rating follows the content or drifts toward whatever
the injected name implies. Base and fine-tuned weights, all three families, 2023.

The name-swap battery ran two arms against one within-region pairing set: **swapped**, where the
framing names a different country, and **correct**, where it names the true one. The same
de-identified evidence goes in both times, so differencing a swap against its own control isolates
what the name alone did, with no panel mean entering and nothing else varying. Panel A reports
that difference, `rating_swapped − rating_correct`, as a mean absolute move in rating points.
Because it *is* a rating-point scale, the ±SESOI band applies and is drawn as the shaded region
from zero to the threshold.

Panel B asks the complementary question — not how far the rating moved, but which of the two
sources it followed. Within the swapped arm, the rating is regressed on both panel means at once
(`swapped_rating ~ named_mean + source_mean`, demeaned within indicator), and the panel plots the
text's share of the two coefficients. The true country's name appears nowhere in a swapped prompt,
so the text channel can only be non-zero if the model is drawing country-specific information out
of the document itself; 0.50 means the two channels pull equally. The predictors are separable
enough to estimate both (correlation +0.364 within indicator, VIF 1.15) because the pairing is a
random derangement within region.

Both panels split on whether that country-indicator's de-identified evidence was re-identified in
Section 6's probe — the project's salience measure, held fixed at the base model's result so the
Base and Fine-Tuned rows divide identical items.

**The name moves ratings well above the threshold that matters, and fine-tuning roughly doubles
its pull.** Mean shifts are 0.301 / 0.230 / 0.261 for the base models (Llama / Qwen / Gemma) and
0.421 / 0.714 / 0.467 fine-tuned — between two and six times the SESOI of 0.113. Since the ratings
are integers, the same manipulation changes the published rating outright in 21–25% of base cases
and 30–41% of fine-tuned ones, Qwen highest. Panel B keeps that in proportion: for the base models
the text still carries about 60% of the pull, so the name matters without dominating. After
fine-tuning it stops dominating — 47% for Llama, 44% for Gemma, and 22% for Qwen, which follows
the injected name more than the document in front of it.

**Salience amplifies the name in every cell.** Solid sits right of open throughout Panel A and left
of open throughout Panel B: where the model could recover the country from the stripped text, the
name does more work. Appendix A11 reports the paired tests.

```{r}
#| label: fig-nameswap-channels
#| fig-cap: "Name-swap outcome by country salience, model family and type (2023). (A) Mean absolute change in the rating when the same de-identified evidence is framed with a different country's name instead of the true one, in rating points; shaded region = below the ±SESOI threshold of 0.113. (B) The text's share of the two channels from `swapped_rating ~ named_mean + source_mean`, demeaned within indicator; dashed 50% = the text and the injected name pull equally. Solid = the source country-indicator's de-identified evidence was correctly re-identified (high salience); open = it was not, with the partition fixed at the base model's result across both weight states so the rows divide identical items. Panel A uses both name-swap arms matched item for item; Panel B uses the swapped arm only. Country-clustered bootstrap 95% CIs on the source country, 2000 draws. Paired tests of the solid–open difference are in Appendix A11."
#| fig-width: 11
#| fig-height: 4.6
#| eval: !expr have_nameswap_channels
fig_nameswap_channels(nameswap_channels_2023, sesoi = boot_bundle_2023$sesoi)
```

## 8. Compression and de-identification

Sections 6 and 7 establish that the model conditions on country identity. This section asks what
the evidence-processing pipeline itself costs, by splitting the distance between raw source text
and the Summarized condition Figures 3–5 report into the two things that change along the way:

    raw text ──[ compression ]──▶ summarized-identified ──[ de-identification ]──▶ summarized

**Compression** (Summarized-Identified − Raw Text) keeps names present on both sides, so the only
thing that moves is the ~400-word rewrite. **De-identification** (Summarized −
Summarized-Identified) keeps the text compressed on both sides, so the only thing that moves is the
country's identity. Sign is later − earlier throughout, so positive MAE means further from the
panel mean. Base models only, 2023, on all three outcomes Figures 3–5 track, with shared
country-clustered draws per family.

The two steps add up to the whole raw-to-Summarized gap, but they are **successive steps, not
shares of a total**: the first is measured with identity present and the second with compression
present, so their magnitudes are conditioned differently. Moving the same two factors in the other
order (via Anonymized) reverses which one looks larger for Gemma. No percentage of either step
against the total is reported anywhere for that reason — see
`notes/identity-mechanism-arc-2026-09-16.md` for the demonstration and
`notes/mockups/identity-and-compression-concept.md` for three percentage presentations that were
built and rejected.

No ±SESOI band, following Figure 9's precedent. SESOI defends against small *random* error — a
change below half a rounding step vanishes in any single published score — whereas these are
consistent directional pushes, where the question is whether a step moves the model and in which
direction. The band also discriminates nothing here: at 0.113 it would swallow every cell except
Llama's de-identification MAE.

**MAE.** De-identification is the larger step for all three families (+0.108 / +0.076 / +0.068,
Llama / Qwen / Gemma), and compression is small: null for Llama (+0.001), +0.045 for Qwen, +0.022
for Gemma.

**Difficulty tracking slope.** De-identification again dominates and excludes zero everywhere
(−0.048 / −0.107 / −0.198), flattening the slope. Compression clears zero only for Llama (−0.042);
Qwen (−0.011) and Gemma (+0.000) are null.

**Signed deviation.** The pattern reverses. **Compression excludes zero for all three families**
(−0.101 / −0.117 / −0.044) — compressing the evidence while keeping every name makes each model
harsher — and it is the *larger* of the two steps for Qwen and Gemma, whose de-identification steps
are −0.044 and a null −0.010. Only Llama's de-identification step (−0.131) exceeds its compression
step.

```{r}
#| label: fig-identity-and-compression
#| fig-cap: "The two successive steps from raw source text to the Summarized condition, on all three outcomes (2023, base models). **Compression** = Summarized-Identified − Raw Text (names present on both sides, so only the rewrite moves); **De-identification** = Summarized − Summarized-Identified (compressed on both sides, so only identity moves). Sign is later − earlier, so positive MAE = further from the panel mean, negative signed deviation = harsher, negative slope = flatter difficulty tracking. Each panel has its own x-scale: the outcomes are in different units and are not comparable across panels. The two steps sum to the raw-to-Summarized gap but are successive steps, not shares of a total (see prose). Country-clustered bootstrap 95% CIs, shared draws per family. No ±SESOI band."
#| fig-width: 12
#| fig-height: 3.6
#| eval: !expr have_identityeffect && have_identityeffect_sd && have_identityeffect_slope
fig_identity_and_compression(identityeffect_bundle, identityeffect_slope_bundle,
                             identityeffect_sd_bundle)
```

## Synthesis: what explains Figures 3–5's de-identification pattern?

<!-- TODO (EJT): the mechanism sentence below ("identity itself carrying informative signal the
model loses when stripped") is contradicted by the task-framing result in Appendix A12: a
contentless label reproduces most of the effect, which makes it a retrieval cue rather than lost
signal. Pointers have been renumbered; the claim itself still needs rewriting. See
notes/narrative-arc-and-figures-2026-09-18.md, "Paper-side edits pending". -->

Two candidate explanations for the original puzzle — de-identifying evidence pushes MAE further
from the human panel, sometimes past it entirely — were tested directly. The mechanism
Appendix A14-A15 establishes, country-level staleness (and, more weakly, prominence), does **not** explain it:
crossing movement and re-identification against raw MAE by condition (rather than against the
difficulty-tracking slope or signed deviation) turns up no consistent relationship — movement
clears zero in only 5 of 18 cells with no pattern by condition, and re-identification in none.
Direct identity removal does explain it: holding text length fixed (Anonymized vs. Raw Text) or
compression fixed (Summarized vs. the new Summarized-Identified condition), removing the
country's name increases error for most model/condition pairs, most consistently at the
compressed level (all three families). See Figure 8. The effects are mostly modest — inside
the ±SESOI band except for Llama's full-text case — so this is real, not decisive, evidence: it
points at identity itself carrying informative signal the model loses when stripped, rather than
at staleness or prominence, but doesn't close the question. See
`notes/mockups/mechanism-tests-summary-2026-09-06.md` for the fuller writeup, including
alternative explanations (rewrite-artifact/distributional-shift from the summarization process,
nuance loss from compression) that remain untested.

## 9. Synthetic-coder deployment: does an AI seat change the panel's verdict?

Figures 3–8 establish performance and mechanism; this section asks the deployment question
underneath Figure 1. Leading-edge cells arrive below V-Dem's floor between a fifth and
two-fifths of the time and take several release cycles to fill, and the practical question is
whether filling one of those seats with a synthetic coder changes what the panel actually
*concludes*.

The outcome is a **verdict change**: a move in `round(panel mean)`, the panel's own rounded
judgement — the same rounding the SESOI derives from. It is **not** V-Dem's published `_ord`,
which comes from the measurement model, so nothing here licenses a claim about how many
published scores would change.

**Human churn is measured, not assumed.** Figures on the *mean* shift can zero-reference,
because refilling a seat from the panel's own coders draws from the distribution the mean was
computed from and so shifts it by zero in expectation. A verdict *change* from ordinary turnover
is emphatically not zero — it runs 16–42% below — so it is estimated and plotted as its own
line.

That benchmark needs spare real coders, which thin panels do not have. So thinness is
**constructed** rather than observed, holding the base panel size fixed (a base of *m−k* would
shrink as k grows and confound the dose-response with panel size). Source panels are the 7,775
cells with ≥8 coders (110 countries, the deepest quarter of 2023); each is randomly permuted,
the first 4 coders form the base panel whose verdict is the baseline, and the next 4 are held
out. This figure **adds** k seats, growing the panel 4 → 4+k; Panel A counts changes, Panel B
gives direction. Both arms start from the identical base panel, so the AI-vs-churn contrast is
within-panel and panel size cannot drive it. Three AI regimes as throughout: **Same AI** (every
seat Qwen FT-raw Evidence-ZS), **Mixed FT** (drawn from the 6-cell fine-tuned pool), **Mixed
pool** (drawn from the full 18-cell grid).

Adding is the mechanic the motivation calls for: at the leading edge the seats were never
filled rather than vacated, so a synthetic coder joins an incomplete panel instead of taking a
departed expert's place. The complementary **replacement** mechanic — swapping k of the base
panel's four seats, holding size at 4 — is the academic counterfactual, and is reported in
Appendix A16. Every result below holds there too, in the same order and at roughly twice the
level; the one finding specific to replacement is flagged where it arises.

**At one seat, a synthetic coder is indistinguishable from ordinary turnover.** Panel A:
14.1–17.4% against churn's 15.7%, with two of the three regimes *below* it. All four arms sit
within about 1.5 points with overlapping intervals.

**The regimes separate as seats accumulate.** At k=4, Same AI disturbs most (38.2%), Mixed FT
least (27.4%) and Mixed pool sits between (28.3%), against churn's 22.4%. The same ordering
holds under replacement (A16). That a *single* well-chosen model is the most disruptive regime, not the
least, is the counterintuitive part: identical ratings in every seat move the mean in concert,
while a mixed pool's disagreements partly cancel.

**Panel B is where the two mixed regimes come apart, and it is the reason direction is in the
main text rather than an appendix.** On the change rate alone, Mixed FT and Mixed pool are nearly
indistinguishable (27.4 vs 28.3 at k=4), which would invite the conclusion that an unfiltered
mix of models behaves like a curated fine-tuned one. On direction they have **opposite signs**.
Mixed pool excludes zero at every k and grows monotonically — +3.4 to +8.5 points — an
increasingly downward push relative to what human turnover would have done. That regime
contains the base cells whose signed deviation Figure 5 shows is strongly negative, so this is
Figure 5's finding arriving at the panel's published judgement. Mixed FT drifts mildly the
other way, −1.4 to −3.3 points, also excluding zero at every k.

**Same AI sits on the churn line** (+0.3 to +1.1, never clearing zero at any k) — the regime
this paper's own results point to, contributing no directional push. That is a property of
*adding* seats rather than of the model: when seats are replaced instead, Same AI does clear
zero at k=3 and k=4, because every Same AI seat carries an identical rating and replacing three
or four of four collapses the panel onto a single deterministic judgement whose offset is then
expressed in full instead of being diluted by the remaining humans (Appendix A16). Putting one
model in many seats is the one deployment configuration these results caution against — and it
is also why Same AI changes the most verdicts here: identical seats move the mean in concert,
while a mixed pool's internal disagreements partly cancel.

Panel B is differenced against the churn arm, paired at the panel level, and that differencing
is load-bearing rather than cosmetic. With a base panel of 4 the mean is a
multiple of 0.25, so exact half-integers are common and `floor(x + 0.5)` rounds them up: a base
sitting at 2.5 has verdict 3, so any added rating below the mean drops it while additions above
merely confirm — a built-in downward push that then alternates with k, because k=1,3 give panel
sizes 5 and 7 where a new mean can never land on .5, while k=2,4 give 6 and 8 where ties are
possible. **Human churn shows the same zigzag** (10.9, 3.3, 10.8, 5.0 raw), which is the tell
that it is arithmetic and not behavior; both arms face identical base panels and identical
parity, so differencing cancels it.

One caveat spans both mechanics: the source panels are, by construction, the deepest quarter
of 2023 rather than the thin panels in the field. This is a controlled experiment on constructed
thinness — the price of measuring human churn rather than assuming it. The observational
counterpart (real thin panels, resampled churn) is described in
`notes/mockups/flip-rate-concept.md`, along with why its benchmark understates churn.

```{r}
#| label: fig-fliprate
#| fig-cap: "Does an added synthetic seat change the panel's verdict, and in which direction? k seats are added to a base panel of four, growing it 4 → 4+k. (A) Share of panels whose `round(panel mean)` changes, against a human-churn benchmark built from real held-out coders (dashed grey). (B) Net downward minus upward changes, in points, differenced against that churn arm and paired at the panel level, so the dashed zero line *is* churn's own directional push; positive = the AI arm pushes verdicts down more than human turnover does. Base panel of 4 constructed by random permutation of source panels with ≥8 coders (7,775 panels, 110 countries, 2023); both arms start from the identical base panel, so panel size cannot drive the contrast. Three AI regimes: Same AI = Qwen FT-raw Evidence-ZS; Mixed FT = the 6-cell fine-tuned pool; Mixed pool = the full 18-cell grid. Regimes offset slightly on k so the intervals do not overlap. Country-clustered bootstrap 95% CIs, 30 Monte Carlo permutations. Appendix A16 reports the replacement mechanic on its own scale; replacement disturbs roughly twice as many verdicts at every k, so the two are compared by reading the axes rather than by shape."
#| fig-width: 11.5
#| fig-height: 4.4
#| eval: !expr have_fliprate
fig_fliprate_quad(fliprate_augmentation_2023, fliprate_degradation_2023, mechanic = "add")
```

# Appendix

## A1. Panel size by country and regime

The cross-sectional and temporal cuts behind Figure 1. Both panels measure the same
quantity — coders per country-indicator-year — so they share a y-axis; only the thing being
varied differs. Panel A varies time and regime type, Panel B varies country.

Panel A's dotted rules mark features of V-Dem's coding schedule rather than of its
workforce. 2005 is where contemporary recruits stop back-coding: a newly recruited expert
codes their country from 2005 to the present, so every later cohort adds to 2005 and after
but not before it. 2013 is where the original coding round ended. As Figure 1B establishes,
the decline to the right of 2013 is years that have not yet accumulated their back-coders,
not a shrinking workforce.

No country's *mean* falls below five — the United Arab Emirates sits lowest at 5.01, on the
line — so sub-floor cells are within-country variation rather than a set of permanently
deficient countries. The colouring shows the two are related: thin countries carry a much
larger share of sub-floor cells.

```{r}
#| label: fig-panelsize-cuts
#| fig-cap: "Panel size by time, regime type and country (205 scored indicators). (A) Mean coders per country-indicator-year by year and V-Dem's Regimes of the World, 1990-2024. Dotted rules mark coding-schedule features, not workforce events: 2005 is where contemporary recruits stop back-coding, 2013 where the original coding round ended. The post-2013 decline is back-coding that has not arrived yet (Figure 1B), not attrition. (B) Mean panel size by country, 2005-2024, ranked, with the 6 thinnest and 6 thickest labelled; colour is the share of that country's cells falling below five, which is related to but not redundant with the mean (r = -0.54). Dashed rule = V-Dem's recommended floor of five. Panels share a y-axis."
#| fig-width: 13
#| fig-height: 5.6
#| eval: !expr have_panelsize
fig_panelsize_cuts(panelsize)
```

## A2. Panel size by survey, and what accounts for the variation

V-Dem recruits at the **survey** level, not the indicator level. C-indicators are organised
into 16 topical surveys grouped into five clusters, and the methodology notes that experts
"code at least one cluster" and "most code one to two clusters" (methodology p. 16). So
indicator-level variation is largely survey-level variation, and the survey is the unit that
corresponds to an actual recruitment decision.

The two thinnest surveys are also the two whose series begin after 1789 — Digital Society
(from 2000) and Civic & Academic Space (from 1900), marked in vermillion. A survey
introduced later has had fewer recruitment rounds to accumulate coders, so its youth and its
thinness are not separable from this figure. Digital Society is also the largest survey by
indicator count, at 32.

Two things Panel B does **not** license, both of which follow from what a one-way R² is.
The single-predictor bars are not orthogonal shares: they sum to 0.764 against a cumulative
fit of 0.648, because country and year overlap. And R² rises mechanically with the number of
levels a grouping has — country 179, indicator 205, year 20, survey 14 — so country's lead
is partly a degrees-of-freedom artifact. Per level, year is by far the strongest predictor,
which is Figure 1B's finding in another form.

```{r}
#| label: fig-panelsize-sources
#| fig-cap: "Where contemporary variation in panel size comes from (2005-2024, 205 scored indicators). (A) Mean panel size by survey, the unit at which V-Dem recruits; indicator counts in parentheses, dashed rule = the recommended floor of five. Vermillion marks surveys whose series begin after 1789, which have had fewer recruitment rounds, so youth and thinness are confounded. (B) One-way R² from `n_coders ~ 1 | <grouping>` for each of four groupings, plus two nested fits (grey). These are not orthogonal shares — the four single-predictor bars sum past the cumulative fit because country and year overlap — and R² rises mechanically with the number of levels (country 179, indicator 205, year 20, survey 14), so the ranking is partly a degrees-of-freedom artifact; per level, year is the strongest."
#| fig-width: 11
#| fig-height: 5
#| eval: !expr have_panelsize
fig_panelsize_sources(panelsize)
```

## A3. Synthetic-coder performance — 2023 holdout

Replication of Figure 3 on the 2023 holdout year. Same design and readout convention, same
Codebook + Raw Text trim on the FT-raw block as Figure 3 (the untrimmed grid is Appendix A7).
All four base conditions are in for all three families, as are all four Llama FT-raw
conditions, at full row counts.

```{r}
#| label: fig-crossmodel-2023
#| fig-cap: "Synthetic coder performance across model families on the 2023 holdout (Llama 70B / Qwen 72B / Gemma 27B). Layout, readout convention (solid = greedy, open = mean; base rows greedy only), and reference lines match Figure 3. FT-raw block: Codebook and Raw Text only (Appendix A7 has the untrimmed grid). Country-clustered 95% CIs."
#| fig-width: 8
#| fig-height: 5.2
#| eval: !expr have_2023
fig_crossmodel_landscape(exp_bundle_2023, boot_bundle_2023, band = "human_loo", base_readout = "greedy",
                         ft_conds = c("codebook", "evidence-zeroshot"))
```

## A4. Difficulty tracking — 2023 holdout

Replication of Figure 4 on the 2023 holdout year. Same two-panel design, featured cells, reference
(slope 1), greedy readout, and Codebook + Raw Text trim on the FT-raw block as Figure 4. Full
model/condition coverage, as in A3.

```{r}
#| label: fig-crossmodel-slope-2023
#| fig-cap: "Difficulty tracking on the 2023 holdout — replication of Figure 4. (A) AI error vs case difficulty for the same featured cells, every cell a faint grey trace; (B) difficulty-tracking slope per cell, long-dash = human reference (slope 1). FT-raw block: Codebook and Raw Text only (Appendix A7 has the untrimmed grid). Greedy readout, country-clustered 95% CIs."
#| fig-width: 12
#| fig-height: 5.4
#| eval: !expr have_2023_slope
fig_crossmodel_slope_2panel(dm_slope_2023, ft_conds = c("codebook", "evidence-zeroshot"))
```

## A5. Signed Deviation

<!-- STOPGAP PROSE (EJT to rewrite): -->
Two companions to Figure 5. First, the same measure on the **democracy-quintile** cut for 2019 —
the continuous version of the regime-type gradient in the main text. Second, the **regime-type**
cut replicated on the **2023 holdout**. Panel B (average signed deviation) is identical across
cuts by construction, so it repeats here; only Panel A (the gradient) changes.

### Democracy quintile — 2019

```{r}
#| label: fig-crossmodel-signeddev
#| fig-cap: "Directional bias by democracy quintile, 2019 (greedy readout) — the continuous companion to Figure 5's regime-type cut. (A) Signed deviation (AI − panel mean) across democracy quintiles for three exemplar cells — best-calibrated, steepest tilt, harshest average deviation — with every cell shown as a faint grey trace; dotted 0 = human panel-member benchmark, dashed = V-Dem IRT `_ord` target, grey band = ±SESOI, country-clustered 95% CI whiskers. (B) Average signed deviation per cell (same as Figure 5B). FT-raw block: Codebook and Raw Text only (Appendix A7 has the untrimmed grid)."
#| fig-width: 12
#| fig-height: 5.4
fig_crossmodel_signeddev_2panel(sd_signed, ft_conds = c("codebook", "evidence-zeroshot"))
```

### Regime type — 2023 holdout

```{r}
#| label: fig-crossmodel-signeddev-2023-regime
#| fig-cap: "Directional bias by regime type on the 2023 holdout — replication of Figure 5. (A) Signed deviation across V-Dem's Regimes of the World for three exemplar cells (best-calibrated, steepest tilt, harshest average deviation), every cell a faint grey trace; dotted 0 = panel-member benchmark, dashed = V-Dem IRT. (B) Average signed deviation per cell. FT-raw block: Codebook and Raw Text only. Greedy readout, country-clustered 95% CIs, grey band = ±SESOI."
#| fig-width: 12
#| fig-height: 5.4
#| eval: !expr have_2023_signed
fig_crossmodel_signeddev_2panel_regime(sd_signed_2023, ft_conds = c("codebook", "evidence-zeroshot"))
```

## A6. Synthetic-coder performance — 2024 FH-only holdout

Loose out-of-sample replication of Figure 3 on the most recent year — data furthest from the
fine-tuning window (through 2018) and not part of any earlier run. Same design and readout
convention, same Codebook + Raw Text trim on the FT-raw block as Figure 3 (no untrimmed 2024
companion is planned — Appendix A7 covers that check on 2019 only). Base ladder (codebook/
evidence/anonymized/summarized) × 3 families (Llama 70B / Qwen 72B / Gemma 27B) = 12 cells;
FT-raw diagonal restricted to Codebook + Raw Text × 3 families = 6 cells. FH-only: State
Department country reports changed format enough for 2024 that they aren't used, so every cell
here rests on Freedom House evidence alone (see `notes/plan-2024-holdout.md`, private notes
repo, for the staging/run plan). Within-2024 comparisons are valid (base vs. FT, condition
ordering, AI vs. human reference); cross-year levels are confounded (FH-only vs. full-source
drops one source), so this is not placed on a shared axis with 2019/2023.

```{r}
#| label: fig-crossmodel-2024
#| fig-cap: "Synthetic coder performance across model families on the 2024 FH-only holdout (Llama 70B / Qwen 72B / Gemma 27B). Layout, readout convention (solid = greedy, open = mean; base rows greedy only), and reference lines match Figure 3. FT-raw block: Codebook and Raw Text only (Appendix A7 has the untrimmed grid on 2019). Country-clustered 95% CIs."
#| fig-width: 8
#| fig-height: 5.2
#| eval: !expr have_2024
fig_crossmodel_landscape(exp_bundle_2024, boot_bundle_2024, band = "human_loo", base_readout = "greedy",
                         ft_conds = c("codebook", "evidence-zeroshot"))
```

## A7. Untrimmed FT-raw evidence-format grid (2019, cross-model)

Figures 3–5's FT-raw block shows Codebook and Raw Text only — the one apples-to-apples comparison
(does reading the matched-training-format evidence help beyond fine-tuning). This section restores
the two conditions cut from there — Anonymized and Summarized — for all three model families, i.e.
the FT-raw adapter tested on input formats it was never fine-tuned on. As established in A8 (Llama
only, but the same logic applies here): this is a generalization/robustness check, not a second
"does evidence help" test, and it should not be read as evidence for or against the matched-format
finding above.

```{r}
#| label: fig-crossmodel-full
#| fig-cap: "Untrimmed companion to Figure 3: FT-raw block restored to all four inputs, cross-model. Same design, readout convention, and reference lines as Figure 3."
#| fig-width: 8
#| fig-height: 5.2
fig_crossmodel_landscape(exp_bundle, boot_bundle, band = "human_loo", base_readout = "greedy")
```

## A8. Full model × input grid

Llama only: Base plus all three Llama fine-tuned adapters (FT-raw, FT-anon, FT-summ), each run on
all four inputs — Qwen and Gemma trained only the raw-text adapter, so there is no matched-vs-
mismatched comparison to make for them here. The ringed cell in each FT block is that adapter's own
matched training input (e.g., FT-anon on Anonymized); it is not consistently the best-performing
cell for that model, but the ranking among matched/mismatched conditions is not consistent across
model families either (Llama's matched condition loses to Codebook; Qwen's and Gemma's FT-raw
matched condition — Raw Text — is their best of the four), and under the greedy readout every
cell in every block falls within ±SESOI of the human reference regardless of condition (under the
mean readout no cell rates worse than human beyond SESOI, and most rate better). Treated here as
noise rather than a format effect.

```{r}
#| label: fig-grid
#| fig-cap: "Every model on every input (the off-diagonal expansion of Figure 3), Llama only. Greedy (mode) vs expectation (mean) MAE with country-clustered 95% CIs. Rings mark each fine-tuned model's own training input (the diagonal) — not consistently its best-performing cell; under the greedy readout every cell falls within ±SESOI of the human reference, and no cell under either readout rates worse than human beyond SESOI. Dashed line = naive persistence model; long-dash line = human reference; grey band = ±SESOI around the human reference."
#| fig-width: 8
#| fig-height: 9
fig_readout_grid(exp_bundle, boot_bundle, band = "human_loo")
```


## A9. Would a sampled rating do better than the greedy mode?

Every rating in this paper is the **greedy mode** of the model's distribution over rating
categories — temperature 0. The reasoning is in `preregistration/finetune-training-target.md`: a
human coder emits one integer, the human reference compares that integer to the mean of the
others, and the mode is the principled point estimate of a distribution. That note also records
the cost, which this section prices. Greedy extraction makes the synthetic coder the *modal*
panel member rather than a typical draw, so it carries none of the case-level variance a real
coder's rating carries, and it can sit near the human reference partly for that reason.

**No new inference is involved.** `code_country_year.py` stores `rating_dist` on every row — the
model's full probability distribution over that indicator's categories, which is the
temperature-1 softmax. Temperature scaling is a closed-form reweighting of a stored distribution,
`q_T(y) ∝ p(y)^(1/T)`, so the *expected* behaviour of a sampled coder at any temperature is
computable exactly: expected MAE is `Σ q(y)·|y − m|`, expected signed deviation is
`Σ q(y)·y − m`. The difficulty slope is exact rather than approximate because OLS is linear in
the outcome, so regressing expected error on difficulty *is* the expected sampled slope. At
T = 0 the panel reproduces `distmatch_slope_2023` to four decimals for Llama and Qwen (Gemma runs
0.01 high, from the few rows whose distribution was not captured).

**MAE and difficulty tracking both favour T = 0, monotonically and in all six cells.** Rising
temperature moves every cell away from the human reference on MAE, and past it by T = 1; the
difficulty slope falls away from the human reference of 1 throughout. The slope result runs
against the natural expectation that sampling would *steepen* difficulty tracking, since the
fine-tuned models' entropy tracks human disagreement by location. The mechanism appears to be a
floor effect: sampling adds roughly a fixed quantum of error to every case, which is most of the
error on an easy case and a small addition on a hard one, compressing the spread of errors and
flattening the regression.

**Signed deviation does not follow the other two.** It rises to a small hump around T ≈ 0.1–0.2,
then falls, ending at T = 1 slightly closer to zero than T = 0 in the Codebook cells
(Qwen 0.171 → 0.074, Gemma 0.153 → 0.067) and roughly unchanged on Raw evidence. So the Codebook
overshoot into generosity that Figure 5 reports is in part a mode artifact rather than a property
of the fitted distribution. The hump is a skew effect — at low temperature the mass that first
appears sits above the mode — and is not interpreted further here.

Read together: temperature 0 is the best available setting on the two criteria the paper leads
on, and the one place sampling helps is the condition where no evidence is supplied at all. What
the ladder cannot settle is the panel-level question — every quantity here measures a single
rater against the panel mean, not a constructed panel against a measurement model — which is
where a synthetic panel of several distinct raters would have to be evaluated instead. See
`notes/mockups/temperature-ladder-concept.md`.

```{r}
#| label: fig-temperature-ladder
#| fig-cap: "Decoding temperature against all three outcomes (2023, fine-tuned raw-text adapters). T = 0 is the greedy mode used throughout the paper, marked with a solid dot; T > 0 is the expected behaviour of a sampled rating, computed in closed form from the stored rating distribution rather than simulated, so no new inference is involved. Rows are the two reported conditions, columns the three outcomes, each on its own y-scale. Dashed rule = the human reference in each panel (MAE 0.725, difficulty slope 1.00, signed deviation 0.00). Country-clustered bootstrap 95% ribbons, 200 draws. MAE and difficulty tracking move away from the human reference monotonically in all six cells; signed deviation does not follow, ending marginally closer to zero at T = 1 in the Codebook cells."
#| fig-width: 11
#| fig-height: 6.2
#| eval: !expr have_temperature_ladder
fig_temperature_ladder(temperature_ladder_2023)
```

## A10. Re-identification — top-3 rate

Companion to Figure 6 (Section 6): the same model × weights × condition grid, but crediting the
model whenever the true country appears anywhere in its top-3 guesses rather than only at top-1.
Same layout, same "no chance-line drawn" treatment, same bar ordering rule (descending Base ×
Anonymized rate) applied independently to this metric.

```{r}
#| label: fig-reid-rates-top3
#| fig-cap: "Top-3 re-identification rate by model family, weights, and de-identification condition (2023) — appendix companion to Figure 6. Plain bars, no CI."
#| fig-width: 9
#| fig-height: 5
#| eval: !expr have_reidrates
fig_reid_rates(reidrates_bundle_2023, metric = "top3")
```

## A11. Name-swap by salience — paired tests

Figure 7 splits both panels by whether the model could recover the country from that
country-indicator's de-identified evidence. This is the test of the gap the figure shows. The
difference is computed **within each bootstrap draw**, both strata carrying the same country
weights, so the shared-country covariance is preserved — a country contributes some of its
indicators to each side, and comparing the two marginal intervals would treat them as independent
samples. Sign convention: re-identified minus not, so positive means the name did more work where
the evidence was identifiable. The text's share is negated and relabelled "Name's share" so that
right-of-zero means the same thing in both panels; raw, it moves the other way by construction,
since more weight on the name is less on the text.

```{r}
#| label: fig-nameswap-salience-tests
#| fig-cap: "Paired salience differences behind Figure 7's solid/open split, re-identified minus not (2023). Each panel has its own x-scale; the two quantities are in different units. Right of zero means the injected name did more work where the evidence was identifiable — the text's share is negated and relabelled \"Name's share\" so the direction is common across panels. Both strata carry the same country weights within each bootstrap draw, so the shared-country covariance is preserved rather than the two being treated as independent samples. Country-clustered bootstrap 95% CIs on the source country, 2000 draws."
#| fig-width: 9
#| fig-height: 4.4
#| eval: !expr have_nameswap_channels
fig_nameswap_salience_tests(nameswap_channels_2023)
```

Mean shift clears zero in five of six cells, all positive; only Gemma's base model does not
(+0.021, p = 0.219). The name's share is significant in the two fine-tuned cells where the name
pull is largest, Qwen (+0.102) and Gemma (+0.112), and in neither of the base cells.

Two caveats on reading these. Twelve tests are reported, and Qwen's base model sits on the line on
mean shift (p = 0.045); nothing else is near it. And re-identification is
not assigned, it is a property of the text and of the model reading it, so a difference here
establishes that the two groups behave differently rather than that identifiability causes the
difference. A permutation that shuffles re-identification within country leaves little residual,
which is what a country-level prior looks like rather than item-level detection of the mismatch;
see `notes/mockups/nameswap-reid-split-concept.md`.

## A12. Task framing or evidence body?

Figure 8 treats de-identification as one manipulation, which is how it is registered
(`preregistration-draft.md:75` defines the treatment as stripping the country's name, named
entities such as officials and institutions, **and datable events**). But the manipulation reaches
the model through two channels at once: the country's name appears in the **task framing** (the
`{FOCAL_COUNTRY}` / `{FOCAL_YEAR}` slots of the user message) *and* throughout the **evidence
body**. This section reports what happens when those channels are varied separately. It is
included as something ruled out, not as a finding: **the two channels do not separate cleanly
enough to build a claim on**, for reasons set out at the end.

### The three summarized arms, measured from raw evidence

Three arms already exist for 2023 base models, all ~400-word rewrites, differing only in where the
country's identity survives:

| arm | evidence body | task framing |
|---|---|---|
| Summ-Identified | names kept | country named |
| Summ + named framing | names stripped | country named |
| Summarized | names stripped | blanked |

The middle arm is the name-swap battery's control condition: name-swap mode forces `hide_identity`
off, so it is the only run in the study carrying de-identified text alongside a real country name
in the framing. Each arm is a paired contrast against **raw evidence**, so all three are distances
from the same origin — no contrast depends on an ordering, and none is a share of a total. That is
deliberate: the earlier nested-tree presentation of these same numbers labelled each node as a
share of a grand total, which reads as attribution when only recovery is licensed (retired; see
`notes/mockups/decomposition-tree-build.R`).

**On MAE and the difficulty slope the arms are ordered**, with the fully de-identified arm furthest
from raw evidence (MAE +0.110 / +0.121 / +0.091; slope −0.090 / −0.116 / −0.197). **On signed
deviation the first two arms coincide** for Llama (−0.102 vs. −0.100) and Qwen (−0.117 vs. −0.120),
while the fully de-identified arm sits well out (−0.233, −0.161) — restoring names to the evidence
body adds essentially nothing once the framing already supplies the country. **This does not hold
for Gemma**, whose three arms sit within 0.012 of each other (−0.044, −0.052, −0.056), so nothing
separates there at all. It is not a three-family result and should not be written as one.

```{r}
#| label: fig-summarization-arms
#| fig-cap: "The three summarized arms as distances from one origin (raw evidence), 2023 base models. Summ-Identified = names kept in the evidence body and the task framing; Summ + named framing = names stripped from the body, country named in the framing; Summarized = names stripped from both. Every arm is a paired contrast against raw evidence on shared country-clustered draws, so each is a distance from the same origin and none is a share of a total. Sign is arm − raw evidence: positive MAE = further from the panel mean, negative signed deviation = harsher, negative slope = flatter difficulty tracking. Each panel has its own x-scale. Country-clustered bootstrap 95% CIs. No ±SESOI band."
#| fig-width: 12
#| fig-height: 4.1
#| eval: !expr have_summarization_arms
fig_summarization_arms(summarization_arms_2023)
```

### Is the task-framing leg just residual leakage?

An immediate objection to the arms above: Figure 6 shows the base model still names the country
24–29% of the time from summarized text, so perhaps naming it in the framing only matters for the
cases where identity was genuinely gone, and the aggregate effect is a partial-leakage artifact.
That has a testable implication. The **task-framing leg** (Summ + named framing − Summarized) holds
the de-identified evidence text identical on both sides, with only `{FOCAL_COUNTRY}` /
`{FOCAL_YEAR}` moving; splitting it by whether the base model's own re-identification probe named
the country correctly gives two competing predictions. *Partial leakage* predicts a **larger** leg
among cases the model could **not** name. *Framing-only attention* — the model uses the
task-specification slot and does not spontaneously condition on identity it could have extracted
from the evidence body — predicts the leg is **flat** across the two groups.

**It is flat, and where it moves it moves the wrong way for partial leakage.** Every MAE and slope
moderation is null. On signed deviation the moderation excludes zero for Llama (+0.041) and Qwen
(+0.039), both in the direction of a *larger* effect among cases the model **could** name; Gemma is
null throughout. Partial leakage is not what produces the task-framing leg.

```{r}
#| label: fig-reid-split
#| fig-cap: "The task-framing leg (Summ + named framing − Summarized), split by whether the base model correctly named the country on the summarized re-identification probe (2023, base models). The same de-identified evidence text appears on both sides of every contrast; only {FOCAL_COUNTRY} / {FOCAL_YEAR} moves. Partial leakage predicts a larger effect among cases the model could NOT name; framing-only attention predicts a flat split. Color encodes the re-identification group, not the model (model is on the facet rows). Country-clustered bootstrap 95% CIs, shared draws, so each subgroup contrast and their difference are computed within draw."
#| fig-width: 11.5
#| fig-height: 6.2
#| eval: !expr have_reid_split
fig_reid_split(reid_split_2023)
```

### Why this is reported as inconclusive

Three reasons the channels cannot be cleanly apportioned from these runs, all of which would have
to be addressed before any "the framing carries most of it" claim could be made.

1. **The fourth cell does not exist.** The arms above vary the evidence body *given* that the
   framing already names the country. The complementary cell — identity present in the evidence
   body, framing blanked — has never been run, so the 2×2 does not close and the two channels
   cannot be shown to partition the effect. Estimated at roughly 30 GPU-hours; it reuses the
   existing `summarized-identified` text cache and needs only `hide_identity` split into two
   booleans in `pipeline/assemble_prompt.py`.
2. **Any two-part split is path-dependent.** Whichever factor moves first absorbs the first-taste
   effect. The alternative ordering of the same two factors exists in the data (via Anonymized) and
   disagrees, with three sign flips and Gemma's story inverting entirely.
3. **The comparison is position-confounded.** `{FOCAL_COUNTRY}` occupies the opening and closing
   of the user message while evidence-body names sit in the middle. Liu et al. (2024) predict a
   U-shaped context-usage asymmetry of exactly this shape from serial position alone, so a framing
   advantage cannot be attributed to the task-specification role of the slot rather than to where
   it sits.

## A13. Movement

The other ingredient the prominence-crossing figures in A14-A15 cross against identifiability:
how much has
a country's democracy level actually changed since roughly when the model's own "impression" of
it would have formed? **Change in V-Dem's Polyarchy score, 2018 to 2023**, one point per country,
ranked. Most countries barely moved; a handful moved a lot, in both directions — exactly the
long-tailed shape that made a plain (non-rank-transformed) version of this measure too
noisy to use directly in a bootstrap (see `notes/mockups/prominence-crossing-fig-concept.md`).
Labels mark exactly the 5 most-backslid and 5 most-improved countries — a fixed rule, not a
visual judgment call — and four of the five steepest declines here are well-documented coups or
crises in this window (Burkina Faso, Myanmar, Mali, Tunisia); the fifth, Mauritius, is a slower
measured democratic erosion under the Jugnauth government. Either way the ranking is a useful
sanity check on the measure itself before it's used as a mechanism-test input.

```{r}
#| label: fig-movement
#| fig-cap: "Change in V-Dem's Polyarchy score (v2x_polyarchy), 2018-2023, one point per country (n=179), ranked. Labels: the 5 most-backslid and 5 most-improved countries."
#| fig-width: 10
#| fig-height: 6.5
#| eval: !expr have_movement
fig_movement(movement_bundle_2023, country_names)
```

## A14. Difficulty tracking under prominence and movement

Does the "tracks difficulty like a human coder would" behavior from Figure 4 hold up equally well
for every country, or does it depend on how identifiable and how recently-changed the country is?
Two separate models (no interaction term — dropped 2026-09-06 in favor of this simpler, more
legible design), each run across all four Base conditions and the trimmed Fine-Tuned pair
(Codebook + Raw Text, matching the FT-raw trim used throughout Figures 3-5):
`AI error ~ h + movement + h:movement` (Panel A) and `AI error ~ h + reid + h:reid` (Panel B),
where `h` is case difficulty (human LOO error). The reported coefficient is the slope-modifying
term in each — does movement/prominence change how well AI error tracks difficulty, not a level
effect (see `notes/mockups/prominence-crossing-fig-concept.md`). Panel A's coefficient is on a
percentile-rank scale (least- to most-moved country, not a real democracy-score unit — see the
same note for why); Panel B's is a literal identified-vs-not comparison, no transform. No SESOI
band is shown; these are slope-modifier units, not MAE, so the paper's rating-point SESOI
threshold doesn't apply here.

**Movement excludes zero in every one of the 18 cells (Panel A), always in the same direction,
including Codebook** — where the model sees no case-specific evidence at all. That last part is
the key to reading it: with no evidence, a rating can only reflect whatever generic impression the
model carries from pretraining. For a stable country that impression is still roughly right, so
whatever error remains still tracks genuine case difficulty; for a country that's recently changed
a lot, the impression is simply wrong, and wrong by about the same amount for every indicator in
that country, which washes out the correlation with difficulty. This holds up whether or not real
evidence is given, and after fine-tuning — a stale prior exerts a baseline pull that in-context
evidence doesn't override. **Prominence (Panel B) is far less consistent** — strong for Gemma at
Base, essentially gone after fine-tuning; the reverse for Qwen (null at Base, consistent once
fine-tuned) — a pattern worth flagging honestly rather than explaining with confidence.

```{r}
#| label: fig-slope-by-condition
#| fig-cap: "Effect of movement (Panel A, percentile rank of |Δv2x_polyarchy|, 2018-2023) and prominence (Panel B, exact-match re-identification, Summarized text) on the difficulty-tracking slope, across all four base conditions and the trimmed Fine-Tuned pair (Codebook, Raw Text). Country-clustered bootstrap 95% CIs; no SESOI band (slope-modifier units, not MAE)."
#| fig-width: 12
#| fig-height: 6
#| eval: !expr have_sbc_movement && have_sbc_reid
fig_slope_by_condition_2panel(sbc_movement, sbc_reid)
```

## A15. Signed deviation under prominence and movement

The same question as A14, but for directional bias (Figure 5) instead of difficulty
tracking: does a country that recently backslid still get a rating that's too generous — still
reflecting its pre-2018 level — and does one that improved get rated too harshly? Two separate
models, same shape as A14 but without the case-difficulty term (signed deviation isn't
itself regressed against difficulty): `signed deviation ~ movement_signed` (Panel A) and
`signed deviation ~ reid` (Panel B), across all four Base conditions and the trimmed Fine-Tuned
pair. Movement uses the *signed*, rank-tamed version — direction is the point here, unlike
A14's magnitude-only measure (see `notes/mockups/prominence-crossing-fig-concept.md`).

**Movement's effect concentrates almost entirely in Codebook** (5 of 6 cells exclude zero, all
in the anchoring direction) and mostly disappears once any real evidence is given. That's a
clean, theory-consistent result: with no case-specific evidence, a rating can only reflect the
model's baked-in impression of the country, so the lag/anchoring signature should be strongest
exactly there — and real evidence, even de-identified or compressed, is enough to mostly correct
the *direction* of the guess, even though A14 shows it doesn't fix the model's ability to
track case-specific *difficulty*. Prominence (Panel B) is scattered and inconsistent — a few
significant cells with no pattern by family or condition, nothing at all once fine-tuned — the
same weak-signal story prominence told in A14.

```{r}
#| label: fig-signeddev-by-condition
#| fig-cap: "Effect of movement (Panel A, signed percentile rank of Δv2x_polyarchy, 2018-2023) and prominence (Panel B, exact-match re-identification, Summarized text) on signed deviation (AI rating - panel mean), across all four base conditions and the trimmed Fine-Tuned pair (Codebook, Raw Text). Country-clustered bootstrap 95% CIs, grey band = ±SESOI."
#| fig-width: 12
#| fig-height: 6
#| eval: !expr have_sdbc_movement && have_sdbc_reid && have_2023
fig_signeddev_by_condition_2panel(sdbc_movement, sdbc_reid, sesoi = boot_bundle_2023$sesoi)
```
## A16. Deployment by replacement: swapping seats rather than adding them

Figure 9's companion mechanic. Instead of adding k seats to a base panel of four, this
**replaces** k of that panel's four seats, holding panel size fixed at 4. Same construction,
same source panels, same three AI regimes, same churn benchmark built from real held-out
coders.

It sits in the appendix because it is the counterfactual rather than the deployment path.
Figure 1 establishes that leading-edge cells arrive below V-Dem's floor and fill in over
subsequent releases — those seats were never occupied rather than vacated — so augmentation is
the mechanic that matches the problem the paper poses. Replacement asks the narrower question
of how well a synthetic coder stands in for an expert already in the seat. It is also the
procedure the synthetic-panel follow-on is built on (`future-papers/02-synthetic-panel.md`),
where correlated error across AI seats is the central risk.

**The findings track Figure 9 in the same order, at roughly twice the level.** At one seat the
three regimes are indistinguishable from churn (19.9–21.3% against 20.9%). At k=4 Same AI
disturbs most (52.5%), Mixed FT least (46.6%) and Mixed pool sits between (48.1%), against
churn's 41.5%. On direction, Mixed pool excludes zero at every k and grows monotonically, +3.9
to +14.9 points; Mixed FT drifts the other way, −1.4 to −4.4 points, also excluding zero at
every k.

**Replacement disturbs far more than augmentation at every k** — churn alone reaches 41.5%
against 22.4% — because a replaced seat removes a human judgement as well as adding a new one.
This figure and Figure 9 are each drawn on their own scale, so the gap is in the axis numbers
rather than in the shape of the curves.

**The one finding with no counterpart in Figure 9 is Same AI's directional push.** Adding Same
AI seats never clears zero at any k; replacing them does, at k=3 (+5.6) and k=4 (+8.1). The
reason is specific to the regime rather than to the model: every Same AI seat carries an
identical rating, so replacing three or four of four collapses the panel onto a single
deterministic judgement, and whatever small offset that judgement carries is then expressed in
full instead of being diluted by the remaining humans. That is a true property of putting one
model in many seats, not a bias particular to Qwen FT.

Panel B's differencing also works differently here. Under replacement the denominator never
changes, so the rounding-parity alternation that makes Figure 9's differencing load-bearing
cannot arise; raw net-down for churn comes out flat at −0.1 to −0.3 points, which makes this
panel a genuine benchmark rather than an artifact-canceller.

```{r}
#| label: fig-fliprate-replace
#| fig-cap: "Does a *replaced* synthetic seat change the panel's verdict, and in which direction? The companion mechanic to Figure 9: k of the base panel's four seats are swapped for synthetic ones, holding panel size at 4. (A) Share of panels whose `round(panel mean)` changes, against the human-churn benchmark (dashed grey). (B) Net downward minus upward changes, in points, differenced against that churn arm and paired at the panel level; positive = the AI arm pushes verdicts down more than human turnover does. Because panel size never changes here, the rounding-parity alternation that makes Figure 9B's differencing load-bearing does not arise. Same construction, source panels, regimes and CIs as Figure 9. Each figure is drawn on its own scale, so note the axis: replacement disturbs roughly twice as much at every k, churn alone reaching 41.5% against 22.4%."
#| fig-width: 11.5
#| fig-height: 4.4
#| eval: !expr have_fliprate
fig_fliprate_quad(fliprate_augmentation_2023, fliprate_degradation_2023, mechanic = "rep")
```
