---
title: "Paper figures & tables — draft gallery"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

**Reminder**: Yihui has released a fig.note option for `knitr`. https://github.com/yihui/knitr/pull/2453. Install development version and try it when you build the PDF.

```{r setup}
library(tidyverse)
library(gt)
library(glue)

proj_root <- rprojroot::find_root(rprojroot::is_git_root)
source(file.path(proj_root, "helpers", "bootstrap_helpers.R"))
source(file.path(proj_root, "helpers", "paper_figures.R"))
derived <- file.path(proj_root, "data", "derived")

boot_bundle <- readRDS(file.path(derived, "bootstrap_2019.rds"))
exp_bundle  <- readRDS(file.path(derived, "expectation_2019.rds"))
dm_slope    <- readRDS(file.path(derived, "distmatch_slope_2019.rds"))  # Fig 2 (Test-3 slope)

# 2023 holdout bundles (appendix replication of Figs 1 & 2). Guarded so the doc still renders if
# they haven't been built yet:
#   Rscript helpers/build_bundles.R    --year 2023 --runs-subdir 2023 --exp-subdir 2023
#   Rscript helpers/build_distmatch.R  --year 2023 --runs-subdir 2023
b23 <- file.path(derived, "bootstrap_2023.rds")
e23 <- file.path(derived, "expectation_2023.rds")
have_2023   <- file.exists(b23) && file.exists(e23)
if (have_2023) {
  boot_bundle_2023 <- readRDS(b23)
  exp_bundle_2023  <- readRDS(e23)
}
ds23 <- file.path(derived, "distmatch_slope_2023.rds")
have_2023_slope <- file.exists(ds23)
if (have_2023_slope) dm_slope_2023 <- readRDS(ds23)

# 2024 FH-only holdout (loose out-of-sample replication of Figure 1; see
# notes/plan-2024-holdout.md — private notes repo). Base ladder (4 conditions) + FT-raw diagonal
# restricted to {codebook, evidence-zeroshot} — FT-raw on anonymized/summarized text is
# appendix-only (2026-09-04 decision), matching the Fine-tuned block's 2-row design here. Guarded
# so the doc still renders if not yet built:
#   Rscript helpers/build_bundles.R --year 2024 --runs-subdir 2024-fhonly --exp-subdir 2024-fhonly
b24 <- file.path(derived, "bootstrap_2024.rds")
e24 <- file.path(derived, "expectation_2024.rds")
have_2024 <- file.exists(b24) && file.exists(e24)
if (have_2024) {
  boot_bundle_2024 <- readRDS(b24)
  exp_bundle_2024  <- readRDS(e24)
}

sd_signed <- readRDS(file.path(derived, "signeddev_xmodel_2019.rds"))  # Fig 3 (signed deviation)
sds23 <- file.path(derived, "signeddev_xmodel_2023.rds")
have_2023_signed <- file.exists(sds23)
if (have_2023_signed) sd_signed_2023 <- readRDS(sds23)

# Re-identification rates (Fig 4; Question 5). Build with: Rscript helpers/build_reidrates.R
rr23 <- file.path(derived, "reidrates_2023.rds")
have_reidrates <- file.exists(rr23)
if (have_reidrates) reidrates_bundle_2023 <- readRDS(rr23)

# Movement illustration (Appendix A9). Build with: Rscript helpers/build_movement.R --year 2023 --baseline-year 2018
mv23 <- file.path(derived, "movement_2023.rds")
have_movement <- file.exists(mv23)
if (have_movement) {
  movement_bundle_2023 <- readRDS(mv23)
  country_names <- read_csv(file.path(proj_root, "data", "processed", "ert.csv"),
                            show_col_types = FALSE) |>
    distinct(country_text_id, country_name)
}

# Movement/Prominence effects on the difficulty-tracking slope, by condition (Appendix A10).
# Superseded
# the single-panel Identity x Compression interaction version (2026-09-06) -- two separate,
# simpler models (movement-only, reid-only, no interaction term), each run across all four base
# conditions + the trimmed Codebook/Raw Text Fine-Tuned pair. Build with:
#   Rscript helpers/build_slope_by_condition.R --moderator movement
#   Rscript helpers/build_slope_by_condition.R --moderator reid
sbc_movement_path <- file.path(derived, "slopebycondition_movement_2023.rds")
sbc_reid_path     <- file.path(derived, "slopebycondition_reid_2023.rds")
have_sbc_movement <- file.exists(sbc_movement_path)
have_sbc_reid     <- file.exists(sbc_reid_path)
if (have_sbc_movement) sbc_movement <- readRDS(sbc_movement_path)
if (have_sbc_reid)     sbc_reid     <- readRDS(sbc_reid_path)

# Movement/Prominence effects on signed deviation, by condition (Appendix A11). Build with:
#   Rscript helpers/build_signeddev_by_condition.R --moderator movement
#   Rscript helpers/build_signeddev_by_condition.R --moderator reid
sdbc_movement_path <- file.path(derived, "signeddevbycondition_movement_2023.rds")
sdbc_reid_path     <- file.path(derived, "signeddevbycondition_reid_2023.rds")
have_sdbc_movement <- file.exists(sdbc_movement_path)
have_sdbc_reid     <- file.exists(sdbc_reid_path)
if (have_sdbc_movement) sdbc_movement <- readRDS(sdbc_movement_path)
if (have_sdbc_reid)     sdbc_reid     <- readRDS(sdbc_reid_path)

# Name-swap, base + fine-tuned, all three families (Fig 5, companion to Section 4). The paper
# renders the RESCALED version (nameswaprescaled); the rating-point bundle is still loaded
# because Section 5's prose quotes its estimates to motivate the rescaling. Build with:
#   Rscript helpers/build_priorreliance.R --panel B
#   Rscript helpers/build_nameswap_rescaled.R
priorreliance_panelB_path <- file.path(derived, "priorreliance_panelB_2023.rds")
have_priorreliance_panelB <- file.exists(priorreliance_panelB_path)
if (have_priorreliance_panelB) priorreliance_panelB <- readRDS(priorreliance_panelB_path)

# Compression and de-identification (Fig 6) -- the two successive steps from raw source text to
# the Summarized condition, each a paired contrast on all three outcomes: MAE, signed deviation,
# and the difficulty-tracking slope. These bundles also still carry the Full Text contrast that
# the retired Appendix A8 rendered (fig_identity_effect_combined), which nothing calls now.
# Build with:
#   Rscript helpers/build_identity_effect.R
#   Rscript helpers/build_identity_effect_signeddev.R
#   Rscript helpers/build_identity_effect_slope.R
identityeffect_path <- file.path(derived, "identityeffect_2023.rds")
have_identityeffect <- file.exists(identityeffect_path)
if (have_identityeffect) identityeffect_bundle <- readRDS(identityeffect_path)

identityeffect_sd_path <- file.path(derived, "identityeffect_signeddev_2023.rds")
have_identityeffect_sd <- file.exists(identityeffect_sd_path)
if (have_identityeffect_sd) identityeffect_sd_bundle <- readRDS(identityeffect_sd_path)

identityeffect_slope_path <- file.path(derived, "identityeffect_slope_2023.rds")
have_identityeffect_slope <- file.exists(identityeffect_slope_path)
if (have_identityeffect_slope) identityeffect_slope_bundle <- readRDS(identityeffect_slope_path)

# Synthetic-swap deployment, MEAN-SHIFT version. No longer rendered: Figure 7 now reports
# verdict flips (build_fliprate.R below) instead. Retained because the mean-shift
# estimator is (signed deviation)/(n+1), i.e. Figure 3 rescaled, and the numbers are still
# quotable in prose; fig_augmentation_combined() and fig_degradation_combined() still work.
# Build with:
#   Rscript helpers/build_augmentation.R
#   Rscript helpers/build_degradation.R
#   Rscript helpers/build_doseresponse_augmentation.R
#   Rscript helpers/build_doseresponse_degradation.R
augmentation_path <- file.path(derived, "augmentation_2023.rds")
degradation_path  <- file.path(derived, "degradation_2023.rds")
dr_aug_path       <- file.path(derived, "doseresponse_augmentation_2023.rds")
dr_deg_path       <- file.path(derived, "doseresponse_degradation_2023.rds")
have_augmentation <- file.exists(augmentation_path)
have_degradation  <- file.exists(degradation_path)
have_dr_aug       <- file.exists(dr_aug_path)
have_dr_deg       <- file.exists(dr_deg_path)
if (have_augmentation) augmentation_bundle_2023 <- readRDS(augmentation_path)
if (have_degradation)  degradation_bundle_2023  <- readRDS(degradation_path)
if (have_dr_aug)        dr_augmentation_2023    <- readRDS(dr_aug_path)
if (have_dr_deg)        dr_degradation_2023     <- readRDS(dr_deg_path)

# Verdict flips and directional push (Figure 7, four panels). A verdict is
# round(panel mean); the benchmark is real held-out-coder churn, measured rather than assumed.
# Two mechanics on one constructed-thinness design (base panel of 4 drawn from source panels
# with >=8 coders). Build with:
#   Rscript helpers/build_fliprate.R --mechanic augmentation
#   Rscript helpers/build_fliprate.R --mechanic degradation
fra23 <- file.path(derived, "fliprate_augmentation_2023.rds")
frd23 <- file.path(derived, "fliprate_degradation_2023.rds")
have_fliprate <- file.exists(fra23) && file.exists(frd23)
if (have_fliprate) {
  fliprate_augmentation_2023 <- readRDS(fra23)
  fliprate_degradation_2023  <- readRDS(frd23)
}

# Name-swap on the rescaled axis (Fig 5). Build with:
#   Rscript helpers/build_nameswap_rescaled.R
nsr23 <- file.path(derived, "nameswaprescaled_2023.rds")
have_nameswap_rescaled <- file.exists(nsr23)
if (have_nameswap_rescaled) nameswap_rescaled_2023 <- readRDS(nsr23)

# Task framing vs. evidence body (Appendix A8): the three summarized arms measured from raw
# evidence, and the task-framing leg split by whether the model re-identified the country.
# Build with:
#   Rscript helpers/build_summarization_arms.R
#   Rscript helpers/build_reid_split.R
sa23 <- file.path(derived, "summarizationarms_2023.rds")
rs23 <- file.path(derived, "reidsplit_2023.rds")
have_summarization_arms <- file.exists(sa23)
have_reid_split         <- file.exists(rs23)
if (have_summarization_arms) summarization_arms_2023 <- readRDS(sa23)
if (have_reid_split)         reid_split_2023         <- readRDS(rs23)
```

Broader question: how does information help an AI become a synthetic coder? 

Thesis: information helps the AI to emulate human coders, but there is a risk of data leakage. Since the AI may have prior knowledge of the country case that could bias its rating, anonymized and summarized texts will help to ensure that the model is taking into account relevant context and not simply applying a memorized score or relying entirely on prior country knowledge. Finetuning should help the model to learn more precisely the specific context relevant for generating human-like country scores. 

We therefore predict that as the AI receives evidence, its MAE relative to human panel consensus should decrease relative to the base model codebook-only condition. We also anticipate that the MAE will drop further as information becomes progressively more abstracted away from named entities related to the country in question, i.e. that the model will learn more from anonymized and summarized text than from raw text.^[finetune-note] Finally we anticipate that finetuning will magnify these effects, with finetuned models showing lower MAE than base models. 

[^finetune-note]:(Drop this in if we decide to relegate the FT-summ and FT-anon models to the appendix) We also tested models that were finetuned on anonymized and summarized evidence. The results suggested that... We explore these results further in the online appendix. 

## 1. Synthetic-coder performance

We hypothesize that more evidence should improve the synthetic coder's ability to emulate expert coder behavior because it will 

```{r}
#| label: fig-crossmodel
#| fig-cap: "Synthetic coder performance across model families (Llama 70B / Qwen 72B / Gemma 27B). Base conditions (top, all four text inputs) and fine-tuned raw-text adapters (bottom, Codebook and Raw Text only — the matched-format check on Anonymized/Summarized is Appendix A5). Color = model, with country-clustered 95% CIs. The base rows show the greedy (mode) readout only — greedy and mean differ by at most 0.045 MAE there (Qwen; ≤0.017 for Llama and Gemma), well inside the SESOI band and with the same condition ordering under either readout; the fine-tuned rows show both greedy (filled) and expectation/mean (open), where the two readouts diverge by 0.16–0.30 MAE. Dashed line = naive persistence; long-dash = human reference; grey band = ±SESOI."
#| fig-width: 8
#| fig-height: 5.2
fig_crossmodel_landscape(exp_bundle, boot_bundle, band = "human_loo", base_readout = "greedy",
                         ft_conds = c("codebook", "evidence-zeroshot"))
```

We will highlight here how the B hypotheses were either null or confirmed the alternative hypotheses. The most surprising finding is that evidence appears to push the models further away from consensus. It is not beyond the SESOI, but it is striking that evidence appears to increase MAE of the models rather than reduce it. The models are paying attention to the evidence but it is making them disagree as much as or more than individual human coders rather than pushing them towards the consensus view.  

The F hypotheses were supported if we look at the expectation readout rather than greedy decoding. But the greedy decoding results are suggestive of something we did not anticipate which is that greedy mode induces the model to better emulate the human coder. However we cannot assess how human-like the AI is by looking at MAE alone. But rather we have to look at the distribution of the AI codings relative to human coders. (transition to results in QMD 07 and 08)

Difficulty slope, attitude, exaggeration gap analysis suggest that the finetuned model is the most human-like and, further, that it is the weights more than the information in the text that does the heavy lifting. The 2x2 suggests that the fine-tuned raw evidence model can perform as well or better without the evidence in the prompt. 

For QMD 09, part I may be underpowered and does not use stale priors. Need to repeat on 2023. The Part II horse race suggests that of the seven models, FT-Raw Text performs the best in terms of mimicking human coders. This is an underpowered version of QMD 07s distributional analysis.

How frame the difference between expectation (mean) and greedy (mode) readouts: 

- Mean readout → an estimator of the panel consensus (a central estimate).
- Greedy readout → a draw from the panel's rating distribution (behaves like one more coder).

## 2. Difficulty tracking

A companion to Figure 1 on the same rows and greedy readout. **Case difficulty** `h_c` is the
within-cell human leave-one-out MAE — how much the real coders disagreed on that country-year-
indicator (the per-case version of Figure 1's human reference line). The **difficulty-tracking
slope** is the weighted OLS slope of a cell's AI error `a_c` on `h_c`. Because `h_c` is defined
as the human error, the human reference lies on the 45° line (slope 1) by construction; a slope
below 1 is flatter than that reference.

Panel A draws the relationship for two cells against the human reference (curves = mean error per
difficulty bin, with country-clustered 95% CI bands), with every cell shown as a faint grey trace
behind them. Panel B is the slope for every cell, with country-clustered 95% CIs, against the
human reference at 1 (slope 1 = tracks difficulty like a human, flatter = leans on a prior).

```{r}
#| label: fig-crossmodel-slope
#| fig-cap: "Difficulty tracking (greedy readout). (A) Mean AI error vs case difficulty h_c for a flat and a steep exemplar, against the human reference (which lies on the 45° line because h_c is defined as human error), every cell a faint grey trace; 20 equal-count difficulty bins, country-clustered 95% CI bands. (B) Weighted OLS slope of AI error on h_c for every cell, country-clustered 95% CIs, long-dash = human reference (slope 1). Base block covers all four inputs; the FT-raw block shows Codebook and Raw Text only (the matched-format check on Anonymized/Summarized is Appendix A5)."
#| fig-width: 12
#| fig-height: 5.4
fig_crossmodel_slope_2panel(dm_slope, ft_conds = c("codebook", "evidence-zeroshot"))
```

<!-- TODO (EJT): interpretation of the difficulty-tracking pattern. Robustness: slopes on the
>=5-coder difficulty pool preserve the cross-model ordering and the FT-vs-base gap (see
build_distmatch.R min_coders arg); the >=2-coder pool is used here for consistency with Fig 1. -->

## 3. Directional bias — signed deviation

<!-- STOPGAP PROSE (EJT to rewrite): -->
The third axis keeps the *direction* MAE discards: `s_c = AI rating − panel mean`. Zero is the
human panel-member benchmark, since a real coder's signed deviations average to zero by
construction. Panel A traces `s_c` across V-Dem's Regimes of the World for three exemplar cells —
the best-calibrated cell, the one with the steepest regime tilt, and the harshest (most negative
average `s_c`) — with every cell drawn faintly behind them, against the dotted zero line and the
dashed V-Dem IRT `_ord` target on the same axis. Panel B is the average `s_c` per cell: negative
means the model rates below the panel on average (harsh), positive means above (generous). The
democracy-quintile cut of the same measure is in Appendix A3.

```{r}
#| label: fig-crossmodel-signeddev-regime
#| fig-cap: "Directional bias by regime type (greedy readout). (A) Signed deviation (AI − panel mean) across V-Dem's Regimes of the World for three exemplar cells — best-calibrated, steepest regime tilt, harshest average deviation — with every cell shown as a faint grey trace; dotted 0 = human panel-member benchmark, dashed = V-Dem IRT `_ord` target on the same axis, grey band = ±SESOI, country-clustered 95% CI whiskers. (B) Average signed deviation per cell (0 = panel-member benchmark; negative = rates below the panel on average), country-clustered 95% CIs, grey band = ±SESOI. Base block covers all four inputs; the FT-raw block shows Codebook and Raw Text only (the matched-format check on Anonymized/Summarized is Appendix A5)."
#| fig-width: 12
#| fig-height: 5.4
fig_crossmodel_signeddev_2panel_regime(sd_signed, ft_conds = c("codebook", "evidence-zeroshot"))
```

<!-- TODO (EJT): interpretation. Llama and Qwen rate near-zero under Codebook; every base cell
with de-identified text is strongly negative (rates below the panel), worst under
Anonymized/Summarized. Gemma's base Codebook is itself strongly negative (-0.25). Fine-tuning
shrinks the large negative deviations but overshoots into mild generosity for Qwen and Gemma
on Codebook. -->

## 4. Re-identification

Even shown only de-identified evidence, how often does the model name the true country anyway?
**Exact-match re-identification rate** by model family (Llama 70B / Qwen 72B / Gemma 27B), by weights
(Base vs. Fine-Tuned raw-text adapter — the only fine-tuned condition in the study now), and by
de-identification condition (Anonymized vs. Summarized), 2023. Bars ordered left to right within
each condition by descending Base × Anonymized rate. This is a descriptive result in its own
right — the rates are far above chance regardless of condition or model (chance ≈ 1/176 countries
≈ 0.6%; not drawn as a reference line since it would sit indistinguishably close to zero at this
scale) — and one of the ingredients the prominence-crossing figures in Appendix A10-A11 build
on: a country has to be
identifiable before the model can lean on an identity-based prior about it. See
`notes/mockups/reid-rates-fig-concept.md` for the full design rationale and a data-integrity fix
(duplicated rows from a job-submission race) this bundle already reflects.

```{r}
#| label: fig-reid-rates
#| fig-cap: "Exact-match re-identification rate by model family, weights, and de-identification condition (2023). Plain bars, no CI (a descriptive proportion, not a paired contrast). Chance level ≈0.6%, not drawn."
#| fig-width: 9
#| fig-height: 5
#| eval: !expr have_reidrates
fig_reid_rates(reidrates_bundle_2023)
```

## 5. Name-swap: does the rating follow the evidence or the label?

Section 4 asks whether the model can still *recognize* a country from de-identified evidence. This
section asks the complementary question directly: hold the real evidence text fixed, swap in a
different country's name, and see whether the rating follows the content or drifts toward whatever
the injected name implies. Base and fine-tuned weights, all three families, 2023.

The underlying metric is `|rating − named_mean| − |rating − source_mean|`
(`analysis/10-nameswap-2019.qmd`'s Metric 1; positive = tracks content, negative = follows the fake
name). In raw rating points the base estimates are +0.070 / +0.062 / +0.057, which read as
negligible. They are not: the metric is **bounded by how far apart the two countries actually
are** — it lies within ±`gap` where `gap = |named_mean − source_mean|`, whose mean is 0.698 — so
the axis has to be normalized before any magnitude can be read off it. Both panels restrict to
swaps where the two countries meaningfully differ (`gap` ≥ 0.5, retaining 53% of rows; 6.8% of
swaps have `gap` = 0 exactly, where the metric is degenerate).

Panel A is the per-swap **position** on that range: +1 = the rating sits at or past the true
country's panel mean, −1 = at or past the injected country's, 0 = midway between the two. Panel B
is the **win rate**, the share of swaps where the rating lands strictly closer to the true
country's mean, against a 0.50 coin flip. The win rate is reported alongside because the per-case
position is strongly bimodal (25th percentile −1.00, median −0.06, 75th percentile +1.00): on an
individual swap the model commits to one country's answer or the other, so a near-zero average
position is an even split rather than a consistent midpoint. Both panels are drawn on their full
theoretical range rather than zoomed to the estimates — what the axis shows is the point of the
rescaling. No ±SESOI band: the paper's rating-point threshold is a level comparison against the
human reference, and this is neither a level nor a rating-point scale (see
`notes/mockups/nameswap-rescaled-concept.md`).

**Every base cell sits above the coin flip and every fine-tuned cell below it**, and all twelve
intervals exclude their reference. Base win rates are 0.550 / 0.545 / 0.535 (Llama / Qwen /
Gemma); fine-tuned are 0.472 / 0.366 / 0.454, Qwen furthest. The position axis shows the same
flip: base +0.101 / +0.089 / +0.072, fine-tuned −0.045 / −0.255 / −0.081. Read against the
endpoints, the base models are roughly a tenth of the way from the midpoint toward the true
country, not at it.

```{r}
#| label: fig-nameswap-rescaled
#| fig-cap: "Name-swap on the metric's achievable range, base vs. fine-tuned, by model family (2023). (A) Position between the two countries: +1 = the rating is at or past the true source country's panel mean, −1 = at or past the injected country's, dashed 0 = midway. (B) Win rate: share of swaps where the rating is strictly closer to the true country's mean, dashed 0.50 = coin flip. Both panels on their full theoretical range. Swaps with |named_mean − source_mean| < 0.5 excluded (53% of rows retained). Country-clustered bootstrap 95% CIs on the source country. No ±SESOI band (not a rating-point scale)."
#| fig-width: 11
#| fig-height: 4.4
#| eval: !expr have_nameswap_rescaled
fig_nameswap_rescaled(nameswap_rescaled_2023)
```

## 6. Compression and de-identification

Sections 4 and 5 establish that the model conditions on country identity. This section asks what
the evidence-processing pipeline itself costs, by splitting the distance between raw source text
and the Summarized condition Figures 1–3 report into the two things that change along the way:

    raw text ──[ compression ]──▶ summarized-identified ──[ de-identification ]──▶ summarized

**Compression** (Summarized-Identified − Raw Text) keeps names present on both sides, so the only
thing that moves is the ~400-word rewrite. **De-identification** (Summarized −
Summarized-Identified) keeps the text compressed on both sides, so the only thing that moves is the
country's identity. Sign is later − earlier throughout, so positive MAE means further from the
panel mean. Base models only, 2023, on all three outcomes Figures 1–3 track, with shared
country-clustered draws per family.

The two steps add up to the whole raw-to-Summarized gap, but they are **successive steps, not
shares of a total**: the first is measured with identity present and the second with compression
present, so their magnitudes are conditioned differently. Moving the same two factors in the other
order (via Anonymized) reverses which one looks larger for Gemma. No percentage of either step
against the total is reported anywhere for that reason — see
`notes/identity-mechanism-arc-2026-09-16.md` for the demonstration and
`notes/mockups/identity-and-compression-concept.md` for three percentage presentations that were
built and rejected.

No ±SESOI band, following Figure 7's precedent. SESOI defends against small *random* error — a
change below half a rounding step vanishes in any single published score — whereas these are
consistent directional pushes, where the question is whether a step moves the model and in which
direction. The band also discriminates nothing here: at 0.113 it would swallow every cell except
Llama's de-identification MAE.

**MAE.** De-identification is the larger step for all three families (+0.108 / +0.076 / +0.068,
Llama / Qwen / Gemma), and compression is small: null for Llama (+0.001), +0.045 for Qwen, +0.022
for Gemma.

**Difficulty tracking slope.** De-identification again dominates and excludes zero everywhere
(−0.048 / −0.107 / −0.198), flattening the slope. Compression clears zero only for Llama (−0.042);
Qwen (−0.011) and Gemma (+0.000) are null.

**Signed deviation.** The pattern reverses. **Compression excludes zero for all three families**
(−0.101 / −0.117 / −0.044) — compressing the evidence while keeping every name makes each model
harsher — and it is the *larger* of the two steps for Qwen and Gemma, whose de-identification steps
are −0.044 and a null −0.010. Only Llama's de-identification step (−0.131) exceeds its compression
step.

```{r}
#| label: fig-identity-and-compression
#| fig-cap: "The two successive steps from raw source text to the Summarized condition, on all three outcomes (2023, base models). **Compression** = Summarized-Identified − Raw Text (names present on both sides, so only the rewrite moves); **De-identification** = Summarized − Summarized-Identified (compressed on both sides, so only identity moves). Sign is later − earlier, so positive MAE = further from the panel mean, negative signed deviation = harsher, negative slope = flatter difficulty tracking. Each panel has its own x-scale: the outcomes are in different units and are not comparable across panels. The two steps sum to the raw-to-Summarized gap but are successive steps, not shares of a total (see prose). Country-clustered bootstrap 95% CIs, shared draws per family. No ±SESOI band."
#| fig-width: 12
#| fig-height: 3.6
#| eval: !expr have_identityeffect && have_identityeffect_sd && have_identityeffect_slope
fig_identity_and_compression(identityeffect_bundle, identityeffect_slope_bundle,
                             identityeffect_sd_bundle)
```

## Synthesis: what explains Figures 1–3's de-identification pattern?

<!-- TODO (EJT): the mechanism sentence below ("identity itself carrying informative signal the
model loses when stripped") is contradicted by the task-framing result in Appendix A8: a
contentless label reproduces most of the effect, which makes it a retrieval cue rather than lost
signal. Pointers have been renumbered; the claim itself still needs rewriting. See
notes/narrative-arc-and-figures-2026-09-18.md, "Paper-side edits pending". -->

Two candidate explanations for the original puzzle — de-identifying evidence pushes MAE further
from the human panel, sometimes past it entirely — were tested directly. The mechanism
Appendix A10-A11 establishes, country-level staleness (and, more weakly, prominence), does **not** explain it:
crossing movement and re-identification against raw MAE by condition (rather than against the
difficulty-tracking slope or signed deviation) turns up no consistent relationship — movement
clears zero in only 5 of 18 cells with no pattern by condition, and re-identification in none.
Direct identity removal does explain it: holding text length fixed (Anonymized vs. Raw Text) or
compression fixed (Summarized vs. the new Summarized-Identified condition), removing the
country's name increases error for most model/condition pairs, most consistently at the
compressed level (all three families). See Figure 6. The effects are mostly modest — inside
the ±SESOI band except for Llama's full-text case — so this is real, not decisive, evidence: it
points at identity itself carrying informative signal the model loses when stripped, rather than
at staleness or prominence, but doesn't close the question. See
`notes/mockups/mechanism-tests-summary-2026-09-06.md` for the fuller writeup, including
alternative explanations (rewrite-artifact/distributional-shift from the summarization process,
nuance loss from compression) that remain untested.

## 7. Synthetic-coder deployment: does an AI seat change the panel's verdict?

Figures 1–6 establish performance and mechanism; this section asks the deployment question
underneath the paper's opening motivation. Median panel size has fallen from 13 coders over
2005–2012 to 6 by 2023, and the practical question is whether filling a seat with a synthetic
coder changes what the panel actually *concludes*.

The outcome is a **verdict flip**: a change in `round(panel mean)`, the panel's own rounded
judgement — the same rounding the SESOI derives from. It is **not** V-Dem's published `_ord`,
which comes from the measurement model, so nothing here licenses a claim about how many
published scores would change.

**Human churn is measured, not assumed.** Figures on the *mean* shift can zero-reference,
because refilling a seat from the panel's own coders draws from the distribution the mean was
computed from and so shifts it by zero in expectation. A verdict *flip* from ordinary turnover
is emphatically not zero — it runs 16–42% below — so it is estimated and plotted as its own
line.

That benchmark needs spare real coders, which thin panels do not have. So thinness is
**constructed** rather than observed, holding the base panel size fixed (a base of *m−k* would
shrink as k grows and confound the dose-response with panel size). Source panels are the 7,775
cells with ≥8 coders (110 countries, the deepest quarter of 2023); each is randomly permuted,
the first 4 coders form the base panel whose verdict is the baseline, and the next 4 are held
out. The **top row adds** k seats, growing the panel 4 → 4+k; the **bottom row replaces** k of the
base panel's seats, holding size at 4. The left column counts flips, the right gives direction. Both arms in both panels start from the identical base panel,
so the AI-vs-churn contrast is within-panel and panel size cannot drive it. Three AI regimes as
throughout: **Same AI** (every seat Qwen FT-raw Evidence-ZS), **Mixed FT** (drawn from the
6-cell fine-tuned pool), **Mixed pool** (drawn from the full 18-cell grid).

**At one seat, a synthetic coder is indistinguishable from ordinary turnover — under both
mechanics.** Adding (Panel A): 14.1–17.4% against churn's 15.7%, with two of the three regimes
*below* it. Replacing (Panel C): 19.9–21.3% against churn's 20.9%. In both, all four arms sit
within about 1.5 points with overlapping intervals.

**The regimes separate as seats accumulate, and they separate in the same order both ways** —
Same AI disturbs most (38.2% adding, 52.5% replacing, at k=4), Mixed FT least (27.4%, 46.6%),
Mixed pool between. That a *single* well-chosen model is the most disruptive regime, not the
least, is the counterintuitive part: identical ratings in every seat move the mean in concert,
while a mixed pool's disagreements partly cancel.

**Replacement disturbs far more than augmentation at every k** — churn alone reaches 41.5%
versus 22.4% — because a replaced seat removes a human judgement as well as adding a new one.
The level gap is why Panels A and C share a y-scale.

**The right-hand column is where the two mixed regimes come apart, and it is the reason
direction is in the main text rather than an appendix.** On flips alone, Mixed FT and Mixed pool are nearly
indistinguishable (27.4 vs 28.3 adding, 46.6 vs 48.1 replacing), which would invite the
conclusion that an unfiltered mix of models behaves like a curated fine-tuned one. On direction
they have **opposite signs**. Mixed pool excludes zero at every k under both mechanics and grows
monotonically — +3.4 to +8.5 points adding, +3.9 to +14.9 replacing — an increasingly downward
push relative to what human turnover would have done. That regime contains the base cells whose
signed deviation Figure 3 shows is strongly negative, so this is Figure 3's finding arriving at
the panel's published judgement. Mixed FT drifts mildly the other way (−1.4 to −4.4 replacing,
clearing zero only at k=4).

**Same AI sits on the churn line when seats are added** (+0.3 to +1.0, never clearing zero) —
the regime this paper's own results point to, contributing no directional push. It does clear
zero when seats are *replaced*, at k=3 (+5.6) and k=4 (+8.1), for a reason specific to that
regime rather than to the model: every Same AI seat carries an identical rating, so replacing
three or four of four collapses the panel onto a single deterministic judgement and whatever
small offset it carries is expressed in full instead of being diluted by the remaining humans.
That is a true property of putting one model in many seats, not a bias particular to Qwen FT —
and it is also why Same AI flips the most verdicts in both panels: identical seats move the mean
in concert, while a mixed pool's internal disagreements partly cancel.

Panels B and D are differenced against the churn arm, paired at the panel level, and in Panel B
that differencing is load-bearing rather than cosmetic. With a base panel of 4 the mean is a
multiple of 0.25, so exact half-integers are common and `floor(x + 0.5)` rounds them up: a base
sitting at 2.5 has verdict 3, so any added rating below the mean drops it while additions above
merely confirm — a built-in downward push that then alternates with k, because k=1,3 give panel
sizes 5 and 7 where a new mean can never land on .5, while k=2,4 give 6 and 8 where ties are
possible. **Human churn shows the same zigzag** (10.9, 3.3, 10.8, 5.0 raw), which is the tell
that it is arithmetic and not behavior; both arms face identical base panels and identical
parity, so differencing cancels it. Under replacement the denominator never changes, the
alternation cannot arise, and raw net-down for churn comes out flat at −0.1 to −0.3 points, so
Panel D's differencing is a genuine benchmark rather than an artifact-canceller.

One caveat spans the whole figure: the source panels are, by construction, the deepest quarter
of 2023 rather than the thin panels in the field. This is a controlled experiment on constructed
thinness — the price of measuring human churn rather than assuming it. The observational
counterpart (real thin panels, resampled churn) is described in
`notes/mockups/flip-rate-concept.md`, along with why its benchmark understates churn.

```{r}
#| label: fig-fliprate
#| fig-cap: "Does a synthetic seat change the panel's verdict, and in which direction? Rows are the two deployment mechanics, columns the two outcomes. Top row adds k seats (panel grows 4 → 4+k); bottom row replaces k seats (size held at 4). (A, C) Share of panels whose `round(panel mean)` changes, against a human-churn benchmark built from real held-out coders (dashed grey). (B, D) Net down-flips minus up-flips, in points, differenced against that churn arm and paired at the panel level, so the dashed zero line *is* churn's own directional push; positive = the AI arm pushes verdicts down more than human turnover does. Base panel of 4 constructed by random permutation of source panels with ≥8 coders (7,775 panels, 110 countries, 2023); both arms start from the identical base panel, so panel size cannot drive the contrast. Three AI regimes: Same AI = Qwen FT-raw Evidence-ZS; Mixed FT = the 6-cell fine-tuned pool; Mixed pool = the full 18-cell grid. Shared y-scale within each column; regimes offset slightly on k so the intervals do not overlap. Country-clustered bootstrap 95% CIs, 30 Monte Carlo permutations."
#| fig-width: 11.5
#| fig-height: 7.6
#| eval: !expr have_fliprate
fig_fliprate_quad(fliprate_augmentation_2023, fliprate_degradation_2023)
```

# Appendix

## A1. Synthetic-coder performance — 2023 holdout

Replication of Figure 1 on the 2023 holdout year. Same design and readout convention, same
Codebook + Raw Text trim on the FT-raw block as Figure 1 (the untrimmed grid is Appendix A5).
All four base conditions are in for all three families, as are all four Llama FT-raw
conditions, at full row counts.

```{r}
#| label: fig-crossmodel-2023
#| fig-cap: "Synthetic coder performance across model families on the 2023 holdout (Llama 70B / Qwen 72B / Gemma 27B). Layout, readout convention (solid = greedy, open = mean; base rows greedy only), and reference lines match Figure 1. FT-raw block: Codebook and Raw Text only (Appendix A5 has the untrimmed grid). Country-clustered 95% CIs."
#| fig-width: 8
#| fig-height: 5.2
#| eval: !expr have_2023
fig_crossmodel_landscape(exp_bundle_2023, boot_bundle_2023, band = "human_loo", base_readout = "greedy",
                         ft_conds = c("codebook", "evidence-zeroshot"))
```

## A2. Difficulty tracking — 2023 holdout

Replication of Figure 2 on the 2023 holdout year. Same two-panel design, featured cells, reference
(slope 1), greedy readout, and Codebook + Raw Text trim on the FT-raw block as Figure 2. Full
model/condition coverage, as in A1.

```{r}
#| label: fig-crossmodel-slope-2023
#| fig-cap: "Difficulty tracking on the 2023 holdout — replication of Figure 2. (A) AI error vs case difficulty for the same featured cells, every cell a faint grey trace; (B) difficulty-tracking slope per cell, long-dash = human reference (slope 1). FT-raw block: Codebook and Raw Text only (Appendix A5 has the untrimmed grid). Greedy readout, country-clustered 95% CIs."
#| fig-width: 12
#| fig-height: 5.4
#| eval: !expr have_2023_slope
fig_crossmodel_slope_2panel(dm_slope_2023, ft_conds = c("codebook", "evidence-zeroshot"))
```

## A3. Signed Deviation

<!-- STOPGAP PROSE (EJT to rewrite): -->
Two companions to Figure 3. First, the same measure on the **democracy-quintile** cut for 2019 —
the continuous version of the regime-type gradient in the main text. Second, the **regime-type**
cut replicated on the **2023 holdout**. Panel B (average signed deviation) is identical across
cuts by construction, so it repeats here; only Panel A (the gradient) changes.

### Democracy quintile — 2019

```{r}
#| label: fig-crossmodel-signeddev
#| fig-cap: "Directional bias by democracy quintile, 2019 (greedy readout) — the continuous companion to Figure 3's regime-type cut. (A) Signed deviation (AI − panel mean) across democracy quintiles for three exemplar cells — best-calibrated, steepest tilt, harshest average deviation — with every cell shown as a faint grey trace; dotted 0 = human panel-member benchmark, dashed = V-Dem IRT `_ord` target, grey band = ±SESOI, country-clustered 95% CI whiskers. (B) Average signed deviation per cell (same as Figure 3B). FT-raw block: Codebook and Raw Text only (Appendix A5 has the untrimmed grid)."
#| fig-width: 12
#| fig-height: 5.4
fig_crossmodel_signeddev_2panel(sd_signed, ft_conds = c("codebook", "evidence-zeroshot"))
```

### Regime type — 2023 holdout

```{r}
#| label: fig-crossmodel-signeddev-2023-regime
#| fig-cap: "Directional bias by regime type on the 2023 holdout — replication of Figure 3. (A) Signed deviation across V-Dem's Regimes of the World for three exemplar cells (best-calibrated, steepest tilt, harshest average deviation), every cell a faint grey trace; dotted 0 = panel-member benchmark, dashed = V-Dem IRT. (B) Average signed deviation per cell. FT-raw block: Codebook and Raw Text only. Greedy readout, country-clustered 95% CIs, grey band = ±SESOI."
#| fig-width: 12
#| fig-height: 5.4
#| eval: !expr have_2023_signed
fig_crossmodel_signeddev_2panel_regime(sd_signed_2023, ft_conds = c("codebook", "evidence-zeroshot"))
```

## A4. Synthetic-coder performance — 2024 FH-only holdout

Loose out-of-sample replication of Figure 1 on the most recent year — data furthest from the
fine-tuning window (through 2018) and not part of any earlier run. Same design and readout
convention, same Codebook + Raw Text trim on the FT-raw block as Figure 1 (no untrimmed 2024
companion is planned — Appendix A5 covers that check on 2019 only). Base ladder (codebook/
evidence/anonymized/summarized) × 3 families (Llama 70B / Qwen 72B / Gemma 27B) = 12 cells;
FT-raw diagonal restricted to Codebook + Raw Text × 3 families = 6 cells. FH-only: State
Department country reports changed format enough for 2024 that they aren't used, so every cell
here rests on Freedom House evidence alone (see `notes/plan-2024-holdout.md`, private notes
repo, for the staging/run plan). Within-2024 comparisons are valid (base vs. FT, condition
ordering, AI vs. human reference); cross-year levels are confounded (FH-only vs. full-source
drops one source), so this is not placed on a shared axis with 2019/2023.

```{r}
#| label: fig-crossmodel-2024
#| fig-cap: "Synthetic coder performance across model families on the 2024 FH-only holdout (Llama 70B / Qwen 72B / Gemma 27B). Layout, readout convention (solid = greedy, open = mean; base rows greedy only), and reference lines match Figure 1. FT-raw block: Codebook and Raw Text only (Appendix A5 has the untrimmed grid on 2019). Country-clustered 95% CIs."
#| fig-width: 8
#| fig-height: 5.2
#| eval: !expr have_2024
fig_crossmodel_landscape(exp_bundle_2024, boot_bundle_2024, band = "human_loo", base_readout = "greedy",
                         ft_conds = c("codebook", "evidence-zeroshot"))
```

## A5. Untrimmed FT-raw evidence-format grid (2019, cross-model)

Figures 1–3's FT-raw block shows Codebook and Raw Text only — the one apples-to-apples comparison
(does reading the matched-training-format evidence help beyond fine-tuning). This section restores
the two conditions cut from there — Anonymized and Summarized — for all three model families, i.e.
the FT-raw adapter tested on input formats it was never fine-tuned on. As established in A6 (Llama
only, but the same logic applies here): this is a generalization/robustness check, not a second
"does evidence help" test, and it should not be read as evidence for or against the matched-format
finding above.

```{r}
#| label: fig-crossmodel-full
#| fig-cap: "Untrimmed companion to Figure 1: FT-raw block restored to all four inputs, cross-model. Same design, readout convention, and reference lines as Figure 1."
#| fig-width: 8
#| fig-height: 5.2
fig_crossmodel_landscape(exp_bundle, boot_bundle, band = "human_loo", base_readout = "greedy")
```

## A6. Full model × input grid

Llama only: Base plus all three Llama fine-tuned adapters (FT-raw, FT-anon, FT-summ), each run on
all four inputs — Qwen and Gemma trained only the raw-text adapter, so there is no matched-vs-
mismatched comparison to make for them here. The ringed cell in each FT block is that adapter's own
matched training input (e.g., FT-anon on Anonymized); it is not consistently the best-performing
cell for that model, but the ranking among matched/mismatched conditions is not consistent across
model families either (Llama's matched condition loses to Codebook; Qwen's and Gemma's FT-raw
matched condition — Raw Text — is their best of the four), and under the greedy readout every
cell in every block falls within ±SESOI of the human reference regardless of condition (under the
mean readout no cell rates worse than human beyond SESOI, and most rate better). Treated here as
noise rather than a format effect.

```{r}
#| label: fig-grid
#| fig-cap: "Every model on every input (the off-diagonal expansion of Figure 1), Llama only. Greedy (mode) vs expectation (mean) MAE with country-clustered 95% CIs. Rings mark each fine-tuned model's own training input (the diagonal) — not consistently its best-performing cell; under the greedy readout every cell falls within ±SESOI of the human reference, and no cell under either readout rates worse than human beyond SESOI. Dashed line = naive persistence model; long-dash line = human reference; grey band = ±SESOI around the human reference."
#| fig-width: 8
#| fig-height: 9
fig_readout_grid(exp_bundle, boot_bundle, band = "human_loo")
```

## A7. Re-identification — top-3 rate

Companion to Figure 4 (Section 4): the same model × weights × condition grid, but crediting the
model whenever the true country appears anywhere in its top-3 guesses rather than only at top-1.
Same layout, same "no chance-line drawn" treatment, same bar ordering rule (descending Base ×
Anonymized rate) applied independently to this metric.

```{r}
#| label: fig-reid-rates-top3
#| fig-cap: "Top-3 re-identification rate by model family, weights, and de-identification condition (2023) — appendix companion to Figure 4. Plain bars, no CI."
#| fig-width: 9
#| fig-height: 5
#| eval: !expr have_reidrates
fig_reid_rates(reidrates_bundle_2023, metric = "top3")
```

## A8. Task framing or evidence body?

Figure 6 treats de-identification as one manipulation, which is how it is registered
(`preregistration-draft.md:75` defines the treatment as stripping the country's name, named
entities such as officials and institutions, **and datable events**). But the manipulation reaches
the model through two channels at once: the country's name appears in the **task framing** (the
`{FOCAL_COUNTRY}` / `{FOCAL_YEAR}` slots of the user message) *and* throughout the **evidence
body**. This section reports what happens when those channels are varied separately. It is
included as something ruled out, not as a finding: **the two channels do not separate cleanly
enough to build a claim on**, for reasons set out at the end.

### The three summarized arms, measured from raw evidence

Three arms already exist for 2023 base models, all ~400-word rewrites, differing only in where the
country's identity survives:

| arm | evidence body | task framing |
|---|---|---|
| Summ-Identified | names kept | country named |
| Summ + named framing | names stripped | country named |
| Summarized | names stripped | blanked |

The middle arm is the name-swap battery's control condition: name-swap mode forces `hide_identity`
off, so it is the only run in the study carrying de-identified text alongside a real country name
in the framing. Each arm is a paired contrast against **raw evidence**, so all three are distances
from the same origin — no contrast depends on an ordering, and none is a share of a total. That is
deliberate: the earlier nested-tree presentation of these same numbers labelled each node as a
share of a grand total, which reads as attribution when only recovery is licensed (retired; see
`notes/mockups/decomposition-tree-build.R`).

**On MAE and the difficulty slope the arms are ordered**, with the fully de-identified arm furthest
from raw evidence (MAE +0.110 / +0.121 / +0.091; slope −0.090 / −0.116 / −0.197). **On signed
deviation the first two arms coincide** for Llama (−0.102 vs. −0.100) and Qwen (−0.117 vs. −0.120),
while the fully de-identified arm sits well out (−0.233, −0.161) — restoring names to the evidence
body adds essentially nothing once the framing already supplies the country. **This does not hold
for Gemma**, whose three arms sit within 0.012 of each other (−0.044, −0.052, −0.056), so nothing
separates there at all. It is not a three-family result and should not be written as one.

```{r}
#| label: fig-summarization-arms
#| fig-cap: "The three summarized arms as distances from one origin (raw evidence), 2023 base models. Summ-Identified = names kept in the evidence body and the task framing; Summ + named framing = names stripped from the body, country named in the framing; Summarized = names stripped from both. Every arm is a paired contrast against raw evidence on shared country-clustered draws, so each is a distance from the same origin and none is a share of a total. Sign is arm − raw evidence: positive MAE = further from the panel mean, negative signed deviation = harsher, negative slope = flatter difficulty tracking. Each panel has its own x-scale. Country-clustered bootstrap 95% CIs. No ±SESOI band."
#| fig-width: 12
#| fig-height: 4.1
#| eval: !expr have_summarization_arms
fig_summarization_arms(summarization_arms_2023)
```

### Is the task-framing leg just residual leakage?

An immediate objection to the arms above: Figure 4 shows the base model still names the country
24–29% of the time from summarized text, so perhaps naming it in the framing only matters for the
cases where identity was genuinely gone, and the aggregate effect is a partial-leakage artifact.
That has a testable implication. The **task-framing leg** (Summ + named framing − Summarized) holds
the de-identified evidence text identical on both sides, with only `{FOCAL_COUNTRY}` /
`{FOCAL_YEAR}` moving; splitting it by whether the base model's own re-identification probe named
the country correctly gives two competing predictions. *Partial leakage* predicts a **larger** leg
among cases the model could **not** name. *Framing-only attention* — the model uses the
task-specification slot and does not spontaneously condition on identity it could have extracted
from the evidence body — predicts the leg is **flat** across the two groups.

**It is flat, and where it moves it moves the wrong way for partial leakage.** Every MAE and slope
moderation is null. On signed deviation the moderation excludes zero for Llama (+0.041) and Qwen
(+0.039), both in the direction of a *larger* effect among cases the model **could** name; Gemma is
null throughout. Partial leakage is not what produces the task-framing leg.

```{r}
#| label: fig-reid-split
#| fig-cap: "The task-framing leg (Summ + named framing − Summarized), split by whether the base model correctly named the country on the summarized re-identification probe (2023, base models). The same de-identified evidence text appears on both sides of every contrast; only {FOCAL_COUNTRY} / {FOCAL_YEAR} moves. Partial leakage predicts a larger effect among cases the model could NOT name; framing-only attention predicts a flat split. Color encodes the re-identification group, not the model (model is on the facet rows). Country-clustered bootstrap 95% CIs, shared draws, so each subgroup contrast and their difference are computed within draw."
#| fig-width: 11.5
#| fig-height: 6.2
#| eval: !expr have_reid_split
fig_reid_split(reid_split_2023)
```

### Why this is reported as inconclusive

Three reasons the channels cannot be cleanly apportioned from these runs, all of which would have
to be addressed before any "the framing carries most of it" claim could be made.

1. **The fourth cell does not exist.** The arms above vary the evidence body *given* that the
   framing already names the country. The complementary cell — identity present in the evidence
   body, framing blanked — has never been run, so the 2×2 does not close and the two channels
   cannot be shown to partition the effect. Estimated at roughly 30 GPU-hours; it reuses the
   existing `summarized-identified` text cache and needs only `hide_identity` split into two
   booleans in `pipeline/assemble_prompt.py`.
2. **Any two-part split is path-dependent.** Whichever factor moves first absorbs the first-taste
   effect. The alternative ordering of the same two factors exists in the data (via Anonymized) and
   disagrees, with three sign flips and Gemma's story inverting entirely.
3. **The comparison is position-confounded.** `{FOCAL_COUNTRY}` occupies the opening and closing
   of the user message while evidence-body names sit in the middle. Liu et al. (2024) predict a
   U-shaped context-usage asymmetry of exactly this shape from serial position alone, so a framing
   advantage cannot be attributed to the task-specification role of the slot rather than to where
   it sits.

## A9. Movement

The other ingredient the prominence-crossing figures in A10-A11 cross against identifiability:
how much has
a country's democracy level actually changed since roughly when the model's own "impression" of
it would have formed? **Change in V-Dem's Polyarchy score, 2018 to 2023**, one point per country,
ranked. Most countries barely moved; a handful moved a lot, in both directions — exactly the
long-tailed shape that made a plain (non-rank-transformed) version of this measure too
noisy to use directly in a bootstrap (see `notes/mockups/prominence-crossing-fig-concept.md`).
Labels mark exactly the 5 most-backslid and 5 most-improved countries — a fixed rule, not a
visual judgment call — and four of the five steepest declines here are well-documented coups or
crises in this window (Burkina Faso, Myanmar, Mali, Tunisia); the fifth, Mauritius, is a slower
measured democratic erosion under the Jugnauth government. Either way the ranking is a useful
sanity check on the measure itself before it's used as a mechanism-test input.

```{r}
#| label: fig-movement
#| fig-cap: "Change in V-Dem's Polyarchy score (v2x_polyarchy), 2018-2023, one point per country (n=179), ranked. Labels: the 5 most-backslid and 5 most-improved countries."
#| fig-width: 10
#| fig-height: 6.5
#| eval: !expr have_movement
fig_movement(movement_bundle_2023, country_names)
```

## A10. Difficulty tracking under prominence and movement

Does the "tracks difficulty like a human coder would" behavior from Figure 2 hold up equally well
for every country, or does it depend on how identifiable and how recently-changed the country is?
Two separate models (no interaction term — dropped 2026-09-06 in favor of this simpler, more
legible design), each run across all four Base conditions and the trimmed Fine-Tuned pair
(Codebook + Raw Text, matching the FT-raw trim used throughout Figures 1-3):
`AI error ~ h + movement + h:movement` (Panel A) and `AI error ~ h + reid + h:reid` (Panel B),
where `h` is case difficulty (human LOO error). The reported coefficient is the slope-modifying
term in each — does movement/prominence change how well AI error tracks difficulty, not a level
effect (see `notes/mockups/prominence-crossing-fig-concept.md`). Panel A's coefficient is on a
percentile-rank scale (least- to most-moved country, not a real democracy-score unit — see the
same note for why); Panel B's is a literal identified-vs-not comparison, no transform. No SESOI
band is shown; these are slope-modifier units, not MAE, so the paper's rating-point SESOI
threshold doesn't apply here.

**Movement excludes zero in every one of the 18 cells (Panel A), always in the same direction,
including Codebook** — where the model sees no case-specific evidence at all. That last part is
the key to reading it: with no evidence, a rating can only reflect whatever generic impression the
model carries from pretraining. For a stable country that impression is still roughly right, so
whatever error remains still tracks genuine case difficulty; for a country that's recently changed
a lot, the impression is simply wrong, and wrong by about the same amount for every indicator in
that country, which washes out the correlation with difficulty. This holds up whether or not real
evidence is given, and after fine-tuning — a stale prior exerts a baseline pull that in-context
evidence doesn't override. **Prominence (Panel B) is far less consistent** — strong for Gemma at
Base, essentially gone after fine-tuning; the reverse for Qwen (null at Base, consistent once
fine-tuned) — a pattern worth flagging honestly rather than explaining with confidence.

```{r}
#| label: fig-slope-by-condition
#| fig-cap: "Effect of movement (Panel A, percentile rank of |Δv2x_polyarchy|, 2018-2023) and prominence (Panel B, exact-match re-identification, Summarized text) on the difficulty-tracking slope, across all four base conditions and the trimmed Fine-Tuned pair (Codebook, Raw Text). Country-clustered bootstrap 95% CIs; no SESOI band (slope-modifier units, not MAE)."
#| fig-width: 12
#| fig-height: 6
#| eval: !expr have_sbc_movement && have_sbc_reid
fig_slope_by_condition_2panel(sbc_movement, sbc_reid)
```

## A11. Signed deviation under prominence and movement

The same question as A10, but for directional bias (Figure 3) instead of difficulty
tracking: does a country that recently backslid still get a rating that's too generous — still
reflecting its pre-2018 level — and does one that improved get rated too harshly? Two separate
models, same shape as A10 but without the case-difficulty term (signed deviation isn't
itself regressed against difficulty): `signed deviation ~ movement_signed` (Panel A) and
`signed deviation ~ reid` (Panel B), across all four Base conditions and the trimmed Fine-Tuned
pair. Movement uses the *signed*, rank-tamed version — direction is the point here, unlike
A10's magnitude-only measure (see `notes/mockups/prominence-crossing-fig-concept.md`).

**Movement's effect concentrates almost entirely in Codebook** (5 of 6 cells exclude zero, all
in the anchoring direction) and mostly disappears once any real evidence is given. That's a
clean, theory-consistent result: with no case-specific evidence, a rating can only reflect the
model's baked-in impression of the country, so the lag/anchoring signature should be strongest
exactly there — and real evidence, even de-identified or compressed, is enough to mostly correct
the *direction* of the guess, even though A10 shows it doesn't fix the model's ability to
track case-specific *difficulty*. Prominence (Panel B) is scattered and inconsistent — a few
significant cells with no pattern by family or condition, nothing at all once fine-tuned — the
same weak-signal story prominence told in A10.

```{r}
#| label: fig-signeddev-by-condition
#| fig-cap: "Effect of movement (Panel A, signed percentile rank of Δv2x_polyarchy, 2018-2023) and prominence (Panel B, exact-match re-identification, Summarized text) on signed deviation (AI rating - panel mean), across all four base conditions and the trimmed Fine-Tuned pair (Codebook, Raw Text). Country-clustered bootstrap 95% CIs, grey band = ±SESOI."
#| fig-width: 12
#| fig-height: 6
#| eval: !expr have_sdbc_movement && have_sdbc_reid && have_2023
fig_signeddev_by_condition_2panel(sdbc_movement, sdbc_reid, sesoi = boot_bundle_2023$sesoi)
```
