---
title: "Paper figures & tables — draft gallery"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

**Reminder**: Yihui has released a fig.note option for `knitr`. https://github.com/yihui/knitr/pull/2453. Install development version and try it when you build the PDF.

```{r setup}
library(tidyverse)
library(gt)
library(glue)

proj_root <- rprojroot::find_root(rprojroot::is_git_root)
source(file.path(proj_root, "helpers", "bootstrap_helpers.R"))
source(file.path(proj_root, "helpers", "paper_figures.R"))
derived <- file.path(proj_root, "data", "derived")

boot_bundle <- readRDS(file.path(derived, "bootstrap_2019.rds"))
exp_bundle  <- readRDS(file.path(derived, "expectation_2019.rds"))
dm_slope    <- readRDS(file.path(derived, "distmatch_slope_2019.rds"))  # Fig 2 (Test-3 slope)

# 2023 holdout bundles (appendix replication of Figs 1 & 2). Guarded so the doc still renders if
# they haven't been built yet:
#   Rscript helpers/build_bundles.R    --year 2023 --runs-subdir 2023 --exp-subdir 2023
#   Rscript helpers/build_distmatch.R  --year 2023 --runs-subdir 2023
b23 <- file.path(derived, "bootstrap_2023.rds")
e23 <- file.path(derived, "expectation_2023.rds")
have_2023   <- file.exists(b23) && file.exists(e23)
if (have_2023) {
  boot_bundle_2023 <- readRDS(b23)
  exp_bundle_2023  <- readRDS(e23)
}
ds23 <- file.path(derived, "distmatch_slope_2023.rds")
have_2023_slope <- file.exists(ds23)
if (have_2023_slope) dm_slope_2023 <- readRDS(ds23)

# 2024 FH-only holdout (loose out-of-sample replication of Figure 1; see
# notes/plan-2024-holdout.md — private notes repo). Base ladder (4 conditions) + FT-raw diagonal
# restricted to {codebook, evidence-zeroshot} — FT-raw on anonymized/summarized text is
# appendix-only (2026-09-04 decision), matching the Fine-tuned block's 2-row design here. Guarded
# so the doc still renders if not yet built:
#   Rscript helpers/build_bundles.R --year 2024 --runs-subdir 2024-fhonly --exp-subdir 2024-fhonly
b24 <- file.path(derived, "bootstrap_2024.rds")
e24 <- file.path(derived, "expectation_2024.rds")
have_2024 <- file.exists(b24) && file.exists(e24)
if (have_2024) {
  boot_bundle_2024 <- readRDS(b24)
  exp_bundle_2024  <- readRDS(e24)
}

sd_signed <- readRDS(file.path(derived, "signeddev_xmodel_2019.rds"))  # Fig 3 (signed deviation)
sds23 <- file.path(derived, "signeddev_xmodel_2023.rds")
have_2023_signed <- file.exists(sds23)
if (have_2023_signed) sd_signed_2023 <- readRDS(sds23)

# Re-identification rates (Fig 5; Question 5). Build with: Rscript helpers/build_reidrates.R
rr23 <- file.path(derived, "reidrates_2023.rds")
have_reidrates <- file.exists(rr23)
if (have_reidrates) reidrates_bundle_2023 <- readRDS(rr23)

# Movement illustration (Fig 4). Build with: Rscript helpers/build_movement.R --year 2023 --baseline-year 2018
mv23 <- file.path(derived, "movement_2023.rds")
have_movement <- file.exists(mv23)
if (have_movement) {
  movement_bundle_2023 <- readRDS(mv23)
  country_names <- read_csv(file.path(proj_root, "data", "processed", "ert.csv"),
                            show_col_types = FALSE) |>
    distinct(country_text_id, country_name)
}

# Movement/Prominence effects on the difficulty-tracking slope, by condition (Fig 6). Superseded
# the single-panel Identity x Compression interaction version (2026-09-06) -- two separate,
# simpler models (movement-only, reid-only, no interaction term), each run across all four base
# conditions + the trimmed Codebook/Raw Text Fine-Tuned pair. Build with:
#   Rscript helpers/build_slope_by_condition.R --moderator movement
#   Rscript helpers/build_slope_by_condition.R --moderator reid
sbc_movement_path <- file.path(derived, "slopebycondition_movement_2023.rds")
sbc_reid_path     <- file.path(derived, "slopebycondition_reid_2023.rds")
have_sbc_movement <- file.exists(sbc_movement_path)
have_sbc_reid     <- file.exists(sbc_reid_path)
if (have_sbc_movement) sbc_movement <- readRDS(sbc_movement_path)
if (have_sbc_reid)     sbc_reid     <- readRDS(sbc_reid_path)

# Movement/Prominence effects on signed deviation, by condition (Fig 7). Build with:
#   Rscript helpers/build_signeddev_by_condition.R --moderator movement
#   Rscript helpers/build_signeddev_by_condition.R --moderator reid
sdbc_movement_path <- file.path(derived, "signeddevbycondition_movement_2023.rds")
sdbc_reid_path     <- file.path(derived, "signeddevbycondition_reid_2023.rds")
have_sdbc_movement <- file.exists(sdbc_movement_path)
have_sdbc_reid     <- file.exists(sdbc_reid_path)
if (have_sdbc_movement) sdbc_movement <- readRDS(sdbc_movement_path)
if (have_sdbc_reid)     sdbc_reid     <- readRDS(sdbc_reid_path)

# Name-swap tracking level, base + fine-tuned, all three families (Appendix, companion to
# Section 5). Build with: Rscript helpers/build_priorreliance.R --panel B
priorreliance_panelB_path <- file.path(derived, "priorreliance_panelB_2023.rds")
have_priorreliance_panelB <- file.exists(priorreliance_panelB_path)
if (have_priorreliance_panelB) priorreliance_panelB <- readRDS(priorreliance_panelB_path)

# Identity's direct effect (Appendix A8) -- does removing identity, holding text length or
# compression fixed, change the outcome? Same paired contrast on all three outcomes: MAE,
# signed deviation, and the difficulty-tracking slope. Build with:
#   Rscript helpers/build_identity_effect.R
#   Rscript helpers/build_identity_effect_signeddev.R
#   Rscript helpers/build_identity_effect_slope.R
identityeffect_path <- file.path(derived, "identityeffect_2023.rds")
have_identityeffect <- file.exists(identityeffect_path)
if (have_identityeffect) identityeffect_bundle <- readRDS(identityeffect_path)

identityeffect_sd_path <- file.path(derived, "identityeffect_signeddev_2023.rds")
have_identityeffect_sd <- file.exists(identityeffect_sd_path)
if (have_identityeffect_sd) identityeffect_sd_bundle <- readRDS(identityeffect_sd_path)

identityeffect_slope_path <- file.path(derived, "identityeffect_slope_2023.rds")
have_identityeffect_slope <- file.exists(identityeffect_slope_path)
if (have_identityeffect_slope) identityeffect_slope_bundle <- readRDS(identityeffect_slope_path)

# Synthetic-swap deployment check. Figure 8 (main text) = augmentation only, the paper's
# actual deployment scenario: does adding one AI rating to a thin panel move its mean, and
# does that hold as more seats are added? Appendix A9 = the degradation companion (healthy
# panel, coder replaced) -- a robustness check, not a real deployment path. Build with:
#   Rscript helpers/build_augmentation.R
#   Rscript helpers/build_degradation.R
#   Rscript helpers/build_doseresponse_augmentation.R
#   Rscript helpers/build_doseresponse_degradation.R
augmentation_path <- file.path(derived, "augmentation_2023.rds")
degradation_path  <- file.path(derived, "degradation_2023.rds")
dr_aug_path       <- file.path(derived, "doseresponse_augmentation_2023.rds")
dr_deg_path       <- file.path(derived, "doseresponse_degradation_2023.rds")
have_augmentation <- file.exists(augmentation_path)
have_degradation  <- file.exists(degradation_path)
have_dr_aug       <- file.exists(dr_aug_path)
have_dr_deg       <- file.exists(dr_deg_path)
if (have_augmentation) augmentation_bundle_2023 <- readRDS(augmentation_path)
if (have_degradation)  degradation_bundle_2023  <- readRDS(degradation_path)
if (have_dr_aug)        dr_augmentation_2023    <- readRDS(dr_aug_path)
if (have_dr_deg)        dr_degradation_2023     <- readRDS(dr_deg_path)
```

Broader question: how does information help an AI become a synthetic coder? 

Thesis: information helps the AI to emulate human coders, but there is a risk of data leakage. Since the AI may have prior knowledge of the country case that could bias its rating, anonymized and summarized texts will help to ensure that the model is taking into account relevant context and not simply applying a memorized score or relying entirely on prior country knowledge. Finetuning should help the model to learn more precisely the specific context relevant for generating human-like country scores. 

We therefore predict that as the AI receives evidence, its MAE relative to human panel consensus should decrease relative to the base model codebook-only condition. We also anticipate that the MAE will drop further as information becomes progressively more abstracted away from named entities related to the country in question, i.e. that the model will learn more from anonymized and summarized text than from raw text.^[finetune-note] Finally we anticipate that finetuning will magnify these effects, with finetuned models showing lower MAE than base models. 

[^finetune-note]:(Drop this in if we decide to relegate the FT-summ and FT-anon models to the appendix) We also tested models that were finetuned on anonymized and summarized evidence. The results suggested that... We explore these results further in the online appendix. 

## 1. Synthetic-coder performance

We hypothesize that more evidence should improve the synthetic coder's ability to emulate expert coder behavior because it will 

```{r}
#| label: fig-crossmodel
#| fig-cap: "Synthetic coder performance across model families (Llama 70B / Qwen 72B / Gemma 27B). Base conditions (top, all four text inputs) and fine-tuned raw-text adapters (bottom, Codebook and Raw Text only — the matched-format check on Anonymized/Summarized is Appendix A5). Color = model, with country-clustered 95% CIs. The base rows show the greedy (mode) readout only — greedy and mean differ by at most 0.045 MAE there (Qwen; ≤0.017 for Llama and Gemma), well inside the SESOI band and with the same condition ordering under either readout; the fine-tuned rows show both greedy (filled) and expectation/mean (open), where the two readouts diverge by 0.16–0.30 MAE. Dashed line = naive persistence; long-dash = human reference; grey band = ±SESOI."
#| fig-width: 8
#| fig-height: 5.2
fig_crossmodel_landscape(exp_bundle, boot_bundle, band = "human_loo", base_readout = "greedy",
                         ft_conds = c("codebook", "evidence-zeroshot"))
```

We will highlight here how the B hypotheses were either null or confirmed the alternative hypotheses. The most surprising finding is that evidence appears to push the models further away from consensus. It is not beyond the SESOI, but it is striking that evidence appears to increase MAE of the models rather than reduce it. The models are paying attention to the evidence but it is making them disagree as much as or more than individual human coders rather than pushing them towards the consensus view.  

The F hypotheses were supported if we look at the expectation readout rather than greedy decoding. But the greedy decoding results are suggestive of something we did not anticipate which is that greedy mode induces the model to better emulate the human coder. However we cannot assess how human-like the AI is by looking at MAE alone. But rather we have to look at the distribution of the AI codings relative to human coders. (transition to results in QMD 07 and 08)

Difficulty slope, attitude, exaggeration gap analysis suggest that the finetuned model is the most human-like and, further, that it is the weights more than the information in the text that does the heavy lifting. The 2x2 suggests that the fine-tuned raw evidence model can perform as well or better without the evidence in the prompt. 

For QMD 09, part I may be underpowered and does not use stale priors. Need to repeat on 2023. The Part II horse race suggests that of the seven models, FT-Raw Text performs the best in terms of mimicking human coders. This is an underpowered version of QMD 07s distributional analysis.

How frame the difference between expectation (mean) and greedy (mode) readouts: 

- Mean readout → an estimator of the panel consensus (a central estimate).
- Greedy readout → a draw from the panel's rating distribution (behaves like one more coder).

## 2. Difficulty tracking

A companion to Figure 1 on the same rows and greedy readout. **Case difficulty** `h_c` is the
within-cell human leave-one-out MAE — how much the real coders disagreed on that country-year-
indicator (the per-case version of Figure 1's human reference line). The **difficulty-tracking
slope** is the weighted OLS slope of a cell's AI error `a_c` on `h_c`. Because `h_c` is defined
as the human error, the human reference lies on the 45° line (slope 1) by construction; a slope
below 1 is flatter than that reference.

Panel A draws the relationship for two cells against the human reference (curves = mean error per
difficulty bin, with country-clustered 95% CI bands), with every cell shown as a faint grey trace
behind them. Panel B is the slope for every cell, with country-clustered 95% CIs, against the
human reference at 1 (slope 1 = tracks difficulty like a human, flatter = leans on a prior).

```{r}
#| label: fig-crossmodel-slope
#| fig-cap: "Difficulty tracking (greedy readout). (A) Mean AI error vs case difficulty h_c for a flat and a steep exemplar, against the human reference (which lies on the 45° line because h_c is defined as human error), every cell a faint grey trace; 20 equal-count difficulty bins, country-clustered 95% CI bands. (B) Weighted OLS slope of AI error on h_c for every cell, country-clustered 95% CIs, long-dash = human reference (slope 1). Base block covers all four inputs; the FT-raw block shows Codebook and Raw Text only (the matched-format check on Anonymized/Summarized is Appendix A5)."
#| fig-width: 12
#| fig-height: 5.4
fig_crossmodel_slope_2panel(dm_slope, ft_conds = c("codebook", "evidence-zeroshot"))
```

<!-- TODO (EJT): interpretation of the difficulty-tracking pattern. Robustness: slopes on the
>=5-coder difficulty pool preserve the cross-model ordering and the FT-vs-base gap (see
build_distmatch.R min_coders arg); the >=2-coder pool is used here for consistency with Fig 1. -->

## 3. Directional bias — signed deviation

<!-- STOPGAP PROSE (EJT to rewrite): -->
The third axis keeps the *direction* MAE discards: `s_c = AI rating − panel mean`. Zero is the
human panel-member benchmark, since a real coder's signed deviations average to zero by
construction. Panel A traces `s_c` across V-Dem's Regimes of the World for three exemplar cells —
the best-calibrated cell, the one with the steepest regime tilt, and the harshest (most negative
average `s_c`) — with every cell drawn faintly behind them, against the dotted zero line and the
dashed V-Dem IRT `_ord` target on the same axis. Panel B is the average `s_c` per cell: negative
means the model rates below the panel on average (harsh), positive means above (generous). The
democracy-quintile cut of the same measure is in Appendix A3.

```{r}
#| label: fig-crossmodel-signeddev-regime
#| fig-cap: "Directional bias by regime type (greedy readout). (A) Signed deviation (AI − panel mean) across V-Dem's Regimes of the World for three exemplar cells — best-calibrated, steepest regime tilt, harshest average deviation — with every cell shown as a faint grey trace; dotted 0 = human panel-member benchmark, dashed = V-Dem IRT `_ord` target on the same axis, grey band = ±SESOI, country-clustered 95% CI whiskers. (B) Average signed deviation per cell (0 = panel-member benchmark; negative = rates below the panel on average), country-clustered 95% CIs, grey band = ±SESOI. Base block covers all four inputs; the FT-raw block shows Codebook and Raw Text only (the matched-format check on Anonymized/Summarized is Appendix A5)."
#| fig-width: 12
#| fig-height: 5.4
fig_crossmodel_signeddev_2panel_regime(sd_signed, ft_conds = c("codebook", "evidence-zeroshot"))
```

<!-- TODO (EJT): interpretation. Llama and Qwen rate near-zero under Codebook; every base cell
with de-identified text is strongly negative (rates below the panel), worst under
Anonymized/Summarized. Gemma's base Codebook is itself strongly negative (-0.25). Fine-tuning
shrinks the large negative deviations but overshoots into mild generosity for Qwen and Gemma
on Codebook. -->

## 4. Movement

The other ingredient the prominence-crossing figures cross against identifiability: how much has
a country's democracy level actually changed since roughly when the model's own "impression" of
it would have formed? **Change in V-Dem's Polyarchy score, 2018 to 2023**, one point per country,
ranked. Most countries barely moved; a handful moved a lot, in both directions — exactly the
long-tailed shape that made a plain (non-rank-transformed) version of this measure too
noisy to use directly in a bootstrap (see `notes/mockups/prominence-crossing-fig-concept.md`).
Labels mark exactly the 5 most-backslid and 5 most-improved countries — a fixed rule, not a
visual judgment call — and four of the five steepest declines here are well-documented coups or
crises in this window (Burkina Faso, Myanmar, Mali, Tunisia); the fifth, Mauritius, is a slower
measured democratic erosion under the Jugnauth government. Either way the ranking is a useful
sanity check on the measure itself before it's used as a mechanism-test input.

```{r}
#| label: fig-movement
#| fig-cap: "Change in V-Dem's Polyarchy score (v2x_polyarchy), 2018-2023, one point per country (n=179), ranked. Labels: the 5 most-backslid and 5 most-improved countries."
#| fig-width: 10
#| fig-height: 6.5
#| eval: !expr have_movement
fig_movement(movement_bundle_2023, country_names)
```

## 5. Re-identification

Even shown only de-identified evidence, how often does the model name the true country anyway?
**Exact-match re-identification rate** by model family (Llama 70B / Qwen 72B / Gemma 27B), by weights
(Base vs. Fine-Tuned raw-text adapter — the only fine-tuned condition in the study now), and by
de-identification condition (Anonymized vs. Summarized), 2023. Bars ordered left to right within
each condition by descending Base × Anonymized rate. This is a descriptive result in its own
right — the rates are far above chance regardless of condition or model (chance ≈ 1/176 countries
≈ 0.6%; not drawn as a reference line since it would sit indistinguishably close to zero at this
scale) — and one of the ingredients the prominence-crossing figures build on: a country has to be
identifiable before the model can lean on an identity-based prior about it. See
`notes/mockups/reid-rates-fig-concept.md` for the full design rationale and a data-integrity fix
(duplicated rows from a job-submission race) this bundle already reflects.

```{r}
#| label: fig-reid-rates
#| fig-cap: "Exact-match re-identification rate by model family, weights, and de-identification condition (2023). Plain bars, no CI (a descriptive proportion, not a paired contrast). Chance level ≈0.6%, not drawn."
#| fig-width: 9
#| fig-height: 5
#| eval: !expr have_reidrates
fig_reid_rates(reidrates_bundle_2023)
```

## 6. Difficulty tracking under prominence and movement

Does the "tracks difficulty like a human coder would" behavior from Figure 2 hold up equally well
for every country, or does it depend on how identifiable and how recently-changed the country is?
Two separate models (no interaction term — dropped 2026-09-06 in favor of this simpler, more
legible design), each run across all four Base conditions and the trimmed Fine-Tuned pair
(Codebook + Raw Text, matching the FT-raw trim used throughout Figures 1-3):
`AI error ~ h + movement + h:movement` (Panel A) and `AI error ~ h + reid + h:reid` (Panel B),
where `h` is case difficulty (human LOO error). The reported coefficient is the slope-modifying
term in each — does movement/prominence change how well AI error tracks difficulty, not a level
effect (see `notes/mockups/prominence-crossing-fig-concept.md`). Panel A's coefficient is on a
percentile-rank scale (least- to most-moved country, not a real democracy-score unit — see the
same note for why); Panel B's is a literal identified-vs-not comparison, no transform. No SESOI
band is shown; these are slope-modifier units, not MAE, so the paper's rating-point SESOI
threshold doesn't apply here.

**Movement excludes zero in every one of the 18 cells (Panel A), always in the same direction,
including Codebook** — where the model sees no case-specific evidence at all. That last part is
the key to reading it: with no evidence, a rating can only reflect whatever generic impression the
model carries from pretraining. For a stable country that impression is still roughly right, so
whatever error remains still tracks genuine case difficulty; for a country that's recently changed
a lot, the impression is simply wrong, and wrong by about the same amount for every indicator in
that country, which washes out the correlation with difficulty. This holds up whether or not real
evidence is given, and after fine-tuning — a stale prior exerts a baseline pull that in-context
evidence doesn't override. **Prominence (Panel B) is far less consistent** — strong for Gemma at
Base, essentially gone after fine-tuning; the reverse for Qwen (null at Base, consistent once
fine-tuned) — a pattern worth flagging honestly rather than explaining with confidence.

```{r}
#| label: fig-slope-by-condition
#| fig-cap: "Effect of movement (Panel A, percentile rank of |Δv2x_polyarchy|, 2018-2023) and prominence (Panel B, exact-match re-identification, Summarized text) on the difficulty-tracking slope, across all four base conditions and the trimmed Fine-Tuned pair (Codebook, Raw Text). Country-clustered bootstrap 95% CIs; no SESOI band (slope-modifier units, not MAE)."
#| fig-width: 12
#| fig-height: 6
#| eval: !expr have_sbc_movement && have_sbc_reid
fig_slope_by_condition_2panel(sbc_movement, sbc_reid)
```

## 7. Signed deviation under prominence and movement

The same question as Section 6, but for directional bias (Figure 3) instead of difficulty
tracking: does a country that recently backslid still get a rating that's too generous — still
reflecting its pre-2018 level — and does one that improved get rated too harshly? Two separate
models, same shape as Section 6 but without the case-difficulty term (signed deviation isn't
itself regressed against difficulty): `signed deviation ~ movement_signed` (Panel A) and
`signed deviation ~ reid` (Panel B), across all four Base conditions and the trimmed Fine-Tuned
pair. Movement uses the *signed*, rank-tamed version — direction is the point here, unlike
Section 6's magnitude-only measure (see `notes/mockups/prominence-crossing-fig-concept.md`).

**Movement's effect concentrates almost entirely in Codebook** (5 of 6 cells exclude zero, all
in the anchoring direction) and mostly disappears once any real evidence is given. That's a
clean, theory-consistent result: with no case-specific evidence, a rating can only reflect the
model's baked-in impression of the country, so the lag/anchoring signature should be strongest
exactly there — and real evidence, even de-identified or compressed, is enough to mostly correct
the *direction* of the guess, even though Section 6 shows it doesn't fix the model's ability to
track case-specific *difficulty*. Prominence (Panel B) is scattered and inconsistent — a few
significant cells with no pattern by family or condition, nothing at all once fine-tuned — the
same weak-signal story prominence told in Section 6.

```{r}
#| label: fig-signeddev-by-condition
#| fig-cap: "Effect of movement (Panel A, signed percentile rank of Δv2x_polyarchy, 2018-2023) and prominence (Panel B, exact-match re-identification, Summarized text) on signed deviation (AI rating - panel mean), across all four base conditions and the trimmed Fine-Tuned pair (Codebook, Raw Text). Country-clustered bootstrap 95% CIs, grey band = ±SESOI."
#| fig-width: 12
#| fig-height: 6
#| eval: !expr have_sdbc_movement && have_sdbc_reid && have_2023
fig_signeddev_by_condition_2panel(sdbc_movement, sdbc_reid, sesoi = boot_bundle_2023$sesoi)
```

## 8. Synthetic-coder deployment: does adding an AI move the panel?

Figures 1–7 establish performance and mechanism; this section asks the literal
deployment question underneath the paper's opening motivation — median panel size has
fallen from roughly eleven coders in the 2010s to 5.8 by 2024, and the practical
question is whether adding a synthetic coder to one of these thinning panels actually
moves its mean, and which way. **Augmentation**: one AI rating added to a thin 2023
panel (2–8 coders), growing it to n+1 — the help a thinning panel would actually
receive. (The companion question — replacing coders on a *healthy* panel — isn't a
real deployment scenario; that robustness check is Appendix A9.) Panel A is the
single-seat result by condition and model; Panel B asks whether it holds up as more
seats are added, k=1..4, for three regimes: **Same AI** (every seat filled by Qwen
FT-raw Evidence-ZS — the fine-tuned, evidence-reading system this paper's own results
point to, and the one cell that lands indistinguishable from zero across every
single-seat analysis run against it), **Mixed FT** (seats drawn from the 6-cell
fine-tuned pool used throughout the paper), and **Mixed pool** (seats drawn from the
full 18-cell grid). Both panels are a signed shift in the panel mean, rating points,
referenced against zero (no shift) — which is also human coder churn's own expected
shift: refilling a seat from the panel's own coders draws from the same distribution
the mean was computed from, so its expected effect is exactly zero, not a separate
line to plot. No ±SESOI band: unlike Figures 1, 3, and 7, this shift is a difference of
two integers over a panel-size denominator, with no forced-rounding floor the way a
single rating compared against a fractional mean has — reusing the SESOI number here
would borrow the number without the reason (see `notes/mockups/augmentation-concept.md`
for the full derivation and the earlier designs — a mismatched remove/add threshold,
then SESOI itself, then a single-panel spread band — that led to this one).

**Panel A repeats Figure 3 and 7's finding independently**: 16 of 18 cells exclude
zero; Base models push the mean down (harsher) under Raw Text, Anonymized, and
Summarized, worst under the de-identified conditions, while Codebook is
mild-to-near-zero, and fine-tuning pulls the shift toward, and often past, zero.
**Panel B shows this is not a one-seat curiosity.** Same AI stays flat and
indistinguishable from ordinary churn all the way to k=4 — the CI barely widens and
never drifts from zero. Mixed FT and Mixed pool are different: both already exclude
zero at k=1, and the effect grows roughly linearly with k, in *opposite* directions —
Mixed pool (which includes the harsher-biased base cells) drifts increasingly negative,
Mixed FT drifts mildly positive. The practical reading: a well-chosen, fine-tuned,
evidence-reading model looks like ordinary coder turnover no matter how many seats it
fills; an unfiltered mix of models and conditions does not, and the gap widens with
every seat added.

```{r}
#| label: fig-augmentation
#| fig-cap: "Does adding a synthetic coder move a real panel's mean? (A) One AI rating added to a thin 2023 panel (2-8 coders), panel grows to n+1, by condition and model. (B) Dose-response: k=1-4 AI seats added, three regimes (Same AI = Qwen FT-raw Evidence-ZS; Mixed FT = drawn from the 6-cell fine-tuned pool; Mixed pool = drawn from the full 18-cell grid). Signed shift in the panel mean, rating points, zero-referenced (dashed) — also human churn's own expected shift, not separately plotted. Panel A's Base block covers all four inputs; the FT-raw block shows Codebook and Raw Text only (the matched-format check on Anonymized/Summarized is Appendix A5). Country-clustered 95% CIs throughout. No ±SESOI band (see prose)."
#| fig-width: 13
#| fig-height: 6
#| eval: !expr have_augmentation && have_dr_aug
fig_augmentation_combined(augmentation_bundle_2023, dr_augmentation_2023)
```

## Synthesis: what explains Figures 1–3's de-identification pattern?

Two candidate explanations for the original puzzle — de-identifying evidence pushes MAE further
from the human panel, sometimes past it entirely — were tested directly. The mechanism Sections
6–7 establish, country-level staleness (and, more weakly, prominence), does **not** explain it:
crossing movement and re-identification against raw MAE by condition (rather than against the
difficulty-tracking slope or signed deviation) turns up no consistent relationship — movement
clears zero in only 5 of 18 cells with no pattern by condition, and re-identification in none.
Direct identity removal does explain it: holding text length fixed (Anonymized vs. Raw Text) or
compression fixed (Summarized vs. the new Summarized-Identified condition), removing the
country's name increases error for most model/condition pairs, most consistently at the
compressed level (all three families). See Appendix A8. The effects are mostly modest — inside
the ±SESOI band except for Llama's full-text case — so this is real, not decisive, evidence: it
points at identity itself carrying informative signal the model loses when stripped, rather than
at staleness or prominence, but doesn't close the question. See
`notes/mockups/mechanism-tests-summary-2026-09-06.md` for the fuller writeup, including
alternative explanations (rewrite-artifact/distributional-shift from the summarization process,
nuance loss from compression) that remain untested.

# Appendix

## A1. Synthetic-coder performance — 2023 holdout

Replication of Figure 1 on the 2023 holdout year. Same design and readout convention, same
Codebook + Raw Text trim on the FT-raw block as Figure 1 (the untrimmed grid is Appendix A5).
All four base conditions are in for all three families, as are all four Llama FT-raw
conditions, at full row counts.

```{r}
#| label: fig-crossmodel-2023
#| fig-cap: "Synthetic coder performance across model families on the 2023 holdout (Llama 70B / Qwen 72B / Gemma 27B). Layout, readout convention (solid = greedy, open = mean; base rows greedy only), and reference lines match Figure 1. FT-raw block: Codebook and Raw Text only (Appendix A5 has the untrimmed grid). Country-clustered 95% CIs."
#| fig-width: 8
#| fig-height: 5.2
#| eval: !expr have_2023
fig_crossmodel_landscape(exp_bundle_2023, boot_bundle_2023, band = "human_loo", base_readout = "greedy",
                         ft_conds = c("codebook", "evidence-zeroshot"))
```

## A2. Difficulty tracking — 2023 holdout

Replication of Figure 2 on the 2023 holdout year. Same two-panel design, featured cells, reference
(slope 1), greedy readout, and Codebook + Raw Text trim on the FT-raw block as Figure 2. Full
model/condition coverage, as in A1.

```{r}
#| label: fig-crossmodel-slope-2023
#| fig-cap: "Difficulty tracking on the 2023 holdout — replication of Figure 2. (A) AI error vs case difficulty for the same featured cells, every cell a faint grey trace; (B) difficulty-tracking slope per cell, long-dash = human reference (slope 1). FT-raw block: Codebook and Raw Text only (Appendix A5 has the untrimmed grid). Greedy readout, country-clustered 95% CIs."
#| fig-width: 12
#| fig-height: 5.4
#| eval: !expr have_2023_slope
fig_crossmodel_slope_2panel(dm_slope_2023, ft_conds = c("codebook", "evidence-zeroshot"))
```

## A3. Signed Deviation

<!-- STOPGAP PROSE (EJT to rewrite): -->
Two companions to Figure 3. First, the same measure on the **democracy-quintile** cut for 2019 —
the continuous version of the regime-type gradient in the main text. Second, the **regime-type**
cut replicated on the **2023 holdout**. Panel B (average signed deviation) is identical across
cuts by construction, so it repeats here; only Panel A (the gradient) changes.

### Democracy quintile — 2019

```{r}
#| label: fig-crossmodel-signeddev
#| fig-cap: "Directional bias by democracy quintile, 2019 (greedy readout) — the continuous companion to Figure 3's regime-type cut. (A) Signed deviation (AI − panel mean) across democracy quintiles for three exemplar cells — best-calibrated, steepest tilt, harshest average deviation — with every cell shown as a faint grey trace; dotted 0 = human panel-member benchmark, dashed = V-Dem IRT `_ord` target, grey band = ±SESOI, country-clustered 95% CI whiskers. (B) Average signed deviation per cell (same as Figure 3B). FT-raw block: Codebook and Raw Text only (Appendix A5 has the untrimmed grid)."
#| fig-width: 12
#| fig-height: 5.4
fig_crossmodel_signeddev_2panel(sd_signed, ft_conds = c("codebook", "evidence-zeroshot"))
```

### Regime type — 2023 holdout

```{r}
#| label: fig-crossmodel-signeddev-2023-regime
#| fig-cap: "Directional bias by regime type on the 2023 holdout — replication of Figure 3. (A) Signed deviation across V-Dem's Regimes of the World for three exemplar cells (best-calibrated, steepest tilt, harshest average deviation), every cell a faint grey trace; dotted 0 = panel-member benchmark, dashed = V-Dem IRT. (B) Average signed deviation per cell. FT-raw block: Codebook and Raw Text only. Greedy readout, country-clustered 95% CIs, grey band = ±SESOI."
#| fig-width: 12
#| fig-height: 5.4
#| eval: !expr have_2023_signed
fig_crossmodel_signeddev_2panel_regime(sd_signed_2023, ft_conds = c("codebook", "evidence-zeroshot"))
```

## A4. Synthetic-coder performance — 2024 FH-only holdout

Loose out-of-sample replication of Figure 1 on the most recent year — data furthest from the
fine-tuning window (through 2018) and not part of any earlier run. Same design and readout
convention, same Codebook + Raw Text trim on the FT-raw block as Figure 1 (no untrimmed 2024
companion is planned — Appendix A5 covers that check on 2019 only). Base ladder (codebook/
evidence/anonymized/summarized) × 3 families (Llama 70B / Qwen 72B / Gemma 27B) = 12 cells;
FT-raw diagonal restricted to Codebook + Raw Text × 3 families = 6 cells. FH-only: State
Department country reports changed format enough for 2024 that they aren't used, so every cell
here rests on Freedom House evidence alone (see `notes/plan-2024-holdout.md`, private notes
repo, for the staging/run plan). Within-2024 comparisons are valid (base vs. FT, condition
ordering, AI vs. human reference); cross-year levels are confounded (FH-only vs. full-source
drops one source), so this is not placed on a shared axis with 2019/2023.

```{r}
#| label: fig-crossmodel-2024
#| fig-cap: "Synthetic coder performance across model families on the 2024 FH-only holdout (Llama 70B / Qwen 72B / Gemma 27B). Layout, readout convention (solid = greedy, open = mean; base rows greedy only), and reference lines match Figure 1. FT-raw block: Codebook and Raw Text only (Appendix A5 has the untrimmed grid on 2019). Country-clustered 95% CIs."
#| fig-width: 8
#| fig-height: 5.2
#| eval: !expr have_2024
fig_crossmodel_landscape(exp_bundle_2024, boot_bundle_2024, band = "human_loo", base_readout = "greedy",
                         ft_conds = c("codebook", "evidence-zeroshot"))
```

## A5. Untrimmed FT-raw evidence-format grid (2019, cross-model)

Figures 1–3's FT-raw block shows Codebook and Raw Text only — the one apples-to-apples comparison
(does reading the matched-training-format evidence help beyond fine-tuning). This section restores
the two conditions cut from there — Anonymized and Summarized — for all three model families, i.e.
the FT-raw adapter tested on input formats it was never fine-tuned on. As established in A10 (Llama
only, but the same logic applies here): this is a generalization/robustness check, not a second
"does evidence help" test, and it should not be read as evidence for or against the matched-format
finding above.

```{r}
#| label: fig-crossmodel-full
#| fig-cap: "Untrimmed companion to Figure 1: FT-raw block restored to all four inputs, cross-model. Same design, readout convention, and reference lines as Figure 1."
#| fig-width: 8
#| fig-height: 5.2
fig_crossmodel_landscape(exp_bundle, boot_bundle, band = "human_loo", base_readout = "greedy")
```

## A6. Re-identification — top-3 rate

Companion to Figure 5 (Section 5): the same model × weights × condition grid, but crediting the
model whenever the true country appears anywhere in its top-3 guesses rather than only at top-1.
Same layout, same "no chance-line drawn" treatment, same bar ordering rule (descending Base ×
Anonymized rate) applied independently to this metric.

```{r}
#| label: fig-reid-rates-top3
#| fig-cap: "Top-3 re-identification rate by model family, weights, and de-identification condition (2023) — appendix companion to Figure 5. Plain bars, no CI."
#| fig-width: 9
#| fig-height: 5
#| eval: !expr have_reidrates
fig_reid_rates(reidrates_bundle_2023, metric = "top3")
```

## A7. Name-swap tracking

A different, more direct test of the same identity-reliance question as Section 5: instead of
asking whether the model can *recognize* a country from de-identified evidence, hold the real
evidence text fixed and swap in a different country's name — does the rating still follow the
real content, or drift toward whatever the fake name implies? Country-clustered bootstrap mean of
`|rating − named_mean| − |rating − source_mean|` (analysis/10-nameswap-2019.qmd's Metric 1;
positive = tracks content, negative/near-zero = follows the fake name), base and fine-tuned, all
three families, 2023 — no crossing against prominence or movement here, unlike Section 5's
figures (that's a separate question this appendix doesn't ask). **All six cells exclude zero, and
every one sits inside the ±SESOI band** — real but small effects, and the direction flips cleanly
between weight states: every base model leans slightly toward content (reads), every fine-tuned
model flips to leaning slightly toward the injected name (recites), consistent across all three
families. Not featured in the main text because it doesn't cross either of the two ingredients
(prominence, movement) the main-text mechanism figures are built around — see
`notes/mockups/prior-reliance-fig-concept.md` for the original figure design this is drawn from.

```{r}
#| label: fig-nameswap-tracking
#| fig-cap: "Name-swap tracking level, base vs. fine-tuned, by model family (2023). Positive = rating tracks the true source country's content over the injected fake name; negative = follows the fake name. Country-clustered bootstrap 95% CIs, grey band = ±SESOI."
#| fig-width: 6.5
#| fig-height: 4.6
#| eval: !expr have_priorreliance_panelB && have_2023
fig_nameswap_tracking(priorreliance_panelB, sesoi = boot_bundle_2023$sesoi)
```

## A8. Identity's direct effect

The direct test behind the Synthesis note above: does removing a country's identity from the
evidence, holding either text length or compression fixed, change the synthetic coder's behavior?
Two paired condition contrasts per family, same shared country-clustered bootstrap draws:
**Full Text** (Anonymized − Raw Text, identity removed with length held fixed) and **Compressed**
(Summarized − Summarized-Identified, identity removed with compression held fixed) — base models
only, 2023. The contrast is run on all three outcomes Figures 1–3 track — MAE, signed deviation,
and the difficulty-tracking slope — the same three-axis view Sections 6–7 take of the movement and
prominence moderators. The **Compressed** contrast is the clean one: both sides are LLM rewrites,
so the raw-vs-rewrite confound cancels; the **Full Text** contrast (Anonymized is a rewrite, Raw
Text is not) does not have that protection. A third row differencing the two, **Full Text −
Compressed**, is reported in prose only (its point estimate is mechanically the difference of the
other two): it excludes zero in opposite directions across families on every outcome — MAE
(Llama +0.04, Gemma −0.06, Qwen ~0), signed deviation (Llama −0.06, Gemma +0.15, Qwen ~0),
slope (all three +0.04 to +0.08) — so identity-removal and rewrite-artifact effects cannot be
fully separated with this data.

**MAE.** The Compressed contrast is positive and excludes zero for all three families; Full Text
excludes zero for Llama and Qwen but not Gemma. Every effect sits inside the ±SESOI band except
Llama's Full Text case — real, fairly consistent, modest. Unlike Sections 6–7's country-level
moderators (movement, prominence), which came back null against raw MAE by condition, this direct
contrast is the one candidate mechanism tested that predicts part of Figures 1–3's
de-identification pattern.

**Signed deviation.** The axis on which the de-identification pattern is largest (base Llama's
average signed deviation moves from +0.03 under Codebook to −0.48 under Summarized). Negative =
removing the name makes the model **harsher** (rates further below the panel). **Llama shows the
strongest, cleanest identity effect in the analysis** — Full Text −0.19 and Compressed −0.13, both
excluding zero, both beyond ±SESOI. Qwen's contrasts are the same direction but small (−0.03,
−0.04, inside the band). Gemma's clean Compressed contrast is null (−0.01, CI crosses zero) and
its Full Text contrast runs the *other* way (+0.14, beyond the band) — the rewrite confound, not
an identity effect.

**Difficulty tracking slope.** No ±SESOI band — slope-modifier units, not MAE, so the paper's
rating-point threshold doesn't apply (same convention as Section 6 / Figure 6). Negative =
removing the name **flattens** difficulty tracking (the model leans harder on a prior). On the
clean Compressed contrast every family is negative and excludes zero — **Gemma most (−0.20)**,
Qwen −0.11, Llama −0.05 — and Gemma is also the family whose aggregate slope drops most across the
de-identification ladder (0.32 → 0.10, Codebook → Summarized). Full Text agrees for Qwen and Gemma
(−0.06, −0.12) and is near-zero for Llama.

```{r}
#| label: fig-identity-effect
#| fig-cap: "Effect of removing country identity — the Full Text (Anonymized − Raw Text) and Compressed (Summarized − Summarized-Identified) paired contrasts — on all three outcomes (2023, base models), side by side. Each panel has its own x-scale: the outcomes are in different units (MAE and signed deviation in rating points, difficulty slope in slope-modifier units) and are not comparable across panels. The ±SESOI band is drawn only on the two rating-point-scale panels, not on Difficulty Slope. Country-clustered bootstrap 95% CIs, shared draws per family. The Full Text − Compressed contrast is reported in prose, not shown."
#| fig-width: 11
#| fig-height: 3.8
#| eval: !expr have_identityeffect && have_identityeffect_sd && have_identityeffect_slope && have_2023
fig_identity_effect_combined(identityeffect_bundle, identityeffect_sd_bundle,
                             identityeffect_slope_bundle, sesoi = boot_bundle_2023$sesoi)
```

## A9. Degradation: does replacing a healthy panel's coders move it?

The companion to Figure 8 that isn't a real deployment scenario on its own — no one is
proposing to swap out a functioning human panel for AI — but is a useful robustness
check: does Figure 8's story depend on starting from a thin panel, or does it hold up
on a healthy one too? Same design as Figure 8, swapped for the degradation mechanic:
Panel A replaces one human coder with the AI on a healthy 2023 panel (n≥9), panel size
held fixed (every coder-in-panel used as the swapped-out seat); Panel B is the k=1..6
dose-response for the same three regimes (Same AI, Mixed FT, Mixed pool). Same
zero-referenced signed shift, no ±SESOI band, same reasoning as Figure 8.

**The story replicates closely.** Panel A excludes zero in 15 of 18 cells, same
de-identification gradient as Figure 8's Panel A. Panel B: Same AI stays flat through
k=6; Mixed FT and Mixed pool already exclude zero at k=1 and diverge in opposite
directions as k grows, reaching roughly ±0.06–0.09 rating points by k=6 — larger in
absolute terms than Figure 8's k=4 ceiling simply because k runs further, not because
degradation behaves differently from augmentation. The deployment risk Figure 8
measures doesn't depend on how thin the panel already was.

```{r}
#| label: fig-degradation
#| fig-cap: "Robustness companion to Figure 8: does replacing a healthy panel's coders with AI move it the same way? (A) One human coder replaced by the AI on a healthy 2023 panel (n≥9), panel size held fixed, by condition and model. (B) Dose-response: k=1-6 seats replaced, same three regimes as Figure 8B. Signed shift in the panel mean, rating points, zero-referenced (dashed). Panel A's Base block covers all four inputs; the FT-raw block shows Codebook and Raw Text only (Appendix A5 has the untrimmed grid). Country-clustered 95% CIs throughout. No ±SESOI band (see Figure 8's prose)."
#| fig-width: 13
#| fig-height: 6
#| eval: !expr have_degradation && have_dr_deg
fig_degradation_combined(degradation_bundle_2023, dr_degradation_2023)
```

## A10. Full model × input grid

Llama only: Base plus all three Llama fine-tuned adapters (FT-raw, FT-anon, FT-summ), each run on
all four inputs — Qwen and Gemma trained only the raw-text adapter, so there is no matched-vs-
mismatched comparison to make for them here. The ringed cell in each FT block is that adapter's own
matched training input (e.g., FT-anon on Anonymized); it is not consistently the best-performing
cell for that model, but the ranking among matched/mismatched conditions is not consistent across
model families either (Llama's matched condition loses to Codebook; Qwen's and Gemma's FT-raw
matched condition — Raw Text — is their best of the four), and under the greedy readout every
cell in every block falls within ±SESOI of the human reference regardless of condition (under the
mean readout no cell rates worse than human beyond SESOI, and most rate better). Treated here as
noise rather than a format effect.

```{r}
#| label: fig-grid
#| fig-cap: "Every model on every input (the off-diagonal expansion of Figure 1), Llama only. Greedy (mode) vs expectation (mean) MAE with country-clustered 95% CIs. Rings mark each fine-tuned model's own training input (the diagonal) — not consistently its best-performing cell; under the greedy readout every cell falls within ±SESOI of the human reference, and no cell under either readout rates worse than human beyond SESOI. Dashed line = naive persistence model; long-dash line = human reference; grey band = ±SESOI around the human reference."
#| fig-width: 8
#| fig-height: 9
fig_readout_grid(exp_bundle, boot_bundle, band = "human_loo")
```
