---
title: "Paper figures & tables — draft gallery"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

**Reminder**: Yihui has released a fig.note option for `knitr`. https://github.com/yihui/knitr/pull/2453. Install development version and try it when you build the PDF.

```{r setup}
library(tidyverse)
library(gt)
library(glue)

proj_root <- rprojroot::find_root(rprojroot::is_git_root)
source(file.path(proj_root, "helpers", "bootstrap_helpers.R"))
source(file.path(proj_root, "helpers", "paper_figures.R"))
derived <- file.path(proj_root, "data", "derived")

boot_bundle <- readRDS(file.path(derived, "bootstrap_2019.rds"))
exp_bundle  <- readRDS(file.path(derived, "expectation_2019.rds"))
dm_slope    <- readRDS(file.path(derived, "distmatch_slope_2019.rds"))  # Fig 2 (Test-3 slope)

# 2023 holdout bundles (appendix replication of Figs 1 & 2). Guarded so the doc still renders if
# they haven't been built yet:
#   Rscript helpers/build_bundles.R    --year 2023 --runs-subdir 2023 --exp-subdir 2023
#   Rscript helpers/build_distmatch.R  --year 2023 --runs-subdir 2023
b23 <- file.path(derived, "bootstrap_2023.rds")
e23 <- file.path(derived, "expectation_2023.rds")
have_2023   <- file.exists(b23) && file.exists(e23)
if (have_2023) {
  boot_bundle_2023 <- readRDS(b23)
  exp_bundle_2023  <- readRDS(e23)
}
ds23 <- file.path(derived, "distmatch_slope_2023.rds")
have_2023_slope <- file.exists(ds23)
if (have_2023_slope) dm_slope_2023 <- readRDS(ds23)

# 2024 FH-only holdout (loose out-of-sample replication of Figure 1; see
# notes/plan-2024-holdout.md — private notes repo). Base ladder (4 conditions) + FT-raw diagonal
# restricted to {codebook, evidence-zeroshot} — FT-raw on anonymized/summarized text is
# appendix-only (2026-09-04 decision), matching the Fine-tuned block's 2-row design here. Guarded
# so the doc still renders if not yet built:
#   Rscript helpers/build_bundles.R --year 2024 --runs-subdir 2024-fhonly --exp-subdir 2024-fhonly
b24 <- file.path(derived, "bootstrap_2024.rds")
e24 <- file.path(derived, "expectation_2024.rds")
have_2024 <- file.exists(b24) && file.exists(e24)
if (have_2024) {
  boot_bundle_2024 <- readRDS(b24)
  exp_bundle_2024  <- readRDS(e24)
}

sd_signed <- readRDS(file.path(derived, "signeddev_xmodel_2019.rds"))  # Fig 3 (signed deviation)
sds23 <- file.path(derived, "signeddev_xmodel_2023.rds")
have_2023_signed <- file.exists(sds23)
if (have_2023_signed) sd_signed_2023 <- readRDS(sds23)
```

Broader question: how does information help an AI become a synthetic coder? 

Thesis: information helps the AI to emulate human coders, but there is a risk of data leakage. Since the AI may have prior knowledge of the country case that could bias its rating, anonymized and summarized texts will help to ensure that the model is taking into account relevant context and not simply applying a memorized score or relying entirely on prior country knowledge. Finetuning should help the model to learn more precisely the specific context relevant for generating human-like country scores. 

We therefore predict that as the AI receives evidence, its MAE relative to human panel consensus should decrease relative to the base model codebook-only condition. We also anticipate that the MAE will drop further as information becomes progressively more abstracted away from named entities related to the country in question, i.e. that the model will learn more from anonymized and summarized text than from raw text.^[finetune-note] Finally we anticipate that finetuning will magnify these effects, with finetuned models showing lower MAE than base models. 

[^finetune-note]:(Drop this in if we decide to relegate the FT-summ and FT-anon models to the appendix) We also tested models that were finetuned on anonymized and summarized evidence. The results suggested that... We explore these results further in the online appendix. 

## 1. Synthetic-coder performance

We hypothesize that more evidence should improve the synthetic coder's ability to emulate expert coder behavior because it will 

```{r}
#| label: fig-crossmodel
#| fig-cap: "Synthetic coder performance across model families (Llama 70B / Qwen 72B / Gemma 27B). Base conditions (top, all four text inputs) and fine-tuned raw-text adapters (bottom, Codebook and Raw Text only — the matched-format check on Anonymized/Summarized is Appendix A7). Color = model, with country-clustered 95% CIs. The base rows show the greedy (mode) readout only — greedy and mean differ by <0.02 MAE there (~1/9 SESOI); the fine-tuned rows show both greedy (open) and expectation/mean (filled), where the two readouts diverge by 0.16–0.30 MAE. Dashed line = naive persistence; long-dash = human reference; grey band = ±SESOI."
#| fig-width: 8
#| fig-height: 5.2
fig_crossmodel_landscape(exp_bundle, boot_bundle, band = "human_loo", base_readout = "greedy",
                         ft_conds = c("codebook", "evidence-zeroshot"))
```

We will highlight here how the B hypotheses were either null or confirmed the alternative hypotheses. The most surprising finding is that evidence appears to push the models further away from consensus. It is not beyond the SESOI, but it is striking that evidence appears to increase MAE of the models rather than reduce it. The models are paying attention to the evidence but it is making them disagree as much as or more than individual human coders rather than pushing them towards the consensus view.  

The F hypotheses were supported if we look at the expectation readout rather than greedy decoding. But the greedy decoding results are suggestive of something we did not anticipate which is that greedy mode induces the model to better emulate the human coder. However we cannot assess how human-like the AI is by looking at MAE alone. But rather we have to look at the distribution of the AI codings relative to human coders. (transition to results in QMD 07 and 08)

Difficulty slope, attitude, exaggeration gap analysis suggest that the finetuned model is the most human-like and, further, that it is the weights more than the information in the text that does the heavy lifting. The 2x2 suggests that the fine-tuned raw evidence model can perform as well or better without the evidence in the prompt. 

For QMD 09, part I may be underpowered and does not use stale priors. Need to repeat on 2023. The Part II horse race suggests that of the seven models, FT-Raw Text performs the best in terms of mimicking human coders. This is an underpowered version of QMD 07s distributional analysis.

How frame the difference between expectation (mean) and greedy (mode) readouts: 

- Mean readout → an estimator of the panel consensus (a central estimate).
- Greedy readout → a draw from the panel's rating distribution (behaves like one more coder).

## 2. Difficulty tracking

A companion to Figure 1 on the same rows and greedy readout. **Case difficulty** `h_c` is the
within-cell human leave-one-out MAE — how much the real coders disagreed on that country-year-
indicator (the per-case version of Figure 1's human reference line). The **difficulty-tracking
slope** is the weighted OLS slope of a cell's AI error `a_c` on `h_c`. Because `h_c` is defined
as the human error, the human reference lies on the 45° line (slope 1) by construction; a slope
below 1 is flatter than that reference.

Panel A draws the relationship for two cells against the human reference (curves = mean error per
difficulty bin, with country-clustered 95% CI bands). Panel B is the slope for every cell, with
country-clustered 95% CIs, against the human reference at 1.

```{r}
#| label: fig-crossmodel-slope
#| fig-cap: "Difficulty tracking (greedy readout). (A) Mean AI error vs case difficulty h_c for a flat and a steep exemplar, against the human reference (which lies on the 45° line because h_c is defined as human error); 20 equal-count difficulty bins, country-clustered 95% CI bands. (B) Weighted OLS slope of AI error on h_c for every cell, country-clustered 95% CIs, long-dash = human reference (slope 1). Base block covers all four inputs; the FT-raw block shows Codebook and Raw Text only (the matched-format check on Anonymized/Summarized is Appendix A7)."
#| fig-width: 12
#| fig-height: 5.4
fig_crossmodel_slope_2panel(dm_slope, ft_conds = c("codebook", "evidence-zeroshot"))
```

<!-- TODO (EJT): interpretation of the difficulty-tracking pattern. Robustness: slopes on the
>=5-coder difficulty pool preserve the cross-model ordering and the FT-vs-base gap (see
build_distmatch.R min_coders arg); the >=2-coder pool is used here for consistency with Fig 1. -->

## 3. Directional bias — signed deviation

The third axis keeps the *direction* MAE discards: `s_c = AI rating − panel mean`, averaged within
regime bins. Zero is the panel-member target (a real coder's signed deviations sum to zero per bin
by construction). The scalar summary is the **exaggeration gap** (positive = the model spreads the
regime gradient wider than the human panel — rates autocracies below / democracies above; negative
= compresses toward the middle), computed two ways below: first on V-Dem's categorical **regime
type** (Regimes of the World), then on the continuous **democracy quintile** cut directly beneath
it, so the two views of the same gradient can be compared side by side.

Panel A in each figure draws the gradient for the flattest and steepest cells against two
references: the dotted zero line (raw panel mean) and the dashed line (V-Dem's IRT `_ord` score
expressed on the same axis, `mean(ord − panel_mean)` per bin). Panel B is the exaggeration gap for
every cell.

### Regime type (V-Dem Regimes of the World)

```{r}
#| label: fig-crossmodel-signeddev-regime
#| fig-cap: "Directional bias by regime type (greedy readout). (A) Signed deviation (AI − panel mean) across V-Dem's Regimes of the World (closed autocracy → liberal democracy) for the flattest and steepest cells; dotted 0 = panel mean, dashed = V-Dem IRT target on the same axis, grey band = ±SESOI, country-clustered 95% CI bands. (B) Exaggeration gap (liberal democracy − closed autocracy signed deviation) for every cell, country-clustered 95% CIs, dashed 0 = no tilt, grey band = ±SESOI. Base block covers all four inputs; the FT-raw block shows Codebook and Raw Text only (the matched-format check on Anonymized/Summarized is Appendix A7)."
#| fig-width: 12
#| fig-height: 5.4
fig_crossmodel_signeddev_2panel_regime(sd_signed, ft_conds = c("codebook", "evidence-zeroshot"))
```

### Democracy quintile

```{r}
#| label: fig-crossmodel-signeddev
#| fig-cap: "Directional bias by democracy quintile (greedy readout). (A) Signed deviation (AI − panel mean) across democracy quintiles for the flattest and steepest cells; dotted 0 = panel mean, dashed = V-Dem IRT target on the same axis, grey band = ±SESOI, country-clustered 95% CI bands. (B) Exaggeration gap (Q5 − Q1 signed deviation) for every cell, country-clustered 95% CIs, dashed 0 = no tilt (positive = exaggerates the gradient, negative = compresses), grey band = ±SESOI. Base block covers all four inputs; the FT-raw block shows Codebook and Raw Text only (the matched-format check on Anonymized/Summarized is Appendix A7)."
#| fig-width: 12
#| fig-height: 5.4
fig_crossmodel_signeddev_2panel(sd_signed, ft_conds = c("codebook", "evidence-zeroshot"))
```

<!-- TODO (EJT): interpretation of the directional-bias pattern (base vs FT tilt; the IRT line's
own slope; Qwen's calibration vs its Fig-2 slope; regime-type vs quintile-cut agreement). -->

## 4. Continuous Movement

**Runs:** No additional runs required.

Test how the relevance of information changes with the staleness of the prior. 

## 5. Re-identificaiton

**Runs:** Re-identification per model (FT Llama: Ev; FT Qwen: Ev; FT Gemma: Ev) on...what condition? Anonymized, summarized or both? We did see that Llama FT-summ had a harder time re-identifying relative to the other models. Briefly dig into that more. However, fundamentally we are trying to explain patterns in the figures pertaining to the FT-evidence models and so our efforts should primarily focus on those. 

## 6. Name swap

**Runs:** Run name-swap on three models (FT Llama: Ev, FT Qwen: Ev; FT Gemma: Ev). Consider whether using the raw text adapter leaks the names. Perhaps do a trial run to see whether you get different results using the raw versus summarized adapters for Llama. 

# Appendix

## A1. Synthetic-coder performance — 2023 holdout

Replication of Figure 1 on the 2023 holdout year. Same design and readout convention, same
Codebook + Raw Text trim on the FT-raw block as Figure 1 (the untrimmed grid is Appendix A7).
Base block: Qwen and Gemma 2023 base runs on Evidence/Anonymized/Summarized are still in
flight (Codebook and, for Qwen, Anonymized/Summarized are in). Llama's FT-raw 2023
`anonymized-zeroshot`/`summarized-zeroshot` runs are pending regardless, which is moot for this
trimmed view since those two conditions aren't shown here.

```{r}
#| label: fig-crossmodel-2023
#| fig-cap: "Synthetic coder performance across model families on the 2023 holdout (Llama 70B / Qwen 72B / Gemma 27B). Layout, readout convention (solid = greedy, open = mean; base rows greedy only), and reference lines match Figure 1. FT-raw block: Codebook and Raw Text only (Appendix A7 has the untrimmed grid). Country-clustered 95% CIs."
#| fig-width: 8
#| fig-height: 5.2
#| eval: !expr have_2023
fig_crossmodel_landscape(exp_bundle_2023, boot_bundle_2023, band = "human_loo", base_readout = "greedy",
                         ft_conds = c("codebook", "evidence-zeroshot"))
```

## A2. Difficulty tracking — 2023 holdout

Replication of Figure 2 on the 2023 holdout year. Same two-panel design, featured cells, reference
(slope 1), greedy readout, and Codebook + Raw Text trim on the FT-raw block as Figure 2. Same
model/condition coverage caveats as A1.

```{r}
#| label: fig-crossmodel-slope-2023
#| fig-cap: "Difficulty tracking on the 2023 holdout — replication of Figure 2. (A) AI error vs case difficulty for the same featured cells; (B) difficulty-tracking slope per cell, long-dash = human reference (slope 1). FT-raw block: Codebook and Raw Text only (Appendix A7 has the untrimmed grid). Greedy readout, country-clustered 95% CIs."
#| fig-width: 12
#| fig-height: 5.4
#| eval: !expr have_2023_slope
fig_crossmodel_slope_2panel(dm_slope_2023, ft_conds = c("codebook", "evidence-zeroshot"))
```

## A3. Directional bias — 2023 holdout

Replication of Figure 3 on the 2023 holdout year. Same two-panel design, featured cells, references
(panel mean at 0, V-Dem IRT dashed), greedy readout, and Codebook + Raw Text trim on the FT-raw
block as Figure 3. Same model/condition coverage caveats as A1–A2. Regime type first, democracy
quintile directly below it — same pairing as Figure 3.

### Regime type (V-Dem Regimes of the World)

```{r}
#| label: fig-crossmodel-signeddev-2023-regime
#| fig-cap: "Directional bias by regime type on the 2023 holdout — replication of the regime-type panel of Figure 3. (A) Signed deviation across V-Dem's Regimes of the World (closed autocracy → liberal democracy) for the flattest and steepest cells, dotted 0 = panel mean, dashed = V-Dem IRT; (B) exaggeration gap (liberal democracy − closed autocracy signed deviation) per cell. FT-raw block: Codebook and Raw Text only (Appendix A7 has the untrimmed grid). Greedy readout, country-clustered 95% CIs, grey band = ±SESOI."
#| fig-width: 12
#| fig-height: 5.4
#| eval: !expr have_2023_signed
fig_crossmodel_signeddev_2panel_regime(sd_signed_2023, ft_conds = c("codebook", "evidence-zeroshot"))
```

### Democracy quintile

```{r}
#| label: fig-crossmodel-signeddev-2023
#| fig-cap: "Directional bias by democracy quintile on the 2023 holdout — replication of Figure 3. (A) Signed deviation across democracy quintiles for the flattest and steepest cells, dotted 0 = panel mean, dashed = V-Dem IRT; (B) exaggeration gap per cell. FT-raw block: Codebook and Raw Text only (Appendix A7 has the untrimmed grid). Greedy readout, country-clustered 95% CIs, grey band = ±SESOI."
#| fig-width: 12
#| fig-height: 5.4
#| eval: !expr have_2023_signed
fig_crossmodel_signeddev_2panel(sd_signed_2023, ft_conds = c("codebook", "evidence-zeroshot"))
```

## A4. Synthetic-coder performance — 2024 FH-only holdout

Loose out-of-sample replication of Figure 1 on the most recent year — data furthest from the
fine-tuning window (through 2018) and not part of any earlier run. Same design and readout
convention, same Codebook + Raw Text trim on the FT-raw block as Figure 1 (no untrimmed 2024
companion is planned — Appendix A7 covers that check on 2019 only). Base ladder (codebook/
evidence/anonymized/summarized) × 3 families (Llama 70B / Qwen 72B / Gemma 27B) = 12 cells;
FT-raw diagonal restricted to Codebook + Raw Text × 3 families = 6 cells. FH-only: State
Department country reports changed format enough for 2024 that they aren't used, so every cell
here rests on Freedom House evidence alone (see `notes/plan-2024-holdout.md`, private notes
repo, for the staging/run plan). Within-2024 comparisons are valid (base vs. FT, condition
ordering, AI vs. human reference); cross-year levels are confounded (FH-only vs. full-source
drops one source), so this is not placed on a shared axis with 2019/2023.

```{r}
#| label: fig-crossmodel-2024
#| fig-cap: "Synthetic coder performance across model families on the 2024 FH-only holdout (Llama 70B / Qwen 72B / Gemma 27B). Layout, readout convention (solid = greedy, open = mean; base rows greedy only), and reference lines match Figure 1. FT-raw block: Codebook and Raw Text only (Appendix A7 has the untrimmed grid on 2019). Country-clustered 95% CIs."
#| fig-width: 8
#| fig-height: 5.2
#| eval: !expr have_2024
fig_crossmodel_landscape(exp_bundle_2024, boot_bundle_2024, band = "human_loo", base_readout = "greedy",
                         ft_conds = c("codebook", "evidence-zeroshot"))
```

## A5. B and F hypothesis tests

```{r}
#| label: fig-bf
#| fig-cap: "Paired ΔMAE for the registered B and F contrasts, with the zero line and ±SESOI band."
#| fig-width: 8
#| fig-height: 4.8
fig_bf(boot_bundle)
```

## A6. Full model × input grid

Llama only: Base plus all three Llama fine-tuned adapters (FT-raw, FT-anon, FT-summ), each run on
all four inputs — Qwen and Gemma trained only the raw-text adapter, so there is no matched-vs-
mismatched comparison to make for them here. The ringed cell in each FT block is that adapter's own
matched training input (e.g., FT-anon on Anonymized); it is not consistently the best-performing
cell for that model, but the ranking among matched/mismatched conditions is not consistent across
model families either (Llama's matched condition loses to Codebook; Qwen's and Gemma's FT-raw
matched condition — Raw Text — is their best of the four), and every cell in every block falls
within ±SESOI of the human reference regardless of condition. Treated here as noise rather than a
format effect.

```{r}
#| label: fig-grid
#| fig-cap: "Every model on every input (the off-diagonal expansion of Figure 1), Llama only. Greedy (mode) vs expectation (mean) MAE with country-clustered 95% CIs. Rings mark each fine-tuned model's own training input (the diagonal) — not consistently its best-performing cell, and every cell here falls within ±SESOI of the human reference. Dashed line = naive persistence model; long-dash line = human reference; grey band = ±SESOI around the human reference."
#| fig-width: 8
#| fig-height: 9
fig_readout_grid(exp_bundle, boot_bundle, band = "human_loo")
```

## A7. Untrimmed FT-raw evidence-format grid (2019, cross-model)

Figures 1–3's FT-raw block shows Codebook and Raw Text only — the one apples-to-apples comparison
(does reading the matched-training-format evidence help beyond fine-tuning). This section restores
the two conditions cut from there — Anonymized and Summarized — for all three model families, i.e.
the FT-raw adapter tested on input formats it was never fine-tuned on. As established in A5 (Llama
only, but the same logic applies here): this is a generalization/robustness check, not a second
"does evidence help" test, and it should not be read as evidence for or against the matched-format
finding above.

```{r}
#| label: fig-crossmodel-full
#| fig-cap: "Untrimmed companion to Figure 1: FT-raw block restored to all four inputs, cross-model. Same design, readout convention, and reference lines as Figure 1."
#| fig-width: 8
#| fig-height: 5.2
fig_crossmodel_landscape(exp_bundle, boot_bundle, band = "human_loo", base_readout = "greedy")
```

```{r}
#| label: fig-crossmodel-slope-full
#| fig-cap: "Untrimmed companion to Figure 2: FT-raw block restored to all four inputs, cross-model. Same design and reference (human self-reference, slope 1) as Figure 2."
#| fig-width: 12
#| fig-height: 5.4
fig_crossmodel_slope_2panel(dm_slope)
```

```{r}
#| label: fig-crossmodel-signeddev-full-regime
#| fig-cap: "Untrimmed companion to Figure 3 (regime-type cut): FT-raw block restored to all four inputs, cross-model. Same design and references as Figure 3's regime-type panel."
#| fig-width: 12
#| fig-height: 5.4
fig_crossmodel_signeddev_2panel_regime(sd_signed)
```

```{r}
#| label: fig-crossmodel-signeddev-full
#| fig-cap: "Untrimmed companion to Figure 3 (democracy-quintile cut): FT-raw block restored to all four inputs, cross-model. Same design and references as Figure 3's quintile panel."
#| fig-width: 12
#| fig-height: 5.4
fig_crossmodel_signeddev_2panel(sd_signed)
```
