---
title: "Paper figures & tables — draft gallery"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

**Reminder**: Yihui has released a fig.note option for `knitr`. https://github.com/yihui/knitr/pull/2453. Install development version and try it when you build the PDF.

```{r setup}
library(tidyverse)
library(gt)
library(glue)

proj_root <- rprojroot::find_root(rprojroot::is_git_root)
source(file.path(proj_root, "helpers", "bootstrap_helpers.R"))
source(file.path(proj_root, "helpers", "paper_figures.R"))
derived <- file.path(proj_root, "data", "derived")

boot_bundle <- readRDS(file.path(derived, "bootstrap_2019.rds"))
exp_bundle  <- readRDS(file.path(derived, "expectation_2019.rds"))
dm_slope    <- readRDS(file.path(derived, "distmatch_slope_2019.rds"))  # Fig 2 (Test-3 slope)

# 2023 holdout bundles (appendix replication of Figs 1 & 2). Guarded so the doc still renders if
# they haven't been built yet:
#   Rscript helpers/build_bundles.R    --year 2023 --runs-subdir 2023 --exp-subdir 2023
#   Rscript helpers/build_distmatch.R  --year 2023 --runs-subdir 2023
b23 <- file.path(derived, "bootstrap_2023.rds")
e23 <- file.path(derived, "expectation_2023.rds")
have_2023   <- file.exists(b23) && file.exists(e23)
if (have_2023) {
  boot_bundle_2023 <- readRDS(b23)
  exp_bundle_2023  <- readRDS(e23)
}
ds23 <- file.path(derived, "distmatch_slope_2023.rds")
have_2023_slope <- file.exists(ds23)
if (have_2023_slope) dm_slope_2023 <- readRDS(ds23)

sd_signed <- readRDS(file.path(derived, "signeddev_xmodel_2019.rds"))  # Fig 3 (signed deviation)
sds23 <- file.path(derived, "signeddev_xmodel_2023.rds")
have_2023_signed <- file.exists(sds23)
if (have_2023_signed) sd_signed_2023 <- readRDS(sds23)
```

Broader question: how does information help an AI become a synthetic coder? 

Thesis: information helps the AI to emulate human coders, but there is a risk of data leakage. Since the AI may have prior knowledge of the country case that could bias its rating, anonymized and summarized texts will help to ensure that the model is taking into account relevant context and not simply applying a memorized score or relying entirely on prior country knowledge. Finetuning should help the model to learn more precisely the specific context relevant for generating human-like country scores. 

We therefore predict that as the AI receives evidence, its MAE relative to human panel consensus should decrease relative to the base model codebook-only condition. We also anticipate that the MAE will drop further as information becomes progressively more abstracted away from named entities related to the country in question, i.e. that the model will learn more from anonymized and summarized text than from raw text.^[finetune-note] Finally we anticipate that finetuning will magnify these effects, with finetuned models showing lower MAE than base models. 

[^finetune-note]:(Drop this in if we decide to relegate the FT-summ and FT-anon models to the appendix) We also tested models that were finetuned on anonymized and summarized evidence. The results suggested that... We explore these results further in the online appendix. 

## 1. Synthetic-coder performance

We hypothesize that more evidence should improve the synthetic coder's ability to emulate expert coder behavior because it will 

```{r}
#| label: fig-crossmodel
#| fig-cap: "Synthetic coder performance across model families (Llama 70B / Qwen 72B / Gemma 27B). Base conditions (top) and fine-tuned raw-text adapters (bottom), each on the four text inputs. Color = model, with country-clustered 95% CIs. The base rows show the greedy (mode) readout only — greedy and mean differ by <0.02 MAE there (~1/9 SESOI); the fine-tuned rows show both greedy (open) and expectation/mean (filled), where the two readouts diverge by 0.16–0.30 MAE. Dashed line = naive persistence; long-dash = human reference; grey band = ±SESOI."
#| fig-width: 8
#| fig-height: 5.2
fig_crossmodel_landscape(exp_bundle, boot_bundle, band = "human_loo", base_readout = "greedy")
```

We will highlight here how the B hypotheses were either null or confirmed the alternative hypotheses. The most surprising finding is that evidence appears to push the models further away from consensus. It is not beyond the SESOI, but it is striking that evidence appears to increase MAE of the models rather than reduce it. The models are paying attention to the evidence but it is making them disagree as much as or more than individual human coders rather than pushing them towards the consensus view.  

The F hypotheses were supported if we look at the expectation readout rather than greedy decoding. But the greedy decoding results are suggestive of something we did not anticipate which is that greedy mode induces the model to better emulate the human coder. However we cannot assess how human-like the AI is by looking at MAE alone. But rather we have to look at the distribution of the AI codings relative to human coders. (transition to results in QMD 07 and 08)

Difficulty slope, attitude, exaggeration gap analysis suggest that the finetuned model is the most human-like and, further, that it is the weights more than the information in the text that does the heavy lifting. The 2x2 suggests that the fine-tuned raw evidence model can perform as well or better without the evidence in the prompt. 

For QMD 09, part I may be underpowered and does not use stale priors. Need to repeat on 2023. The Part II horse race suggests that of the seven models, FT-Raw Text performs the best in terms of mimicking human coders. This is an underpowered version of QMD 07s distributional analysis.

How frame the difference between expectation (mean) and greedy (mode) readouts: 

- Mean readout → an estimator of the panel consensus (a central estimate).
- Greedy readout → a draw from the panel's rating distribution (behaves like one more coder).

## 2. Difficulty tracking

A companion to Figure 1 on the same rows and greedy readout. **Case difficulty** `h_c` is the
within-cell human leave-one-out MAE — how much the real coders disagreed on that country-year-
indicator (the per-case version of Figure 1's human reference line). The **difficulty-tracking
slope** is the weighted OLS slope of a cell's AI error `a_c` on `h_c`. Because `h_c` is defined
as the human error, the human reference lies on the 45° line (slope 1) by construction; a slope
below 1 is flatter than that reference.

Panel A draws the relationship for two cells against the human reference (curves = mean error per
difficulty bin, with country-clustered 95% CI bands). Panel B is the slope for every cell, with
country-clustered 95% CIs, against the human reference at 1.

```{r}
#| label: fig-crossmodel-slope
#| fig-cap: "Difficulty tracking (greedy readout). (A) Mean AI error vs case difficulty h_c for a flat and a steep exemplar, against the human reference (which lies on the 45° line because h_c is defined as human error); 20 equal-count difficulty bins, country-clustered 95% CI bands. (B) Weighted OLS slope of AI error on h_c for every cell, country-clustered 95% CIs, long-dash = human reference (slope 1). Base block shows Llama only pending the Qwen/Gemma base reruns; the FT-raw block covers all three models."
#| fig-width: 12
#| fig-height: 5.4
fig_crossmodel_slope_2panel(dm_slope)
```

<!-- TODO (EJT): interpretation of the difficulty-tracking pattern. Robustness: slopes on the
>=5-coder difficulty pool preserve the cross-model ordering and the FT-vs-base gap (see
build_distmatch.R min_coders arg); the >=2-coder pool is used here for consistency with Fig 1. -->

## 3. Directional bias — signed deviation

The third axis keeps the *direction* MAE discards: `s_c = AI rating − panel mean`, averaged within
democracy quintiles. Zero is the panel-member target (a real coder's signed deviations sum to zero
per bin by construction). The scalar summary is the **exaggeration gap** = (Q5 − Q1): positive =
the model spreads the regime gradient wider than the human panel (rates autocracies below /
democracies above), negative = compresses toward the middle.

Panel A draws the gradient for the flattest and steepest cells against two references: the dotted
zero line (raw panel mean) and the dashed line (V-Dem's IRT `_ord` score expressed on the same
axis, `mean(ord − panel_mean)` per bin). Panel B is the exaggeration gap for every cell.

```{r}
#| label: fig-crossmodel-signeddev
#| fig-cap: "Directional bias (greedy readout). (A) Signed deviation (AI − panel mean) across democracy quintiles for the flattest and steepest cells; dotted 0 = panel mean, dashed = V-Dem IRT target on the same axis, grey band = ±SESOI, country-clustered 95% CI bands. (B) Exaggeration gap (Q5 − Q1 signed deviation) for every cell, country-clustered 95% CIs, dashed 0 = no tilt (positive = exaggerates the gradient, negative = compresses), grey band = ±SESOI. Base block shows Llama only pending the Qwen/Gemma base reruns; the FT-raw block covers all three models."
#| fig-width: 12
#| fig-height: 5.4
fig_crossmodel_signeddev_2panel(sd_signed)
```

<!-- TODO (EJT): interpretation of the directional-bias pattern (base vs FT tilt; the IRT line's
own slope; Qwen's calibration vs its Fig-2 slope). -->

## 4. Continuous Movement

**Runs:** No additional runs required.

Test how the relevance of information changes with the staleness of the prior. 

## 5. Re-identificaiton

**Runs:** Re-identification per model (FT Llama: Ev; FT Qwen: Ev; FT Gemma: Ev) on...what condition? Anonymized, summarized or both? We did see that Llama FT-summ had a harder time re-identifying relative to the other models. Briefly dig into that more. However, fundamentally we are trying to explain patterns in the figures pertaining to the FT-evidence models and so our efforts should primarily focus on those. 

## 6. Name swap

**Runs:** Run name-swap on three models (FT Llama: Ev, FT Qwen: Ev; FT Gemma: Ev). Consider whether using the raw text adapter leaks the names. Perhaps do a trial run to see whether you get different results using the raw versus summarized adapters for Llama. 

## 7. 2024 FH only

# Appendix

## A1. Synthetic-coder performance — 2023 holdout

Replication of Figure 1 on the 2023 holdout year. Same design and readout convention. The
base block (Llama) is complete; among the fine-tuned raw-text adapters, Qwen and Gemma cover
all four inputs while Llama covers codebook and raw text only (the 2023 `anonymized-zeroshot`
and `summarized-zeroshot` raw-adapter runs are pending).

```{r}
#| label: fig-crossmodel-2023
#| fig-cap: "Synthetic coder performance across model families on the 2023 holdout (Llama 70B / Qwen 72B / Gemma 27B). Layout, readout convention (solid = greedy, open = mean; base rows greedy only), and reference lines match Figure 1. Country-clustered 95% CIs."
#| fig-width: 8
#| fig-height: 5.2
#| eval: !expr have_2023
fig_crossmodel_landscape(exp_bundle_2023, boot_bundle_2023, band = "human_loo", base_readout = "greedy")
```

## A2. Difficulty tracking — 2023 holdout

Replication of Figure 2 on the 2023 holdout year. Same two-panel design, featured cells, reference
(slope 1), and greedy readout. The base block (Llama) is complete; among the fine-tuned raw-text
adapters, Qwen and Gemma cover all four inputs while Llama covers codebook and raw text only (the
2023 `anonymized-zeroshot` and `summarized-zeroshot` raw-adapter runs are pending — same gap as A1).

```{r}
#| label: fig-crossmodel-slope-2023
#| fig-cap: "Difficulty tracking on the 2023 holdout — replication of Figure 2. (A) AI error vs case difficulty for the same featured cells; (B) difficulty-tracking slope per cell, long-dash = human reference (slope 1). Greedy readout, country-clustered 95% CIs."
#| fig-width: 12
#| fig-height: 5.4
#| eval: !expr have_2023_slope
fig_crossmodel_slope_2panel(dm_slope_2023)
```

## A3. Directional bias — 2023 holdout

Replication of Figure 3 on the 2023 holdout year. Same two-panel design, featured cells, references
(panel mean at 0, V-Dem IRT dashed), and greedy readout. Same model/condition coverage as A1–A2.

```{r}
#| label: fig-crossmodel-signeddev-2023
#| fig-cap: "Directional bias on the 2023 holdout — replication of Figure 3. (A) Signed deviation across democracy quintiles for the flattest and steepest cells, dotted 0 = panel mean, dashed = V-Dem IRT; (B) exaggeration gap per cell. Greedy readout, country-clustered 95% CIs, grey band = ±SESOI."
#| fig-width: 12
#| fig-height: 5.4
#| eval: !expr have_2023_signed
fig_crossmodel_signeddev_2panel(sd_signed_2023)
```

## A4. B and F hypothesis tests

```{r}
#| label: fig-bf
#| fig-cap: "Paired ΔMAE for the registered B and F contrasts, with the zero line and ±SESOI band."
#| fig-width: 8
#| fig-height: 4.8
fig_bf(boot_bundle)
```

## A5. Full model × input grid

```{r}
#| label: fig-grid
#| fig-cap: "Every model on every input (the off-diagonal expansion of Figure 1). Greedy (mode) vs expectation (mean) MAE with country-clustered 95% CIs. Rings mark each fine-tuned model's own training input (the diagonal). Dashed line = naive persistence model; long-dash line = human reference; grey band = ±SESOI around the human reference."
#| fig-width: 8
#| fig-height: 9
fig_readout_grid(exp_bundle, boot_bundle, band = "human_loo")
```
