---
title: "2019 results — main-text figures and descriptive tables"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: true
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
---

Main-text deliverable for the 2019 results. Everything here reads the precomputed bootstrap
bundle written by `analysis/06-bootstrap-metrics-2019.qmd`; **no bootstrap is rerun**. The
two figures are the reader-facing core — **Figure 1** (the MAE landscape) and **Figure 2**
(divergence from the codebook-only prior) — followed by the descriptive tables that frame the
opening discussion. Formal per-hypothesis tests live in `paper/appendix-2019.qmd`.

Primary metric is **AI MAE against the panel mean**, read against the human leave-one-out
(LOO) MAE. SESOI = half the 2019 rounding floor; a delta whose CI sits inside ±SESOI is not
substantively meaningful.

## Setup — load helpers and the bundle

```{r setup}
library(tidyverse)
library(gt)
library(glue)

# paper/ lives inside the panel-member git repo, so the root is the enclosing .git.
proj_root <- rprojroot::find_root(rprojroot::is_git_root)

# Static vocabulary + paired-delta machinery (single source of truth).
source(file.path(proj_root, "helpers", "bootstrap_helpers.R"))

# Unpack the precomputed results into this environment: boot_ci, boot_results, human_ref,
# persist_ref, sesoi, rounding_floor, movement_bins. paired_delta()/cell_series() read
# `boot_results` and `sesoi` from here by default, so unpacking is what makes them run.
list2env(readRDS(file.path(proj_root, "data", "derived", "bootstrap_2019.rds")),
         environment())

glue("bundle loaded · SESOI ±{round(sesoi, 3)} · human LOO {round(human_ref$human_mae, 3)} · ",
     "persistence(round) {round(persist_ref$persist_mae, 3)}")
```

---

## Figure 1 — Synthetic coder performance (the MAE landscape)

One panel, all seven cells on a common AI-MAE axis: the base condition ladder (Cb→Ev→An→Su)
and the fine-tuned diagonal, against the **Naive model** (persistence copy-forward) and
**human LOO** rails. Two dynamics read straight off it — the *centrifugal* base ladder (adding
evidence and stripping identity pushes the base model rightward, away from consensus) and the
*centripetal* fine-tuned cluster (training pulls the models onto the human LOO line).

```{r}
#| label: fig-1
#| fig-width: 8
#| fig-height: 4.8

land <- bind_rows(
  boot_ci |>
    filter(model_key == "llama-70b",
           condition %in% c("codebook", "evidence", "anonymized", "summarized")) |>
    mutate(family = "Base",
           row = recode(condition, codebook = "Codebook", evidence = "Raw Text",
                        anonymized = "Anonymized", summarized = "Summarized")),
  boot_ci |>
    semi_join(ft_diag, by = c("model_key", "condition")) |>
    mutate(family = "Fine-tuned",
           row = recode(model_key,
             "llama-70b-ft-raw"  = "FT · Raw Text",
             "llama-70b-ft-anon" = "FT · Anonymized",
             "llama-70b-ft-summ" = "FT · Summarized"))
) |>
  mutate(
    family = factor(family, levels = c("Base", "Fine-tuned")),
    # first level plots at the bottom: FT block below, base ladder above
    row = factor(row, levels = c("FT · Summarized", "FT · Anonymized", "FT · Raw Text",
                                 "Summarized", "Anonymized", "Raw Text", "Codebook"))
  )

ggplot(land, aes(ai_mae, row, color = family)) +
  # ±SESOI of the human LOO level: the band within which a synthetic cell's error is not
  # substantively distinguishable from the human coder's. Reference lines and bands are capped
  # just above the top row (y = 7.3) so the two labels sit in the clean strip above the plot.
  annotate("rect", xmin = human_ref$human_mae - sesoi, xmax = human_ref$human_mae + sesoi,
           ymin = -Inf, ymax = 7.3, fill = "grey85", alpha = 0.55) +
  annotate("segment", x = persist_ref$persist_mae, xend = persist_ref$persist_mae,
           y = -Inf, yend = 7.3, linetype = "dashed", color = "grey40") +
  annotate("segment", x = human_ref$human_mae, xend = human_ref$human_mae,
           y = -Inf, yend = 7.3, linetype = "longdash", color = "grey40") +
  annotate("text", x = persist_ref$persist_mae, y = 6.8, label = "Naive model",
           color = "grey40", size = 2.9, hjust = -0.07, vjust = 0.5) +
  annotate("text", x = human_ref$human_mae, y = 7.5, label = "±SESOI of human LOO MAE",
           color = "grey40", size = 2.9, hjust = 0.5, vjust = 0.5) +
  # faint separator between the base block (top 4) and the FT block (bottom 3)
  geom_hline(yintercept = 3.5, color = "grey85", linewidth = 0.4) +
  geom_errorbar(aes(xmin = ai_lo, xmax = ai_hi), orientation = "y",
                width = 0.20, linewidth = 0.7) +
  geom_point(size = 2) +
  scale_color_manual(values = family_pal, name = NULL) +
  scale_y_discrete() +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.08), add = c(0.02, 0))) +
  coord_cartesian(clip = "off") +
  labs(x = "AI Mean Absolute Error", y = NULL,
       title = "Figure 1 — Synthetic coder performance on V-Dem ratings",
       subtitle = "2019 · mean absolute error relative to panel mean (lower = closer to consensus)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", legend.justification = "left",
        panel.grid.major.y = element_blank(), plot.title.position = "plot",
        plot.title = element_text(face = "bold"))
```

---

## Figure 2 — Divergence from the codebook-only prior

A single ruler: the codebook-only base condition is the reference (0), and every other thing —
the base de-identified conditions, the fine-tuned diagonal, and the human coder — is its own
paired ΔMAE from that pure-prior floor, read against ±SESOI. Reading top to bottom: the base
conditions fan *right* as identity is stripped (An, Su past SESOI), the fine-tuned models sit
near the prior, and the human coder marks where a real held-out rater lands relative to the
same anchor. (The delta for each cell is formed *within* each bootstrap draw and then quantiled,
so these CIs are the honest paired test — not the overlap of two level bars.)

```{r}
#| label: fig-2
#| fig-width: 8
#| fig-height: 4.6

vs_cb_all <- bind_rows(
  paired_delta("llama-70b", "evidence",   "llama-70b", "codebook", "Ev") |> mutate(kind = "Base"),
  paired_delta("llama-70b", "anonymized", "llama-70b", "codebook", "An") |> mutate(kind = "Base"),
  paired_delta("llama-70b", "summarized", "llama-70b", "codebook", "Su") |> mutate(kind = "Base"),
  paired_delta("llama-70b-ft-raw",  "evidence-zeroshot",   "llama-70b", "codebook", "FT-raw")  |> mutate(kind = "Fine-tuned"),
  paired_delta("llama-70b-ft-anon", "anonymized-zeroshot", "llama-70b", "codebook", "FT-anon") |> mutate(kind = "Fine-tuned"),
  paired_delta("llama-70b-ft-summ", "summarized-zeroshot", "llama-70b", "codebook", "FT-summ") |> mutate(kind = "Fine-tuned")
)

# human LOO − codebook-only, paired per draw off boot_results (human_mae = the LOO for that draw).
loo_draws <- boot_results |>
  filter(model_key == "llama-70b", condition == "codebook") |>
  transmute(id, delta = human_mae - ai_mae)
loo_vs_cb <- tibble(
  label = "human LOO", kind = "Human coder",
  est   = loo_draws$delta[loo_draws$id == "Apparent"],
  lo    = quantile(loo_draws$delta[loo_draws$id != "Apparent"], 0.025),
  hi    = quantile(loo_draws$delta[loo_draws$id != "Apparent"], 0.975),
  p_gt  = mean(loo_draws$delta[loo_draws$id != "Apparent"] > 0),
  sesoi_out = abs(loo_draws$delta[loo_draws$id == "Apparent"]) > sesoi
)

vs_cb_all <- bind_rows(vs_cb_all, loo_vs_cb) |>
  mutate(
    kind  = factor(kind, levels = c("Base", "Fine-tuned", "Human coder")),
    # first level plots at the bottom: human at the very bottom, FT block, then base ladder on top
    label = factor(label, levels = rev(c("Ev", "An", "Su",
                                         "FT-raw", "FT-anon", "FT-summ", "human LOO")))
  )

ggplot(vs_cb_all, aes(est, label, color = kind)) +
  annotate("rect", xmin = -sesoi, xmax = sesoi, ymin = -Inf, ymax = Inf,
           fill = "grey85", alpha = 0.55) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  # separators: base ladder (top 3) | FT block (middle 3) | human coder (bottom 1)
  geom_hline(yintercept = 4.5, color = "grey85", linewidth = 0.4) +
  geom_hline(yintercept = 1.5, color = "grey85", linewidth = 0.4) +
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y", width = 0.18, linewidth = 0.7) +
  geom_point(size = 3) +
  scale_color_manual(values = c(family_pal, "Human coder" = "grey25"), name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0.04, 0.06))) +
  labs(x = "ΔMAE vs. codebook-only prior", y = NULL,
       title = "Figure 2 — Divergence from the codebook-only prior",
       subtitle = glue("2019 · paired ΔMAE from the codebook-only baseline ",
                       "(right = more divergent than the pure prior) · grey band = ±SESOI {round(sesoi, 3)}")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", legend.justification = "left",
        panel.grid.major.y = element_blank(), plot.title.position = "plot",
        plot.title = element_text(face = "bold"))
```

Companion numbers for Figure 2 (the divergence-from-prior estimates plotted above):

```{r fig-2-readout}
vs_cb_all |>
  arrange(desc(as.integer(label))) |>
  transmute(
    Cell     = label,
    Family   = kind,
    ΔMAE     = est,
    `95% CI` = glue("[{round(lo, 3)}, {round(hi, 3)}]"),
    `Beyond SESOI?` = if_else(sesoi_out, "yes", "no")
  ) |>
  gt() |>
  tab_header(
    title    = "Divergence from the codebook-only prior",
    subtitle = glue("Paired country-clustered bootstrap · SESOI = ±{round(sesoi, 3)}")
  ) |>
  fmt_number(columns = ΔMAE, decimals = 3)
```

**Reading these as preliminary.** Both figures are descriptive, single-year (2019) estimates on
a metric — static MAE against the panel mean — that rewards central tendency: the naive
copy-forward sits closest of all, and the pool is dominated by country-years that barely move
between waves, where no independent rater can beat a copy-forward. Whether the synthetic coders
earn their keep on the country-years that actually *move* — where a persistence copier is
structurally blind and independent reading should pay — is what the movement split below
motivates, and what the preregistered R2 transition test decides on 2023 with a pre-fixed flag.

---

## Descriptive tables (opening discussion)

### Table 1 — Nobody beats copy-forward on the full 2019 pool

The natural naive baseline is **persistence**: predict the 2019 panel mean with the (rounded)
2018 panel mean. V-Dem panel means are very sticky year to year, so on the full pool a trivial
copy-forward sits closest of all — below both the base model and a typical human coder. This is
the fact that reframes the whole exercise: MAE against a sticky aggregate rewards redundant
predictors and penalizes independent observation.

```{r table-1-leaderboard}
persist_cont <- weighted.mean(movement_bins$persist_cont, movement_bins$n)  # continuous persistence

leaderboard <- bind_rows(
  tibble(Predictor = "Persistence (2018 mean, rounded)",
         MAE = persist_ref$persist_mae, lo = persist_ref$lo, hi = persist_ref$hi),
  boot_ci |> filter(model_key == "llama-70b", condition == "codebook") |>
    transmute(Predictor = "Base : Codebook-only", MAE = ai_mae, lo = ai_lo, hi = ai_hi),
  boot_ci |> filter(model_key == "llama-70b", condition == "evidence") |>
    transmute(Predictor = "Base : Evidence", MAE = ai_mae, lo = ai_lo, hi = ai_hi),
  tibble(Predictor = "Human LOO (typical held-out coder)",
         MAE = human_ref$human_mae, lo = human_ref$lo, hi = human_ref$hi)
)

leaderboard |>
  transmute(Predictor, MAE, `95% CI` = glue("[{round(lo, 3)}, {round(hi, 3)}]")) |>
  gt() |>
  tab_header(
    title    = "Panel-mean MAE, 2019 (lower = closer to the panel mean)",
    subtitle = glue("Persistence beats every rater on the full pool · ",
                    "continuous persistence MAE ≈ {round(persist_cont, 3)} (integer-fair shown)")
  ) |>
  fmt_number(columns = MAE, decimals = 3)
```

### Table 2 — The reversal is a stable-cell artifact (exploratory)

Split the pool by how far the panel mean actually moved 2018→2019. On the ~87% of cells that
barely move, persistence is unbeatable and reading evidence only adds noise that pulls the
rating off the sticky mean (base:Ev > base:Cb). On the cells that genuinely move (|Δ| ≥ 1.0),
the base model *beats* persistence and evidence helps most exactly there. The aggregate
"evidence hurts" is a stable-cell artifact.

**Exploratory / hypothesis-generating only.** The bin edge is the realized 2018→2019 change, so
persistence's per-bin MAE is mechanical (it is just the bin's mean |Δ|), and the split is
endogenous. The confirmatory movement test is preregistered R2 on 2023 with the ERT transition
flag fixed in advance — this table is not a confirmatory 2019 result.

```{r table-2-movement}
movement_bins |>
  transmute(
    `|Δ| bin (2018→2019)` = bin,
    n,
    `Persistence (cont.)` = persist_cont,
    `Base : Codebook`     = base_cb_mae,
    `Base : Evidence`     = base_ev_mae
  ) |>
  gt() |>
  tab_header(
    title    = "Panel-mean MAE by 2018→2019 movement bin (≥2-coder pool)",
    subtitle = "Exploratory. Persistence (cont.) = the bin's mean |Δ|; evidence helps only where the mean moves"
  ) |>
  fmt_number(columns = c(`Persistence (cont.)`, `Base : Codebook`, `Base : Evidence`), decimals = 3) |>
  fmt_number(columns = n, decimals = 0)
```

The throughline: a rater's value is being an *independent observation*, not minimizing MAE
against a quantity the past already predicts. "Worse than persistence" is not a deployment
failure — it is what any honest independent coder looks like on a low-variance target, and it is
why the study's real question is decided conditional on movement (R2), not on the full pool.
